---
title: "Revision Simulation R3: two-stage dynamic discovery and functional testing"
author: "Ziang Zhang"
date: "2026-09-17"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9.4,
  fig.height = 5.8,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = TRUE
)
project_root <- if (
  file.exists("code/revision_simulations/shared/simulation_functions.R")
) "." else ".."
experiment <- file.path(
  project_root,
  "code/revision_simulations/internal/r3_two_stage_functional_testing_pilot10"
)
source(file.path(experiment, "pilot10_helpers.R"))
source(file.path(experiment, "reporting.R"))
comparison_experiment <- file.path(
  project_root, "code/revision_simulations/internal/r3_stage1_method_comparison"
)
source(file.path(comparison_experiment, "comparison_helpers.R"))
source(file.path(comparison_experiment, "reporting.R"))
report <- r3c_load_report(project_root)
stage1_summary <- report$tables$stage1_mc_summary
stage1_alpha005 <- report$tables$stage1_alpha005_summary
stage2_summary <- report$tables$functional_mc_summary
mechanism_labels <- report$baseline$config$contract$mechanism_labels

stage1_value <- function(mechanism, field, digits = 3L, method = "FASH-IWP1-BF") {
  row <- stage1_alpha005[
    stage1_alpha005$truth_mechanism == mechanism &
    stage1_alpha005$method == method,
    ,
    drop = FALSE
  ]
  stopifnot(nrow(row) == 1L)
  formatC(row[[field]], format = "f", digits = digits)
}
```

<style>
p.caption {
  font-weight: 600;
  color: #333;
}
.r3-two-stage-table {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.r3-two-stage-table table {
  width: 100%;
  margin-bottom: 1em;
}
.r3-two-stage-table caption {
  font-weight: 600;
  color: #333;
}
.main-container p code {
  white-space: normal;
  overflow-wrap: anywhere;
  word-break: break-word;
}
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: hidden;
}
</style>

## Overview

This simulation evaluates dynamic eQTL screening followed by temporal
classification. It uses ten Monte Carlo seeds and two truth mechanisms: broad
random B-spline trajectories and compact raised-cosine trajectories. The
analysis proceeds in two stages:

1. Compare BF-adjusted FASH-IWP1, BF-adjusted FASH-linear, and direct linear
   and quadratic interaction tests for dynamic discovery across nominal FDR
   levels from `0.005` through `0.20`, using all `6,362` units.
2. At the fixed threshold $q_{\mathrm{dyn}}=0.05$, classify the FASH-IWP1
   dynamic discoveries as Early, Middle, Late, or Switch, separately for each
   functional, using $q_{\mathrm{func}}=0.05$.

For donor $i$, unit $j$, and observed time $t$,

$$
Y_{ijt}=\alpha_{jt}+G_{ij}\beta_j(t)+X_i^T\gamma_{jt}+\epsilon_{ijt},
\qquad \epsilon_{ijt}\sim N(0,1).
$$

For each seed, one tested cis variant is sampled uniformly for each of the
`6,362` tested genes, and its YRI dosage vector is retained for the same `19`
donors. Each replicate has `16` time points and five standardized covariates.
Per-time regressions include an intercept, genotype, and five covariates;
their t-adjusted standard errors use 12 residual degrees of freedom.
The dynamic, constant, and zero-effect counts are exactly
`1,272/2,545/2,545`, with truth classes balanced across empirical-MAF deciles.

The dynamic effects have Early/Middle/Late probabilities `0.29/0.42/0.29`.
Switch status is balanced within each location, giving six exact cell counts
`185/184/267/267/185/184`. The final trajectory, including its random genetic
main effect, must satisfy the assigned location and Switch margins. Early is
$t\leq3$, Middle is $3<t<12$, and Late is $t\geq12$ on the dense 0.1-day grid;
Switch uses threshold 0.25. The two mechanisms retain the same truth contract:

- Broad effects are cubic B-splines with six independent standard-normal
  coefficients, centered over the observed times and scaled to maximum absolute
  deviation 2.
- Compact effects contain one, two, or three raised-cosine peaks with half-width
  1.5 and are scaled to centered RMS 0.9.

## Stage 1: dynamic discovery across alpha

FASH-IWP1 and FASH-linear share the unrestricted constant null, the one-day
predictive-SD grid (zero plus 51 positive values from $\exp(-5)$ to 1), and
`penalty = 10`. FASH-linear uses a finite mixture of Gaussian linear slopes.
Both fits use the pinned BF update, which retains the conditional alternative
mixture and updates its total weight against the constant null.

The direct linear LRT tests the genotype-by-time term; the quadratic LRT
jointly tests the linear and quadratic interaction terms. Their eFDR uses
100 donor-level permutations, with expression trajectories and covariates
permuted together relative to genotype. The
direct tests are supplied with the true $\pi_0=5090/6362$; FASH estimates its
null weight. All methods use the same simulated data. FASH receives per-time
regression summaries, while the direct tests receive individual-level data.

Let $D_m(\alpha)$ be the discoveries for method $m$. FASH uses cumulative mean
lfdr, and direct tests use monotone permutation eFDR q-values. For each replicate,

$$
\mathrm{Power}_m(\alpha)=\frac{|D_m(\alpha)\cap H_1|}{|H_1|},
\qquad
\mathrm{FDP}_m(\alpha)=
\frac{|D_m(\alpha)\cap H_0|}{\max\{|D_m(\alpha)|,1\}}.
$$

The curves show the ten-seed means of power and FDP (empirical FDR). Ribbons
are pointwise replicate minima and maxima. Each panel is one truth mechanism;
the vertical line marks $\alpha=0.05$, and the FDR diagonal marks nominal
calibration. Both FASH methods also retain Raw results in the cache for later
Raw-versus-BF comparisons.

```{r stage1-selection, eval=FALSE}
# FASH: rank all-unit lfdr values and select by their cumulative mean.
lfdr_order <- order(lfdr, unit_index)
fdr <- cumsum(lfdr[lfdr_order]) / seq_along(lfdr_order)
fash_calls <- lfdr_order[fdr <= alpha]

# Direct LRTs: apply the same donor permutations to both interaction degrees.
permutation_null <- compute_direct_interaction_permutation_null(
  out = individual_level_data,
  n_permutations = 100L,
  interaction_degrees = c(1L, 2L),
  seed = seed + 10000L,
  permute_covariates_with_expression = TRUE,
  num_cores = num_cores
)
direct_qvalues <- empirical_fdr_qvalues(
  observed_pvalues = observed_lrt$pvalue,
  null_pvalues = as.vector(permutation_null$null_pvalues[[degree_label]]),
  pi0_method = "fixed",
  fixed_pi0 = 5090 / 6362
)$qvalue
direct_calls <- which(direct_qvalues <= alpha)
```

### Power as a function of alpha

```{r stage1-power, fig.width=10.4, fig.height=5.2, fig.cap="Figure 1. Dynamic discovery power for BF-adjusted FASH-IWP1, BF-adjusted FASH-linear, and direct linear and quadratic interaction tests. Panels show broad random B-spline and compact raised-cosine truth; curves are ten-seed means and ribbons are pointwise replicate minima and maxima. Direct-test eFDR uses the true null proportion and 100 donor permutations."}
# Select the four main-comparison methods; Raw FASH results remain in the cache.
power_curves <- stage1_summary[
  stage1_summary$method %in% report$contract$display_methods,
  c("truth_mechanism", "method", "alpha", "mean_power", "power_min", "power_max")
]
power_curves <- power_curves[order(
  power_curves$truth_mechanism,
  match(power_curves$method, report$contract$display_methods),
  power_curves$alpha
), ]
r3c_plot_stage1(power_curves, metric = "power")
```

### Empirical FDR as a function of alpha

```{r stage1-fdr, fig.width=10.4, fig.height=5.2, fig.cap="Figure 2. Empirical dynamic FDR for the same four methods and paired datasets. Curves average replicate-specific false-discovery proportions across ten seeds; ribbons are pointwise replicate minima and maxima. The diagonal marks nominal FDR, and the vertical line marks alpha = 0.05."}
# Use the same truth, method order, and alpha grid as the power comparison.
fdr_curves <- stage1_summary[
  stage1_summary$method %in% report$contract$display_methods,
  c("truth_mechanism", "method", "alpha", "mean_empirical_fdr",
    "empirical_fdr_min", "empirical_fdr_max")
]
fdr_curves <- fdr_curves[order(
  fdr_curves$truth_mechanism,
  match(fdr_curves$method, report$contract$display_methods),
  fdr_curves$alpha
), ]
r3c_plot_stage1(fdr_curves, metric = "fdr")
```

At $\alpha=0.05$, mean power for BF-adjusted FASH-IWP1 and direct quadratic
is `r stage1_value("random_bspline", "mean_power")` and
`r stage1_value("random_bspline", "mean_power", method = "Direct-quadratic-LRT-eFDR-true-pi0")`
under broad truth, and `r stage1_value("raised_cosine", "mean_power")` and
`r stage1_value("raised_cosine", "mean_power", method = "Direct-quadratic-LRT-eFDR-true-pi0")`
under compact truth. Under compact truth, their mean empirical FDR values are
`r stage1_value("raised_cosine", "mean_empirical_fdr")` and
`r stage1_value("raised_cosine", "mean_empirical_fdr", method = "Direct-quadratic-LRT-eFDR-true-pi0")`,
respectively; power should therefore be read together with empirical FDR.

## Stage 2: conditional functional classification at 0.05

For the remainder of the analysis, $D(0.05)$ denotes the BF-adjusted
FASH-IWP1 discoveries at 0.05. This candidate set remains fixed for Stage 2.
The fitted prior remains the full-data prior. For each candidate and each
functional, 3,000 posterior draws are evaluated on the dense time grid. The
candidate lfsr values are then ranked separately for Early, Middle, Late, and
Switch, and calls are made while their cumulative FSR is at most 0.05. The four
classifications are evaluated separately and need not be mutually exclusive.

For target $k$, let $S_k\subseteq D(0.05)$ be its final calls and let
$T_k\{\beta_j\}>0$ denote true membership. We report

$$
\mathrm{FSR}_k=
\frac{\sum_{j\in S_k}\mathbf{1}[T_k\{\beta_j\}\leq0]}
{\max(|S_k|,1)},
$$

and conditional classification power

$$
\mathrm{Power}_{k\mid\mathrm{Stage\ 1}}=
\frac{\sum_{j\in S_k}\mathbf{1}[T_k\{\beta_j\}>0]}
{\sum_{j\in D(0.05)}\mathbf{1}[T_k\{\beta_j\}>0]}.
$$

The tables average each replicate-specific rate and call count across the ten
seeds.

```{r stage2-selection, eval=FALSE}
dynamic_candidates <- r3ts_stage1(
  fdr_table,
  n_units = 6362L,
  q = 0.05
)
lfsr_by_target <- compute_functional_lfsr(
  fit = fit_iwp1_bf,
  functionals = temporal_functionals,
  indices = dynamic_candidates,
  smooth_var = seq(0, 15, by = 0.1),
  num_cores = num_cores,
  seed = functional_posterior_seed
)
stage2_calls <- r3ts_stage2(
  dynamic_candidates,
  lfsr_by_target,
  n_units = 6362L,
  q = 0.05
)
```

```{r stage2-broad-table, results='asis'}
# Select the truth mechanism and retain only the three displayed estimands.
broad_table <- stage2_summary[
  stage2_summary$truth_mechanism == "random_bspline",
  c("truth_mechanism", "target", "mean_empirical_fsr",
    "mean_conditional_classification_power", "mean_calls")
]
broad_table <- broad_table[
  match(c("early", "middle", "late", "switch"), broad_table$target),
]
r3ts10_show_table(
  r3ts10_format_stage2_table(broad_table, "random_bspline"),
  paste(
    "Table 1. Ten-seed mean Stage-2 results under broad random B-spline truth.",
    "Classification power is conditional on true target members retained by",
    "Stage 1 at q = 0.05."
  )
)
```

```{r stage2-compact-table, results='asis'}
# Select the truth mechanism and retain only the three displayed estimands.
compact_table <- stage2_summary[
  stage2_summary$truth_mechanism == "raised_cosine",
  c("truth_mechanism", "target", "mean_empirical_fsr",
    "mean_conditional_classification_power", "mean_calls")
]
compact_table <- compact_table[
  match(c("early", "middle", "late", "switch"), compact_table$target),
]
r3ts10_show_table(
  r3ts10_format_stage2_table(compact_table, "raised_cosine"),
  paste(
    "Table 2. Ten-seed mean Stage-2 results under compact raised-cosine truth.",
    "Classification power is conditional on true target members retained by",
    "Stage 1 at q = 0.05."
  )
)
```

Stage-1 screening and Stage-2 classification answer distinct questions. The
curves quantify which true dynamic effects reach the candidate set; the tables
then quantify functional classification among true target members in that set.
The two nominal 0.05 thresholds are operating rules, and the ten-seed empirical
rates describe performance under the two retained truth mechanisms.

## Reproducibility

The additional comparison is implemented in
`code/revision_simulations/internal/r3_stage1_method_comparison/` and reuses
the completed ten-seed R3 inputs and IWP1 results. Its input-pairing checks
validate genotype, truth, regression summaries, individual-level expression,
and covariates before fitting the additional methods. The original two-stage
runner remains in
`code/revision_simulations/internal/r3_two_stage_functional_testing_pilot10/`.
All FASH fits use `fashr` 0.1.43 at Git SHA
`bf223df75da6e41ae48607a56b4cd12d7c3b24e7`.

The compact comparison cache retains all-unit scores, Raw/BF prior weights,
per-seed curves, and direct-test diagnostics. Full Raw/BF fit objects and
permutation null matrices remain in server checkpoints. Stage-2 tables are
copied unchanged from the validated two-stage cache. This page reads only
completed, hash-validated caches. To run the additional methods on Midway3:

```bash
sbatch /project/mstephens/ziangzhang/fash/workspace/slurm/run_r3_stage1_method_comparison.sbatch
```

```{r cache-identity, results='asis', echo=FALSE}
cat("Comparison cache ID: `", report$contract$result_id, "`.\n", sep = "")
```
