---
title: "Revision Simulation R3: two-stage dynamic discovery and functional testing"
author: "Ziang Zhang"
date: "2026-09-05"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9.4,
  fig.height = 5.8,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = TRUE
)
project_root <- if (
  file.exists("code/revision_simulations/shared/simulation_functions.R")
) "." else ".."
experiment <- file.path(
  project_root,
  "code/revision_simulations/internal/r3_two_stage_functional_testing_pilot10"
)
source(file.path(experiment, "pilot10_helpers.R"))
source(file.path(experiment, "reporting.R"))
report <- r3ts10_load_report(project_root)
stage1_summary <- report$tables$stage1_mc_summary
stage1_alpha005 <- report$tables$stage1_alpha005_summary
stage2_summary <- report$tables$functional_mc_summary
mechanism_labels <- report$contract$mechanism_labels

stage1_value <- function(mechanism, field, digits = 3L) {
  row <- stage1_alpha005[
    stage1_alpha005$truth_mechanism == mechanism,
    ,
    drop = FALSE
  ]
  stopifnot(nrow(row) == 1L)
  formatC(row[[field]], format = "f", digits = digits)
}
```

<style>
.r3-two-stage-table {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.r3-two-stage-table table {
  width: 100%;
  margin-bottom: 1em;
}
.r3-two-stage-table caption {
  font-weight: 600;
  color: #333;
}
.main-container p code {
  white-space: normal;
  overflow-wrap: anywhere;
  word-break: break-word;
}
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: hidden;
}
</style>

## Overview

This simulation evaluates dynamic eQTL screening followed by temporal
classification. It uses ten Monte Carlo seeds and two truth mechanisms: broad
random B-spline trajectories and compact raised-cosine trajectories. The
analysis proceeds in two stages:

1. Fit BF-adjusted IWP1 to all `6,362` units and evaluate dynamic discovery
   power and empirical FDR across nominal levels from `0.005` through `0.20`.
2. At the fixed threshold $q_{\mathrm{dyn}}=0.05$, classify the retained
   dynamic discoveries as Early, Middle, Late, or Switch, separately for each
   functional, using $q_{\mathrm{func}}=0.05$.

For donor $i$, unit $j$, and observed time $t$,

$$
Y_{ijt}=\alpha_{jt}+G_{ij}\beta_j(t)+X_i^T\gamma_{jt}+\epsilon_{ijt},
\qquad \epsilon_{ijt}\sim N(0,1).
$$

For each seed, one tested cis variant is sampled uniformly for each of the
`6,362` tested genes, and its YRI dosage vector is retained for the same `19`
donors. Each replicate has `16` time points and five standardized covariates.
The regression summaries use the same t-adjusted standard errors as R1 and R2.
The dynamic, constant, and zero-effect counts are exactly
`1,272/2,545/2,545`, with truth classes balanced across empirical-MAF deciles.

The dynamic effects have Early/Middle/Late probabilities `0.29/0.42/0.29`.
Switch status is balanced within each location, giving six exact cell counts
`185/184/267/267/185/184`. The final trajectory, including its random genetic
main effect, must satisfy the assigned location and Switch margins. Early is
$t\leq3$, Middle is $3<t<12$, and Late is $t\geq12$ on the dense 0.1-day grid;
Switch uses threshold 0.25. The two mechanisms retain the same truth contract:

- Broad effects are cubic B-splines with six independent standard-normal
  coefficients, centered over the observed times and scaled to maximum absolute
  deviation 2.
- Compact effects contain one, two, or three raised-cosine peaks with half-width
  1.5 and are scaled to centered RMS 0.9.

## Stage 1: dynamic discovery across alpha

Let $D(\alpha)$ contain the BF-adjusted IWP1 discoveries whose cumulative mean
lfdr is at most $\alpha$. For each replicate,

$$
\mathrm{Power}_{\mathrm{dyn}}(\alpha)=
\frac{|D(\alpha)\cap H_1|}{|H_1|},
\qquad
\mathrm{FDR}_{\mathrm{emp}}(\alpha)=
\frac{|D(\alpha)\cap H_0|}{\max\{|D(\alpha)|,1\}}.
$$

The curves average these replicate-specific rates across ten seeds. Shaded
bands are pointwise minima and maxima across those seeds. The vertical line
marks $\alpha=0.05$; the FDR panel also shows the nominal diagonal.

```{r stage1-selection, eval=FALSE}
fdr_table <- get_fash_fdr_table(fit_iwp1_bf)
stage1_by_alpha <- lapply(
  seq(0.005, 0.20, by = 0.005),
  function(alpha) {
    dynamic_calls <- r3ts_stage1(
      fdr_table,
      n_units = 6362L,
      q = alpha
    )
    c(
      power = sum(true_dynamic[dynamic_calls]) / sum(true_dynamic),
      empirical_fdr = sum(!true_dynamic[dynamic_calls]) /
        max(length(dynamic_calls), 1L)
    )
  }
)
```

### Power as a function of alpha

```{r stage1-power, fig.cap="Figure 1. BF-adjusted IWP1 dynamic discovery power across nominal Stage-1 FDR levels under broad random B-spline and compact raised-cosine truth. Curves are ten-seed Monte Carlo means, and ribbons are pointwise replicate minima and maxima."}
# Plot replicate-mean power and the pointwise replicate range for each truth.
power_curves <- stage1_summary[
  order(stage1_summary$truth_mechanism, stage1_summary$alpha),
  c("truth_mechanism", "alpha", "mean_power", "power_min", "power_max")
]
r3ts10_plot_stage1(power_curves, metric = "power", alpha_reference = 0.05)
```

### Empirical FDR as a function of alpha

```{r stage1-fdr, fig.cap="Figure 2. BF-adjusted IWP1 empirical dynamic FDR across nominal Stage-1 FDR levels under broad random B-spline and compact raised-cosine truth. Curves are ten-seed Monte Carlo means, ribbons are pointwise replicate minima and maxima, and the diagonal marks nominal calibration."}
# Plot replicate-mean empirical FDR and its range against nominal alpha.
fdr_curves <- stage1_summary[
  order(stage1_summary$truth_mechanism, stage1_summary$alpha),
  c("truth_mechanism", "alpha", "mean_empirical_fdr",
    "empirical_fdr_min", "empirical_fdr_max")
]
r3ts10_plot_stage1(fdr_curves, metric = "fdr", alpha_reference = 0.05)
```

At $\alpha=0.05$, mean dynamic power is
`r stage1_value("random_bspline", "mean_power")` under broad truth and
`r stage1_value("raised_cosine", "mean_power")` under compact truth; the
corresponding mean empirical FDR values are
`r stage1_value("random_bspline", "mean_empirical_fdr")` and
`r stage1_value("raised_cosine", "mean_empirical_fdr")`.

## Stage 2: conditional functional classification at 0.05

For the remainder of the analysis, the candidate set is fixed at $D(0.05)$.
The fitted prior remains the full-data prior. For each candidate and each
functional, 3,000 posterior draws are evaluated on the dense time grid. The
candidate lfsr values are then ranked separately for Early, Middle, Late, and
Switch, and calls are made while their cumulative FSR is at most 0.05. The four
classifications are evaluated separately and need not be mutually exclusive.

For target $k$, let $S_k\subseteq D(0.05)$ be its final calls and let
$T_k\{\beta_j\}>0$ denote true membership. We report

$$
\mathrm{FSR}_k=
\frac{\sum_{j\in S_k}\mathbf{1}[T_k\{\beta_j\}\leq0]}
{\max(|S_k|,1)},
$$

and conditional classification power

$$
\mathrm{Power}_{k\mid\mathrm{Stage\ 1}}=
\frac{\sum_{j\in S_k}\mathbf{1}[T_k\{\beta_j\}>0]}
{\sum_{j\in D(0.05)}\mathbf{1}[T_k\{\beta_j\}>0]}.
$$

The tables average each replicate-specific rate and call count across the ten
seeds.

```{r stage2-selection, eval=FALSE}
dynamic_candidates <- r3ts_stage1(
  fdr_table,
  n_units = 6362L,
  q = 0.05
)
lfsr_by_target <- compute_functional_lfsr(
  fit = fit_iwp1_bf,
  functionals = temporal_functionals,
  indices = dynamic_candidates,
  smooth_var = seq(0, 15, by = 0.1),
  num_cores = num_cores,
  seed = functional_posterior_seed
)
stage2_calls <- r3ts_stage2(
  dynamic_candidates,
  lfsr_by_target,
  n_units = 6362L,
  q = 0.05
)
```

```{r stage2-broad-table, results='asis'}
# Select the truth mechanism and retain only the three displayed estimands.
broad_table <- stage2_summary[
  stage2_summary$truth_mechanism == "random_bspline",
  c("truth_mechanism", "target", "mean_empirical_fsr",
    "mean_conditional_classification_power", "mean_calls")
]
broad_table <- broad_table[
  match(c("early", "middle", "late", "switch"), broad_table$target),
]
r3ts10_show_table(
  r3ts10_format_stage2_table(broad_table, "random_bspline"),
  paste(
    "Table 1. Ten-seed mean Stage-2 results under broad random B-spline truth.",
    "Classification power is conditional on true target members retained by",
    "Stage 1 at q = 0.05."
  )
)
```

```{r stage2-compact-table, results='asis'}
# Select the truth mechanism and retain only the three displayed estimands.
compact_table <- stage2_summary[
  stage2_summary$truth_mechanism == "raised_cosine",
  c("truth_mechanism", "target", "mean_empirical_fsr",
    "mean_conditional_classification_power", "mean_calls")
]
compact_table <- compact_table[
  match(c("early", "middle", "late", "switch"), compact_table$target),
]
r3ts10_show_table(
  r3ts10_format_stage2_table(compact_table, "raised_cosine"),
  paste(
    "Table 2. Ten-seed mean Stage-2 results under compact raised-cosine truth.",
    "Classification power is conditional on true target members retained by",
    "Stage 1 at q = 0.05."
  )
)
```

Stage-1 screening and Stage-2 classification answer distinct questions. The
curves quantify which true dynamic effects reach the candidate set; the tables
then quantify functional classification among true target members in that set.
The two nominal 0.05 thresholds are operating rules, and the ten-seed empirical
rates describe performance under the two retained truth mechanisms.

## Reproducibility

The runner, validation gates, and reporting functions are in
`code/revision_simulations/internal/r3_two_stage_functional_testing_pilot10/`.
The completed cache reuses the ten previously validated mechanism-by-seed
records for the original five seeds and computes ten new records for the five
added seeds. All fits use the frozen `fashr` 0.1.43 build at Git SHA
`bf223df75da6e41ae48607a56b4cd12d7c3b24e7`. Large input and fit checkpoints
remain on Midway3; this page reads only completed, hash-validated result and
genotype caches. To regenerate on Midway3 using the staged inputs and pinned
environment:

```bash
sbatch /project/mstephens/ziangzhang/fash/workspace/slurm/run_r3_two_stage_pilot10.sbatch
```

```{r cache-identity, results='asis', echo=FALSE}
cat("Result cache ID: `", report$contract$result_id, "`.\n", sep = "")
```
