---
title: "Appendix B: Simulation"
author: "Ziang Zhang"
date: "2026-08-22"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9,
  fig.height = 6,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

library(ggplot2)
source("code/revision_simulations/appendix_b/reporting.R")

appendix_b_cache <- load_appendix_b_reporting_cache()
grid_summary <- appendix_b_cache$grid_summary
focused_example <- appendix_b_cache$focused_example
focused_alpha <- appendix_b_cache$focused_alpha
grid_long <- prepare_appendix_b_grid_long(grid_summary)
alpha005 <- prepare_appendix_b_alpha005(focused_alpha)
alpha_long <- prepare_appendix_b_alpha_long(focused_alpha)
prior_table <- prepare_appendix_b_prior_table(focused_example)
trajectory_examples <- prepare_appendix_b_trajectory_examples(
  focused_example,
  examples_per_class = 3L
)
cumulative_lfdr <- do.call(rbind, lapply(c("IWP1", "IWP2"), function(order) {
  do.call(rbind, lapply(c("Raw EB", "BF updated"), function(stage) {
    prepare_appendix_b_cumulative_lfdr(
      focused_example,
      order = order,
      stage = stage,
      alpha = 0.05
    )
  }))
}))
cumulative_lfdr$order <- factor(
  cumulative_lfdr$order,
  levels = c("IWP1", "IWP2")
)
cumulative_lfdr$stage <- factor(
  cumulative_lfdr$stage,
  levels = c("Raw EB", "BF updated")
)

format_integer <- function(x) format(x, big.mark = ",", scientific = FALSE)
format_decimal <- function(x, digits = 3L) {
  formatC(x, format = "f", digits = digits)
}
format_percent <- function(x, digits = 1L) {
  paste0(format_decimal(100 * x, digits), "%")
}

stage_colors <- c("Raw EB" = "#0072B2", "BF updated" = "#D55E00")
class_colors <- c(
  "A: Constant" = "#666666",
  "B: Linear dynamic" = "#009E73",
  "C: Nonlinear dynamic" = "#CC79A7"
)

grid_mae <- aggregate(
  abs(grid_long$estimated_pi0 - grid_long$true_pi0),
  list(order = grid_long$order, stage = grid_long$stage),
  mean
)
names(grid_mae)[[3L]] <- "mean_absolute_error"
```

<style>
html,
body {
  max-width: 100%;
  overflow-x: hidden;
}
.r0-key-result {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.main-container .figure {
  max-width: 980px;
  margin: 1.4em auto;
  text-align: center;
}
.main-container .figure img {
  display: block;
  max-width: 100%;
  height: auto;
  margin: 0 auto;
}
.main-container table {
  width: auto;
  max-width: 100%;
}
.main-container table caption,
.main-container p.caption {
  color: #333333;
  font-weight: 600;
  text-align: left;
}
pre {
  box-sizing: border-box;
  max-width: 100%;
  overflow-x: auto;
}
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: hidden;
}
@media (max-width: 767px) {
  .main-container table {
    display: block;
    width: 100%;
    overflow-x: auto;
    -webkit-overflow-scrolling: touch;
  }
}
</style>

## Overview

This page evaluates FASH's exact-null-weight estimation and discovery behavior
in a controlled three-class simulation. It has been rebuilt from the completed
versioned cache generated with `fashr` `r appendix_b_cache$manifest$package_provenance$version`
at commit `r appendix_b_cache$manifest$package_provenance$remote_sha`. Routine
workflowr builds read this cache and do not refit FASH.

For unit $j$, the true effect belongs to one of three ordered classes:

- **A: constant**, which is null for both IWP$_1$ and IWP$_2$ testing;
- **B: linear dynamic**, which is non-null for IWP$_1$ but null for IWP$_2$;
- **C: nonlinear dynamic**, which is non-null for both tests.

Let

$$
\rho_{\mathrm{dynamic}} = P(B \cup C),
\qquad
\rho_{\mathrm{nonlinear}} = P(C).
$$

The exact-null proportions estimated by the two FASH orders are therefore

$$
\pi_{0,\mathrm{IWP1}} = 1-\rho_{\mathrm{dynamic}},
\qquad
\pi_{0,\mathrm{IWP2}} = 1-\rho_{\mathrm{nonlinear}}.
$$

<div class="r0-key-result">
<strong>Observed result.</strong> Across the 92 grid conditions, BF updating
removes all seven raw-IWP1 boundary collapses in which the maximum-likelihood
exact-null weight is zero. The mean absolute exact-null-weight error changes
from `r format_decimal(grid_mae$mean_absolute_error[grid_mae$order == "IWP1" & grid_mae$stage == "Raw EB"], 3)`
to `r format_decimal(grid_mae$mean_absolute_error[grid_mae$order == "IWP1" & grid_mae$stage == "BF updated"], 3)`
for IWP1, and from
`r format_decimal(grid_mae$mean_absolute_error[grid_mae$order == "IWP2" & grid_mae$stage == "Raw EB"], 3)`
to `r format_decimal(grid_mae$mean_absolute_error[grid_mae$order == "IWP2" & grid_mae$stage == "BF updated"], 3)`
for IWP2. Thus BF updating materially stabilizes IWP1 here, but it does not
improve every order or every realized dataset.
</div>

## Simulated effect classes

The retained focused replication contains `r format_integer(focused_example$parameters$J)`
units with standard errors sampled from `r paste(focused_example$parameters$sigma_vec, collapse = ", ")`.
Figure 1 shows the first three units in each truth class. Points are noisy
effect estimates; vertical bars span two reported standard errors; orange
curves are the true functions.

```{r representative-trajectories, fig.width=10, fig.height=8.4, out.width='96%', fig.cap="Figure 1. Representative simulated trajectories from the focused replication. Columns follow the fixed A/B/C truth definitions: constant, linear dynamic, and nonlinear dynamic. Points are estimated effects, vertical bars span two standard errors, and orange curves are the true effects."}
# Select exactly three retained units from each ordered truth class.
trajectory_plot_data <- trajectory_examples

ggplot(
  trajectory_plot_data,
  aes(x = time, y = estimate)
) +
  geom_errorbar(
    aes(
      ymin = estimate - 2 * standard_error,
      ymax = estimate + 2 * standard_error
    ),
    width = 0,
    color = "grey55",
    linewidth = 0.35
  ) +
  geom_point(size = 1.15, color = "black") +
  geom_line(aes(y = truth), color = "#D55E00", linewidth = 0.8) +
  facet_wrap(~ panel, ncol = 3, scales = "free_y") +
  labs(x = "Time", y = "Effect estimate") +
  theme_bw(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 8.5),
    axis.text = element_text(size = 8.5)
  )
```

## Simulation and fitting design

The grid experiment uses the same random-number stream for each grid setting.
For every $\rho_{\mathrm{dynamic}}$,
$\rho_{\mathrm{nonlinear}}=\rho_{\mathrm{dynamic}}/2$. This pairs the original
and denser-grid fits at the simulated-data level; it does not create repeated
Monte Carlo replicates.

```{r design-table, results='asis'}
# Record the fixed simulation, estimand, and fit choices shown in Table 1.
design_table <- data.frame(
  Item = c(
    "Package",
    "Grid experiment units",
    "Dynamic proportion",
    "Nonlinear proportion",
    "Predictive-SD grids",
    "Grid-fit penalty",
    "Focused replication",
    "Focused-fit penalty",
    "Basis and prediction step",
    "Random-number stream",
    "IWP1 exact-null estimand",
    "IWP2 exact-null estimand"
  ),
  Value = c(
    paste0(
      "fashr ", appendix_b_cache$manifest$package_provenance$version,
      " at ", appendix_b_cache$manifest$package_provenance$remote_sha
    ),
    format_integer(appendix_b_cache$manifest$grid_design$J),
    "0.05 to 0.50 by 0.01",
    "rho_dynamic / 2",
    "Original spacing 0.2; denser spacing 0.1; both include exact zero",
    format_integer(appendix_b_cache$manifest$grid_design$penalty),
    paste0(
      "J = ", format_integer(focused_example$parameters$J),
      ", rho_dynamic = 0.20, rho_nonlinear = 0.10"
    ),
    format_integer(focused_example$parameters$penalty),
    paste0(
      focused_example$parameters$num_basis,
      " basis functions; pred_step = 1"
    ),
    format_integer(focused_example$parameters$stream_seed),
    "P(A) = 1 - rho_dynamic",
    "P(A or B) = 1 - rho_nonlinear"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

knitr::kable(
  design_table,
  align = c("l", "l"),
  caption = paste(
    "Table 1. Fixed design and estimands for the versioned Appendix B",
    "simulation. The grid experiment contains one paired random-number",
    "stream, not repeated Monte Carlo replicates."
  )
)
```

## Exact-null-weight estimation across the grid

Figure 2 compares each fitted exact-null weight with its order-specific truth.
The dashed line is equality. Crosses mark the seven raw IWP1 fits whose exact
null weight reached the simplex boundary at zero; `fashr` explicitly warned
that their raw lfdr values were consequently zero. The BF-updated fits retain
the exact-null component and are shown as a separate stage.

```{r grid-null-weight, fig.width=10, fig.height=7.4, out.width='96%', fig.cap="Figure 2. Estimated versus true exact-null weight over 46 dynamic proportions and two predictive-SD grids. Panels separate IWP order and grid spacing; colors separate raw empirical-Bayes and BF-updated fits. The dashed line denotes equality, and crosses identify the seven valid raw-IWP1 boundary solutions with estimated exact-null weight zero."}
# Map each order to its own null estimand before plotting both fit stages.
grid_plot_data <- grid_long[
  order(grid_long$order, grid_long$setting, grid_long$stage, grid_long$true_pi0),
]

ggplot(
  grid_plot_data,
  aes(
    x = true_pi0,
    y = estimated_pi0,
    color = stage,
    group = stage,
    shape = boundary_warning
  )
) +
  geom_abline(
    intercept = 0,
    slope = 1,
    color = "grey35",
    linetype = "dashed",
    linewidth = 0.55
  ) +
  geom_line(linewidth = 0.55, alpha = 0.65) +
  geom_point(size = 1.8, stroke = 0.8) +
  facet_grid(order ~ setting) +
  scale_color_manual(values = stage_colors, drop = FALSE) +
  scale_shape_manual(
    values = c(`FALSE` = 16, `TRUE` = 4),
    labels = c(`FALSE` = "Regular fit", `TRUE` = "Raw null-weight boundary")
  ) +
  scale_x_continuous(limits = c(0.48, 1), breaks = seq(0.5, 1, by = 0.1)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
  labs(
    x = "True exact-null weight",
    y = "Estimated exact-null weight",
    color = "Fit stage",
    shape = "Fit status"
  ) +
  theme_bw(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## Focused replication

The focused replication fixes $J=1200$,
$\rho_{\mathrm{dynamic}}=0.2$, and
$\rho_{\mathrm{nonlinear}}=0.1$. Hence IWP1 has 240 true dynamic units and
IWP2 has 120 true nonlinear units. Calls are the longest prefix of lfdr-ranked
units whose mean lfdr does not exceed $\alpha$.

Table 2 reports realized false-discovery proportion (FDP) and power for this
single replication at $\alpha=0.05$. These realized quantities are not
repeated-simulation FDR or expected power.

```{r focused-alpha005-table, results='asis'}
# Restrict the retained alpha curve to the prespecified alpha = 0.05 rows.
alpha005_table <- data.frame(
  Order = as.character(alpha005$order),
  Target = alpha005$target,
  Stage = as.character(alpha005$stage),
  Discoveries = format_integer(alpha005$discoveries),
  `True discoveries` = format_integer(alpha005$true_discoveries),
  `False discoveries` = format_integer(alpha005$false_discoveries),
  `Realized FDP` = format_percent(alpha005$realized_fdp, 1),
  Power = format_percent(alpha005$power, 1),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

knitr::kable(
  alpha005_table,
  align = c("l", "l", "l", "r", "r", "r", "r", "r"),
  caption = paste(
    "Table 2. Focused-replication cumulative-lfdr calls at alpha = 0.05.",
    "FDP and power are realized values for this one retained replication."
  )
)
```

Figure 3 extends the same calculation across $\alpha=0.01,\ldots,0.20$.
The dashed diagonal in the FDP panels is the nominal reference, not a claim of
calibration from one replication.

```{r focused-alpha-curves, fig.width=9.6, fig.height=6.8, out.width='94%', fig.cap="Figure 3. Focused-replication realized FDP and power across cumulative-lfdr thresholds. Columns separate IWP1 dynamic testing and IWP2 nonlinear testing; colors separate raw and BF-updated fits. The dashed diagonal in the FDP row denotes nominal alpha. Curves summarize one retained replication, not Monte Carlo expectations."}
# Keep the realized-FDP and power estimands distinct before faceting.
alpha_plot_data <- alpha_long[
  order(alpha_long$metric, alpha_long$order, alpha_long$stage, alpha_long$alpha),
]
nominal_reference <- data.frame(
  metric = factor("Realized FDP", levels = levels(alpha_plot_data$metric)),
  order = factor(c("IWP1", "IWP2"), levels = levels(alpha_plot_data$order))
)

ggplot(
  alpha_plot_data,
  aes(x = alpha, y = value, color = stage, group = stage)
) +
  geom_abline(
    data = nominal_reference,
    aes(intercept = 0, slope = 1),
    inherit.aes = FALSE,
    color = "grey35",
    linetype = "dashed",
    linewidth = 0.55
  ) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.35) +
  facet_grid(metric ~ order, scales = "free_y") +
  scale_color_manual(values = stage_colors, drop = FALSE) +
  scale_x_continuous(breaks = seq(0, 0.2, by = 0.05)) +
  scale_y_continuous(labels = function(x) paste0(round(100 * x), "%")) +
  labs(
    x = expression(alpha),
    y = NULL,
    color = "Fit stage"
  ) +
  theme_bw(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

Figure 4 displays the actual lfdr-ranked prefixes underlying Table 2. A unit
is called at $\alpha=0.05$ while its cumulative mean lfdr remains below the
horizontal reference. The vertical dotted line is the resulting final call
count for that panel.

```{r focused-cumulative-lfdr, fig.width=10, fig.height=7.2, out.width='96%', fig.cap="Figure 4. Focused-replication cumulative lfdr after ranking units from smallest to largest lfdr. Panels separate order and fit stage; point colors show the fixed A/B/C truth classes. The horizontal dashed line is alpha = 0.05, and each vertical dotted line marks the final called rank for that panel."}
# Compute the final called prefix independently within each order and stage.
cumulative_plot_data <- cumulative_lfdr
call_cutoffs <- aggregate(
  called_at_alpha ~ order + stage,
  data = cumulative_plot_data,
  FUN = sum
)
names(call_cutoffs)[[3L]] <- "called_rank"

ggplot(
  cumulative_plot_data,
  aes(x = rank, y = cumulative_lfdr, color = class)
) +
  geom_hline(
    yintercept = 0.05,
    color = "grey35",
    linetype = "dashed",
    linewidth = 0.55
  ) +
  geom_vline(
    data = call_cutoffs,
    aes(xintercept = called_rank),
    inherit.aes = FALSE,
    color = "grey35",
    linetype = "dotted",
    linewidth = 0.55
  ) +
  geom_point(size = 0.7, alpha = 0.72) +
  facet_grid(order ~ stage) +
  scale_color_manual(values = class_colors, drop = FALSE) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
  labs(
    x = "lfdr rank",
    y = "Cumulative mean lfdr",
    color = "Truth class"
  ) +
  theme_bw(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

The exact-null weights for the same focused replication are reported in Table
3. BF updating shifts both weights upward. In this one realization that
reduces realized FDP with a small decrease in power, but it does not make both
point estimates closer to their true values.

```{r focused-prior-table, results='asis'}
# Compare each fitted exact-null component with its order-specific truth.
prior_display <- data.frame(
  Order = as.character(prior_table$order),
  Target = prior_table$target,
  Stage = as.character(prior_table$stage),
  `True pi0` = format_decimal(prior_table$true_pi0, 3),
  `Estimated pi0` = format_decimal(prior_table$estimated_pi0, 3),
  `Estimate minus truth` = format_decimal(prior_table$estimation_error, 3),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

knitr::kable(
  prior_display,
  align = c("l", "l", "l", "r", "r", "r"),
  caption = paste(
    "Table 3. Focused-replication exact-null weights before and after BF",
    "updating. The estimand differs by IWP order as defined above."
  )
)
```

## Provenance and reproducibility

The retained family completed at
`r appendix_b_cache$completion[["completed_at"]]` on Midway3. Its manifest
records R `r appendix_b_cache$manifest$package_provenance$r_version` on
`r appendix_b_cache$manifest$package_provenance$platform`. The local reporting
loader verifies the package version and commit, helper and runner SHA-256
values, all retained output SHA-256 values, the 92-row grid contract, the
80-row focused alpha curve, and the seven explicitly identified raw-IWP1
boundary warnings. All older `data/appendixB/*.RData` files remain separate
and unchanged.

The expensive server run used the following entry point. This chunk is shown
for reproducibility and is not evaluated by workflowr.

```{bash appendix-b-server-run, eval=FALSE}
cd /project/mstephens/ziangzhang/fash/workspace
sbatch slurm/10_appendix_b_simulation.sbatch
```

After downloading the completed result family, validate its package, source,
output, historical-cache, and numerical contracts with:

```{bash appendix-b-cache-validation, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/appendix_b/validate_appendix_b_cache.R
```
