---
title: "Simulation: prior misspecification and BF adjustment"
author: "Ziang Zhang"
date: "2026-09-18"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10.4, fig.height = 5.2,
  dpi = 180, fig.retina = 2, fig.align = "center", message = FALSE, warning = TRUE)
library(ggplot2)
source("code/revision_simulations/internal/appendix_c1_mc10/reporting.R")
report <- c1_load_report()
display <- c1_load_display(report)
grid_summary <- report$tables$grid_summary
alpha_summary <- report$tables$alpha_summary
at005 <- alpha_summary[abs(alpha_summary$alpha - 0.05) < 1e-12, ]
value005 <- function(order, stage, metric) {
  row <- at005[at005$order == order & at005$stage == stage, ]
  stopifnot(nrow(row) == 1L)
  sprintf("%.3f", row[[metric]])
}
grid_results <- report$tables$grid_long
raw_zero <- sum(grid_results$stage == "Raw EB" & grid_results$order == "IWP1" & grid_results$pi0 == 0)
bf_grid <- grid_results[grid_results$stage == "BF updated", ]
bf_below <- sum(bf_grid$pi0 < bf_grid$true_pi0)
```

<style>
p.caption, table caption { font-weight: 600; color: #333; }
.main-container p code { white-space: normal; overflow-wrap: anywhere; }
.MathJax_Display { max-width: 100%; overflow-x: auto; overflow-y: hidden; }
.c1-table { width: 100%; overflow-x: auto; }
</style>

## Overview

This experiment evaluates null-weight estimation, discovery power, and empirical
FDR when constant, linear, and nonlinear effect functions are fitted with FASH.

1. Generate ten independent repetitions of each simulation setting.
2. Fit FASH-IWP1 and FASH-IWP2 before and after BF adjustment, pairing the
   simulated inputs across the original and denser shrinkage grids.
3. Average null-weight estimates, power, and false-discovery proportions across
   repetitions. Figure 2 shows pointwise replicate ranges; Figures 3 and 4
   show means as points connected by lines.

All fits use the pinned `fashr` 0.1.43 build. The ten seeds are
`seq(12345L, by = 10000L, length.out = 10L)`.

## Data generation and fitting

At sixteen observed times $t=1,\ldots,16$, each unit belongs to one of three classes:

- Constant: $\beta_j(t)=c_j$, with $c_j\sim N(0,1)$.
- Linear dynamic: $\beta_j(t)=c_j+b_jt$, with independent
  $c_j\sim N(0,1)$ and $b_j\sim N(0,1/16^2)$.
- Nonlinear dynamic: an IWP2 function with predictive standard deviation 5
  over sixteen time units.

Observed effects are independent Gaussian measurements of these functions,
with standard errors drawn from $\{0.1,0.3,0.5\}$ with equal probabilities.
The grid experiment has $J=1000$ units per dataset. The dynamic fraction
$\rho$ ranges from 0.05 to 0.50 in increments of 0.01; half of the dynamic
units are linear and half nonlinear. The exact-null proportions are
$1-\rho$ for IWP1 and $1-\rho/2$ for IWP2.

Both models use an unrestricted polynomial baseline and 20 basis functions.
The positive shrinkage grid is equally spaced from 0 to 10 on the
log-precision scale, with spacing 0.2 or 0.1, together with an exact-null
component. Null-weight comparisons use penalty 1. The power/FDR experiment
uses $J=1200$, $\rho=0.2$, the original grid, and penalty 10.

```{r scientific-settings, eval=FALSE}
seeds <- seq(12345L, by = 10000L, length.out = 10L)
rho <- seq(0.05, 0.50, by = 0.01)
original_grid <- sort(c(0, exp(-0.5 * seq(0, 10, by = 0.2))))
denser_grid <- sort(c(0, exp(-0.5 * seq(0, 10, by = 0.1))))
grid_data <- build_appendix_b_datasets(J = 1000L,
  rho_dynamic = rho_value, rho_nonlinear = rho_value / 2,
  sigma_vec = c(0.1, 0.3, 0.5))
focused_data <- build_appendix_b_datasets(J = 1200L,
  rho_dynamic = 0.2, rho_nonlinear = 0.1,
  sigma_vec = c(0.1, 0.3, 0.5))
# Exact Gaussian integration uses the package's basis and prior precision.
# Package EB optimization and BF updating are unchanged.
focused_fit <- c1_fit_models(focused_data, original_grid,
  penalty = 10, num_basis = 20L, num_cores = 1L)
```

```{r trajectories, fig.width=10, fig.height=8.4, fig.cap="Figure 1. Simulated observations and true effects from the first prespecified repetition. Columns show constant, linear dynamic, and nonlinear dynamic truth. Black points and gray bars show observed estimates and +/-2 standard errors; orange curves show the true effects."}
# Choose the first three units per truth class, independently of discoveries.
truth <- report$focused$data_bundle$truth
chosen <- unlist(lapply(c("nondynamic", "linear", "nonlinear"), function(class) {
  head(which(truth$class == class), 3L)
}))
example_data <- do.call(rbind, lapply(chosen, function(index) {
  cbind(unit_id = truth$unit_id[index], class = truth$class[index],
        report$focused$data_bundle$data_list[[index]])
}))
panel_ids <- as.vector(t(matrix(truth$unit_id[chosen], nrow = 3)))
example_data$panel <- factor(example_data$unit_id, levels = panel_ids)
ggplot(example_data, aes(x, y)) +
  geom_errorbar(aes(ymin = y - 2 * sd, ymax = y + 2 * sd), width = 0,
    color = "grey55", linewidth = 0.35) + geom_point(size = 1.15) +
  geom_line(aes(y = truef), color = "#D55E00", linewidth = 0.8) +
  facet_wrap(~panel, ncol = 3, scales = "free_y") +
  labs(x = "Time", y = "Effect") + c1_plot_theme()
```

## Null-weight estimation

There are 92 combinations of dynamic fraction and shrinkage grid, each with
ten repetitions and both FASH orders. Raw IWP1 has `r raw_zero` zero-boundary
null-weight estimates among its 920 fits. BF-adjusted estimates fall below the
true null proportion in `r bf_below` of the 1,840 order-specific fits.

```{r null-weights, fig.width=10.4, fig.height=8, fig.cap="Figure 2. Mean estimated null weight over ten repetitions per setting. Columns show IWP1 and IWP2; rows show the original and denser grids. Black denotes Raw EB, blue BF adjustment, and the red dashed line equality. Ribbons show pointwise replicate minima and maxima, including zero-boundary estimates."}
null_curves <- grid_summary[, c("setting", "rho_dynamic", "order", "stage",
  "true_pi0", "mean_pi0", "pi0_min", "pi0_max", "n_replicates")]
stopifnot(all(null_curves$n_replicates == 10L))
null_curves$stage <- factor(null_curves$stage, c("Raw EB", "BF updated"))
null_curves$panel <- factor(paste(null_curves$setting, null_curves$order, sep = ":"),
  c("original:IWP1", "original:IWP2", "denser:IWP1", "denser:IWP2"),
  c("A  IWP1: original grid", "B  IWP2: original grid",
    "C  IWP1: denser grid", "D  IWP2: denser grid"))
stopifnot(!anyNA(null_curves$panel))
ggplot(null_curves, aes(true_pi0, mean_pi0, color = stage, fill = stage)) +
  geom_ribbon(aes(ymin = pi0_min, ymax = pi0_max), alpha = 0.10,
    color = NA, show.legend = FALSE) + geom_line(linewidth = 0.7) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", linewidth = 0.45) +
  facet_wrap(~panel, ncol = 2, scales = "free_x") +
  scale_color_manual(values = c("Raw EB" = "black", "BF updated" = "#0072B2")) +
  scale_fill_manual(values = c("Raw EB" = "black", "BF updated" = "#0072B2")) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = expression("True null proportion " * pi[0]),
    y = expression("Mean estimated null proportion " * hat(pi)[0]), color = NULL, fill = NULL) +
  c1_plot_theme()
```

## Discovery power and empirical FDR

For each repetition $b$, discoveries $D_b(\alpha)$ are selected by cumulative
mean lfdr. With true null and alternative sets $H_{0b}$ and $H_{1b}$, the
plotted empirical FDR is

$$
\widehat{\mathrm{FDR}}(\alpha)=\frac{1}{10}\sum_{b=1}^{10}
\frac{|D_b(\alpha)\cap H_{0b}|}{\max\{|D_b(\alpha)|,1\}}.
$$

Power is the mean of $|D_b(\alpha)\cap H_{1b}|/|H_{1b}|$ across the same
ten repetitions. Both rates include every repetition. Points connected by
lines show the ten-repetition means.

```{r power, fig.cap="Figure 3. Mean discovery power over ten repetitions (J = 1,200). Panels A and B show dynamic and nonlinear testing. Points connected by lines show ten-repetition mean power. The vertical line marks nominal FDR 0.05."}
power_curves <- alpha_summary[, c("order", "stage", "alpha", "mean_power",
  "power_min", "power_max", "n_replicates")]
stopifnot(all(power_curves$n_replicates == 10L))
power_curves$stage <- factor(power_curves$stage, c("Raw EB", "BF updated"))
power_curves$panel <- factor(power_curves$order, c("IWP1", "IWP2"),
  c("A  Dynamic effects", "B  Nonlinear effects"))
ggplot(power_curves, aes(alpha, mean_power, color = stage)) +
  geom_line(linewidth = 0.65) + geom_point(size = 1.4) +
  geom_vline(xintercept = 0.05, color = "grey50", linetype = "dashed", linewidth = 0.4) +
  facet_wrap(~panel, nrow = 1) +
  scale_color_manual(values = c("Raw EB" = "#0072B2", "BF updated" = "#D55E00")) +
  scale_y_continuous(breaks = seq(0.75, 0.90, by = 0.05)) +
  coord_cartesian(ylim = c(0.75, 0.90), expand = FALSE) +
  labs(x = "Nominal FDR level", y = "Mean power", color = NULL, fill = NULL) + c1_plot_theme()
```

```{r empirical-fdr, fig.cap="Figure 4. Empirical FDR over the same ten repetitions. Panels A and B show dynamic and nonlinear testing. Points connected by lines show the mean of replicate false-discovery proportions. The diagonal marks nominal calibration and the vertical line nominal FDR 0.05."}
fdr_curves <- alpha_summary[, c("order", "stage", "alpha", "mean_empirical_fdr",
  "empirical_fdr_min", "empirical_fdr_max", "n_replicates")]
stopifnot(all(fdr_curves$n_replicates == 10L))
fdr_curves$stage <- factor(fdr_curves$stage, c("Raw EB", "BF updated"))
fdr_curves$panel <- factor(fdr_curves$order, c("IWP1", "IWP2"),
  c("A  Dynamic effects", "B  Nonlinear effects"))
ggplot(fdr_curves, aes(alpha, mean_empirical_fdr, color = stage)) +
  geom_line(linewidth = 0.65) + geom_point(size = 1.4) +
  geom_abline(slope = 1, intercept = 0, color = "grey35", linetype = "dotted", linewidth = 0.4) +
  geom_vline(xintercept = 0.05, color = "grey50", linetype = "dashed", linewidth = 0.4) +
  facet_wrap(~panel, nrow = 1) +
  scale_color_manual(values = c("Raw EB" = "#0072B2", "BF updated" = "#D55E00")) +
  coord_cartesian(ylim = c(0, max(0.2, 1.05 * fdr_curves$empirical_fdr_max))) +
  labs(x = "Nominal FDR level", y = "Empirical FDR", color = NULL, fill = NULL) + c1_plot_theme()
```

At nominal FDR 0.05, BF adjustment changes empirical FDR from
`r value005("IWP1", "Raw EB", "mean_empirical_fdr")` to
`r value005("IWP1", "BF updated", "mean_empirical_fdr")` for dynamic testing,
and from `r value005("IWP2", "Raw EB", "mean_empirical_fdr")` to
`r value005("IWP2", "BF updated", "mean_empirical_fdr")` for nonlinear testing.
The corresponding power changes are
`r value005("IWP1", "Raw EB", "mean_power")` to
`r value005("IWP1", "BF updated", "mean_power")` and
`r value005("IWP2", "Raw EB", "mean_power")` to
`r value005("IWP2", "BF updated", "mean_power")`.

```{r alpha005-table, results='asis'}
tab <- at005[order(match(at005$order, c("IWP1", "IWP2")),
                  match(at005$stage, c("Raw EB", "BF updated"))),
  c("order", "stage", "mean_discoveries", "mean_power", "mean_empirical_fdr")]
names(tab) <- c("Model", "Prior", "Mean discoveries", "Mean power", "Empirical FDR")
knitr::asis_output(paste0('<div class="c1-table">',
  knitr::kable(tab, format = "html", row.names = FALSE, digits = c(0, 0, 1, 3, 3),
    caption = "Table 1. Ten-repetition mean results at nominal FDR 0.05."), '</div>'))
```

## Posterior examples

Figure 5 retains four prespecified examples from seed 12345. All four are
selected at nominal FDR 0.05 by their respective BF-adjusted models.
Posterior summaries use 3,000 draws evaluated at time-grid step 0.1.

```{r posterior-examples, fig.width=10.4, fig.height=7.8, fig.cap="Figure 5. Posterior examples from seed 12345. Top: IWP1 dynamic testing; bottom: IWP2 nonlinear testing. Black points and bars show estimates and +/-2 standard errors. Red curves and bands show posterior means and pointwise 95% credible intervals; blue dashed curves show truth."}
predictions <- display$predictions; observations <- display$observations
labels <- c("A  C109 (IWP1)", "B  B75 (IWP1)", "C  C24 (IWP2)", "D  C115 (IWP2)")
predictions$panel <- factor(predictions$panel, LETTERS[1:4], labels)
observations$panel <- factor(observations$panel, LETTERS[1:4], labels)
stopifnot(!anyNA(predictions$panel), !anyNA(observations$panel))
ggplot(predictions, aes(x, mean)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "red", alpha = 0.10) +
  geom_line(color = "red", linewidth = 0.8) +
  geom_errorbar(data = observations, aes(x = x, ymin = y - 2 * sd, ymax = y + 2 * sd),
    inherit.aes = FALSE, width = 0, linewidth = 0.35) +
  geom_point(data = observations, aes(x, y), inherit.aes = FALSE, size = 1.1) +
  geom_line(data = observations, aes(x, truef), inherit.aes = FALSE,
    color = "blue", linetype = "dashed", linewidth = 0.6) +
  facet_wrap(~panel, ncol = 2, scales = "free_y") +
  labs(x = "Time", y = "Effect") + c1_plot_theme()
```

## Reproducibility

The runner, equivalence tests, and reporting gate are in
`code/revision_simulations/internal/appendix_c1_mc10/`.
The original generator and pinned package EB/BF routines are retained.
Exact Gaussian integration uses the package's basis and precision matrices;
likelihood and posterior-score equivalence with TMB are checked before
production. Original and denser grids have identical input hashes within
each seed and dynamic fraction.

```{bash, eval=FALSE}
Rscript --vanilla code/revision_simulations/internal/appendix_c1_mc10/test_mc10.R
FASH_C1_WORKERS=4 Rscript --vanilla code/revision_simulations/internal/appendix_c1_mc10/run_appendix_c1_mc10.R
Rscript --vanilla code/revision_simulations/internal/appendix_c1_mc10/build_display_cache.R
```

Completed results are in
`output/revision_simulations/internal/appendix_c1_mc10_fashr0143_20260918/`.
Page builds validate and read this cache without rerunning simulations.
