---
title: "Example: Switch Marginal Effect Due to LD Tagging"
author: "Ziang Zhang"
date: "2026-07-31"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 11,
  fig.height = 4.8,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

## Question

Suppose the marginal effect estimated at a lead or tag SNP changes sign over
time. Does this imply that one shared causal variant changes its direction of
effect?

Not necessarily. A non-causal tag SNP can be correlated with distinct causal
variants that are active at different times. If the signed linkage
disequilibrium (LD) between the tag SNP and the two causal variants has
opposite directions, the marginal effect at the tag SNP can appear to switch
sign even though neither causal effect switches.

## Statistical mechanism

Let $A$ and $B$ be causal variants and let $C$ be a non-causal tag SNP. At day
$t$, generate expression as

$$
Y(t) = \beta_A(t)G_A + \beta_B(t)G_B + \epsilon(t),
$$

where $\beta_A(t)$ is a positive early effect and $\beta_B(t)$ is a positive
late effect. Regressing $Y(t)$ on $G_C$ alone gives the population marginal
coefficient

$$
\beta_C^{\mathrm{marg}}(t)
=
\frac{\operatorname{Cov}(G_C,G_A)}{\operatorname{Var}(G_C)}\beta_A(t)
+
\frac{\operatorname{Cov}(G_C,G_B)}{\operatorname{Var}(G_C)}\beta_B(t).
$$

Thus, positive signed LD between $C$ and $A$ and negative signed LD between
$C$ and $B$ can produce a positive marginal effect early and a negative
marginal effect late. The sign of LD depends on allele alignment; here it is
represented by the signed genotype correlation.

## Sixteen-day example

The simulation uses `10000` individuals to make the population mechanism
visually clear. It is a proof-of-possibility example rather than a model of the
sample size or uncertainty in the iPSC study.

We generate two haplotypes per individual. Variants $A$ and $B$ are
approximately independent. The allele probability for $C$ is constructed so
that $C$ is positively correlated with $A$ and negatively correlated with $B$.

```{r simulate-genotypes}
set.seed(20260731)
n <- 10000L
days <- 0:15

h_a <- rbinom(2L * n, size = 1L, prob = 0.5)
h_b <- rbinom(2L * n, size = 1L, prob = 0.5)
p_c <- 0.5 + 0.45 * (h_a - h_b)
h_c <- rbinom(2L * n, size = 1L, prob = p_c)

combine_haplotypes <- function(x) {
  x[seq_len(n)] + x[n + seq_len(n)]
}

genotype <- data.frame(
  A = combine_haplotypes(h_a),
  B = combine_haplotypes(h_b),
  C = combine_haplotypes(h_c)
)
```

```{r show-ld-matrix}
knitr::kable(
  round(cor(genotype), 3),
  caption = "Signed genotype correlations among the three variants."
)
```

Variant $A$ has a smooth positive effect concentrated early in the time
course, while variant $B$ has a smooth positive effect concentrated late.
Variant $C$ has no direct causal effect at any day.

```{r simulate-expression}
true_effect_a <- exp(-0.5 * ((days - 2) / 2.5)^2)
true_effect_b <- exp(-0.5 * ((days - 13) / 2.5)^2)

expression <- outer(genotype$A, true_effect_a) +
  outer(genotype$B, true_effect_b) +
  matrix(rnorm(n * length(days)), nrow = n)
```

At each day, we first estimate the marginal association of $C$ by regressing
expression on $C$ alone. We then estimate the coefficient of $C$ conditional
on both causal variants.

```{r estimate-tag-effects}
fit_c_effect <- function(day_index, conditional = FALSE) {
  response <- expression[, day_index]
  fit <- if (conditional) {
    lm(response ~ C + A + B, data = genotype)
  } else {
    lm(response ~ C, data = genotype)
  }
  c(
    estimate = coef(fit)[["C"]],
    standard_error = sqrt(vcov(fit)["C", "C"])
  )
}

marginal_fit_c <- t(vapply(seq_along(days), fit_c_effect, numeric(2)))
conditional_fit_c <- t(vapply(
  seq_along(days), fit_c_effect, numeric(2), conditional = TRUE
))

marginal_effect_c <- marginal_fit_c[, "estimate"]
marginal_se_c <- marginal_fit_c[, "standard_error"]
conditional_effect_c <- conditional_fit_c[, "estimate"]
conditional_se_c <- conditional_fit_c[, "standard_error"]

result <- data.frame(
  day = days,
  true_effect_a = true_effect_a,
  true_effect_b = true_effect_b,
  marginal_effect_c = marginal_effect_c,
  marginal_se_c = marginal_se_c,
  conditional_effect_c = conditional_effect_c,
  conditional_se_c = conditional_se_c
)

stopifnot(
  marginal_effect_c[which.max(true_effect_a)] > 0,
  marginal_effect_c[which.max(true_effect_b)] < 0,
  max(abs(conditional_effect_c)) < 0.1,
  all(marginal_se_c > 0),
  all(conditional_se_c > 0)
)
```

```{r plot-ld-tagging-switch, fig.cap="Distinct positive early and late causal effects can induce an apparent sign switch at a non-causal tag SNP. Error bars show the estimate plus or minus one standard error. Conditioning on both causal variants removes the apparent effect at the tag SNP."}
variant_colors <- c(C = "#2a78d6", A = "#eb6834", B = "#1baf7a")

par(mfrow = c(1, 2), mar = c(4.2, 4.4, 2.5, 1.0))

plot(days, true_effect_a,
  type = "l", lwd = 3, col = variant_colors[["A"]],
  ylim = c(0, 1.05), xlab = "Day", ylab = "True causal effect",
  main = "Distinct early and late causal variants"
)
lines(days, true_effect_b, lwd = 3, lty = 2, col = variant_colors[["B"]])
lines(days, rep(0, length(days)), lwd = 2, lty = 3, col = "#777777")
legend("topright",
  legend = c("A: early causal", "B: late causal", "C: no direct effect"),
  col = c(variant_colors[["A"]], variant_colors[["B"]], "#777777"),
  lty = c(1, 2, 3), lwd = c(3, 3, 2), bty = "n"
)

y_limit <- 1.15 * max(abs(c(
  marginal_effect_c - marginal_se_c,
  marginal_effect_c + marginal_se_c,
  conditional_effect_c - conditional_se_c,
  conditional_effect_c + conditional_se_c
)))
plot(days, marginal_effect_c,
  type = "n",
  ylim = c(-y_limit, y_limit), xlab = "Day", ylab = "Effect estimate for C",
  main = "Tag SNP C appears to switch"
)
abline(h = 0, col = "#c3c2b7")
arrows(
  days, marginal_effect_c - marginal_se_c,
  days, marginal_effect_c + marginal_se_c,
  angle = 90, code = 3, length = 0.035, lwd = 1.5,
  col = adjustcolor(variant_colors[["C"]], alpha.f = 0.75)
)
arrows(
  days, conditional_effect_c - conditional_se_c,
  days, conditional_effect_c + conditional_se_c,
  angle = 90, code = 3, length = 0.035, lwd = 1.25,
  col = adjustcolor("#777777", alpha.f = 0.75)
)
lines(days, marginal_effect_c, lwd = 3, col = variant_colors[["C"]])
points(days, marginal_effect_c, pch = 16, col = variant_colors[["C"]])
lines(days, conditional_effect_c, lwd = 2, lty = 2, col = "#777777")
points(days, conditional_effect_c, pch = 1, col = "#777777")
legend("topright",
  legend = c("Marginal C", "C conditional on A and B"),
  col = c(variant_colors[["C"]], "#777777"),
  pch = c(16, 1), lty = c(1, 2), lwd = c(3, 2), bty = "n"
)
```

The marginal coefficient of $C$ is
`r sprintf("%.3f", marginal_effect_c[days == 2])` at the peak of the early
effect and `r sprintf("%.3f", marginal_effect_c[days == 13])` at the peak of
the late effect. After conditioning on $A$ and $B$, the maximum absolute
coefficient of $C$ across all 16 days is only
`r sprintf("%.3f", max(abs(conditional_effect_c)))`.
Error bars in the right panel show the corresponding estimate plus or minus
one regression standard error.

```{r show-selected-days}
selected_result <- result[result$day %in% c(2, 7, 8, 13), ]
names(selected_result) <- c(
  "Day",
  "True effect A",
  "True effect B",
  "Marginal effect C",
  "Marginal SE C",
  "Conditional effect C",
  "Conditional SE C"
)

knitr::kable(
  selected_result,
  digits = 3,
  row.names = FALSE,
  caption = "Selected days around the two causal peaks and the apparent switch."
)
```

## Interpretation

This example shows that a sign-changing marginal effect at a lead or tag SNP
does not by itself establish that one shared causal variant reverses its
biological effect. The same observed pattern can arise when distinct causal
variants act at different times and are tagged with opposite signed LD.

The example establishes only that this mechanism is possible. Distinguishing
it from a genuinely shared, sign-changing causal effect in real data requires
locus-level conditional analysis or multi-condition fine-mapping.
