---
title: "Internal R3 diagnostic: ideal-Gaussian effect estimates with fixed adjusted SE"
author: "Ziang Zhang"
date: "2026-08-27"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 10,
  fig.height = 8,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
project_root <- if (file.exists(
  "code/revision_simulations/shared/simulation_functions.R"
)) {
  "."
} else if (file.exists(
  "../code/revision_simulations/shared/simulation_functions.R"
)) {
  ".."
} else {
  stop("Could not locate the workflowr project root.")
}
source(file.path(
  project_root,
  "code", "revision_simulations", "shared", "simulation_functions.R"
))
source(file.path(
  project_root,
  "code", "revision_simulations", "shared", "real_genotype_one_per_gene.R"
))
source(file.path(
  project_root,
  "code", "revision_simulations", "internal",
  "r3_ideal_gaussian_measurement", "reporting.R"
))
```

This is an internal analysis record and is not linked from the workflowr index.

## Question

The broad random B-spline simulation in R3 shows increasing Middle empirical
false sign rate (FSR) above the nominal level after approximately
$\alpha=0.05$. The current R3 observation pipeline simulates individual-level
expression, fits a separate genotype regression at each time point, and maps
the resulting t-based standard errors onto a normal scale. Does the same
pattern remain when the summary statistics follow the Gaussian likelihood used
by FASH exactly?

## Fixed comparison

The comparison changes one layer only:

1. Regenerate the current R3 genotype sample, covariates, true effect curves,
   individual-level expression, regressions, and final t-adjusted standard
   errors from the same frozen component seeds.
2. Treat the regenerated final adjusted standard-error matrix as fixed and
   known.
3. Draw a new effect estimate directly from the ideal Gaussian observation
   model below, using a separate deterministic seed and iid errors across units
   and time points.
4. Apply the same IWP1 Raw and BF-adjusted fits, common posterior samples,
   full-universe functional testing, and alpha grid used by the current R3 page.

Both truth mechanisms use five seeds and 6,362 units per seed. Broad random
B-spline truth is the primary diagnostic; compact raised-cosine truth is the
same-design control. The truth definitions, class counts, relative clearance
margins, open Middle interval $3<t<12$, and all FASH settings are unchanged.
Early/Middle/Late probabilities are `0.29/0.42/0.29`, giving exact location
counts `369/534/369`; the six location-by-switch cells contain
`185/184/267/267/185/184` dynamic units.

```{r fixed-design-table}
fixed_design <- data.frame(
  Feature = c(
    "Units per seed",
    "Seeds",
    "Truth mechanisms",
    "Early/Middle/Late counts",
    "Middle definition",
    "Candidate universe",
    "FASH version",
    "Adjusted-SE role"
  ),
  Value = c(
    format(ideal_configuration$J, big.mark = ","),
    paste(ideal_configuration$seed_list, collapse = ", "),
    "Broad random B-spline; compact raised cosine",
    "369 / 534 / 369",
    ideal_configuration$middle_expression,
    "All simulated units",
    paste0(
      ideal_manifest$package_provenance$version,
      " at ",
      substr(ideal_manifest$package_provenance$remote_sha, 1, 12)
    ),
    "Regenerated from current R3, then treated as fixed"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(
  fixed_design,
  align = "ll",
  caption = "Table 1. Inputs held fixed in the observation-model isolation experiment."
)
```

## Observation models

For the current R3 pipeline, let
$\widehat\beta^{\mathrm{reg}}_{jt}$ and $s^{\mathrm{adj}}_{jt}$ denote the
per-time regression estimate and its final t-adjusted standard error. Its
standardized estimation error is

$$
Z^{\mathrm{reg}}_{jt}
=
\frac{\widehat\beta^{\mathrm{reg}}_{jt}-\beta^{\mathrm{true}}_{jt}}
     {s^{\mathrm{adj}}_{jt}}.
$$

The ideal experiment conditions on the same regenerated
$s^{\mathrm{adj}}_{jt}$ and draws

$$
\widehat\beta^{\mathrm{ideal}}_{jt}
=
\beta^{\mathrm{true}}_{jt}
+s^{\mathrm{adj}}_{jt}Z^{\mathrm{ideal}}_{jt},
\qquad
Z^{\mathrm{ideal}}_{jt}\overset{\mathrm{iid}}{\sim}N(0,1).
$$

No second t-based correction is applied to
$\widehat\beta^{\mathrm{ideal}}$. This is therefore an oracle experiment
conditional on the realized adjusted-SE matrix. It isolates the summary-level
observation likelihood; it does not remove the mismatch between broad
B-spline truth and the IWP1 working prior, and it does not validate the
estimated-SE procedure itself.

## Standardized-error diagnostics

```{r standardized-error-summary}
diagnostic_display <- standardized_error_summary
diagnostic_display$truth_mechanism <- NULL
for (column in c(
  "mean", "sd", "skewness", "excess_kurtosis", "tail_abs_1_96",
  "tail_abs_3"
)) {
  diagnostic_display[[column]] <- formatC(
    diagnostic_display[[column]], format = "f", digits = 4
  )
}
names(diagnostic_display) <- c(
  "Truth mechanism",
  "Observation model",
  "Mean",
  "SD",
  "Skewness",
  "Excess kurtosis",
  "Pr(|Z| > 1.96)",
  "Pr(|Z| > 3)"
)
knitr::kable(
  diagnostic_display,
  align = "llrrrrrr",
  caption = paste(
    "Table 2. Standardized estimation-error summaries.",
    "Each entry is the mean of the corresponding seed-level statistic."
  )
)
```

```{r standardized-error-histogram, fig.height=7, fig.cap="Figure 1. Pooled standardized estimation-error distributions under the regression-derived and ideal-Gaussian observation models. Columns identify observation models, rows identify truth mechanisms, and the orange curve is the standard-normal density."}
plot_standardized_error_histogram()
```

```{r standardized-error-lag-correlation, fig.height=7, fig.cap="Figure 2. Lag-specific correlations of standardized estimation errors. Curves are means over five seeds and over all time-point pairs at each lag."}
plot_standardized_error_lag_correlation()
```

These diagnostics verify what the intervention changes. The ideal arm has
standard-normal marginal errors and no imposed time correlation by
construction; the regenerated regression arm records the corresponding errors
after the full current R3 regression and t-adjustment pipeline.

## Functional FSR comparison

The primary statistic is the maximum, over
$\alpha\in\{0.05,0.055,\ldots,0.20\}$, of the five-seed mean BF-adjusted
Middle empirical FSR minus nominal alpha. The comparison was fixed before the
ideal-Gaussian results were observed; no post hoc pass/fail threshold is used.

```{r primary-middle-summary}
primary_display <- primary_comparison[c(
  "truth_mechanism_label",
  "maximum_mean_fsr_excess_current",
  "alpha_at_maximum_current",
  "maximum_mean_fsr_excess_ideal",
  "alpha_at_maximum_ideal",
  "current_minus_ideal_excess"
)]
for (column in setdiff(names(primary_display), "truth_mechanism_label")) {
  primary_display[[column]] <- formatC(
    primary_display[[column]], format = "f", digits = 4
  )
}
names(primary_display) <- c(
  "Truth mechanism",
  "Current maximum excess",
  "Current alpha",
  "Ideal maximum excess",
  "Ideal alpha",
  "Current minus ideal excess"
)
knitr::kable(
  primary_display,
  align = "lrrrrr",
  caption = paste(
    "Table 3. Primary BF-adjusted Middle calibration comparison.",
    "Positive current-minus-ideal excess indicates attenuation under the",
    "ideal observation model."
  )
)
```

### Broad random B-spline truth

```{r bspline-fsr-comparison, fig.height=8, fig.cap="Figure 3. BF-adjusted functional empirical FSR under broad random B-spline truth. Curves compare five-seed means from the current regression-derived and ideal-Gaussian observation models; the dotted diagonal is nominal alpha."}
plot_fsr_comparison("random_bspline")
```

### Compact raised-cosine truth

```{r raised-cosine-fsr-comparison, fig.height=8, fig.cap="Figure 4. BF-adjusted functional empirical FSR under compact raised-cosine truth. Curves compare five-seed means from the current regression-derived and ideal-Gaussian observation models; the dotted diagonal is nominal alpha."}
plot_fsr_comparison("raised_cosine")
```

In this five-seed run, idealizing the observation model does not attenuate the
broad random B-spline Middle deviation: its maximum mean FSR excess is `0.0638`
under ideal-Gaussian measurements and `0.0582` under the current regression
pipeline. For compact raised-cosine truth, the corresponding values are
`0.0021` and `-0.0101`. The regression and t-adjustment pipeline is therefore
not a sufficient explanation for the broad random B-spline pattern.

```{r selected-alpha-middle-table}
middle_display <- middle_selected_summary
middle_display$truth_mechanism <- NULL
for (column in c(
  "alpha", "mean_empirical_fsr", "minimum_seed_fsr", "maximum_seed_fsr",
  "mean_power"
)) {
  middle_display[[column]] <- formatC(
    middle_display[[column]], format = "f", digits = 3
  )
}
for (column in c("mean_calls", "mean_false_calls")) {
  middle_display[[column]] <- formatC(
    middle_display[[column]], format = "f", digits = 1
  )
}
names(middle_display) <- c(
  "Truth mechanism",
  "Observation model",
  "Alpha",
  "Mean FSR",
  "Minimum seed FSR",
  "Maximum seed FSR",
  "Mean power",
  "Mean calls",
  "Mean false calls"
)
knitr::kable(
  middle_display,
  align = "llrrrrrrr",
  caption = paste(
    "Table 4. BF-adjusted Middle results at selected nominal levels.",
    "Seed minima and maxima describe the five Monte Carlo replications."
  )
)
```

## Power comparison

### Broad random B-spline truth

```{r bspline-power-comparison, fig.width=12, fig.height=6, fig.cap="Figure 5. Functional-testing power under broad random B-spline truth. Rows identify Raw and BF-adjusted FASH, columns identify functional targets, and curves compare the two observation models."}
plot_power_comparison("random_bspline")
```

### Compact raised-cosine truth

```{r raised-cosine-power-comparison, fig.width=12, fig.height=6, fig.cap="Figure 6. Functional-testing power under compact raised-cosine truth. Rows identify Raw and BF-adjusted FASH, columns identify functional targets, and curves compare the two observation models."}
plot_power_comparison("raised_cosine")
```

## Reproducibility

The production experiment code is contained in
`code/revision_simulations/internal/r3_ideal_gaussian_measurement/`. The runner
pins the frozen R3 simulation functions, real-genotype helper, genotype cache,
temporal-mixture contract, fashr version and git SHA, genotype-content digests,
and the dedicated ideal measurement seed rule. Each mechanism-by-seed result
is saved atomically and validated before reuse. The production summary applies
a tolerance of $10^{-12}$ at the nominal-alpha boundaries. The final directory
is promoted only after all ten replicates, summaries, manifest, and completion
flag pass their contracts.

The Midway3 job is submitted manually from the workspace:

```bash
cd /project/mstephens/ziangzhang/fash/workspace
sbatch slurm/25_r3_ideal_gaussian_measurement_iwp1_mixture.sbatch
```

The immutable server output directory is
`workspace/results/revision_simulations/r3_ideal_gaussian_known_t_adjusted_se_matched_truth_open_middle_3_12_center_aligned_iwp1_geometry_mixture_full_universe_fashr0143_pilot5/`.
The page reads the identical validated local cache under
`output/revision_simulations/internal/` with the same result ID.
