---
title: "Internal analysis: manuscript impact of switching to FASH-CL"
author: "Ziang Zhang"
date: "2026-08-10"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 11,
  fig.height = 8,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(file.path(
  "code",
  "revision_simulations",
  "internal",
  "fash_cl_manuscript_impact",
  "reporting.R"
))
```

<style>
.impact-banner {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border: 1px solid #d9a441;
  border-left: 6px solid #d9a441;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.impact-answer {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.impact-caution {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #cc7a00;
  border-radius: 4px;
  background: #fff8ec;
  color: #563a0c;
}
.impact-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.impact-table-scroll table {
  margin-bottom: 1em;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
.MathJax_Display,
pre {
  max-width: 100%;
  overflow-x: auto;
}
</style>

::: {.impact-banner}
<strong>Internal decision page.</strong> This page is intentionally absent
from the workflowr homepage, navigation, manuscript, and response letter. It
compares the retained current analysis with a matched FASH-CL sensitivity and
does not itself change the primary analysis.
:::

## Executive answer

::: {.impact-answer}
<strong>Switching to FASH-CL would require a substantial rewrite, not a simple
number update.</strong> Dynamic discoveries fall from
`r format_integer(current_dynamic_pairs)` to
`r format_integer(cl_dynamic_pairs)` pairs and from
`r format_integer(current_dynamic_genes)` to
`r format_integer(cl_dynamic_genes)` genes. Only
`r format_integer(method_overlap$intersection[method_overlap$unit == "Gene-variant pairs"])`
pairs are shared. The flexible functional modeling story remains valid, but
the stronger claim that FASH identifies <em>many more</em> dynamic pairs than
the Strober parametric analyses is not supported by FASH-CL.
:::

The sensitivity changes the regression inputs before FASH is fitted: current
FASH uses PCs estimated separately at each time point, whereas FASH-CL reuses
donor-level PCs at every time point. The FASH prior, BF adjustment, and
thresholds are otherwise matched. Therefore, changes in discoveries cannot be
attributed solely to the flexible EB-LGP model.

```{r claim-impact-table, results='asis'}
render_scrollable_table(
  claim_impact,
  align = "lll",
  minimum_width = "1180px"
)
```

```{r discovery-change-table, results='asis'}
render_scrollable_table(
  discovery_change_table,
  digits = 1L,
  minimum_width = "980px"
)
```

## Global lfdr and IWP1 prior diagnostics

### Pair-level lfdr comparison

FASH estimates lfdr for gene-variant pairs, so the left panel includes all
`r format_integer(nrow(lfdr_scatter_all))` tested pairs rather than deduplicated
rsIDs. For the right panel, the pair with the smallest current-FASH lfdr is
selected separately within each gene; variant ID and pair key provide stable
tie-breakers. The same selected pair is then located in FASH-CL, as requested.
Colors indicate whether the pair belongs to either method's cumulative-lfdr
FDR-0.05 discovery set.

```{r lfdr-scatter, fig.width=11.5, fig.height=6.1, fig.cap="Global IWP1 lfdr comparison. The left panel contains every tested gene-variant pair. The right panel contains one pair per gene, selected solely by the minimum current-FASH lfdr, with the same pair evaluated under FASH-CL. Dashed lines mark equality."}
plot_lfdr_scatter()
```

The all-pair Spearman correlation is
`r format_decimal(lfdr_comparison_summary$spearman_correlation[lfdr_comparison_summary$selection == "All tested gene-variant pairs"], 3L)`;
after current-FASH top-pair selection it is
`r format_decimal(lfdr_comparison_summary$spearman_correlation[lfdr_comparison_summary$selection == "Current-FASH top pair per gene"], 3L)`.

### IWP1 prior weights before and after BF adjustment

The raw panel shows the mixture prior stored by each original IWP1 fit. The
BF-adjusted panel uses the retained current-FASH prior and the matched
memory-bounded BF update for FASH-CL. Both axes use pseudo-log transforms so
the null component at PSD zero and components assigned exactly zero weight
remain visible.

```{r iwp1-prior-weights, fig.width=11.2, fig.height=5.8, fig.cap="IWP1 prior weights for current FASH and FASH-CL before and after BF adjustment. PSD zero is the exact-null component. Pseudo-log axes retain zero-valued components while expanding the small non-null weights."}
plot_iwp1_prior_weights()
```

```{r iwp1-prior-summary-table, results='asis'}
render_scrollable_table(
  prior_weight_summary[, c(
    "method", "adjustment", "null_weight", "nonzero_component_count"
  )],
  digits = 6L,
  minimum_width = "760px"
)
```

## Figure 3: dynamic examples and the flexibility claim

Only `r figure3_retained_count` of the six exact Figure 3 units remains a
FASH-CL discovery. The table and first plot keep the original units fixed so
the sensitivity is visible rather than concealed by replacement.

```{r figure3-status-table, results='asis'}
render_scrollable_table(
  figure3_status_table,
  digits = 5L,
  minimum_width = "980px"
)
```

```{r figure3-same-units, fig.width=12, fig.height=11, fig.cap="Figure 1. The six paper Figure 3 units under current FASH and FASH-CL. Red curves and ribbons are the FASH posterior mean and 95% interval. Green and purple dashed curves are weighted linear and quadratic fits."}
plot_figure3_same_units()
```

### A possible FASH-CL replacement figure

The replacement rule preserves the original roles rather than selecting genes
for a preferred biological interpretation:

1. retain GINM1 and add the lowest-lfdr FASH-CL unit detected by Strober;
2. among FASH-CL-only units with both Strober p-values at least 0.20, select
   two strong discoveries with the largest standardized quadratic misfit;
3. repeat within the upper 20% of discovered lfdr values for two
   threshold-near examples.

```{r figure3-replacement-table, results='asis'}
render_scrollable_table(
  figure3_replacement_table,
  digits = 5L,
  minimum_width = "1100px"
)
```

```{r figure3-proposed, fig.width=10.8, fig.height=11, fig.cap="Figure 2. Deterministically selected FASH-CL replacements for paper Figure 3. Replacement examples demonstrate model flexibility, but they do not restore the original discovery-count claim."}
plot_figure3_proposed()
```

::: {.impact-caution}
Replacing examples can preserve the visual demonstration that an EB-LGP model
captures shapes missed by linear or quadratic curves. It cannot support the
original causal wording that model flexibility explains the larger discovery
count, because the discovery advantage disappears under the CL PC design.
:::

## Figure 4: overlap with Strober

FASH-only pairs decline from `r format_integer(current_fash_only_pairs)` to
`r format_integer(cl_fash_only_pairs)`. FASH-CL is more concordant with the
Strober sets, but its total pair count is comparable to, rather than much
larger than, the linear and quadratic analyses.

The following panels preserve the paper's three-set Venn representation. Each
row compares the original current-FASH diagram with the corresponding FASH-CL
diagram. In addition to the paper's gene-level and gene-variant-pair views, a
variant-level view is shown so repeated variants assigned to different genes
can be distinguished from genuinely distinct discovered variants.

```{r overlap-summary-table, results='asis'}
render_scrollable_table(
  method_overlap,
  digits = 4L,
  minimum_width = "820px"
)
```

```{r venn-gene-plot, fig.width=11.5, fig.height=5.8, fig.cap="Paper Figure 4 sensitivity at the gene level. Current FASH is shown on the left and FASH-CL on the right."}
plot_venn_diagram_comparison("Genes")
```

```{r venn-pair-plot, fig.width=11.5, fig.height=5.8, fig.cap="Paper Figure 4 sensitivity at the gene-variant-pair level. Current FASH is shown on the left and FASH-CL on the right."}
plot_venn_diagram_comparison("Gene-variant pairs")
```

```{r venn-variant-plot, fig.width=11.5, fig.height=5.8, fig.cap="Additional paper Figure 4 sensitivity at the unique-variant level. Current FASH is shown on the left and FASH-CL on the right."}
plot_venn_diagram_comparison("Variants")
```

```{r venn-region-table, results='asis'}
render_scrollable_table(
  venn_regions[, c("method", "unit", "region", "count")],
  minimum_width = "980px"
)
```

## Figure 5: nonlinear dynamic examples

Only `r figure5_retained_count` of the four exact nonlinear examples remains
significant under FASH-CL. The IWP2 result itself does not disappear: FASH-CL
finds 60 pairs in 6 genes, compared with 44 pairs in 9 genes under current
FASH. The identities of the genes, however, change sharply.

```{r figure5-status-table, results='asis'}
render_scrollable_table(
  figure5_status_table,
  digits = 5L,
  minimum_width = "820px"
)
```

```{r figure5-same-units, fig.width=10.8, fig.height=9.5, fig.cap="Figure 4. The four paper Figure 5 units under current FASH and FASH-CL IWP2 fits."}
plot_figure5_same_units()
```

The proposed FASH-CL version retains surviving paper units and fills the
remaining panels with the lowest-lfdr discoveries from distinct new genes.

```{r figure5-replacement-table, results='asis'}
render_scrollable_table(
  figure5_replacement_table,
  digits = 6L,
  minimum_width = "760px"
)
```

```{r figure5-proposed, fig.width=10.8, fig.height=7.8, fig.cap="Figure 5. Proposed FASH-CL nonlinear examples selected by a deterministic lowest-lfdr, distinct-gene rule."}
plot_figure5_proposed()
```

## Table 1: functional classifications

The CL classifications below were recomputed from the BF-adjusted FASH-CL
posterior. Each of the `r format_integer(cl_dynamic_pairs)` dynamic units was
sampled once with 3,000 posterior trajectories; the same draws were used for
all four functionals before cumulative FSR control at 0.05.

```{r functional-settings, eval=FALSE}
smooth_var <- seq(0, 15, by = 0.1)
posterior_sample_size <- 3000L
switch_threshold <- 0.25
classification_seed <- 20260810L
classification_alpha <- 0.05
```

```{r category-change-table, results='asis'}
render_scrollable_table(
  category_change_table,
  digits = 1L,
  minimum_width = "1080px"
)
```

Because categories are not mutually exclusive, these changes should be read
category by category rather than summed. The early category disappears at the
paper threshold, middle increases, and switch falls from 984 pairs/250 genes
to 594 pairs/150 genes. These membership changes directly affect Figure 6 and
Table 2.

## Figure 6: category examples

Figure 6 is not in the minimal list in the question, but it is included here
because it is the visual support for Table 1. Only
`r figure6_retained_count` of its four exact units remains significant in the
same FASH-CL category.

```{r figure6-status-table, results='asis'}
render_scrollable_table(
  figure6_status_table,
  digits = 5L,
  minimum_width = "1120px"
)
```

```{r figure6-same-units, fig.width=10.8, fig.height=9.8, fig.cap="Figure 6. The four paper category examples under current FASH and FASH-CL IWP1 fits. The blue dashed line marks zero."}
plot_figure6_same_units()
```

For a proposed CL version, an original unit is retained only if it remains
significant in the same category. Otherwise, the lowest-lfsr significant unit
from a distinct gene is used. If a category has no FASH-CL discoveries at FSR
0.05, the closest nonsignificant unit is displayed only as a diagnostic and is
explicitly not treated as a valid replacement.

```{r figure6-replacement-table, results='asis'}
render_scrollable_table(
  figure6_replacement_table,
  digits = 5L,
  minimum_width = "980px"
)
```

```{r figure6-proposed, fig.width=10.8, fig.height=7.8, fig.cap="Figure 7. Proposed FASH-CL category panels selected by the same-category lowest-lfsr rule. A category with no FSR-0.05 discovery shows its closest nonsignificant unit and cannot retain a significant-example panel."}
plot_figure6_proposed()
```

## Table 2: Hallmark enrichment and the biological story

The first table reproduces the values currently printed in the manuscript.
The second recomputes both methods using the same installed Hallmark annotation
(`msigdbr` `r configuration$msigdbr_version`) and the same tested-gene
universe. This separates PC sensitivity from annotation-version drift.

```{r manuscript-hallmark-table, results='asis'}
render_scrollable_table(
  manuscript_hallmark_display,
  digits = 5L,
  minimum_width = "940px"
)
```

```{r matched-hallmark-table, results='asis'}
render_scrollable_table(
  hallmark_display,
  digits = 6L,
  minimum_width = "1060px"
)
```

The matched-version result separates two biological conclusions. Among all
dynamic genes, Hypoxia strengthens from q =
`r format_decimal(current_all_hypoxia_q, 3L)` to
`r format_decimal(cl_all_hypoxia_q, 3L)`. Among switch genes, however, Hypoxia
weakens from q = `r format_decimal(current_switch_hypoxia_q, 3L)` to
`r format_decimal(cl_switch_hypoxia_q, 3L)`, and KRAS signaling weakens from q
= `r format_decimal(current_switch_kras_q, 3L)` to
`r format_decimal(cl_switch_kras_q, 3L)`. Thus the general Hypoxia signal is
preserved, but the manuscript's switch-specific Hypoxia/KRAS story is not.

## How much rewriting would be needed?

- **Major rewrite:** the discovery-count paragraph, the interpretation of
  Figure 4, and the claim that flexibility leads to many more detected pairs.
- **Figure replacement plus text revision:** Figures 3 and 5. The general
  flexible-trajectory message survives, but most named loci change.
- **Full numerical and biological update:** Table 1, Figure 6, Table 2, and the
  switch/hypoxia/KRAS paragraph.
- **Mostly preserved:** the methodological description of FASH, BF adjustment,
  arbitrary functional testing, and the fact that flexible IWP priors can model
  nonparametric temporal effects.

The safest FASH-CL framing would be: the model finds a partly distinct set of
dynamic and nonlinear eQTLs, including thousands not called by the parametric
analyses, while providing flexible posterior estimation and functional
classification. It should not say that FASH-CL finds *many more pairs* than
both parametric approaches.

## Reproducibility

Routine builds of this page read `analysis_cache.rds`; they do not rerun
posterior sampling. The retained classification used fixed per-unit seeds and
checkpointed every 64 units.

```{r reproduction-command, eval=FALSE}
Sys.setenv(FASH_CL_CLASSIFICATION_CORES = 16)
source(file.path(
  "code",
  "revision_simulations",
  "internal",
  "fash_cl_manuscript_impact",
  "run_fash_cl_manuscript_impact.R"
))

workflowr::wflow_build(
  "analysis/revision_internal_fash_cl_manuscript_impact.rmd",
  local = TRUE
)
```

```{r runtime-table, results='asis'}
render_scrollable_table(
  runtime_summary,
  digits = 1L,
  minimum_width = "760px"
)
```

<details>
<summary><strong>Input provenance</strong></summary>

```{r provenance-table, results='asis'}
render_scrollable_table(
  input_provenance,
  minimum_width = "1420px"
)
```

</details>
