---
title: "Temporary internal simulation: functional testing with targeted broad B-spline effects"
author: "Ziang Zhang"
date: "2026-07-29"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 10,
  fig.height = 6,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source("code/revision_simulations/shared/simulation_functions.R")

mc_root <- file.path("output", "revision_simulations", "mc")
targeted_id <- "internal_targeted_broad_calibration_pilot5"
original_id <- "functional_targeted_broad_baseline075_pilot5"
targeted_dir <- file.path(mc_root, targeted_id)
original_dir <- file.path(mc_root, original_id)
diagnostic_dir <- file.path(
  "output",
  "revision_simulations",
  "internal",
  "targeted_broad_functional_calibration_diagnostic"
)
figure_dir <- file.path(diagnostic_dir, "figures")
summary_dir <- file.path(targeted_dir, "summary")

configuration <- readRDS(file.path(targeted_dir, "configuration.rds"))
example_curves <- readRDS(file.path(targeted_dir, "example_curves.rds"))
mc_alpha <- read.csv(
  file.path(summary_dir, "functional_testing_mc_alpha_curve.csv"),
  stringsAsFactors = FALSE
)
mc_alpha_005 <- read.csv(
  file.path(summary_dir, "functional_testing_mc_alpha005_summary.csv"),
  stringsAsFactors = FALSE
)
mc_pi0 <- read.csv(
  file.path(summary_dir, "functional_testing_mc_pi0_summary.csv"),
  stringsAsFactors = FALSE
)
original_vs_reproduced <- read.csv(
  file.path(diagnostic_dir, "original_vs_reproduced_alpha005.csv"),
  stringsAsFactors = FALSE
)
false_switch_counts <- read.csv(
  file.path(diagnostic_dir, "bf_false_switch_counts_by_seed.csv"),
  stringsAsFactors = FALSE
)
geometry_summary <- read.csv(
  file.path(diagnostic_dir, "dynamic_truth_geometry_summary.csv"),
  stringsAsFactors = FALSE
)
call_lfsr_summary <- read.csv(
  file.path(diagnostic_dir, "bf_switch_call_lfsr_by_truth.csv"),
  stringsAsFactors = FALSE
)

expected_class_probs <- c(
  dynamic_bspline = 0.20,
  constant = 0.40,
  zero = 0.40
)
method_order <- c("FASH-IWP1-Raw", "FASH-IWP1-BF")
target_order <- c("early", "middle", "late", "switch")

if (!identical(
      configuration$scenario,
      "genotype_functional_targeted_local_bspline"
    ) ||
    !isTRUE(all.equal(configuration$J, 1000L)) ||
    !isTRUE(all.equal(configuration$n_donors, 19L)) ||
    !isTRUE(all.equal(configuration$n_covariates, 5L)) ||
    !isTRUE(all.equal(configuration$class_probs, expected_class_probs)) ||
    !identical(configuration$truth_mechanism, "targeted_local_bspline") ||
    !identical(configuration$targeted_profile, "broad") ||
    !isTRUE(all.equal(configuration$dynamic_amplitude, 2)) ||
    !isTRUE(all.equal(configuration$switch_threshold, 0.25)) ||
    !isTRUE(all.equal(configuration$minimum_location_margin, 0.60)) ||
    !isTRUE(all.equal(configuration$minimum_location_ratio, 1.4)) ||
    !isTRUE(all.equal(configuration$non_switch_baseline_fraction, 0.75)) ||
    !isTRUE(all.equal(configuration$non_switch_background_fraction, 0.05)) ||
    length(configuration$seed_list) != 5L ||
    any(mc_alpha$n_replications != 5L) ||
    any(mc_alpha_005$n_replications != 5L) ||
    !all(method_order %in% mc_alpha$method) ||
    !all(target_order %in% mc_alpha$target)) {
  stop("The internal targeted-broad cache does not match the intended setting.")
}

required_figures <- c(
  "non_switch_truth_geometry_comparison.png",
  "bf_switch_call_lfsr_by_truth.png",
  "bf_false_switch_call_examples.png"
)
if (!all(file.exists(file.path(figure_dir, required_figures)))) {
  stop("One or more internal diagnostic figures are missing.")
}
if (!requireNamespace("png", quietly = TRUE)) {
  stop("The png package is required to embed cached diagnostic figures.")
}

plot_cached_png <- function(path) {
  image <- png::readPNG(path)
  grid::grid.newpage()
  grid::grid.raster(
    image,
    width = grid::unit(1, "npc"),
    height = grid::unit(1, "npc"),
    interpolate = TRUE
  )
}

format_decimal <- function(x, digits = 3) {
  formatC(x, format = "f", digits = digits)
}

format_mc_interval <- function(mean, lower, upper, digits = 3) {
  paste0(
    format_decimal(mean, digits),
    " [",
    format_decimal(lower, digits),
    ", ",
    format_decimal(upper, digits),
    "]"
  )
}

format_functional_table <- function(summary_table) {
  table <- summary_table
  table$target <- factor(table$target, levels = target_order)
  table$method <- factor(table$method, levels = method_order)
  table <- table[order(table$target, table$method), , drop = FALSE]
  data.frame(
    Target = tools::toTitleCase(as.character(table$target)),
    Method = as.character(table$method),
    `Mean calls` = format_decimal(table$mean_discoveries, 1),
    `Power (95% MC CI)` = format_mc_interval(
      table$mean_power,
      table$power_ci_lower,
      table$power_ci_upper
    ),
    `Empirical FSR (95% MC CI)` = format_mc_interval(
      table$mean_empirical_fsr,
      table$empirical_fsr_ci_lower,
      table$empirical_fsr_ci_upper
    ),
    check.names = FALSE
  )
}

format_pi0_table <- function(summary_table) {
  summary_table$fit <- factor(
    summary_table$fit,
    levels = c("Raw", "BF-corrected")
  )
  summary_table <- summary_table[order(summary_table$fit), , drop = FALSE]
  data.frame(
    Fit = as.character(summary_table$fit),
    `Mean estimated dynamic-null proportion (95% MC CI)` =
      format_mc_interval(
        summary_table$mean_estimated_pi0,
        summary_table$pi0_ci_lower,
        summary_table$pi0_ci_upper
      ),
    check.names = FALSE
  )
}

plot_truth_examples <- function(examples) {
  group_order <- c(
    "early / switch",
    "early / non-switch",
    "middle / switch",
    "middle / non-switch",
    "late / switch",
    "late / non-switch"
  )
  panels <- lapply(examples[group_order], `[[`, 1L)

  old_par <- par(no.readonly = TRUE)
  on.exit(par(old_par), add = TRUE)
  par(mfrow = c(3, 2), mar = c(3.6, 3.8, 3.0, 0.8), oma = c(0, 0, 2.3, 0))

  for (panel in panels) {
    observed <- panel$observed
    time_group <- sub(" /.*", "", panel$group)
    target_window <- switch(
      time_group,
      early = c(0, 3),
      middle = c(4, 11),
      late = c(12, 15)
    )
    window_color <- switch(
      time_group,
      early = "#d7eaf7",
      middle = "#dff0d8",
      late = "#fde6c9"
    )
    y_limits <- range(c(
      observed$estimate - 2 * observed$se,
      observed$estimate + 2 * observed$se,
      panel$true_curve$true_effect,
      -configuration$switch_threshold,
      configuration$switch_threshold
    ))
    y_padding <- max(0.15, 0.07 * diff(y_limits))
    y_limits <- y_limits + c(-y_padding, y_padding)

    plot(
      observed$time,
      observed$estimate,
      type = "n",
      ylim = y_limits,
      xlim = range(configuration$time_grid),
      xlab = "Time",
      ylab = "Genetic effect",
      main = paste(panel$group, "\n", panel$variant_id)
    )
    rect(
      target_window[1],
      y_limits[1],
      target_window[2],
      y_limits[2],
      col = window_color,
      border = NA
    )
    abline(
      h = c(-configuration$switch_threshold, configuration$switch_threshold),
      col = "gray70",
      lty = 3
    )
    abline(h = 0, col = "gray55", lty = 3)
    arrows(
      observed$time,
      observed$estimate - 2 * observed$se,
      observed$time,
      observed$estimate + 2 * observed$se,
      angle = 90,
      code = 3,
      length = 0.035,
      col = "gray45"
    )
    points(observed$time, observed$estimate, pch = 19)
    lines(
      panel$true_curve$time,
      panel$true_curve$true_effect,
      col = "#E64B35",
      lwd = 2
    )
  }
  mtext(
    paste(
      "Per-time estimates with two-SE bars; red lines are true effects;",
      "shading marks the target time window"
    ),
    outer = TRUE,
    cex = 1.02
  )
}

functional_table <- format_functional_table(mc_alpha_005)
pi0_table <- format_pi0_table(mc_pi0)

reproduction_table <- original_vs_reproduced[
  original_vs_reproduced$target == "switch" &
    original_vs_reproduced$method == "FASH-IWP1-BF",
  ,
  drop = FALSE
]
reproduction_table <- data.frame(
  Run = reproduction_table$run,
  `Mean dynamic discoveries` =
    format_decimal(reproduction_table$mean_dynamic_discoveries, 1),
  `Mean switch calls` =
    format_decimal(reproduction_table$mean_discoveries, 1),
  `Switch power` = format_decimal(reproduction_table$mean_power),
  `Empirical switch FSR` =
    format_decimal(reproduction_table$mean_empirical_fsr),
  check.names = FALSE
)

false_source_table <- aggregate(
  n_false_calls ~ false_call_source,
  data = false_switch_counts,
  FUN = sum
)
false_source_table <- data.frame(
  Source = false_source_table$false_call_source,
  `False switch calls across five seeds` =
    false_source_table$n_false_calls,
  check.names = FALSE
)

non_switch_geometry <- geometry_summary[
  geometry_summary$switch_status == "non-switch",
  ,
  drop = FALSE
]
non_switch_geometry$setting <- factor(
  non_switch_geometry$setting,
  levels = c(
    "Targeted broad; baseline 0.75",
    "Compact peaks; zero background"
  )
)
non_switch_geometry <- non_switch_geometry[
  order(non_switch_geometry$setting),
  ,
  drop = FALSE
]
geometry_table <- data.frame(
  Setting = as.character(non_switch_geometry$setting),
  Curves = non_switch_geometry$n_curves,
  `Median near-zero fraction` =
    format_decimal(non_switch_geometry$near_zero_fraction_median),
  `Median same-sign margin` =
    format_decimal(non_switch_geometry$same_sign_margin_median),
  `Median maximum absolute effect` =
    format_decimal(non_switch_geometry$maximum_absolute_effect_median),
  check.names = FALSE
)

call_lfsr_summary$call_truth_class <- factor(
  call_lfsr_summary$call_truth_class,
  levels = c(
    "True switch",
    "True-dynamic non-switch",
    "First-stage dynamic null"
  )
)
call_lfsr_summary <- call_lfsr_summary[
  order(call_lfsr_summary$call_truth_class),
  ,
  drop = FALSE
]
lfsr_table <- data.frame(
  `Truth class among BF switch calls` =
    as.character(call_lfsr_summary$call_truth_class),
  Calls = call_lfsr_summary$n_calls,
  `Median individual lfsr` =
    format_decimal(call_lfsr_summary$lfsr_median),
  `Individual lfsr range` = paste0(
    format_decimal(call_lfsr_summary$lfsr_minimum),
    " - ",
    format_decimal(call_lfsr_summary$lfsr_maximum)
  ),
  check.names = FALSE
)
```

## Overview

> **Internal temporary analysis.** This page is not part of the formal
> reviewer-facing workflowr navigation.

This genotype-level simulation evaluates early, middle, late, and switch
functional testing under a labelled non-IWP B-spline truth. For donor $i$,
variant $j$, and time $t$, expression is generated as

$$
Y_{ijt} = \alpha_{jt} + G_{ij}\beta_j(t) +
X_i^T\gamma_{jt} + \epsilon_{ijt},
\qquad \epsilon_{ijt} \sim N(0,1).
$$

Each dataset has `19` donors, `16` time points, `1000` variants, and five
PC-like covariates. Exactly `200` variants are dynamic, `400` have constant
effects, and `400` have zero effects. Per-time effects and standard errors are
estimated from `Y ~ G + PC1 + PC2 + PC3 + PC4 + PC5`, followed by the
real-data t-statistic correction with

$$
\mathrm{df} = 19 - 7 = 12.
$$

The results average five independently simulated datasets with seeds `12345`,
`22345`, `32345`, `42345`, and `52345`.

## Functional truth

Dynamic effects are broad local cubic B-spline curves with maximum absolute
effect `2`. They are split as evenly as possible among the six strata formed
by early, middle, or late timing crossed with switch or non-switch status.

Early, middle, and late mean that the largest absolute effect occurs on days
0--3, 4--11, or 12--15. A switch curve has both a positive and a negative
effect exceeding the manuscript threshold $c = 0.25$. For non-switch curves,
a same-signed baseline equal to `0.75` times the primary amplitude keeps the
trajectory away from zero. All labels are validated on a refined time grid
before expression data are generated.

## Example simulated eQTLs

One representative trajectory is shown for each truth stratum. Points are the
per-time effect estimates used by FASH, bars show two corrected standard
errors, and red lines are the continuous true effects. The shaded region marks
the time window used by the early, middle, or late functional.

```{r labelled-truth-examples, echo=FALSE, fig.width=11, fig.height=11}
plot_truth_examples(example_curves)
```

## Inference procedure

At each nominal level $\alpha$, both `FASH-IWP1-Raw` and `FASH-IWP1-BF` use
the same three-stage rule:

1. Select dynamic eQTLs using the global FASH FDR procedure.
2. Evaluate early, middle, late, and switch functionals with
   `testing_functional()` using posterior samples.
3. Rank the functional lfsr values and call the largest set whose cumulative
   FSR is at most $\alpha$.

Power is the fraction of true members of a functional class that are called.
Empirical FSR is the fraction of all reported calls whose true functional is
non-positive. The empirical FSR therefore includes both functional
misclassification among dynamic eQTLs and errors inherited from the first
dynamic-eQTL selection stage.

## Functional-testing results

### Results at alpha 0.05

```{r functional-results-table, echo=FALSE}
knitr::kable(functional_table, align = "l")
```

For BF-updated FASH, mean switch power is `0.925` and mean empirical switch FSR
is `0.021`. Early, middle, and late empirical FSR estimates are zero in these
five replications.

### Power as a function of alpha

```{r functional-power, echo=FALSE, fig.width=11, fig.height=9}
plot_mc_functional_curve_grid(
  mc_curve = mc_alpha,
  metric = "power",
  title = "Functional-testing power for targeted broad B-spline effects"
)
```

### Empirical FSR as a function of alpha

```{r empirical-fsr, echo=FALSE, fig.width=13, fig.height=9.5}
plot_mc_functional_curve_grid(
  mc_curve = mc_alpha,
  metric = "empirical_fsr",
  title = "Empirical FSR for targeted broad B-spline effects"
)
```

The diagonal is the nominal level. The ribbons are 95% Monte Carlo confidence
intervals across the five seeds. The near-zero early, middle, and late curves
mean that no false calls occurred for those targets in this pilot; they should
not be interpreted as high-precision estimates of a zero error rate.

## Dynamic-null proportion

The true first-stage dynamic-null proportion is $\pi_0 = 0.8$. The table shows
the fitted constant-component weight before and after BF updating.

```{r pi0-table, echo=FALSE}
knitr::kable(pi0_table, align = "l")
```

## Why switch FSR is calibrated here

### Non-switch truth geometry

```{r geometry-table, echo=FALSE}
knitr::kable(geometry_table, align = "l")
```

```{r geometry-figure, echo=FALSE, fig.height=3.65, out.width="100%"}
plot_cached_png(
  file.path(figure_dir, "non_switch_truth_geometry_comparison.png")
)
```

The maximum effect sizes are similar in the targeted-broad and compact-peak
settings. The main geometric difference is sign separation: targeted
non-switch curves have a large same-sign margin, whereas compact
zero-background curves spend most of the evaluation grid inside
$[-0.25,0.25]$. Consequently, isolated posterior excursions across zero are
much less likely to turn a true non-switch trajectory into a switch call in
the targeted-broad setting.

### False-call decomposition

At $\alpha = 0.05$, BF-updated FASH makes 10 false switch calls across the five
replications.

```{r false-source-table, echo=FALSE}
knitr::kable(false_source_table, align = "l")
```

```{r lfsr-table, echo=FALSE}
knitr::kable(lfsr_table, align = "l")
```

```{r lfsr-figure, echo=FALSE, fig.width=7, fig.height=5.6, out.width="70%"}
plot_cached_png(
  file.path(figure_dir, "bf_switch_call_lfsr_by_truth.png")
)
```

The six called dynamic non-switch variants have a median individual switch
lfsr of `0.982`. They enter only because the cumulative FSR rule averages
their high lfsr values with many highly confident true-switch calls. Their
realized frequency nevertheless remains within the aggregate error budget.

### Example false switch calls

The red lines are the true effects; titles report the individual lfsr,
cumulative FSR at the call boundary, and true switch functional.

```{r false-switch-examples, echo=FALSE, fig.height=12.7, out.width="100%"}
plot_cached_png(
  file.path(figure_dir, "bf_false_switch_call_examples.png")
)
```

The six true-dynamic non-switch examples remain one-signed. The other four
errors are constant or zero variants that passed the first-stage dynamic-eQTL
screen.

## Reproduction check

The current-code run closely reproduces the original five-seed cache.

```{r reproduction-table, echo=FALSE}
knitr::kable(reproduction_table, align = "l")
```

The one-call difference is consistent with posterior Monte Carlo variation
near the cumulative FSR boundary.

## Interpretation

In this targeted-broad setting, BF-updated FASH has mean empirical FSR below
the nominal level across the evaluated alpha grid and retains high switch
power. Raw FASH is also conservative for early, middle, and late calls, but
its switch FSR exceeds the nominal level as alpha increases. Calibration is
especially favorable after BF updating because true non-switch trajectories
are clearly same-signed; this does not imply that controlling the global
dynamic-null proportion alone guarantees calibration for every functional
hypothesis.

## Reproducibility

The five-seed cache was generated with:

```{bash, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/archived_experiments/run_functional_testing_bspline_mc_replication.R \
  --J 1000 \
  --seed-list 12345,22345,32345,42345,52345 \
  --truth-mechanism targeted_local_bspline \
  --targeted-profile broad \
  --dynamic-amplitude 2 \
  --minimum-location-margin 0.60 \
  --minimum-location-ratio 1.4 \
  --non-switch-baseline-fraction 0.75 \
  --non-switch-background-fraction 0.05 \
  --output-id internal_targeted_broad_calibration_pilot5 \
  --num-cores 2 \
  --cache-full-fits true
```

The diagnostic tables and figures were generated with:

```{bash, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/diagnostics/summarize_internal_targeted_broad_calibration_diagnostic.R
```

Cached outputs are stored under:

```text
output/revision_simulations/mc/internal_targeted_broad_calibration_pilot5/
output/revision_simulations/internal/targeted_broad_functional_calibration_diagnostic/
```
