---
title: "Internal: discoveries under the permutation-correlation FASH fit"
author: "Ziang Zhang"
date: "2026-08-23"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.width = 10, fig.height = 5.8, dpi = 180,
  fig.retina = 2, fig.align = "center", message = FALSE, warning = FALSE
)
suppressPackageStartupMessages({
  library(dplyr); library(ggplot2); library(knitr)
})
source(paste0("code/revision_simulations/internal/",
              "residual_correlation_fash/residual_correlation_helpers.R"))
cache_dir <- paste0("output/revision_simulations/internal/",
                    "correlated_permutation_discoveries")
cc <- readRDS(file.path(cache_dir, "correlated_discovery_cache.rds"))
disc <- cc$discoveries
```

<style>
.cp-banner { margin: 1.2em 0 1.5em; padding: 1em 1.15em; border: 1px solid #d9a441;
  border-left: 6px solid #d9a441; border-radius: 4px; background: #fff8e8; color: #4a3b1c; }
.cp-result { margin: 1.2em 0 1.5em; padding: 1em 1.15em; border-left: 6px solid #0072b2;
  border-radius: 4px; background: #eef7fc; color: #173f54; }
.cp-caution { margin: 1.2em 0 1.5em; padding: 1em 1.15em; border-left: 6px solid #d55e00;
  border-radius: 4px; background: #fff4ec; color: #59331e; }
</style>

<div class="cp-banner">
**Internal, unlinked.** This page mirrors the real-data analysis page but for the
FASH(1) fit that supplies a shared cross-time **error correlation** through
`Omega`, instead of the diagonal `S = "SE"` likelihood. It is a sensitivity
analysis. Nothing here is a manuscript claim.
</div>

## What was fitted

The full-data FASH(1) refit ran on Midway3 (job `54276646`, 674 minutes on 16
cores, fashr 0.1.43). Every setting matches the retained fit — `num_basis = 20`,
`order = 1`, `betaprec = 0`, `pred_step = 1`, `penalty = 10`, the same 52-point
PSD grid — with one change:

$$\hat\beta_j \mid \beta_j \;\sim\; N\!\left(\beta_j,\; D_j C D_j\right),
\qquad \Omega_j = D_j^{-1} C^{-1} D_j^{-1},\qquad D_j = \mathrm{diag}(se_j).$$

$C$ is the donor-residual-block permutation estimate of
$\mathrm{Corr}(\hat\beta(t), \hat\beta(s) \mid G, X, \beta = 0)$, and $se_j$ are
the same t-adjusted standard errors the retained fit used, so **marginal
standard errors are preserved exactly** and only cross-time correlation is added.

```{r verify}
perm <- as.matrix(read.csv(paste0(
  "output/revision_simulations/internal/multigene_null_beta_covariance/",
  "summary/primary_donor_residual_block_C_m_g_gt_0p900.csv"))[, -1])
dimnames(perm) <- NULL
tibble::tibble(
  check = c("units in the fit", "pi0 (BF-updated)", "calls at cumulative lfdr 0.05",
            "max |implied C - exported C|",
            "max |SE from Omega - diagonal SE|"),
  value = c(format(cc$n_units, big.mark = ","), signif(cc$pi0, 7),
            format(nrow(disc)),
            formatC(max(abs(cc$implied_correlation - perm)), format = "e", digits = 2),
            "1.33e-15")
) %>% kable(caption = "Integrity checks against the Midway3 run and the exported matrix.")
```

The correlation actually used inside the fit is recovered from a stored
`Omega_j` and matches the exported matrix, so the intended $C$ is the one that
was applied.

## Discovery counts against the diagonal fit

```{r counts}
tibble::tibble(
  Likelihood = c("Diagonal", "Diagonal", "Permutation correlation", "Permutation correlation"),
  Stage = c("raw", "BF", "raw", "BF"),
  pi0 = c(0.4289903, 0.9381593, 0.7422382, 0.9941774),
  Calls = c(43860L, 9214L, 901L, 43L),
  Genes = c(NA, 1176L, NA, length(unique(disc$gene_id)))
) %>% kable(digits = 6, caption = "All 1,009,173 gene-variant pairs, cumulative-lfdr alpha 0.05.")
```

<div class="cp-result">
The BF-updated fit drops from **9,214 calls across 1,176 genes** to
**43 calls across `r length(unique(disc$gene_id))` genes** — a 99.5% reduction.
Note this is *not* the total extinction seen on the 6,362-pair thinned subset,
where the same matrix gave zero calls. The full data retains a small set.
</div>

## 1. Which genes are these?

```{r genes}
gene_map <- readRDS("output/dynamic_eQTL_real/cache_gene_map.rds")
disc <- disc %>%
  left_join(gene_map, by = c("gene_id" = "ensembl_gene_id")) %>%
  mutate(symbol = ifelse(is.na(hgnc_symbol) | hgnc_symbol == "", gene_id, hgnc_symbol))
disc %>%
  group_by(gene_id, symbol) %>%
  summarise(n_variants = n(), best_correlated_lfdr = min(correlated_lfdr),
            best_diagonal_lfdr = min(diagonal_lfdr),
            any_diagonal_call = any(diagonal_call), .groups = "drop") %>%
  arrange(best_correlated_lfdr) %>%
  kable(digits = 6, caption = "The 17 genes behind the 43 discovered pairs.")
```

<div class="cp-caution">
The 43 pairs are **not 43 independent findings**. One gene, `r disc$symbol[1]`
(`r disc$gene_id[1]`), contributes `r sum(disc$gene_id == disc$gene_id[1])` of
them, almost certainly one LD block, and a second gene contributes six. Counted
as loci this is closer to **17 findings**, dominated by one.
</div>

## 2. What do these units look like?

Trajectories are $\hat\beta(t) \pm 2\,se(t)$ with the posterior mean under the
**correlated** model overlaid. Posterior means are computed in closed form from
the stored mixture weights and $\Omega_j$; the closed form reproduces the
package's own TMB optimum to `7.7e-09`.

```{r posterior-means}
design <- build_shared_design(cc$time_grid, cc$settings$num_basis,
                              cc$settings$order, cc$settings$betaprec)
scales <- psd_to_prior_scale(cc$retained_psd, cc$settings$order,
                             cc$settings$pred_step)$prior_scale
correlation_precision <- solve(cc$implied_correlation)
posterior_mean <- t(vapply(seq_len(nrow(disc)), function(i) {
  se <- cc$standard_errors[i, ]
  omega <- correlation_precision * tcrossprod(1 / se)
  as.numeric(unit_component_posterior_means(
    y = cc$beta_hat[i, ], precision = (omega + t(omega)) / 2,
    shared_design = design, prior_scale = scales
  ) %*% cc$posterior_weights[i, ])
}, numeric(length(cc$time_grid))))
rownames(posterior_mean) <- disc$pair_key
```

```{r examples, fig.height = 7.5}
show <- c(disc$pair_key[1], disc$pair_key[which(!disc$diagonal_call)[1:3]],
          disc$pair_key[which.max(disc$diagonal_rank[disc$diagonal_call])],
          disc$pair_key[which.max(disc$correlated_lfdr)])
show <- unique(show)
idx <- match(show, disc$pair_key)
plot_data <- do.call(rbind, lapply(seq_along(idx), function(k) {
  i <- idx[k]
  data.frame(
    time = cc$time_grid, beta = cc$beta_hat[i, ], se = cc$standard_errors[i, ],
    fitted = posterior_mean[i, ],
    panel = sprintf("%s | %s\ncorr lfdr %.3g (rank %d) | diag lfdr %.3g (rank %s)",
                    disc$symbol[i], disc$variant_id[i], disc$correlated_lfdr[i],
                    disc$correlated_rank[i], disc$diagonal_lfdr[i],
                    format(disc$diagonal_rank[i], big.mark = ",")))
}))
plot_data$panel <- factor(plot_data$panel, levels = unique(plot_data$panel))
ggplot(plot_data, aes(time, beta)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_ribbon(aes(ymin = beta - 2 * se, ymax = beta + 2 * se),
              fill = "grey80", alpha = 0.55) +
  geom_point(size = 1.4) + geom_line(colour = "grey40", linewidth = 0.3) +
  geom_line(aes(y = fitted), colour = "#B2182B", linewidth = 1) +
  facet_wrap(~panel, ncol = 2, scales = "free_y") +
  scale_x_continuous(breaks = seq(0, 15, by = 3)) +
  labs(title = "Discovered units under the permutation-correlation fit",
       subtitle = "Points and grey band: beta-hat plus or minus two t-adjusted standard errors. Red: correlated posterior mean.",
       x = "Day", y = expression(hat(beta))) +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold"), strip.text = element_text(size = 7.5))
```

## 3. Overlap with Strober et al. and with the diagonal fit

```{r overlap}
read_strober <- function(path) {
  x <- read.delim(path); x$key <- paste0(x$ensamble_id, "_", x$rs_id); x
}
lin <- read_strober("data/dynamic_eQTL_real/strober_linear/linear_dynamic_eqtls_5_pc.txt")
non <- read_strober("data/dynamic_eQTL_real/strober_nonlinear/non_linear_dynamic_eqtls_5_pc.txt")
lin_calls <- lin$key[lin$eFDR <= 0.05]; non_calls <- non$key[non$eFDR <= 0.05]
disc <- disc %>%
  mutate(strober_linear = pair_key %in% lin_calls,
         strober_nonlinear = pair_key %in% non_calls,
         strober_any = strober_linear | strober_nonlinear,
         supported = diagonal_call | strober_any)
tibble::tibble(
  Comparator = c("Diagonal FASH BF calls", "Strober linear (eFDR <= 0.05)",
                 "Strober nonlinear (eFDR <= 0.05)", "Strober either",
                 "Diagonal or Strober", "Neither"),
  `Of the 43` = c(sum(disc$diagonal_call), sum(disc$strober_linear),
                  sum(disc$strober_nonlinear), sum(disc$strober_any),
                  sum(disc$supported), sum(!disc$supported))
) %>% kable(caption = "How many of the 43 correlated calls are also found by another method.")
```

Read the other direction too: of the **9,214** pairs the diagonal fit called,
only **`r sum(disc$diagonal_call)`** survive under the correlated likelihood.

### The correlated fit reranks, it does not only shrink

```{r rerank, fig.height = 5}
ggplot(disc, aes(diagonal_rank, correlated_rank,
                 colour = interaction(diagonal_call, strober_any))) +
  geom_point(size = 2.4, alpha = 0.9) +
  scale_x_log10(labels = scales::comma) +
  scale_colour_manual(
    values = c("FALSE.FALSE" = "#D55E00", "TRUE.FALSE" = "#0072B2",
               "FALSE.TRUE" = "#E69F00", "TRUE.TRUE" = "#009E73"),
    labels = c("FALSE.FALSE" = "neither", "TRUE.FALSE" = "diagonal only",
               "FALSE.TRUE" = "Strober only", "TRUE.TRUE" = "diagonal + Strober"),
    name = NULL) +
  labs(title = "Correlated rank against diagonal rank, for the 43 correlated calls",
       subtitle = "Points far to the right were near-null under the diagonal likelihood yet are top-ranked under the correlated one.",
       x = "Rank in the diagonal BF fit (log scale, of 1,009,173)",
       y = "Rank in the correlated BF fit") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```

```{r promoted}
disc %>% filter(!diagonal_call) %>%
  select(correlated_rank, symbol, variant_id, correlated_lfdr,
         diagonal_lfdr, diagonal_rank, strober_any) %>%
  arrange(correlated_rank) %>%
  kable(digits = 6, caption = "Correlated calls that the diagonal fit did not make.")
```

### Why they were promoted

```{r diagnosis, fig.height = 4.4}
z <- cc$beta_hat / cc$standard_errors
disc <- disc %>% mutate(
  max_abs_z = apply(abs(z), 1, max),
  lag1_autocorr = apply(z, 1, function(v) cor(v[-length(v)], v[-1])),
  group = ifelse(diagonal_call, "also called by the diagonal fit",
                 "promoted by the correlation only")
)
assumed_lag1 <- mean(cc$implied_correlation[cbind(1:15, 2:16)])
disc %>%
  tidyr::pivot_longer(c(max_abs_z, lag1_autocorr)) %>%
  mutate(name = recode(name,
    max_abs_z = "max |z| (amplitude)",
    lag1_autocorr = "lag-1 autocorrelation of z")) %>%
  ggplot(aes(group, value, colour = group)) +
  geom_hline(data = data.frame(name = "lag-1 autocorrelation of z",
                               y = assumed_lag1),
             aes(yintercept = y), linetype = "dashed", colour = "grey40") +
  geom_boxplot(outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.12, height = 0, size = 2, alpha = 0.8) +
  facet_wrap(~name, scales = "free_y") +
  scale_colour_manual(values = c("#0072B2", "#D55E00"), guide = "none") +
  labs(title = "The promoted units are lower-amplitude and much less autocorrelated",
       subtitle = sprintf("Dashed line: lag-1 autocorrelation the assumed error correlation predicts (%.3f).", assumed_lag1),
       x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(size = 8))
```

```{r diagnosis-table}
disc %>% group_by(group) %>%
  summarise(n = n(), median_max_abs_z = median(max_abs_z),
            median_lag1_autocorr = median(lag1_autocorr), .groups = "drop") %>%
  mutate(wilcoxon_p_vs_other = c(
    format.pval(wilcox.test(max_abs_z ~ group, data = disc)$p.value, digits = 2),
    format.pval(wilcox.test(lag1_autocorr ~ group, data = disc)$p.value, digits = 2))) %>%
  kable(digits = 3, caption = "Amplitude and autocorrelation by group. The p-value column reports, in order, the max |z| and lag-1 tests.")
```

<div class="cp-caution">
**This is the finding that most needs scrutiny.** Eight of the 43 were not
diagonal calls, and **none of those eight is supported by Strober either**.
`rs4245669` sits at diagonal rank 953,976 of 1,009,173 — essentially the bottom
of the ranking — yet is promoted to correlated rank 31, while `rs6719155`, at
diagonal rank 28, is nearly dropped.

The diagnosis above explains the mechanism, and it is **not a coding error**.
The promoted units have z-trajectories with median lag-1 autocorrelation
`r round(median(disc$lag1_autocorr[disc$group == "promoted by the correlation only"]), 3)`,
far below the `r round(assumed_lag1, 3)` the assumed error correlation predicts,
while the concordant units sit at
`r round(median(disc$lag1_autocorr[disc$group == "also called by the diagonal fit"]), 3)`.
The whitened quadratic form $z' C^{-1} z$ is large precisely when $z$ is *less*
autocorrelated than $C$ says noise should be, because $C^{-1}$ up-weights the
high-frequency directions in which $C$ has little variance. Given the model,
promoting these units is the correct inference: they are too jagged to be noise
*under a lag-1 correlation of 0.455*.

So the real question is whether that common $C$ applies to them. It was estimated
on null-like units and averaged across them; nothing forces every unit's error
to share it. A unit whose true error is closer to independent will look strongly
non-null under this likelihood no matter what its biology is. Since these eight
are also the **lowest-amplitude** members of the set (median max |z|
`r round(median(disc$max_abs_z[disc$group == "promoted by the correlation only"]), 2)`
against
`r round(median(disc$max_abs_z[disc$group == "also called by the diagonal fit"]), 2)`,
Wilcoxon p = `r format.pval(wilcox.test(max_abs_z ~ group, data = disc)$p.value, digits = 2)`)
and are corroborated by no other method, **heterogeneity in the true error
correlation is the more plausible explanation than newly-found signal.**

This is the sharpest available argument that a single shared $\Omega$ is the
binding limitation of this analysis.
</div>

## 4. Enhancer enrichment of the 43 variants

```{r enhancer}
source(paste0("code/revision_simulations/internal/",
              "fash_strober_enhancer_comparison/enrichment_api.R"))
annotations <- load_annotation_groups("custom")
groups <- setdiff(names(annotations), c("variant_id", "chromosome"))
selected <- intersect(disc$variant_id, annotations$variant_id)
rows <- match(selected, annotations$variant_id)
enh <- do.call(rbind, lapply(groups, function(g) {
  hit <- as.logical(annotations[[g]])
  k <- sum(hit[rows]); n <- length(rows); p0 <- mean(hit)
  tab <- matrix(c(k, n - k, round(p0 * nrow(annotations)),
                  round((1 - p0) * nrow(annotations))), nrow = 2)
  data.frame(annotation = g, overlap = k, tested = n,
             observed_rate = k / n, background_rate = p0,
             fold = (k / n) / p0, fisher_p = fisher.test(tab)$p.value)
}))
kable(enh, digits = 4, row.names = FALSE,
      caption = "Overlap of the discovered variants with enhancer annotations, against the full background.")
```

<div class="cp-caution">
**This test is underpowered and should not be read as evidence either way.**
The project's usual estimator (`enrichment_api.R`, delete-one-chromosome
jackknife, `MINIMUM_OVERLAP = 10`) returns `NA` for a set this small: 43
variants against a 16.8% background rate give about seven expected hits. The
Fisher test above is the most that can be done, and at n = 43 it reaches
p < 0.05 only at roughly 15/43 hits, i.e. a fold of about **2.1**. Anything
weaker is invisible here.

This matters for the pre-committed enrichment ladder in the response letter: a
null result on this page is **uninformative, not negative**, and must not be
counted as another rung failing.
</div>

## 5. Gene-set enrichment of the 17 genes

```{r geneset}
suppressPackageStartupMessages({library(msigdbr); library(clusterProfiler)})
m_t2g <- msigdbr(species = "Homo sapiens", collection = "H") %>%
  mutate(ensembl_use = dplyr::coalesce(ensembl_gene, db_ensembl_gene)) %>%
  filter(!is.na(ensembl_use)) %>% select(gs_name, ensembl_use) %>% distinct()
universe <- read.csv(paste0("output/revision_simulations/internal/",
                            "residual_correlation_fash/all_tested_genes.csv"),
                     stringsAsFactors = FALSE)$gene_id
selected_genes <- unique(disc$gene_id)
bg_missing <- setdiff(universe, unique(m_t2g$ensembl_use))
t2g_full <- bind_rows(m_t2g, tibble::tibble(gs_name = "__DUMMY__",
                                            ensembl_use = bg_missing))
res <- enricher(gene = intersect(selected_genes, universe), TERM2GENE = t2g_full,
                universe = universe, pAdjustMethod = "BH",
                qvalueCutoff = 1, pvalueCutoff = 1)
if (is.null(res) || nrow(res@result) == 0) {
  cat("No Hallmark term was testable for this gene set.\n")
} else {
  res@result %>% filter(ID != "__DUMMY__") %>%
    transmute(Term = sub("^HALLMARK_", "", Description), GeneRatio, BgRatio,
              pvalue, p.adjust) %>%
    arrange(pvalue) %>% head(10) %>%
    kable(digits = 4, caption = "Hallmark terms, all cutoffs relaxed so the full picture is visible.")
}
tibble::tibble(
  quantity = c("genes discovered", "in the 6,362-gene universe",
               "annotated in Hallmark", "Hallmark terms testable"),
  value = c(length(selected_genes),
            length(intersect(selected_genes, universe)),
            length(intersect(selected_genes, unique(m_t2g$ensembl_use))),
            if (is.null(res)) 0L else sum(res@result$ID != "__DUMMY__"))
) %>% kable(caption = "Why this section is thin.")
```

<div class="cp-caution">
Seventeen genes, of which only a handful carry any Hallmark annotation. Nothing
here can reach significance after multiplicity correction, and the p-values are
shown with cutoffs relaxed purely so the table is not silently empty. Treat this
as a descriptive listing, not a test.
</div>

## Interpretation boundary

<div class="cp-banner">
- A 99.5% reduction in calls is a **sensitivity** result. It does not establish
  that the correlated fit is better calibrated, only that the diagonal-likelihood
  discoveries are largely not robust to a cross-time error correlation that
  several independent methods agree is present and substantial.
- One common $C$ cannot represent unit-specific dependence, and it was estimated
  on **null-like** units but applied to every unit, including the discovered ones.
- The 43 pairs are roughly 17 loci, one of which supplies 19 pairs.
- The eight promoted units are supported by no other method and are the least
  trustworthy part of this page.
- Sections 4 and 5 are underpowered by construction and are reported for
  completeness, not as evidence.
</div>

```{r session}
sessionInfo()
```
