---
title: "Internal pilot: Set A trajectories and covariance estimators"
author: "Ziang Zhang"
date: "2026-08-07"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 10,
  fig.height = 5.8,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(
  "code/revision_simulations/r4_correlated_errors/real_data_correlation_helpers.R"
)
source(
  paste0(
    "code/revision_simulations/internal/covariance_estimation/",
    "observational_null_set_helpers.R"
  )
)
source(
  paste0(
    "code/revision_simulations/internal/covariance_estimation/",
    "observational_null_set_reporting.R"
  )
)
```

<style>
.null-pilot-banner {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border: 1px solid #d9a441;
  border-left: 6px solid #d9a441;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.null-pilot-result {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.null-pilot-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.null-pilot-table-scroll table {
  margin-bottom: 1em;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
pre {
  max-width: 100%;
  overflow-x: auto;
}
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: hidden;
}
</style>

<div class="null-pilot-banner">
<strong>Internal small-scale pilot.</strong> This page is not linked from the
workflowr homepage, reviewer-facing R4 page, manuscript, or response letter.
It does not use the donor-permutation covariance estimate, and it does not
modify any existing R4 analysis or cache.
</div>

## Question and short answer

For the 200-pair Set A pilot, what do the selected fitted trajectories actually
look like, and how different is the empirical covariance after within-unit
centering from the pairwise-difference estimate?

<div class="null-pilot-result">
<strong>Answer.</strong> The difference is not slight. The pairwise-difference
lag-1 covariance is
`r format_decimal(subset(set_a_key_lag_table, Estimator == "Pairwise difference" & Lag == 1)$Covariance, 3)`,
whereas the raw within-unit centered covariance is
`r format_decimal(subset(set_a_key_lag_table, Estimator == "Within-unit centered covariance" & Lag == 1)$Covariance, 3)`.
At lag 15 the estimates are
`r format_decimal(subset(set_a_key_lag_table, Estimator == "Pairwise difference" & Lag == 15)$Covariance, 3)`
and
`r format_decimal(subset(set_a_key_lag_table, Estimator == "Within-unit centered covariance" & Lag == 15)$Covariance, 3)`,
respectively. Thus within-unit centering removes most of the persistent positive
component and produces negative mean covariance at longer lags. The two
procedures should not be treated as approximately interchangeable estimates of
the same matrix without an explicit centering correction.
</div>

This is still an observational null-enrichment exercise. A gene with no FASH
dynamic discovery is not known to be a true dynamic null, and the trajectory
gallery below is included specifically to audit that concern.

## Set A design

The analysis starts from the complete BF-adjusted FASH(1) fit with
`r format_integer(pilot_configuration$n_fash_pairs)` gene-variant pairs from
`r format_integer(pilot_configuration$n_genes)` genes. Set A excludes every
gene with at least one FASH dynamic discovery under the same cumulative-lfdr
FDR rule at alpha 0.05 as the real-data analysis.

```{r selection-table, results='asis'}
render_scrollable_table(
  selection_table,
  align = "llrrrr",
  minimum_width = "1120px",
  digits = 0L
)
```

For each eligible gene, candidate variants are restricted to variants tested
with only that one gene in the complete FASH catalog. A fixed random score then
chooses one variant per gene and 200 genes. The score does not use beta
estimates, standard errors, lfdr values, p-values, or eFDR values. There are no
duplicated genes or variants in Set A.

```{r discovery-table, results='asis'}
render_scrollable_table(
  discovery_table,
  align = "lr",
  minimum_width = "720px",
  digits = 4L
)
```

The selected pairs were not required to have high individual lfdr values.
Their lfdr distribution is shown as a selection audit.

```{r lfdr-summary, results='asis'}
render_scrollable_table(
  selection_lfdr_summary,
  align = "lrrrrrr",
  minimum_width = "920px",
  digits = 4L
)
```

## Set A trajectory gallery {.tabset .tabset-pills}

The 200 selected units are ranked from highest to lowest BF-adjusted lfdr and
split into four equal groups. Each page shows a fixed-random sample of 25 units
from one group: page 1 is sampled from ranks 1--50 and page 4 from ranks
151--200. Blue points and lines are fitted $\widehat\beta_j(t)$ values, grey
bars are plus or minus one adjusted SE, and the orange dashed line is the
inverse-variance weighted constant fit. Each panel has its own y-axis, so the
gallery is intended for shape inspection rather than cross-panel amplitude
comparison.

### Page 1: highest-lfdr quartile

```{r set-a-gallery-1, fig.width=13, fig.height=13, fig.cap="Twenty-five fixed-random Set A units from the highest BF-adjusted-lfdr quartile."}
plot_set_a_gallery_page(1L)
```

```{r set-a-gallery-table-1, results='asis'}
render_scrollable_table(
  set_a_gallery_table(1L),
  align = "rrllrr",
  minimum_width = "960px",
  digits = 4L
)
```

### Page 2: second lfdr quartile

```{r set-a-gallery-2, fig.width=13, fig.height=13, fig.cap="Twenty-five fixed-random Set A units from the second BF-adjusted-lfdr quartile."}
plot_set_a_gallery_page(2L)
```

```{r set-a-gallery-table-2, results='asis'}
render_scrollable_table(
  set_a_gallery_table(2L),
  align = "rrllrr",
  minimum_width = "960px",
  digits = 4L
)
```

### Page 3: third lfdr quartile

```{r set-a-gallery-3, fig.width=13, fig.height=13, fig.cap="Twenty-five fixed-random Set A units from the third BF-adjusted-lfdr quartile."}
plot_set_a_gallery_page(3L)
```

```{r set-a-gallery-table-3, results='asis'}
render_scrollable_table(
  set_a_gallery_table(3L),
  align = "rrllrr",
  minimum_width = "960px",
  digits = 4L
)
```

### Page 4: lowest-lfdr quartile

```{r set-a-gallery-4, fig.width=13, fig.height=13, fig.cap="Twenty-five fixed-random Set A units from the lowest BF-adjusted-lfdr quartile."}
plot_set_a_gallery_page(4L)
```

```{r set-a-gallery-table-4, results='asis'}
render_scrollable_table(
  set_a_gallery_table(4L),
  align = "rrllrr",
  minimum_width = "960px",
  digits = 4L
)
```

On visual inspection, the highest-lfdr page contains many trajectories that
are compatible with noisy fluctuations relative to their SE bars, but it is
not uniformly featureless. The lower-lfdr pages contain several visibly
structured, monotone, or endpoint-driven trajectories; some of the most abrupt
patterns also have very large SEs. This reinforces the intended interpretation:
Set A is null-enriched, not an observed collection of true constant-effect
nulls.

## Two Set A estimators

Write the fitted time-specific effect estimate as

$$
\widehat\beta_j(t)=\beta_j(t)+s_j(t)\varepsilon_j(t),
\qquad
\operatorname{Cov}\{\varepsilon_j(t),\varepsilon_j(s)\}=C_{ts}.
$$

For Set A, the working dynamic null is $\beta_j(t)=\mu_j$. With

$$
x_{j,ts}=\frac{2s_{jt}s_{js}}{s_{jt}^2+s_{js}^2},
\qquad
q_{j,ts}=\frac{
\{\widehat\beta_j(t)-\widehat\beta_j(s)\}^2
}{s_{jt}^2+s_{js}^2},
$$

the constant $\mu_j$ cancels and
$E(1-q_{j,ts})=x_{j,ts}C_{ts}$. The pairwise estimate is

$$
\widehat C_{ts}^{\mathrm{pair}}
=\frac{\sum_jx_{j,ts}(1-q_{j,ts})}{\sum_jx_{j,ts}^2},
\qquad \widehat C_{tt}^{\mathrm{pair}}=1.
$$

The within-unit method first estimates and removes a weighted constant,

$$
\widehat\mu_j=
\frac{\sum_t\widehat\beta_j(t)/s_j^2(t)}
{\sum_t1/s_j^2(t)},
\qquad
r_j(t)=\frac{\widehat\beta_j(t)-\widehat\mu_j}{s_j(t)},
$$

and then computes the raw empirical second-moment matrix

$$
\widehat C^{\mathrm{center}}
=\frac{1}{M}\sum_{j=1}^M r_jr_j^T.
$$

This second matrix is intentionally not rescaled to unit diagonal. The
correlation matrix of the columns of $r$ is also shown separately, but it is
not called a covariance estimate.

The centering transformation changes the estimand. If
$D_j=\operatorname{diag}\{s_j(t)\}$ and $w_j$ is the normalized
inverse-variance weight vector, then under a constant true effect

$$
r_j=A_j\varepsilon_j,
\qquad
A_j=D_j^{-1}(I-\mathbf 1w_j^T)D_j,
$$

so $E(r_jr_j^T)=A_j C A_j^T$, not $C$. This operation can remove a common
residual mode as well as the unknown intercept. The discrepancy therefore need
not equal a small universal negative offset.

All matrices below are raw. No nearest-positive-definite projection was
applied.

## Set A covariance comparison

```{r set-a-matrix-comparison, fig.width=12, fig.height=4.4, fig.cap="Set A raw matrix estimates. Pairwise covariance has a fixed unit diagonal; centered covariance retains its empirical diagonal; centered correlation rescales that same centered residual matrix to unit diagonal."}
plot_set_a_matrix_comparison()
```

Both covariance matrices happen to be positive definite in this pilot, although
positive definiteness was not imposed.

```{r set-a-centered-diagnostics, results='asis'}
render_scrollable_table(
  set_a_centered_diagnostic_table,
  align = "llrrrrrrrr",
  minimum_width = "1220px",
  digits = 4L
)
```

The off-diagonal patterns are ordered similarly, but their levels differ
substantially: their entrywise Pearson correlation is
`r format_decimal(set_a_method_agreement$Value[1], 3)`, while the mean
centered-minus-pairwise difference is
`r format_decimal(set_a_method_agreement$Value[2], 3)`.

```{r set-a-entry-agreement, fig.width=7.4, fig.height=6.4, fig.cap="Off-diagonal Set A covariance entries. Color denotes time lag; the dashed line is equality."}
plot_set_a_covariance_entry_agreement()
```

```{r set-a-agreement-table, results='asis'}
render_scrollable_table(
  set_a_method_agreement,
  align = "lr",
  minimum_width = "760px",
  digits = 4L
)
```

## Set A lag covariance and variogram

For either raw covariance matrix, the lag-$h$ semivariogram is

$$
\widehat\gamma(h)
=\frac{1}{16-h}\sum_{t=0}^{15-h}
\frac{\widehat C_{tt}+\widehat C_{t+h,t+h}
-2\widehat C_{t,t+h}}{2}.
$$

The shaded bands are percentile 95% intervals from the same 200 gene-level
bootstrap resamples for both estimators.

```{r set-a-lag-covariance, fig.width=10, fig.height=6, fig.cap="Set A lag-averaged standardized covariance with paired gene-bootstrap 95% intervals."}
plot_set_a_covariance_comparison()
```

```{r set-a-variogram, fig.width=10, fig.height=6, fig.cap="Set A semivariograms with paired gene-bootstrap 95% intervals. The centered variogram uses the empirical centered diagonal rather than one minus centered covariance."}
plot_set_a_variogram_comparison()
```

The variograms look closer than the covariance curves because the centered
matrix also has a smaller diagonal: its mean diagonal is
`r format_decimal(mean(diag(set_a_centered_covariance)), 3)`, rather than one.
The centered variogram therefore combines a lower diagonal with lower
off-diagonal covariance; it is not $1$ minus the orange covariance curve.

```{r set-a-key-lag-table, results='asis'}
render_scrollable_table(
  set_a_key_lag_table,
  align = "lrrrrrrrr",
  minimum_width = "1160px",
  digits = 4L
)
```

For reference only, rescaling the centered residual columns to unit variance
gives the following lag-averaged correlations. This rescaling does not undo the
centering projection.

```{r set-a-centered-correlation-table, results='asis'}
render_scrollable_table(
  set_a_centered_correlation_table,
  align = "rrr",
  minimum_width = "720px",
  digits = 4L
)
```

## Legacy Set B context (not interpreted as harmonized evidence)

<div class="null-pilot-banner">
Set B was constructed from existing time-specific eQTL tables that used three
time-specific PCs. The current FASH workflow uses five time-specific PCs. Set B
is retained here only to preserve the earlier pilot record; it is not used as
primary evidence until the time-specific screen is rerun under the matched
five-PC workflow.
</div>

```{r standardized-covariance-heatmaps, fig.width=12, fig.height=4.3, fig.cap="Earlier pilot context: Set A pairwise and legacy non-harmonized Set B estimates."}
plot_standardized_covariance_heatmaps()
```

```{r legacy-variograms, fig.width=10, fig.height=6, fig.cap="Earlier pilot lag semivariograms. Set B remains visible only as legacy non-harmonized context."}
plot_variograms()
```

```{r legacy-lag-covariances, fig.width=10, fig.height=6, fig.cap="Earlier pilot lag-averaged standardized covariance."}
plot_lag_covariances()
```

```{r legacy-set-b-agreement, fig.width=7.4, fig.height=6.4, fig.cap="Earlier Set B direct-versus-pairwise comparison; not interpreted as harmonized evidence."}
plot_set_b_estimator_agreement()
```

## Interpretation and limitations

1. The gallery is an audit of the selected observations, not a proof that Set A
   contains only true dynamic nulls. Nondiscovery can include weak or
   underpowered dynamic effects.

2. Within-unit centering does not merely introduce a small generic negative
   correlation. It applies a unit-specific projection because the adjusted SEs
   differ across time and units. In these data, that projection removes most of
   the persistent positive component estimated by pairwise differences.

3. The two covariance profiles are therefore evidence about estimator
   sensitivity, not two nearly identical confirmations of one residual matrix.
   The pairwise result could reflect genuine residual covariance, violations of
   the constant-effect null among selected units, or both.

4. This remains a 200-pair internal pilot. No result on this page has been
   moved into R4, the manuscript, or the response letter.

## Reproduction

```{bash run-observational-null-set-pilot, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/covariance_estimation/run_observational_null_set_pilot.R \
  --pilot-size 200 \
  --bootstrap-reps 200 \
  --selection-seed 20260806 \
  --bootstrap-seed 20260816 \
  --output-id observational_null_set_pilot
```

```{bash test-observational-null-set-pilot, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/covariance_estimation/test_observational_null_set_helpers.R
```

```{bash build-observational-null-set-pilot-page, eval=FALSE}
Rscript --vanilla -e \
  'workflowr::wflow_build("analysis/revision_internal_observational_null_set_pilot.rmd")'
```
