---
title: "Temporary internal analysis: R3 truth-specification pilot"
author: "Ziang Zhang"
date: "2026-08-24"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 10,
  fig.height = 6,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

project_root <- if (file.exists(
  "code/revision_simulations/shared/simulation_functions.R"
)) {
  "."
} else if (file.exists(
  "../code/revision_simulations/shared/simulation_functions.R"
)) {
  ".."
} else {
  stop("Could not locate the workflowr project root.")
}
pilot_root <- file.path(
  project_root,
  "output",
  "revision_simulations",
  "internal"
)
seed_list <- c(12345L, 22345L, 32345L, 42345L, 52345L)
designs <- c(
  r3a_targeted_broad_bspline_j300 =
    "R3A: functionally separated broad cubic B-spline",
  r3a_constrained_global_bspline_j300 =
    "R3A: constrained global random cubic B-spline",
  r3b_interior_single_lobe_j300 =
    "R3B: target-contained single raised cosine"
)

read_design <- function(prefix) {
  paths <- file.path(
    pilot_root,
    paste0(prefix, "_seed", seed_list),
    "pilot_result.rds"
  )
  if (any(!file.exists(paths))) {
    stop("The five-seed internal pilot cache is incomplete for ", prefix, ".")
  }
  results <- lapply(paths, readRDS)
  configurations <- lapply(results, `[[`, "configuration")
  reference <- configurations[[1L]]
  if (!all(vapply(configurations, function(configuration) {
    identical(configuration$J, reference$J) &&
      identical(configuration$truth_settings, reference$truth_settings) &&
      identical(configuration$middle_definition, reference$middle_definition) &&
      identical(configuration$package_provenance, reference$package_provenance)
  }, logical(1)))) {
    stop("The seed-level pilot configurations are not comparable.")
  }
  alpha <- do.call(rbind, Map(function(result, seed) {
    rows <- result$functional_alpha
    rows$seed <- seed
    rows
  }, results, seed_list))
  list(configuration = reference, alpha = alpha)
}

pilot <- lapply(names(designs), read_design)
names(pilot) <- names(designs)

summarize_alpha <- function(alpha, design_name, alpha_value = 0.05) {
  rows <- alpha[
    alpha$method == "FASH-IWP1-BF" &
      abs(alpha$alpha - alpha_value) < 1e-12,
    ,
    drop = FALSE
  ]
  summary <- do.call(rbind, lapply(split(rows, rows$target), function(target_rows) {
    discoveries <- sum(target_rows$n_discoveries)
    false_discoveries <- sum(target_rows$false_discoveries)
    data.frame(
      design = design_name,
      target = target_rows$target[[1L]],
      mean_power = mean(target_rows$power),
      mean_empirical_fsr = mean(target_rows$empirical_fsr),
      maximum_seed_empirical_fsr = max(target_rows$empirical_fsr),
      pooled_discoveries = discoveries,
      pooled_false_discoveries = false_discoveries,
      pooled_empirical_fsr = if (discoveries == 0L) 0 else {
        false_discoveries / discoveries
      },
      stringsAsFactors = FALSE
    )
  }))
  summary[match(c("early", "middle", "late", "switch"), summary$target), ]
}

alpha005_summary <- do.call(rbind, Map(
  function(design_name, object) summarize_alpha(object$alpha, design_name),
  names(designs),
  pilot
))
alpha005_summary$design <- designs[alpha005_summary$design]

r3a_seed_middle <- pilot$r3a_constrained_global_bspline_j300$alpha
r3a_seed_middle <- r3a_seed_middle[
  r3a_seed_middle$method == "FASH-IWP1-BF" &
    r3a_seed_middle$target == "middle" &
    abs(r3a_seed_middle$alpha - 0.05) < 1e-12,
  c("seed", "n_discoveries", "false_discoveries", "empirical_fsr", "power")
]
```

## Scope

> **Internal exploratory analysis.** This page is a prespecified-truth pilot,
> not a formal replacement for the R3 cache and not evidence of general FSR
> control. Its purpose is to determine whether a full five-seed, real-genotype
> `J = 6362` calibration run is scientifically justified.

The two designs use the same five R3 genotype-cache seeds, `19` donors, `16`
observed times, five covariates, expression noise standard deviation `1`, and
the pinned `fashr` 0.1.43 revision. For computational triage only, the pilot
uses the first `J = 300` gene--variant pairs from each cached seed. The dynamic,
constant, and zero-effect class proportions are `0.20`, `0.40`, and `0.40`.

## R3B: a strictly target-contained compact sensitivity

The raised-cosine **deviation** has one lobe only. Its center ranges are
Early `1.5`, Middle `4.5--10.5`, and Late `13.5`; the half-width is `1.2`.
The resulting lobe supports are Early `[0.3, 2.7]`, Middle `[3.3, 11.7]` at
the two extreme Middle centers, and Late `[12.3, 14.7]`. Thus all primary
deviation support is strictly inside the target window. The full genetic
effect is not compactly supported because it also includes a sampled main
effect; this distinction is intentional and reported explicitly.

The proposal therefore resolves a semantic ambiguity in the compact-truth
construction. It does not make a calibration claim. In particular, allowing
two or three lobes would contradict a claim that the complete deviation is
target-contained, so this sensitivity fixes the lobe count to one.

## R3A alternatives: local and global broad cubic B-spline truth

The R3A candidate remains a non-IWP cubic B-spline construction, but uses a
prespecified acceptance rule on the dense grid rather than assigning a label
solely because one broad random curve happened to have a favorable maximum.
For each dynamic curve, generation requires:

- target functional contrast at least `0.60`;
- target-to-outside absolute peak ratio at least `1.40`;
- exactly one effective sign transition for a switch truth;
- no effective sign transition for a non-switch truth;
- a same-signed non-switch baseline of at least `0.75` of the primary
  amplitude, with a secondary background at most `0.05` of that amplitude.

These constraints were fixed before the `J = 300` run. Two R3A mechanisms are
evaluated under the same contract. The first is a targeted local broad lobe,
which is useful for a deliberately label-clean sensitivity but can be accused
of making the timing task easier. The second first samples a global random
cubic B-spline with six iid coefficient draws, rescales its centered RMS to
`0.90`, and accepts it only if it satisfies the same functional-separation
conditions. For the global mechanism, a switch requires at least one effective
sign transition rather than exactly one: this matches the stated FASH switch
functional, which requires sufficiently large effects of both signs but does
not define a unique crossing count.

The constrained global version is the more conservative primary candidate: it
retains broad global B-spline geometry, while avoiding labels that are only
weakly separated from the competing timing functionals. The original R3A
remains necessary as a misspecification stress test.

## Five-seed pilot results at alpha 0.05

```{r alpha005-table, echo=FALSE}
display <- alpha005_summary
display$design <- sub("R3[A-B]: ", "", display$design)
display$target <- tools::toTitleCase(display$target)
display$mean_power <- formatC(display$mean_power, format = "f", digits = 3)
display$mean_empirical_fsr <- formatC(
  display$mean_empirical_fsr,
  format = "f",
  digits = 3
)
display$maximum_seed_empirical_fsr <- formatC(
  display$maximum_seed_empirical_fsr,
  format = "f",
  digits = 3
)
display$pooled_empirical_fsr <- formatC(
  display$pooled_empirical_fsr,
  format = "f",
  digits = 3
)
names(display) <- c(
  "Design", "Target", "Mean power", "Mean seed-level FSR",
  "Maximum seed FSR", "Pooled calls", "Pooled false calls", "Pooled FSR"
)
knitr::kable(
  display,
  align = "l",
  caption = paste(
    "Table 1. Five-seed real-genotype J=300 exploratory BF results.",
    "Pooled quantities are descriptive; Monte Carlo replication remains the seed."
  )
)
```

The strict R3B sensitivity has only three Middle calls in total, one of which
is false. It is therefore unsuitable as a primary calibration scenario despite
its cleaner compact-support geometry. The local broad R3A pilot has `51`
Middle calls and one false call. The more conservative constrained-global R3A
pilot has `79` Middle calls and two false calls, with mean seed-level empirical
FSR `0.0251` and mean Middle power `0.770`. Its switch FSR is `0.0448`, close
to nominal `0.05`; this is why a full-scale confirmation remains necessary.

```{r r3a-middle-by-seed, echo=FALSE}
r3a_seed_middle$empirical_fsr <- formatC(
  r3a_seed_middle$empirical_fsr,
  format = "f",
  digits = 3
)
r3a_seed_middle$power <- formatC(r3a_seed_middle$power, format = "f", digits = 3)
names(r3a_seed_middle) <- c(
  "Seed", "Middle calls", "False Middle calls", "Empirical FSR", "Power"
)
knitr::kable(
  r3a_seed_middle,
  align = "r",
  caption = paste(
    "Table 2. R3A constrained-global-B-spline Middle results by seed",
    "at alpha 0.05."
  )
)
```

## Prespecified full-scale decision rule

No truth parameter will be tuned further using this pilot. The next analysis,
if submitted, will use all `6,362` pairs and the same five seeds. It will
report Raw and BF results for every target and alpha in `0.005, 0.010, ..., 0.200`.

The candidate will be rejected as a primary calibration scenario if, for the
BF fit, the maximum across Early, Middle, Late, Switch, and alpha values from
`0.05` through `0.20` of

$$
\overline{\operatorname{FSR}}_{\mathrm{empirical}} - \alpha
$$

exceeds `0.03`. Power is descriptive rather than an acceptance gate. The
single-lobe R3B design remains a labelled low-power sensitivity regardless of
this decision. The full-scale primary candidate is the constrained-global R3A
mechanism, not the local targeted-lobe sensitivity.

## Reproduction

The local five-seed pilot is produced with
`code/revision_simulations/internal/pilots/run_r3_truth_specification_pilot.R`
and summarized with
`code/revision_simulations/internal/pilots/summarize_r3_truth_specification_pilot.R`.
The corresponding Slurm array entry point is
`code/revision_simulations/internal/pilots/run_r3_truth_specification_pilot.sbatch`.
