---
title: "Internal analysis: FASH versus FASH-CL variant enrichment"
author: "Ziang Zhang"
date: "2026-08-08"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 10.5,
  fig.height = 8.0,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(file.path(
  "code",
  "revision_simulations",
  "internal",
  "fash_cl_variant_enrichment_comparison",
  "reporting.R"
))
```

<style>
.fash-cl-internal-banner {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border: 1px solid #d9a441;
  border-left: 6px solid #d9a441;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.fash-cl-key-result {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.fash-cl-caution {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #cc7a00;
  border-radius: 4px;
  background: #fff8ec;
  color: #563a0c;
}
.fash-cl-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.fash-cl-table-scroll table {
  margin-bottom: 1em;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
pre {
  max-width: 100%;
  overflow-x: auto;
}
</style>

<div class="fash-cl-internal-banner">
<strong>Internal exploratory page.</strong> This page is intentionally absent
from the workflowr homepage, navigation, manuscript, response letter, and the
two existing internal enrichment pages. It reads retained caches and does not
replace the current FASH analysis.
</div>

## Question and provisional answer

This analysis asks two related questions:

1. within FASH-CL, is enhancer enrichment stronger after selecting one lead
   discovered variant per gene rather than retaining all discovered variants?;
2. under each selection strategy, does current FASH or FASH-CL have stronger
   enhancer enrichment?

Current FASH adjusts expression for five PCs estimated separately at each time
point. FASH-CL adjusts for five PCs estimated after collapsing expression to
the cell-line level. Both comparisons use the order-1 FASH dynamic-eQTL test,
the BF-based prior correction, and cumulative-lfdr FDR 0.05.

<div class="fash-cl-key-result">
<strong>Provisional readout.</strong>
`r lead_pattern_answer`
`r method_comparison_answer`
Thus the one-lead pattern extends to FASH-CL, but the head-to-head method result
is mixed rather than a uniform FASH-CL or current-FASH advantage.
</div>

No pre-specified enhancer reaches within-set BH q below 0.05; the smallest is
`r format_decimal(smallest_enhancer_q, 3L)`. These are exploratory point-estimate
patterns, not thresholded discoveries or formal tests of a difference between
methods.

## Discovery sets and method overlap

For each method, the all-discovery set is collapsed to unique rsIDs. The
one-lead set retains the discovered pair with the lowest lfdr for each gene,
using rsID as the deterministic tie-breaker, and is then deduplicated to unique
rsIDs for enrichment.

```{r discovery-table, results='asis'}
render_scrollable_table(
  discovery_table,
  align = "llrrr",
  minimum_width = "880px"
)
```

```{r overlap-table, results='asis'}
render_scrollable_table(
  overlap_table,
  align = "lrrrrr",
  minimum_width = "820px"
)
```

FASH-CL yields fewer calls: 5,393 pairs and 5,361 unique variants, compared
with 9,205 pairs and 9,139 unique variants for current FASH. The one-lead sets
contain 686 and 1,170 unique variants, respectively. Their low Jaccard overlap
shows that this is not merely the same discoveries with slightly different
scores; the PC construction materially changes set membership.

## BF-adjustment validation

The original FASH-CL fit did not retain a BF-adjusted object. The analysis
therefore recomputes the same two mixture-optimization stages used by
`fashr::BF_update()`, while discarding unused portions of the 2.4-GiB fit before
likelihood expansion. This keeps the calculation laptop-scale.

The stored current-FASH lfdr is reconstructed from its retained null and
conditional-alternative weights to numerical precision. Re-running the
historical mixture optimizer produces nearly identical, but not bitwise
identical, threshold-edge calls; the retained current adjusted fit remains the
authoritative current-FASH result.

```{r bf-validation-table, results='asis'}
render_scrollable_table(
  bf_validation_table,
  align = "lr",
  minimum_width = "760px"
)
```

## Annotation coverage

The custom regulatory matrix covers the entire tested-variant universe.
baselineLD v2.2 uses its 1000G EUR rsID coverage, so every discovery set and
the matched-control universe are restricted before matching.

```{r coverage-table, results='asis'}
render_scrollable_table(
  coverage_table,
  align = "lllrrr",
  minimum_width = "980px"
)
```

## Pre-specified enhancer comparison

The custom panel contains ENCODE enhancer-like cCRE plus E020, E013, and E095
Roadmap enhancers. The baselineLD panel contains Andersson enhancer, Hoffman
enhancer, Hoffman weak enhancer, Hnisz super-enhancer, and Villar human
enhancer. These categories were fixed before examining the FASH-CL results.

Each selected variant is matched to five non-selected tested variants on
chromosome, cohort MAF, minimum target-gene TSS distance, local tested-variant
density, and number of tested genes. Matching is repeated over 100 fixed seeds.
Intervals are delete-one-autosome jackknife 95% intervals.

```{r enhancer-enrichment, fig.width=11.2, fig.height=8.6, fig.cap="Figure 1. Pre-specified enhancer enrichment for current FASH and FASH-CL, using all discoveries or one lead variant per gene. Enrichment is relative to separately matched tested variants. Error bars are delete-one-autosome 95% intervals. The FASH-CL one-lead Andersson estimate is omitted because fewer than ten selected variants overlap that sparse annotation."}
plot_enhancer_enrichment()
```

For FASH-CL, the one-lead strategy raises the median enhancer point estimate in
both systems. In baselineLD, FASH-CL Hoffman enhancer is
`r format_decimal(enhancer_results$enrichment[enhancer_results$annotation_system == "baselineLD v2.2" & enhancer_results$method == "FASH-CL" & enhancer_results$selection_strategy == "One lead variant per gene" & enhancer_results$annotation == "Enhancer_Hoffman"], 2L)`-fold
with a fold-scale interval of
`r format_decimal(2^enhancer_results$ci_lower[enhancer_results$annotation_system == "baselineLD v2.2" & enhancer_results$method == "FASH-CL" & enhancer_results$selection_strategy == "One lead variant per gene" & enhancer_results$annotation == "Enhancer_Hoffman"], 2L)`--`r format_decimal(2^enhancer_results$ci_upper[enhancer_results$annotation_system == "baselineLD v2.2" & enhancer_results$method == "FASH-CL" & enhancer_results$selection_strategy == "One lead variant per gene" & enhancer_results$annotation == "Enhancer_Hoffman"], 2L)`
and q =
`r format_decimal(enhancer_results$q_value_within_set[enhancer_results$annotation_system == "baselineLD v2.2" & enhancer_results$method == "FASH-CL" & enhancer_results$selection_strategy == "One lead variant per gene" & enhancer_results$annotation == "Enhancer_Hoffman"], 3L)`.

The method comparison depends on the panel. FASH-CL is descriptively stronger
for all discoveries in both panels and for baselineLD under one-lead selection.
In the custom one-lead panel, current FASH and FASH-CL are close: current FASH
is slightly higher by the panel median and for E013, whereas FASH-CL is higher
for ENCODE enhancer-like cCRE and E020.

```{r enhancer-summary-table, results='asis'}
render_scrollable_table(
  enhancer_summary_table,
  digits = 4L,
  minimum_width = "1260px"
)
```

### Exact enhancer estimates

```{r enhancer-numeric-table, results='asis'}
render_scrollable_table(
  enhancer_numeric_table,
  digits = 4L,
  minimum_width = "1580px"
)
```

## Whole-annotation comparison

The identity line below provides a descriptive current-versus-CL comparison
for every annotation. Points above the line have a larger FASH-CL estimate;
points below have a larger current-FASH estimate. Because the discovery sets
differ in size and membership, this plot is not a paired significance test.

```{r method-scatter, fig.width=10.8, fig.height=8.2, fig.cap="Figure 2. Current-FASH versus FASH-CL point estimates across all retained annotations. Orange points are the pre-specified enhancer categories. The dashed identity line marks equal point estimates."}
plot_method_scatter()
```

<details>
<summary><strong>Complete numerical results for both annotation systems</strong></summary>

```{r full-results-table, results='asis'}
render_scrollable_table(
  full_results_table,
  digits = 5L,
  minimum_width = "1580px"
)
```

</details>

## Matching diagnostics and local scale

```{r matching-balance, fig.width=10.8, fig.height=7.4, fig.cap="Figure 3. Matching balance over 100 deterministic seeds for both methods, selection strategies, and annotation systems. Dotted lines at +/-0.1 are diagnostic references."}
plot_matching_balance()
```

The maximum absolute standardized mean difference is
`r format_decimal(cache$maximum_absolute_smd, 4L)`, below the pre-specified
0.10 diagnostic boundary.

```{r runtime-table, results='asis'}
render_scrollable_table(
  runtime_table,
  align = "lr",
  minimum_width = "700px"
)
```

The complete analysis through cache assembly took
`r format_decimal(lookup_runtime("analysis_through_cache_assembly"), 1L)`
seconds locally. Existing annotation caches avoid re-downloading or reparsing
the baselineLD archive.

## Interpretation boundaries

<div class="fash-cl-caution">
<strong>Important limitations.</strong> FASH and FASH-CL use the same biological
study and tested pair universe, and the two annotation systems evaluate the
same discovery sets. They are robustness views, not independent replications.
The method sets have different sizes and low overlap, and each is matched to
its own controls. Consequently, a larger point estimate does not by itself
establish that one method is significantly more enriched than the other.
</div>

- The one-lead analysis changes the estimand from variant-weighted discovery
  enrichment toward equal weighting of discovered genes. It also reduces
  repeated-locus influence and increases sampling uncertainty.
- The baselineLD Andersson category is too sparse for a finite FASH-CL
  one-lead estimate under the pre-specified minimum of ten selected overlaps.
- None of the pre-specified enhancer categories passes within-set BH q below
  0.05. The leading results should therefore be described as suggestive.
- Annotation overlap supports functional plausibility; it does not establish
  causal variants or validate discovery FDR.

## Reproduction

The scientific settings below are shown but not evaluated while rendering.
The page reads the completed retained cache.

```{r reproduction, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/fash_cl_variant_enrichment_comparison/\
test_fash_cl_variant_enrichment_helpers.R

Rscript --vanilla \
  code/revision_simulations/internal/fash_cl_variant_enrichment_comparison/\
run_fash_cl_variant_enrichment_comparison.R

Rscript --vanilla -e \
  'workflowr::wflow_build("analysis/revision_internal_fash_cl_variant_enrichment_comparison.rmd", local = TRUE)'
```

<details>
<summary><strong>Retained input provenance</strong></summary>

```{r provenance-table, results='asis'}
render_scrollable_table(
  input_provenance,
  align = "lrrl",
  minimum_width = "1280px"
)
```

</details>
