---
title: "Revision internal: FASH target-decoy (fash-knockoff) permutation check"
author: "Ziang Zhang"
date: "2026-08-24"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 2
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9.6,
  fig.height = 6.0,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(file.path(
  "code", "revision_simulations", "internal", "fash_knockoff",
  "fash_knockoff_helpers.R"
))
source(file.path(
  "code", "revision_simulations", "internal", "fash_knockoff", "reporting.R"
))
library(ggplot2)

statistics <- load_fash_knockoff_statistics(".")
q_grid <- c(0.01, 0.05, 0.10, 0.15, 0.20, 0.30, 0.50)
show_table <- function(x, caption, digits = 4) {
  knitr::kable(x, caption = caption, digits = digits, row.names = FALSE)
}
```

<style>
.fk-banner {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border: 1px solid #d9a441;
  border-left: 6px solid #d9a441;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.fk-result {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.fk-warn {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #b02418;
  border-radius: 4px;
  background: #fdeeec;
  color: #5c1a13;
}
p.caption { color: #243447; font-weight: 650; text-align: left; }
.main-container img { max-width: 100%; height: auto; }
pre { max-width: 100%; overflow-x: auto; }
</style>

<div class="fk-banner">
**Internal exploratory page.** This is an empirical calibration check, not a
formal reviewer deliverable, and it is deliberately absent from
`analysis/_site.yml` and `analysis/index.Rmd`. It does not claim
finite-sample FDR control: exact target-decoy exchangeability is not
established here, so the procedure is a knockoff-*inspired* permutation
check rather than a knockoff procedure.
</div>

## What the procedure does

For each of the tested gene-variant pairs a **matched permuted decoy** is
built, and target and decoy units are fitted **jointly**, so a single
empirical-Bayes prior scores both sides of every pair.

1. **One global donor permutation.** A single donor row map is applied to the
   entire genotype matrix, `G_pi = P G`, shared by all 745,867 variants and all
   16 time points. Row permutation leaves every variant's dosage multiset, and
   therefore its MAF, exactly unchanged, and leaves the full LD structure
   unchanged because `(PG)' (PG) = G' P' P G = G' G` identically.
2. **The map is a derangement.** A fixed point would leave one donor's decoy
   genotype equal to its own, so that donor would keep feeding real signal into
   the decoy. The realised map is a 17-cycle plus a 2-cycle, with no fixed
   points.
3. **Expression PCs are not permuted.** PC1-PC5 are functions of the expression
   matrix; permuting them relative to the expression would make the decoy
   regression a different model from the target regression.
4. **The original model is refit** at each time point,
   `Y ~ 1 + G_permuted + PC1 + ... + PC5`, and the raw standard errors are
   converted to the same t-to-normal scale used by the primary analysis.
5. **Target likelihood rows are reused, never recomputed**, from the published
   fit. `BF_update()` does not modify `L_matrix`, so those rows are
   bit-identical whichever fit is supplied. Only the decoy units required new
   likelihood evaluations.
6. **The statistic** is `W_j = log BF_j - log BF_j,pi`, both Bayes factors
   scored by the merged prior. Selection uses knockoff+,
   `T_q = min{t > 0 : (1 + #{W <= -t}) / (#{W >= t} v 1) <= q}`.

Because the two sides share one prior, swapping a pair leaves the
`2J`-unit multiset unchanged, so the fitted prior is unchanged and only that
pair's `W` flips sign. This antisymmetry is asserted in
`test_fash_knockoff_helpers.R` rather than assumed.

The decoy is **not** a simulated diagonal-Gaussian null. It is built from the
same longitudinal expression measurements, so it retains the gene-specific
temporal dependence induced by repeated measurements on the same cell lines.
That is the point of the design, and, as the results below show, also the
reason it has limited power.

## Headline result: where the competition statistic sits

```{r headline, fig.height = 5.8, fig.cap = "Competition statistic by published discovery status. Panel a draws the outlying points, because the tails on both sides are where knockoff+ actually operates."}
plot_w_by_discovery(statistics)
```

```{r headline-table}
show_table(
  fash_knockoff_group_summary(statistics),
  "Distribution of W by published discovery status."
)
```

<div class="fk-result">
Two things matter in this figure.

**The non-discovered box is centred on zero.** Median `W` is
`r sprintf("%.5f", median(statistics$W_updated_weights[!statistics$discovered]))`
and `W > 0` in
`r sprintf("%.2f%%", 100 * mean(statistics$W_updated_weights[!statistics$discovered] > 0))`
of the
`r format(sum(!statistics$discovered), big.mark = ",")`
pairs the published rule did not call. This is the strongest available check on
the whole construction: on pairs with no claimed signal, target and decoy are
behaving exchangeably. For contrast, earlier residual-block decoy constructions
sat at 0.582 on a near-null pair set, about 13 sd from 0.5.

**But the non-discovered tail reaches further than the discovered one.** The
decoy beats the target by as much as
`r sprintf("%.1f", -min(statistics$W_updated_weights[!statistics$discovered]))`
log units among pairs that were never called. Those are pure artefact, because
the decoy is null by construction, and they are what consumes the FDR budget.
</div>

<div class="fk-warn">
**This figure does not validate the calls.** The grouping conditions on the
target: the published rule selects pairs with a high target Bayes factor, and a
high target BF mechanically forces `W > 0` unless the decoy is also high. A null
pair that reached a high BF by chance would sit in the same box. The figure
locates `W` by discovery status; it is descriptive, not confirmatory. The same
caveat applies to any statistic computed after conditioning on the published
calls.
</div>

## The lfdr distributions side by side

Both sets of lfdr values come from the same merged fit, so any difference
between the arms is a difference in evidence, not in calibration.

```{r lfdr-bands, fig.height = 5.0, fig.cap = "Left tail of the lfdr distribution for both arms. Separation exists only below about 1e-9."}
plot_lfdr_bands(statistics)
```

```{r lfdr-band-table}
show_table(
  fash_knockoff_lfdr_bands(statistics),
  "lfdr band counts and the decoy-to-target ratio.",
  digits = 3
)
```

<div class="fk-result">
For every 100 target pairs below `lfdr = 0.05`, the permuted decoy produces
about
`r sprintf("%.0f", 100 * sum(statistics$decoy_lfdr < 0.05) / sum(statistics$target_lfdr < 0.05))`.
The two arms separate only below roughly `1e-9`; between `1e-6` and `1e-4` the
decoy is marginally ahead.
</div>

## How much non-circular evidence is there for real signal?

The one quantity that conditions on nothing is the global excess of target
wins. Treating genes as independent clusters, and assuming nothing about
within-gene correlation:

```{r sign-evidence}
show_table(
  fash_knockoff_sign_evidence(statistics),
  "Sign-test evidence for a global excess of target wins.",
  digits = 4
)
```

<div class="fk-warn">
The naive `sqrt(n)` sign test badly overstates this. Pairs within a gene share
one expression trajectory and are in LD, so they are not independent: the
measured design effect is about 25, implying a within-gene sign correlation near
0.16. The cluster-robust value is the one to quote, and it agrees with the
independent gene-level test.
</div>

## knockoff+ selection

```{r selection}
show_table(
  fash_knockoff_selection_table(statistics, q_grid),
  "knockoff+ selection at pair level and at gene level.",
  digits = 4
)
```

Nothing survives `q = 0.05` at either level. The pair-level floor is reached at
`q = 0.15`; note that it is bounded below by `1/R`, so a 17-pair selection
cannot report a rate under about 0.06 whatever the data say.

```{r selected-pairs}
show_table(
  fash_knockoff_selected_pairs(statistics, 0.15),
  "The pairs selected at q = 0.15. Every one is a published call.",
  digits = 4
)
```

<div class="fk-result">
These are clean: target evidence of `10^10` to `10^15` against decoy Bayes
factors of about 1. The permutation does not move them.

They are also concentrated. The
`r nrow(fash_knockoff_selected_pairs(statistics, 0.15))`
pairs fall in only
`r length(unique(fash_knockoff_selected_pairs(statistics, 0.15)$gene_id))`
genes, and one gene contributes nine of them.
</div>

## Why the estimated FDR does not fall: the tails are LD-clustered

```{r tails}
show_table(
  fash_knockoff_tail_concentration(statistics, c(5, 10, 15, 20, 25)),
  "Tail counts as pairs and as distinct genes.",
  digits = 3
)
```

<div class="fk-warn">
Counting genes rather than pairs is the honest way to count independent events,
and it removes most of the apparent pair-level advantage. At a threshold of 10
there are **more** decoy genes than target genes. The pair-level estimate at
`q = 0.15` rests on a single decoy win, so its sampling variability is large;
collapsed to genes the same threshold gives roughly 0.33 rather than 0.118.
</div>

## Does a different gene-level statistic help?

Any antisymmetric function of the two Bayes factors fixes the sign pattern, so
only the ordering of `|W|` can change. Aggregating within a gene by a maximum
is a difference of two extreme-value variables and is the noisiest choice.

```{r gene-statistics}
show_table(
  fash_knockoff_gene_statistic_comparison(statistics, c(0.05, 0.20)),
  "Five gene-level aggregations of the same pair-level evidence.",
  digits = 3
)
```

Replacing the maximum with a mean roughly doubles the global `t` statistic, so
the aggregation choice does matter for detecting that signal exists. It does not
rescue selection: no aggregation reaches `q = 0.05`.

## Diagnostics that had to pass first

```{r gates}
paths <- fash_knockoff_paths(".")
show_table(
  read.csv(file.path(paths$summary, "genotype_covariate_alignment.csv"),
           check.names = FALSE),
  "Genotype variance explained by PC1-PC5, both arms, all 16 time points.",
  digits = 4
)
show_table(
  read.csv(file.path(paths$summary, "uninformative_variant_screen.csv")),
  "Variants carrying no residual genotype information, by time point and arm.",
  digits = 0
)
```

- **The cached target likelihood rows reproduce exactly** in the environment
  that computed the decoys: maximum absolute deviation `0` against a scale of
  13.1, on a 50-unit slice. Target and decoy rows therefore sit on identical
  numerical footing.
- **The regression pipeline reproduces the published beta and SE** from raw
  expression, PC and VCF inputs: maximum deviations `2.8e-13` and `1.6e-12` on
  a 2,000-pair slice. The decoy is built by the code path that reproduces the
  target.
- **The genotype-PC alignment gate is clean.** Mean `R^2` is 0.2847 for the
  observed genotype and 0.2863 for the permuted genotype, against a chance
  expectation near `5/18 = 0.278`. Had the observed arm been systematically
  higher, decoy standard errors would have been deflated and the competition
  would have been anti-conservative in a way nothing downstream could reveal.
- **The informativeness screen dropped 23 variants and 37 pairs**, 0.0037% of
  the total, all on the permuted arm and all at day 4. The screen is applied
  symmetrically to both arms and depends only on the design, never on an
  outcome; filtering on the decoy alone would break exchangeability.

## Limitations

- **No claim of finite-sample FDR control.** LD and shared-gene dependence
  across units are real, and the gene-level maximum only removes the
  within-gene part. This is an empirical calibration check.
- **A single permutation.** One realisation, not an average, so accidental
  alignment at 19 donors cannot cancel. This is why the alignment gate above
  matters. Repeating with further seeds is a stability check, not an average.
- **Cell-line missingness is not specially handled.** Three time points are
  incomplete: day 2 and day 4 have 16 donors, day 13 has 18. At those points
  the target uses the genotypes of the donors present while the decoy uses the
  genotypes of a *different* subset, so the two arms do not use the same dosage
  multiset. Measured, the decoy's median residual genotype sum of squares is
  0.9916 of the target's at those three points against 0.9980 elsewhere: a
  roughly 1% asymmetry, in the conservative direction, at 3 of 16 time points.
  A permutation restricted to matching observation patterns would remove it,
  but the four patterns have sizes 13, 3, 2 and 1, so the singleton donor would
  be frozen and would leak real signal into the decoy for every variant. The
  current scheme is the lesser distortion, but it is an approximation.
- **Anything conditioned on the published calls is circular** and is reported
  here only as description.

## Reproduction

Code lives in `code/revision_simulations/internal/fash_knockoff/`. The analysis
ran on Midway3 under R 4.4.1 and `fashr` 0.1.43
(`bf223df75da6e41ae48607a56b4cd12d7c3b24e7`); the decoy likelihood took 4.34
hours on 32 cores.

```
sbatch slurm/18_fash_knockoff.sbatch
```

```{r session}
sessionInfo()
```
