---
title: "Internal analysis: baselineLD functional enrichment of FASH variants"
author: "Ziang Zhang"
date: "2026-08-07"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 10,
  fig.height = 7.2,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(
  "code/revision_simulations/internal/baseline_ld_variant_enrichment/reporting.R"
)
```

<style>
.baseline-ld-internal-banner {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border: 1px solid #d9a441;
  border-left: 6px solid #d9a441;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.baseline-ld-key-result {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.baseline-ld-caution {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #cc7a00;
  border-radius: 4px;
  background: #fff8ec;
  color: #563a0c;
}
.baseline-ld-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.baseline-ld-table-scroll table {
  margin-bottom: 1em;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
pre {
  max-width: 100%;
  overflow-x: auto;
}
</style>

<div class="baseline-ld-internal-banner">
<strong>Internal exploratory page.</strong> This page is intentionally absent
from the workflowr homepage, navigation, manuscript, response letter, and the
first internal annotation-enrichment page. It reads completed retained caches
and does not modify existing FASH results.
</div>

## Question and provisional answer

This analysis asks whether FASH discoveries are enriched in the standardized
functional categories distributed with baselineLD v2.2. It examines the three
sets requested for this follow-up:

1. all FASH discoveries;
2. one top FASH variant per discovered gene; and
3. variants participating in gene-variant pairs discovered by FASH but by
   neither Strober linear nor Strober nonlinear.

<div class="baseline-ld-key-result">
<strong>Provisional readout.</strong>
`r provisional_answer`
`r lead_answer`
`r fash_only_answer`
</div>

The one-lead enhancer estimate is the most interesting hint in this analysis,
but the current result does not support a thresholded claim that FASH variants
are broadly enriched in baselineLD functional categories.

## Data source and analysis scale

The canonical source is the
[S-LDSC reference-files record on Zenodo](https://zenodo.org/records/10515792),
specifically `1000G_Phase3_baselineLD_v2.2_ldscores.tgz`. That archive contains
both annotation files and precomputed annotation-specific LD scores. This page
uses the chromosome-specific `.annot.gz` columns as variant annotations; it
does **not** run S-LDSC or interpret the result as partitioned heritability.

The Zenodo archive was slow to transfer during this interactive run. The same
22 chromosome annotation files were therefore retrieved individually from the
Song Lab mirror, with exact byte sizes and SHA256 hashes recorded. LD-score,
PLINK, log, and count files were not downloaded.

```{r runtime-table, results='asis'}
render_scrollable_table(
  runtime_table,
  align = "lrl",
  minimum_width = "760px"
)
```

```{r resource-table, results='asis'}
render_scrollable_table(
  resource_table,
  align = "lr",
  minimum_width = "700px"
)
```

The annotation-only download was `r format_decimal(lookup_resource("Downloaded annotation MiB"), 1L)`
MiB and completed in approximately 47 seconds. Annotation parsing and
subsetting took `r format_decimal(lookup_runtime("annotation_parse_and_subset"), 1L)`
seconds; the 100-seed matched analysis took
`r format_decimal(lookup_runtime("matched_enrichment"), 1L)` seconds. The full
analysis through cache assembly took
`r format_decimal(lookup_runtime("analysis_through_cache_assembly"), 1L)`
seconds. The retained annotation matrix occupied approximately
`r format_decimal(lookup_resource("Binary annotation matrix bytes in R") / 1024^2, 1L)`
MiB in R. This is a routine laptop-scale analysis rather than a cluster-scale
job.

## Discovery sets

FASH calls use cumulative-lfdr FDR at 0.05 after the BF update. Strober linear
and nonlinear calls use their reported eFDR at 0.05. The first and third sets
are collapsed to unique rsIDs before enrichment so that a variant is not
counted repeatedly merely because it was tested against several genes.

```{r pair-summary, results='asis'}
render_scrollable_table(
  pair_summary_table,
  align = "lrrrrr",
  minimum_width = "900px"
)
```

The FASH-only set begins with 8,062 exact gene-variant pairs, representing
8,001 unique variants and 1,119 genes. A variant is retained if it participates
in at least one FASH-only pair; the same variant may still participate in a
different pair shared with a Strober method.

## Coverage in the baselineLD reference

The 1000G EUR baselineLD annotation universe does not contain every variant in
the YRI study VCF. The matched universe and all three discovery sets were
therefore restricted to rsIDs present in baselineLD before controls were
sampled.

```{r coverage-table, results='asis'}
render_scrollable_table(
  coverage_table,
  align = "lrrrr",
  minimum_width = "860px"
)
```

The tested background coverage is
`r format_percent(coverage_by_set$coverage_proportion[coverage_by_set$discovery_set == "tested_variant_universe"])`.
Coverage is modestly higher for each FASH set, by 2.0--3.7 percentage points,
but remains inside the pre-specified 10-point imbalance boundary. Matching is
performed anew within the covered universe.

## Core functional-annotation enrichment

Each selected variant is matched to five non-selected tested variants on
chromosome, cohort MAF, minimum distance to a tested target-gene TSS, local
tested-variant density within +/-1 Mb, and the number of tested genes. Matching
is repeated over 100 fixed seeds. Points pool matched controls across seeds;
intervals are delete-one-autosome jackknife 95% intervals. Filled points would
mark within-set BH q below 0.05 across all 83 analyzed binary annotations.

```{r core-enrichment, fig.width=10.5, fig.height=11.2, fig.cap="Figure 1. baselineLD v2.2 functional-annotation enrichment for all FASH discoveries, one FASH lead variant per gene, and variants participating in exact FASH-only gene-variant pairs. Enrichment is relative to matched variants from the baselineLD-covered tested universe. Error bars are delete-one-chromosome 95% intervals."}
plot_core_enrichment()
```

No point is filled because no annotation passed the within-set BH threshold.
Three aspects are worth distinguishing:

- **All FASH discoveries:** the estimates for promoter, DHS, histone marks,
  and Hoffman enhancer are generally close to one. The lowest adjusted
  q-value occurs for a non-core ancient-sequence-age enhancer category, at
  `r format_decimal(all_fash_top$enrichment, 2L)`-fold and
  q = `r format_decimal(all_fash_top$q_value_within_set, 3L)`.
- **One lead variant per gene:** Hoffman enhancer is
  `r format_decimal(lead_enhancer$enrichment, 2L)`-fold, with a fold-scale 95%
  interval of `r format_decimal(2^lead_enhancer$ci_lower, 2L)`--`r format_decimal(2^lead_enhancer$ci_upper, 2L)`
  and q = `r format_decimal(lead_enhancer$q_value_within_set, 3L)`. This is a
  suggestive sensitivity result, not a thresholded discovery.
- **FASH-only pairs:** after exact pair subtraction and unique-variant
  collapse, no baselineLD category passes BH q below 0.05. The result therefore
  does not show a broad functional-enrichment advantage unique to FASH.

### Exact core estimates

```{r core-numeric-table, results='asis'}
render_scrollable_table(
  core_numeric_table,
  digits = 4L,
  minimum_width = "1380px"
)
```

<details>
<summary><strong>Complete results for all 83 binary annotations</strong></summary>

```{r full-results-table, results='asis'}
render_scrollable_table(
  full_results_table,
  digits = 5L,
  minimum_width = "1480px"
)
```

</details>

## Matching diagnostics

```{r matching-balance, fig.width=10.2, fig.height=5.6, fig.cap="Figure 2. Matching balance over 100 deterministic seeds. Boxes summarize standardized mean differences between each FASH set and its matched controls; dotted lines at +/-0.1 are visual references."}
plot_matching_balance()
```

The maximum absolute standardized mean difference over every set, seed, and
matching covariate is
`r format_decimal(lookup_resource("Maximum absolute matching SMD"), 4L)`.
No matching pool crosses chromosome or cohort-MAF bins.

## Annotation schema

The baselineLD files contain 97 annotation columns in addition to chromosome,
position, rsID, and genetic-map position. On chromosome 1, 84 annotation
columns are binary and 13 are continuous. The constant `base` column is removed,
leaving 83 binary annotations for enrichment. Flanking-500 and MAF-bin columns
remain in the complete table but are not included in Figure 1.

Continuous quantities are documented below but deliberately not converted to
binary categories. A separate matched continuous-score analysis would require
a distinct estimand and multiplicity plan.

```{r continuous-annotation-table, results='asis'}
render_scrollable_table(
  continuous_annotation_table,
  digits = 5L,
  minimum_width = "760px"
)
```

## Interpretation boundaries

<div class="baseline-ld-caution">
<strong>What this page does not establish.</strong> baselineLD v2.2 is built
around a 1000G EUR reference, whereas the study cohort is YRI. Restricting both
selected variants and controls to the covered rsID universe avoids treating
missing annotation as absence, but it does not make the reference ancestry
matched. The analysis uses raw annotation columns and therefore is not an
S-LDSC heritability analysis. FASH-only status is exact at the gene-variant-pair
level and is not LD-aware. Finally, annotation enrichment is evidence about a
set, not proof that an individual selected variant is causal.
</div>

The present result is best treated as a standardized functional-annotation
sensitivity analysis alongside, rather than instead of, the iPSC/mesoderm/cCRE
analysis on the first internal page.

## Reproduction

```{bash, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/baseline_ld_variant_enrichment/download_baseline_ld_annotations.R

Rscript --vanilla \
  code/revision_simulations/internal/baseline_ld_variant_enrichment/test_baseline_ld_variant_enrichment_helpers.R

Rscript --vanilla \
  code/revision_simulations/internal/baseline_ld_variant_enrichment/run_baseline_ld_variant_enrichment.R

Rscript --vanilla -e \
  'workflowr::wflow_build("analysis/revision_internal_baseline_ld_variant_enrichment.rmd", local = TRUE)'
```

<details>
<summary><strong>Validated download manifest</strong></summary>

```{r download-manifest, results='asis'}
render_scrollable_table(
  download_table,
  digits = 0L,
  minimum_width = "1500px"
)
```

</details>

<details>
<summary><strong>Input provenance</strong></summary>

```{r input-provenance, results='asis'}
render_scrollable_table(
  provenance_table,
  digits = 0L,
  minimum_width = "1260px"
)
```

</details>

<details>
<summary><strong>Session information</strong></summary>

```{r session-info, echo=FALSE}
cat(session_info_text)
```

</details>
