---
title: "Internal: diagonal, common-C and per-unit likelihoods on one panel"
author: "Ziang Zhang"
date: "2026-08-23"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.width = 10, fig.height = 5.6, dpi = 180,
  fig.retina = 2, fig.align = "center", message = FALSE, warning = FALSE
)
suppressPackageStartupMessages({library(dplyr); library(ggplot2); library(knitr)})
cache <- "output/revision_simulations/internal/correlated_permutation_discoveries"
u <- read.csv(file.path(cache, "three_likelihood_units.csv"), stringsAsFactors = FALSE)
fits <- readRDS(file.path(cache, "three_likelihood_fits.rds"))
alpha <- 0.05
call_set <- function(l, k) { o <- order(l, method = "radix")
  k[o[cumsum(l[o]) / seq_along(o) <= alpha]] }
sets <- list(diagonal = call_set(u$diagonal_bf_lfdr, u$pair_key),
             common   = call_set(u$common_bf_lfdr,   u$pair_key),
             per_unit = call_set(u$per_unit_bf_lfdr,  u$pair_key))
u$cond <- vapply(fits$per_unit_C, function(C) {
  e <- eigen(C, symmetric = TRUE, only.values = TRUE)$values; max(e)/min(e) }, numeric(1))
```

<style>
.tl-banner { margin: 1.2em 0 1.5em; padding: 1em 1.15em; border: 1px solid #d9a441;
  border-left: 6px solid #d9a441; border-radius: 4px; background: #fff8e8; color: #4a3b1c; }
.tl-result { margin: 1.2em 0 1.5em; padding: 1em 1.15em; border-left: 6px solid #0072b2;
  border-radius: 4px; background: #eef7fc; color: #173f54; }
.tl-caution { margin: 1.2em 0 1.5em; padding: 1em 1.15em; border-left: 6px solid #d55e00;
  border-radius: 4px; background: #fff4ec; color: #59331e; }
</style>

<div class="tl-banner">
**Internal, unlinked.** Three observation models fitted to one fixed panel of
1,169 gene-variant pairs, with everything else held identical. The focus is
whether the corrected fits merely shift lfdr or **reorder** it.
</div>

## The panel, and why one row of the table is near-tautological

Built from the diagonal full-data BF fit:

1. take its 9,214 calls at cumulative-lfdr `r alpha` — these span 1,176 genes;
2. keep, in each gene, the variant with the smallest lfdr (verified: all 1,169 are
   exactly their gene's minimum);
3. drop 7 genes whose top variant was also another gene's top variant, keeping
   the smaller-lfdr gene.

Result: **1,169 units, 1,169 genes, 1,169 distinct variants**, one variant per gene.

All three fits use the same units, the same $\hat\beta$, the same t-adjusted
standard errors, and identical FASH settings (`order = 1`, `penalty = 10`,
`num_basis = 20`, `betaprec = 0`, `pred_step = 1`, the 52-point PSD grid). Only
$\Omega$ differs:

- **diagonal** — $S = \mathrm{diag}(se_j)$, the manuscript's likelihood;
- **common C** — $\Omega_j = D_j^{-1} \bar{C}^{-1} D_j^{-1}$, one shared matrix;
- **per-unit** — $\Omega_j = D_j^{-1} C_j^{-1} D_j^{-1}$, each unit's own matrix
  from its own 400 donor-residual-block permutation draws.

<div class="tl-caution">
**The panel is selected on the outcome, twice over:** on having been discovered,
and on being the most significant variant within its gene. So the diagonal row
below, 1,169 of 1,169, is **near-tautological** — these units were chosen because
the diagonal fit found them significant, so refitting the mixture on only them
drives $\pi_0$ to 0.003 and calls everything. It is not evidence that the
diagonal likelihood is well behaved. The only interpretable contrast is
**per-unit against common**.
</div>

```{r summary}
read.csv(file.path(cache, "three_likelihood_summary.csv"), stringsAsFactors = FALSE) %>%
  filter(stage == "bf") %>%
  transmute(model, pi0, mean_lfdr, calls,
            share_called = sprintf("%.1f%%", 100 * calls / nrow(u))) %>%
  kable(digits = 5, caption = "BF-updated stage. Cumulative-lfdr alpha 0.05 on 1,169 units.")
```

## lfdr consistency

The diagonal lfdr on this panel is uniformly tiny (median `r sprintf("%.1e", median(u$diagonal_bf_lfdr))`,
maximum `r sprintf("%.4f", max(u$diagonal_bf_lfdr))`), so both axes are shown on
$\log_{10}$. Points are coloured by each unit's own lag-1 error autocorrelation,
the quantity shown previously to drive the reranking.

```{r scatter-main, fig.height = 5.4}
long <- bind_rows(
  u %>% transmute(pair_key, diagonal = diagonal_bf_lfdr, corrected = common_bf_lfdr,
                  lag1 = per_unit_lag1, model = "common C"),
  u %>% transmute(pair_key, diagonal = diagonal_bf_lfdr, corrected = per_unit_bf_lfdr,
                  lag1 = per_unit_lag1, model = "per-unit Omega")
)
ggplot(long, aes(diagonal, corrected, colour = lag1)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey55", linetype = "dashed") +
  geom_point(size = 1.5, alpha = 0.75) +
  scale_x_log10(labels = scales::label_log()) +
  scale_y_log10(labels = scales::label_log()) +
  scale_colour_viridis_c(name = "unit lag-1\nerror autocorr.", option = "C") +
  facet_wrap(~model) +
  labs(title = "Corrected lfdr against diagonal lfdr, same 1,169 units",
       subtitle = "Dashed line is equality. A pure conservatism shift would sit parallel to it; instead both panels are nearly flat,\nmeaning the corrected lfdr carries almost no information about the diagonal lfdr.",
       x = "diagonal BF lfdr", y = "corrected BF lfdr") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), legend.position = "right")
```

```{r consistency-table}
cons <- function(a, b, la, lb) data.frame(
  comparison = paste(la, "vs", lb),
  spearman = cor(a, b, method = "spearman"),
  pearson_log10 = cor(log10(a), log10(b)),
  mean_abs_difference = mean(abs(a - b)))
bind_rows(
  cons(u$per_unit_bf_lfdr, u$diagonal_bf_lfdr, "per-unit", "diagonal"),
  cons(u$common_bf_lfdr,   u$diagonal_bf_lfdr, "common",   "diagonal"),
  cons(u$per_unit_bf_lfdr, u$common_bf_lfdr,   "per-unit", "common")
) %>% kable(digits = 3, caption = "Agreement of the BF-updated lfdr across models.")
```

<div class="tl-result">
The corrected fits do **not** simply shift lfdr upward. Against the diagonal
lfdr, the per-unit lfdr has Spearman `r round(cor(u$per_unit_bf_lfdr, u$diagonal_bf_lfdr, method="spearman"),3)`
and a $\log_{10}$ Pearson correlation of only
`r round(cor(log10(u$per_unit_bf_lfdr), log10(u$diagonal_bf_lfdr)),3)` — that is
**essentially no relationship at all**. The common-C fit is only slightly better
at `r round(cor(u$common_bf_lfdr, u$diagonal_bf_lfdr, method="spearman"),3)`.
Even the two corrected fits agree with each other only at
`r round(cor(u$per_unit_bf_lfdr, u$common_bf_lfdr, method="spearman"),3)`.

Whatever these likelihoods are ranking on, it is largely orthogonal to what the
diagonal likelihood ranks on, and to each other.
</div>

## Rank-rank view, and the two corrected fits against each other

```{r scatter-rank, fig.height = 5.4}
u <- u %>% mutate(across(ends_with("_bf_lfdr"), ~rank(.x), .names = "rank_{.col}"))
bind_rows(
  u %>% transmute(x = rank_diagonal_bf_lfdr, y = rank_per_unit_bf_lfdr,
                  lag1 = per_unit_lag1, panel = "per-unit rank vs diagonal rank"),
  u %>% transmute(x = rank_common_bf_lfdr, y = rank_per_unit_bf_lfdr,
                  lag1 = per_unit_lag1, panel = "per-unit rank vs common rank")
) %>%
  ggplot(aes(x, y, colour = lag1)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey55", linetype = "dashed") +
  geom_point(size = 1.4, alpha = 0.7) +
  scale_colour_viridis_c(name = "unit lag-1", option = "C") +
  facet_wrap(~panel) +
  labs(title = "Scale-free view: rank of lfdr, 1 = most significant",
       subtitle = "Near-total diffusion. Colour shows each unit's error autocorrelation; note that no gradient is visible, which is itself the point below.",
       x = "rank under the reference model", y = "rank under the per-unit model") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

```{r overlap}
j <- function(a, b) length(intersect(a, b)) / length(union(a, b))
data.frame(
  pair = c("per-unit & diagonal", "common & diagonal", "per-unit & common"),
  n_a = c(length(sets$per_unit), length(sets$common), length(sets$per_unit)),
  n_b = c(length(sets$diagonal), length(sets$diagonal), length(sets$common)),
  intersection = c(length(intersect(sets$per_unit, sets$diagonal)),
                   length(intersect(sets$common, sets$diagonal)),
                   length(intersect(sets$per_unit, sets$common))),
  jaccard = c(j(sets$per_unit, sets$diagonal), j(sets$common, sets$diagonal),
              j(sets$per_unit, sets$common)),
  a_nested_in_b = c(all(sets$per_unit %in% sets$diagonal),
                    all(sets$common %in% sets$diagonal),
                    all(sets$per_unit %in% sets$common))
) %>% kable(digits = 3, caption = "Call-set overlap at cumulative-lfdr 0.05.")
```

`common` is nested in `diagonal`, and `per_unit` is nested in `diagonal`, but
**`per_unit` is not nested in `common`**: `r sum(!(sets$per_unit %in% sets$common))`
of its `r length(sets$per_unit)` calls are not common calls. The per-unit
likelihood is reordering relative to the common one, not just tightening it.

## What drives the reranking

```{r lag1-vs-lfdr, fig.height = 4.6}
bind_rows(
  u %>% transmute(lag1 = per_unit_lag1, lfdr = diagonal_bf_lfdr, model = "diagonal"),
  u %>% transmute(lag1 = per_unit_lag1, lfdr = common_bf_lfdr,   model = "common C"),
  u %>% transmute(lag1 = per_unit_lag1, lfdr = per_unit_bf_lfdr, model = "per-unit Omega")
) %>% mutate(model = factor(model, c("diagonal", "common C", "per-unit Omega"))) %>%
  ggplot(aes(lag1, lfdr)) +
  geom_point(size = 1.2, alpha = 0.5, colour = "#0072B2") +
  geom_smooth(method = "loess", se = FALSE, colour = "#B2182B", linewidth = 0.9) +
  scale_y_log10() + facet_wrap(~model, scales = "free_y") +
  labs(title = "lfdr against the unit's own error autocorrelation",
       subtitle = "The diagonal fit finds high-correlation units MORE significant; the corrections push against that, and overshoot.",
       x = "unit lag-1 error autocorrelation", y = "BF lfdr") +
  theme_minimal(base_size = 11) + theme(plot.title = element_text(face = "bold"))
```

```{r lag1-table}
u$min_eig <- vapply(fits$per_unit_C, function(C)
  min(eigen(C, symmetric = TRUE, only.values = TRUE)$values), numeric(1))
u$logdet <- vapply(fits$per_unit_C, function(C)
  determinant(C, logarithm = TRUE)$modulus, numeric(1))
predictors <- c("per_unit_lag1", "cond", "min_eig", "logdet", "diagonal_full_lfdr")
do.call(rbind, lapply(predictors, function(v) data.frame(
  predictor = v,
  diagonal = cor(u$diagonal_bf_lfdr, u[[v]], method = "spearman"),
  common_C = cor(u$common_bf_lfdr,   u[[v]], method = "spearman"),
  per_unit = cor(u$per_unit_bf_lfdr, u[[v]], method = "spearman")
))) %>% kable(digits = 3,
  caption = "Spearman correlation of each model's lfdr with candidate explanators. For the error autocorrelation, negative means a more correlated unit looks more significant.")
```

The diagonal fit is **negatively** associated with a unit's error correlation
(`r round(cor(u$diagonal_bf_lfdr, u$per_unit_lag1, method="spearman"),3)`): the
more correlated a unit's errors, the more significant it looks. That is the
expected direction if correlated errors inflate apparent smooth signal, and it is
independent evidence that the diagonal likelihood is affected. It is also
associated with the conditioning of that unit's matrix.

The **common-C** fit flips every one of those signs, which is what a correction
should do: it pushes back against exactly the units the diagonal fit was most
liable to over-call.

<div class="tl-caution">
The **per-unit** fit does neither. Its lfdr is essentially uncorrelated with all
five candidates — error autocorrelation
`r sprintf("%+.3f", cor(u$per_unit_bf_lfdr, u$per_unit_lag1, method="spearman"))`,
condition number `r sprintf("%+.3f", cor(u$per_unit_bf_lfdr, u$cond, method="spearman"))`,
minimum eigenvalue `r sprintf("%+.3f", cor(u$per_unit_bf_lfdr, u$min_eig, method="spearman"))`,
log-determinant `r sprintf("%+.3f", cor(u$per_unit_bf_lfdr, u$logdet, method="spearman"))`,
and the full-data diagonal lfdr
`r sprintf("%+.3f", cor(u$per_unit_bf_lfdr, u$diagonal_full_lfdr, method="spearman"))`.

So the per-unit ordering is not explained by the strength of the unit's error
correlation, nor by how badly conditioned its matrix is, nor by how strong its
signal looked originally. **We cannot presently attribute it to anything**, and
an ordering with no identifiable driver is not one to trust. This is a stronger
reason for caution than the conservatism itself.
</div>

## Why the per-unit fit is so conservative

```{r conditioning}
data.frame(
  quantity = c("common C condition number",
               "per-unit condition number, median",
               "per-unit condition number, 90th percentile",
               "condition number expected from sampling noise alone",
               "free residual dimensions per day (19 or 16 donors minus rank-6 covariates)"),
  value = c(sprintf("%.1f", {e <- eigen(fits$common_C, symmetric = TRUE, only.values = TRUE)$values; max(e)/min(e)}),
            sprintf("%.1f", median(u$cond)), sprintf("%.1f", quantile(u$cond, 0.9)),
            "18.9 (90% range 16.1-22.2)", "10 to 13")
) %>% kable(caption = "The per-unit matrices are genuinely, reproducibly ill-conditioned.")
```

<div class="tl-caution">
The per-unit matrices are **not** noisy — split-half reproducibility of their
minimum eigenvalue is 0.93 and of their lag-1 is 0.98. They are genuinely
near-singular, because after projecting out the intercept and five PCs each day
leaves only 10 to 13 free donor dimensions to support 16 time points. Inverting
such a matrix puts enormous precision on directions the data barely constrain,
and the common matrix escapes this only because averaging over hundreds of units
acts as regularisation.

So the 118-call result should not yet be read as sharper inference. The
diagnostic that would separate "genuine per-unit structure" from "near-singular
directions" is a **regularised** per-unit fit, shrinking each $C_j$ until its
condition number matches the common matrix. That has not been run.
</div>

## Boundaries

<div class="tl-banner">
- The panel is outcome-selected. No $\pi_0$ or call count here is an FDR
  statement, and the diagonal row is near-tautological by construction.
- Two orthogonal checks run separately — Hallmark gene sets and Strober
  corroboration — were flat across all three sets, so there is no external
  evidence that any of these orderings is the better one.
- The near-singularity of the per-unit matrices is unresolved, so the per-unit
  column is a diagnostic, not a proposed analysis.
- Nothing here is a manuscript claim.
</div>

```{r session}
sessionInfo()
```
