---
title: "Revision Simulation R2: genotype-level eQTLs with spiky transient effects"
author: "Ziang Zhang"
date: "2026-07-30"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9,
  fig.height = 6.5,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source("code/revision_simulations/shared/simulation_functions.R")
source("code/revision_simulations/r2_spiky_transient/reporting.R")
```

## Overview

This genotype-level simulation evaluates dynamic-eQTL detection when the true
effects are continuous but sharply localized and are not generated from the
IWP prior used by FASH. It uses the same individual-level pipeline as the
[broad random B-spline simulation](revision_genotype_level_simulation.html).

For donor $i$, variant $j$, and time $t$,

$$
Y_{ijt} = \alpha_{jt} + G_{ij}\beta_j(t) +
X_i^T\gamma_{jt} + \epsilon_{ijt},
\qquad \epsilon_{ijt} \sim N(0,1).
$$

For each variant, the minor-allele frequency is drawn from
$p_j\sim\mathrm{Uniform}(0.1,0.5)$ and
$G_{ij}\sim\mathrm{Binomial}(2,p_j)$; monomorphic draws are repeated.
The five columns of $X$ are generated independently from $N(0,1)$ and
standardized across donors. Independently for each covariate, variant, and
time, $\gamma_{kjt}\sim N(0,0.5^2)$. The current simulation sets
$\alpha_{jt}=0$; each fitted regression still includes an intercept. Genotype,
covariate, covariate-effect, and noise draws are mutually independent.

```{r r2-simulate-genotypes-and-covariates, eval=FALSE}
seed <- 12345
component_seeds <- revision_component_seeds(seed)

genotype_sim <- simulate_genotype_matrix(
  n_donors = 19,
  n_variants = 1000,
  maf_range = c(0.1, 0.5),
  seed = component_seeds[["genotype"]]
)
covariates <- simulate_covariate_matrix(
  n_donors = 19,
  n_covariates = 5,
  seed = component_seeds[["covariates"]]
)
```

Each replicate has `19` donors, `16` time points, `1000` variants, and five
PC-like covariates. There are `200` dynamic eQTLs, `400` constant eQTLs with
effects drawn from $N(0,1)$, and `400` zero-effect variants. Therefore, the
true dynamic-null proportion is $\pi_0=0.8$.

For a dynamic eQTL,

$$
\beta_j(t) = \mu_j + h_j(t),
\qquad \mu_j \sim N(0,1),
$$

where $h_j(t)$ is a sum of one, two, or three compact raised-cosine peaks,

$$
h(t;c,w)=
\frac{1+\cos\{\pi(t-c)/w\}}{2}
\mathbf{1}(|t-c|\leq w),
\qquad w=1.5.
$$

The largest peak is placed at a random location in an early, middle, or late
window. Additional peaks, when present, occur in different windows and have
random amplitudes between `35%` and `75%` of the primary peak. Multi-peak
curves include both same-sign and alternating-sign patterns. The deviation
$h_j$ is not mean-centered; it is scaled to centered RMS `0.9`, and the
independent genetic main effect $\mu_j$ remains as the background effect
outside the compact supports.

```{r r2-simulate-effects-and-expression, eval=FALSE}
class_probs <- c(
  dynamic_bspline = 0.20,
  constant = 0.40,
  zero = 0.40
)
shape_cell_probs <- c(
  k1__spiky__single = 1 / 3,
  `k2__spiky__same-sign` = 1 / 6,
  `k2__spiky__alternating-sign` = 1 / 6,
  `k3__spiky__same-sign` = 1 / 6,
  `k3__spiky__alternating-sign` = 1 / 6
)
evaluation_grid <- seq(0, 15, by = 0.1)

effect_sim <- simulate_raised_cosine_multipeak_effect_set(
  n_variants = 1000,
  time_grid = 0:15,
  evaluation_grid = evaluation_grid,
  class_probs = class_probs,
  width_levels = c(spiky = 1.5),
  spike_counts = 1:3,
  shape_cell_probs = shape_cell_probs,
  primary_time_groups = c("early", "middle", "late"),
  relative_amplitude_range = c(0.35, 0.75),
  center_by_observed_mean = FALSE,
  validate_functional_labels = FALSE,
  switch_threshold = 0.25,
  target_centered_rms = 0.9,
  baseline_sd = 1,
  constant_sd = 1,
  dynamic_baseline_sd = 1,
  exact_class_counts = TRUE,
  seed = seed,
  class_seed = component_seeds[["classes"]],
  constant_seed = component_seeds[["constant_effects"]],
  shape_seed = component_seeds[["functional_truth"]],
  scenario = "r2_genotype_timed_cosine_one_two_three_peak_main_effect_dynamic_eqtl"
)
expression_sim <- simulate_eqtl_expression_from_genotypes(
  G = genotype_sim$G,
  beta_matrix = effect_sim$beta_matrix,
  time_grid = 0:15,
  covariates = covariates,
  expression_noise_sd = 1,
  covariate_effect_sd = 0.5,
  intercept_sd = 0,
  seed = component_seeds[["expression"]]
)
```

The `200` dynamic variants are approximately balanced across one, two, and
three peaks. Alternating peak signs describe geometry only; this page tests
global dynamic eQTL status and does not equate an alternating pattern with the
manuscript's switch functional. Functional testing is evaluated separately.

### Dynamic shape and timing allocation

```{r allocation-table}
knitr::kable(
  allocation_table,
  align = "l",
  caption = paste(
    "Table R2.1: Allocation of dynamic eQTLs across peak counts, sign",
    "patterns, and primary timing windows in the spiky simulation."
  )
)
```

At each time, effects and standard errors are estimated with
`Y ~ G + PC1 + PC2 + PC3 + PC4 + PC5`. The t-statistic correction uses
`df = 19 - 7 = 12`. FASH receives only these per-time estimates and corrected
standard errors. The direct interaction tests use the full individual-level
data.

```{r r2-estimate-and-fit-fash, eval=FALSE}
eqtl_summary <- estimate_eqtl_summaries_from_genotypes(
  G = genotype_sim$G,
  expression = expression_sim$expression,
  covariates = covariates,
  apply_t_se_correction = TRUE
)
datasets <- make_fash_datasets_from_eqtl_summary(
  beta_hat = eqtl_summary$beta_hat,
  se = eqtl_summary$se,
  true_beta = effect_sim$beta_matrix,
  time_grid = 0:15,
  unit_info = effect_sim$unit_info,
  scenario = "r2_genotype_timed_cosine_one_two_three_peak_main_effect_dynamic_eqtl"
)

fash_iwp1_raw <- fashr::fash(
  Y = "y",
  smooth_var = "x",
  S = "sd",
  data_list = datasets,
  order = 1,
  grid = default_revision_grid(),
  num_basis = 20,
  penalty = 10,
  num_cores = 4,
  verbose = FALSE
)
fash_iwp1_bf <- fashr::BF_update(fash_iwp1_raw, plot = FALSE)

linear_sigma_grid <- exp(seq(log(0.05), log(5), length.out = 25))
fash_linear_raw <- fit_simplified_fash(
  datasets = datasets,
  estimate_sigma = TRUE,
  sigma_beta_grid = linear_sigma_grid,
  scale_time = TRUE
)
fash_linear_bf <- BF_update_simplified_fash(fash_linear_raw)
```

All tables and curves below average five independently simulated datasets with
seeds `12345`, `22345`, `32345`, `42345`, and `52345`. The seed-12345 full fit
is used only for representative trajectories.

## Example simulated eQTLs

The first figure shows zero-effect and constant dynamic-null examples. Points
are per-time effect estimates, bars show two corrected standard errors, and
the colored line is the true effect.

```{r null-examples, fig.width=10, fig.height=5.2, fig.cap="Figure R2.1: Representative zero-effect and constant-effect dynamic-null eQTLs from seed 12345. Points are estimated genetic effects, vertical bars span two corrected standard errors, and colored curves are the true effects."}
plot_genotype_eqtl_examples(
  out,
  classes = c("zero", "constant"),
  n_per_class = 3,
  seed = 1,
  true_curve_n = 500
)
```

The second figure shows one representative from each available
shape-by-primary-window cell. Truth curves are drawn on a dense time grid so
that compact peaks between the 16 observed time points remain visible.

```{r dynamic-examples, fig.width=12, fig.height=18, fig.cap="Figure R2.2: Representative compact dynamic eQTLs across peak-count, sign-pattern, and primary-window cells. Points and error bars show the 16 estimated effects and two corrected standard errors; colored dense curves show the continuous truths."}
plot_cosine_peak_cell_examples(
  out,
  seed = 1,
  include_time_groups = TRUE
)
```

## Comparison 1: IWP versus linear FASH

`FASH-IWP1` uses the first-order IWP alternative. `FASH-linear` uses a
constant-versus-linear model on time scaled to $[0,1]$: its alternative adds a
slope $b\sim N(0,\sigma_\beta^2)$ to the unrestricted intercept. Within each
seed, the fit profiles $\pi_0$ separately at each of 25 log-spaced candidate
values of $\sigma_\beta$ from `0.05` to `5`, then selects the pair with the
largest marginal likelihood. The BF adjustment updates the mixture weight
while holding the selected $\sigma_\beta$ fixed. Across the five seeds, the
selected slope SD has mean
`r format_decimal(linear_sigma_summary$mean_selected_sigma)` and observed range
[`r format_decimal(linear_sigma_summary$min_selected_sigma)`,
`r format_decimal(linear_sigma_summary$max_selected_sigma)`]. Both raw and
BF-corrected fits use the same summary statistics and discovery rule.

### Estimated dynamic-null proportion

```{r pi0-table}
knitr::kable(
  pi0_table,
  align = "l",
  caption = paste(
    "Table R2.2: Estimated dynamic-null proportions and selected linear slab",
    "scales across five spiky-effect simulation replicates."
  )
)
```

### Results at alpha 0.05

```{r fash-table}
knitr::kable(
  fash_table,
  align = "l",
  caption = paste(
    "Table R2.3: Monte Carlo power and empirical FDR at alpha 0.05 for IWP",
    "and linear FASH under compact spiky effects."
  )
)
```

### Power as a function of alpha

```{r fash-power, fig.cap="Figure R2.3: Monte Carlo power across nominal alpha for IWP and linear FASH under compact one-, two-, and three-peak effects. Curves identify raw and BF-updated methods in the in-figure legend; uncertainty summarizes five seeds."}
plot_mc_alpha_curves(
  fash_alpha,
  metric = "power",
  title = "IWP versus linear FASH: Monte Carlo power",
  subtitle = "Five seeds; compact one-, two-, and three-peak effects",
  style_profile = "combined"
)
```

### Empirical FDR as a function of alpha

```{r fash-fdr, fig.cap="Figure R2.4: Monte Carlo empirical FDR across nominal alpha for IWP and linear FASH under compact spiky effects. Curves identify methods in the in-figure legend, and the diagonal reference marks nominal calibration."}
plot_mc_alpha_curves(
  fash_alpha,
  metric = "fdr",
  title = "IWP versus linear FASH: Monte Carlo empirical FDR",
  subtitle = "Five seeds; compact one-, two-, and three-peak effects",
  legend_position = "topleft",
  style_profile = "combined"
)
```

At $\alpha=0.05$, IWP-BF has mean power
`r format_decimal(metric_at(fash_alpha_005, "FASH-IWP1-BF", "mean_power"))`,
compared with
`r format_decimal(metric_at(fash_alpha_005, "FASH-linear-BF", "mean_power"))`
for linear-BF. Their empirical FDR estimates are
`r format_decimal(metric_at(fash_alpha_005, "FASH-IWP1-BF", "mean_fdr"))`
and
`r format_decimal(metric_at(fash_alpha_005, "FASH-linear-BF", "mean_fdr"))`.

## Comparison 2: IWP FASH versus direct interaction tests

The direct linear LRT tests a linear genotype-by-time interaction. The direct
quadratic LRT jointly tests linear and quadratic interaction terms. Their
permutation eFDR calculations use `B=100` donor-level permutations and are
supplied with the true $\pi_0=0.8$, deliberately removing null-proportion
estimation uncertainty from the direct baselines.

```{r r2-direct-interaction-efdr, eval=FALSE}
out <- add_direct_interaction_efdr_results_to_genotype_output(
  out = out,
  n_permutations = 100,
  alpha = 0.05,
  seed = component_seeds[["permutations"]],
  lambda = 0.5,
  pi0_method = "conservative",
  true_pi0 = 0.8,
  include_true_pi0 = TRUE,
  permute_covariates_with_expression = TRUE,
  num_cores = 4,
  overwrite = TRUE,
  verbose = FALSE
)
```

### Results at alpha 0.05

```{r direct-table}
knitr::kable(
  direct_table,
  align = "l",
  caption = paste(
    "Table R2.4: Monte Carlo power and empirical FDR at alpha 0.05 for IWP",
    "FASH and direct interaction tests under compact spiky effects."
  )
)
```

### Power as a function of alpha

```{r direct-power, fig.cap="Figure R2.5: Monte Carlo power across nominal alpha for IWP FASH and direct linear or quadratic interaction tests under compact spiky effects. The in-figure legend identifies methods; direct-test eFDR uses the true pi0 and 100 permutations."}
plot_mc_alpha_curves(
  direct_alpha,
  metric = "power",
  title = "IWP FASH versus direct interaction: Monte Carlo power",
  subtitle = "Five seeds; direct eFDR uses true pi0 = 0.8 and B = 100",
  style_profile = "combined"
)
```

### Empirical FDR as a function of alpha

```{r direct-fdr, fig.cap="Figure R2.6: Monte Carlo empirical FDR across nominal alpha for IWP FASH and direct interaction tests under compact spiky effects. The in-figure legend identifies methods, and the diagonal reference marks nominal calibration."}
plot_mc_alpha_curves(
  direct_alpha,
  metric = "fdr",
  title = "IWP FASH versus direct interaction: Monte Carlo empirical FDR",
  subtitle = "Five seeds; direct eFDR uses true pi0 = 0.8 and B = 100",
  legend_position = "bottomleft",
  style_profile = "combined"
)
```

At $\alpha=0.05$, mean power is
`r format_decimal(metric_at(direct_alpha_005, "FASH-IWP1-BF", "mean_power"))`
for IWP-BF,
`r format_decimal(metric_at(direct_alpha_005, "Direct-linear-LRT-eFDR-true-pi0", "mean_power"))`
for direct linear, and
`r format_decimal(metric_at(direct_alpha_005, "Direct-quadratic-LRT-eFDR-true-pi0", "mean_power"))`
for direct quadratic. The corresponding empirical FDR estimates are
`r format_decimal(metric_at(direct_alpha_005, "FASH-IWP1-BF", "mean_fdr"))`,
`r format_decimal(metric_at(direct_alpha_005, "Direct-linear-LRT-eFDR-true-pi0", "mean_fdr"))`,
and
`r format_decimal(metric_at(direct_alpha_005, "Direct-quadratic-LRT-eFDR-true-pi0", "mean_fdr"))`.

## Power by number of peaks

The global discovery threshold is fitted once across all `1000` variants.
Power is then summarized within the true one-, two-, and three-peak dynamic
subgroups; no subgroup-specific threshold is estimated.

### Results at alpha 0.05

```{r peak-table}
knitr::kable(
  peak_table,
  align = "l",
  caption = paste(
    "Table R2.5: Monte Carlo power at alpha 0.05 stratified by the true",
    "number of compact peaks; all calls use the same global threshold."
  )
)
```

### Power as a function of alpha

```{r peak-power, fig.width=10, fig.height=6, fig.cap="Figure R2.7: Monte Carlo discovery power across nominal alpha, stratified by one-, two-, and three-peak truth. Panels denote peak count, curves identify methods in the in-figure legend, and the FDR threshold is fitted globally before subgrouping."}
plot_mc_shape_power_grid(
  peak_alpha[peak_alpha$method %in% peak_methods, ],
  shape_order = c("1-peak", "2-peak", "3-peak"),
  title = "Power by number of compact peaks",
  subtitle = "Five seeds; global FDR threshold applied before subgrouping",
  style_profile = "combined",
  width = 2600
)
```

## Interpretation

For continuous but compact transient effects, IWP-BF is more powerful than
linear-BF while maintaining comparable or lower empirical FDR. Relative to
the direct baselines, the table and peak-stratified analysis show how
performance changes as the trajectory contains one, two, or three localized
features. The direct methods are given the true null proportion, whereas FASH
estimates its mixture weights from the summary statistics.

## Reproducibility

The
[R2 simulation driver](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/r2_spiky_transient/run_spiky_transient_mc_replication.R)
uses the
[shared simulation functions](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/shared/simulation_functions.R).
The page-specific cache validation and formatting are in
[R2 reporting code](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/r2_spiky_transient/reporting.R).
The following command regenerates the full five-seed cache. It is shown for
reproducibility but is intentionally not evaluated during workflowr builds;
the table and figure chunks above read the saved cache.

The full five-seed cache can be regenerated with:

```{bash regenerate-r2-cache, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/r2_spiky_transient/run_spiky_transient_mc_replication.R \
  --J 1000 \
  --n-donors 19 \
  --n-covariates 5 \
  --noise-sd 1 \
  --dynamic-main-effect-sd 1 \
  --width-half 1.5 \
  --target-centered-rms 0.9 \
  --efdr-permutations 100 \
  --seed-list 12345,22345,32345,42345,52345 \
  --output-id r2_timed_cosine_one_two_three_peak_main_effect_profile_sigma_pilot5 \
  --num-cores 4
```

Cached replicate metrics, summaries, plots, and the seed-12345 full fit are
stored under:

```text
output/revision_simulations/mc/r2_timed_cosine_one_two_three_peak_main_effect_profile_sigma_pilot5/
```
