---
title: "Internal sensitivity analysis: 20 random variants per eligible gene"
author: "Ziang Zhang"
date: "2026-08-07"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9,
  fig.height = 5.4,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(
  "code/revision_simulations/r5_balanced_thinning/reporting.R"
)
```

<style>
.r5-key-result {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.r5-note {
  margin: 1.2em 0 1.5em;
  padding: 0.9em 1.05em;
  border-left: 5px solid #e69f00;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.r5-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.r5-table-scroll table {
  margin-bottom: 1em;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
pre {
  max-width: 100%;
  overflow-x: auto;
}
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: hidden;
}
</style>

<div class="r5-note">
<strong>Internal supporting analysis.</strong> This is the former
reviewer-facing R5 analysis. It retains the completed 10-seed design with 20
uniformly sampled variants for each gene having at least 20 tested variants.
It is no longer linked from the revision index and is preserved only as
supporting provenance for the more direct one-variant-per-gene R5 analysis.
</div>

## Question and main result

The complete FASH(1) analysis treats each tested gene-variant pair as one
empirical-Bayes unit. Because genes contribute different numbers of tested cis
variants, genes with more variants have greater representation in the fitted
mixture. This real-data sensitivity analysis asks whether the BF-updated prior
weights and pair-level local false discovery rates (lfdr) change materially
when that representation is balanced.

Every eligible gene is represented by exactly 20 uniformly sampled variants
in each refit. This retains substantially more information than a
one-variant-per-gene analysis while giving every included gene the same number
of units.

<div class="r5-key-result">
<strong>Main result.</strong> Across ten prespecified random selections, the
BF-updated null weight is `r format_range(seed_summary$thinned_pi0, 4)` versus
`r format_decimal(full_pi0, 4)` in the full analysis. Paired lfdr Spearman
correlations are `r format_range(lfdr_spearman_range, 4)`, and mean absolute
lfdr differences are `r format_range(lfdr_mae_range, 4)`. Thus, the fitted
null weight and the lfdr assigned to retained pairs are practically unchanged
by balancing the number of variants per gene at 20.
</div>

This page reports only the paper's BF-updated fit. The raw empirical-Bayes
penalty remains fixed at its original value of 10 before every BF update.

## Balanced-thinning design

```{r design-table, results='asis'}
render_scrollable_table(
  design_table,
  caption = paste(
    "Table R5.1: Design of the balanced-variant thinning sensitivity",
    "analysis. All ten refits use the BF-updated reporting version."
  ),
  align = "ll",
  minimum_width = "900px"
)
```

### Full-data FASH settings and matched penalty

Yes: the full-data fit and every balanced-thinning refit use the same
empirical-Bayes penalty, `penalty = 10`. The following call makes all settings
used to construct the full-data likelihood matrix explicit. It matches the
code on the main
[dynamic eQTL analysis page](https://stephenslab.github.io/fashr-paper/dynamic_eQTL_real.html);
the Gaussian likelihood was the default in the original call and is written
explicitly here for clarity.

```{r full-data-fash-settings, eval=FALSE}
log_prec <- seq(0, 10, by = 0.2)
fine_grid <- sort(c(0, exp(-0.5 * log_prec)))

fash_fit1 <- fashr::fash(
  Y = "beta",
  smooth_var = "time",
  S = "SE",
  data_list = datasets,
  num_basis = 20,
  order = 1,
  betaprec = 0,
  pred_step = 1,
  likelihood = "gaussian",
  penalty = 10,
  grid = fine_grid,
  num_cores = num_cores,
  verbose = TRUE
)

fash_fit1_update <- fashr::BF_update(fash_fit1, plot = FALSE)
```

The retained raw object records these settings as `num_basis = 20`,
`order = 1`, `betaprec = 0`, `pred_step = 1`,
`likelihood = "gaussian"`, and `penalty = 10`. Its PSD grid has 52 values,
including the null value zero.

### Exact sampling procedure

For each seed, the selection is performed as follows.

1. Read the ordered gene-variant keys from the current full FASH object and
   separate each key into its Ensembl gene ID and variant ID.
2. Partition all `r format_integer(expected_full_units)` pair indices by gene.
3. Keep the `r format_integer(expected_eligible_genes)` genes having at least
   20 tested variants. The remaining `r format_integer(expected_excluded_genes)`
   genes are omitted; none is among the genes discovered by the full BF-updated
   analysis.
4. Within each eligible gene, sample exactly 20 pair indices uniformly without
   replacement. Sampling is independent across genes and does not use beta,
   SE, likelihood, lfdr, or significance.
5. Subset the corresponding rows of the already computed likelihood matrix,
   re-estimate the mixture weights with the original penalty of 10, and apply
   the same BF update used in the paper.

```{r balanced-selection-code, eval=FALSE}
full_raw_fit <- fash_fit1
full_bf_fit <- fash_fit1_update
pair_keys <- names(full_raw_fit$fash_data$data_list)

gene_index <- split(
  seq_along(pair_keys),
  sub("_.*$", "", pair_keys)
)
eligible_gene_index <- gene_index[lengths(gene_index) >= 20L]

set.seed(seed)
selected_indices <- unlist(
  lapply(
    eligible_gene_index,
    function(indices) sample(indices, size = 20L, replace = FALSE)
  ),
  use.names = FALSE
)

stopifnot(
  full_raw_fit$settings$num_basis == 20,
  full_raw_fit$settings$order == 1,
  full_raw_fit$settings$betaprec == 0,
  full_raw_fit$settings$pred_step == 1,
  identical(full_raw_fit$settings$likelihood, "gaussian"),
  full_raw_fit$settings$penalty == 10
)

selected_pair_keys <- pair_keys[selected_indices]
selected_likelihood <- full_raw_fit$L_matrix[
  selected_indices,
  ,
  drop = FALSE
]
rownames(selected_likelihood) <- selected_pair_keys

# Use exactly the same penalty as the full-data fit.
thinned_eb <- fashr::fash_eb_est(
  L_matrix = selected_likelihood,
  grid = full_raw_fit$psd_grid,
  penalty = full_raw_fit$settings$penalty
)
rownames(thinned_eb$posterior_weight) <- selected_pair_keys
null_column <- which(thinned_eb$prior_weight$psd == 0)

thinned_raw_fit <- structure(
  list(
    prior_weights = thinned_eb$prior_weight,
    posterior_weights = thinned_eb$posterior_weight,
    psd_grid = full_raw_fit$psd_grid,
    lfdr = thinned_eb$posterior_weight[, null_column],
    L_matrix = selected_likelihood
  ),
  class = "fash"
)
thinned_bf_fit <- fashr::BF_update(thinned_raw_fit, plot = FALSE)

paired_lfdr <- data.frame(
  pair_key = selected_pair_keys,
  full_lfdr = full_bf_fit$lfdr[selected_indices],
  thinned_lfdr = thinned_bf_fit$lfdr
)
```

Reusing likelihood rows is exact for this comparison: beta estimates, standard
errors, basis settings, and the PSD grid are fixed, so thinning only changes
which likelihood contributions enter the mixture-weight
optimization. It does not approximate or recompute those per-pair likelihoods.
The penalty is used when estimating the raw empirical-Bayes mixture in both
analyses; the same `BF_update(..., plot = FALSE)` step is then applied to the
full and thinned raw fits. No seed-specific penalty is introduced.

## Number of variants per gene

```{r variant-count-summary, results='asis'}
render_scrollable_table(
  variant_count_summary_table,
  caption = paste(
    "Table R5.2: Distribution summary for the number of tested variants",
    "contributed by each of the 6,362 genes in the full FASH analysis."
  ),
  align = "lr",
  minimum_width = "680px"
)
```

The median gene contributes `r format_integer(variant_count_median)` tested
variants (IQR `r format_integer(variant_count_q1)`--`r format_integer(variant_count_q3)`),
whereas the minimum is `r format_integer(variant_count_min)`. Exactly
`r format_integer(expected_excluded_genes)` genes fall below the target of 20.

```{r variants-per-gene-histogram, fig.width=9.2, fig.height=5.4, fig.cap="Figure R5.1: Distribution of tested variants per gene in the full FASH analysis. Blue bars count genes; the orange dashed line marks the retention target of 20 variants. The horizontal axis is logarithmic to display both the main distribution and the long right tail."}
plot_variants_per_gene_histogram()
```

Each balanced dataset contains `r format_integer(expected_selected_units)`
pairs, or `r format_decimal(100 * expected_selected_units / expected_full_units, 1)`%
of the full pair-level analysis, while retaining
`r format_decimal(100 * expected_eligible_genes / expected_full_genes, 1)`% of
its genes.

## A concrete draw: seed 12345

```{r seed-12345-table, results='asis'}
render_scrollable_table(
  specific_seed_table,
  caption = paste(
    "Table R5.3: BF-updated full-versus-thinned comparison for the concrete",
    "balanced draw generated with seed 12345. Pair-level calls on both sides",
    "use cumulative-lfdr FDR 0.05 within the same retained-pair universe."
  ),
  align = "rrrrrrrrrr",
  minimum_width = "1700px"
)
```

Both axes below refer to exactly the same retained gene-variant pairs. The
horizontal coordinate is their lfdr in the fixed full-data BF-updated fit; the
vertical coordinate is their lfdr after the balanced refit and BF update.

```{r seed-12345-lfdr, fig.width=7.8, fig.height=6.5, fig.cap="Figure R5.2: Paired lfdr agreement for seed 12345. Each blue point is one of the 127,040 retained gene-variant pairs, and the grey dashed line denotes exact equality between the full-data and balanced-thinning lfdr values."}
plot_seed_12345_lfdr()
```

For this draw, the lfdr Spearman correlation is
`r format_decimal(specific_seed_metrics$spearman_lfdr, 4)` and the mean
absolute difference is
`r format_decimal(specific_seed_metrics$mean_absolute_lfdr_difference, 4)`.

## Estimated null weight

The full BF-updated fit estimates
$\widehat\pi_0 = `r format_decimal(full_pi0, 4)`$. The ten balanced refits give
`r format_range(thinned_pi0_range, 4)`, corresponding to changes of only
`r format_range(pi0_difference_range, 4, signed = TRUE)`.

```{r null-weight-stability, fig.width=7.6, fig.height=5.2, fig.cap="Figure R5.3: Estimated null weight across ten balanced thinnings. Blue points are the ten prespecified seeds, the blue box summarizes their distribution, and the orange dashed line is the single fixed full-data BF-updated estimate. The y-axis is fixed at 0.93 to 0.94 so the small absolute differences remain visible without visually overstating them."}
plot_null_weight_stability()
```

## Paired lfdr agreement across seeds

```{r across-seed-table, results='asis'}
render_scrollable_table(
  across_seed_summary_table,
  caption = paste(
    "Table R5.4: Range of BF-updated prior, paired-lfdr, and gene-discovery",
    "summaries across the ten balanced-thinning seeds."
  ),
  align = "lll",
  minimum_width = "1180px"
)
```

```{r paired-lfdr-agreement, fig.width=9.0, fig.height=5.2, fig.cap="Figure R5.4: Paired lfdr agreement across ten balanced thinnings. Each point summarizes all 127,040 retained pairs from one seed. The left facet shows the mean absolute full-versus-thinned lfdr difference; the right facet shows their Spearman correlation. Boxes summarize variability across seeds."}
plot_paired_lfdr_agreement()
```

The paired comparison conditions on each seed's retained pairs; it therefore
isolates the effect of re-estimating the FASH mixture after balancing gene
representation. The extremely small absolute differences and near-unit rank
correlations support the conclusion that the BF-updated lfdr calibration is
not sensitive to this thinning.

## Cumulative replication of discovered genes

The full BF-updated analysis discovers
`r format_integer(full_discovery_summary$discovered_pairs)` gene-variant pairs
from `r format_integer(full_discovered_gene_count)` unique genes at
cumulative-lfdr FDR 0.05. A single balanced refit discovers
`r format_integer(per_seed_discovery_range[1])`--`r format_integer(per_seed_discovery_range[2])`
genes because only 20 random variants are retained for each gene.

To evaluate repeated recovery, seeds are added in the prespecified order
12345, 22345, ..., 102345. At each step, the count is the union of all genes
discovered so far: a gene is counted once when it first appears and never
removed. Both cumulative curves are therefore monotonically nondecreasing by
construction.

```{r cumulative-discovered-genes, fig.width=9.2, fig.height=5.8, fig.cap="Figure R5.5: Cumulative unique gene discoveries as balanced-thinning seeds are added. The blue curve is the union of all genes discovered by the thinned refits; the orange curve is the subset also discovered in the full analysis. The grey dashed line marks the 1,177 genes discovered in the full analysis. The blue shaded band is the minimum-to-maximum cumulative union over every subset of the indicated number of seeds, so it shows how much the result depends on seed order."}
plot_cumulative_discovered_genes()
```

```{r cumulative-discovery-table, results='asis'}
render_scrollable_table(
  cumulative_discovery_table,
  caption = paste(
    "Table R5.5: Cumulative unique gene discoveries in the fixed prespecified",
    "seed order. Counts can only stay unchanged or increase as another seed",
    "is added."
  ),
  align = "rrrrr",
  minimum_width = "1260px"
)
```

After all ten seeds, the union contains
`r format_integer(final_cumulative_unique)` unique genes. Of these,
`r format_integer(final_cumulative_recovered)` are among the
`r format_integer(full_discovered_gene_count)` full-data discoveries, a
recovery rate of `r format_decimal(100 * final_recovery_fraction, 1)`%; the
remaining `r format_integer(final_additional_unique)` are not full-data gene
discoveries. This cumulative union is a descriptive replication diagnostic,
not a new multiple-testing procedure or a proposed discovery list.

## Interpretation and limitations

Balancing at 20 variants per eligible gene sharply reduces the unequal
representation of genes while preserving nearly every gene and retaining
within-gene genotype structure among the sampled variants. The BF-updated
prior weight and paired lfdr results are highly stable, which supports the
robustness of the main fit to the number of variants contributed by each gene.

This experiment does not remove LD among the 20 selected variants, establish
independence between gene-variant units, or address temporal correlation in
the estimated effect trajectories. Temporal correlation is examined
separately in Revision Real-data Sensitivity R4. In addition, gene-level
discovery counts can change because a finite random subset of variants is
tested for each gene, and unions across repeated seeds have increasing
opportunity to include a gene.

## Reproducibility

All displayed results are read from retained caches. The runner
validates the exact source-object sizes and SHA-256 hashes, refuses to
overwrite an existing seed cache, and records warnings and elapsed time for
every refit.

```{r reproduction-command, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/r5_balanced_thinning/run_balanced_thinning.R \
  --seeds 12345,22345,32345,42345,52345,62345,72345,82345,92345,102345 \
  --target-per-gene 20 \
  --alpha 0.05 \
  --output-id r5_balanced_thinning

Rscript --vanilla \
  code/revision_simulations/r5_balanced_thinning/test_reporting.R

Rscript --vanilla -e \
  'workflowr::wflow_build("analysis/revision_balanced_variant_thinning.rmd")'
```

The primary cache directory is
`output/revision_simulations/r5_balanced_thinning/`. The earlier internal
one-variant-per-gene page remains retained as experimental provenance; it is
not linked from the formal revision index and is not used for any result on
this page.
