---
title: "Revision Real-data Analysis R6: enhancer enrichment of FASH dynamic eQTLs"
author: "Ziang Zhang"
date: "2026-08-25"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 2
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9.5,
  fig.height = 3.9,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(file.path(
  "code",
  "revision_simulations",
  "internal",
  "fash_strober_enhancer_comparison",
  "reporting.R"
))
```

<style>
html,
body {
  max-width: 100%;
  overflow-x: hidden;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
pre {
  max-width: 100%;
  overflow-x: auto;
}
</style>

## Question

The reviewer asked whether the dynamic eQTLs that FASH discovers show enrichment
in functionally relevant enhancers. This page is a short, orthogonal check on
that question: it asks whether the discovered variants fall inside enhancer
regions of the cell types involved in the differentiation more often than the
variants tested in this study do. The estimates below are descriptive and
exploratory. They are not a fully adjusted functional-enrichment analysis, and
they are not used anywhere as evidence that individual discoveries are correct.

Two sections, each comparing all discovered variants against one lead variant
per gene:

* **Section 1** compares current FASH with Strober's linear and
  quadratic/nonlinear analyses.
* **Section 2** compares all FASH discoveries with the subset at variants that
  neither Strober analysis identified.

## Enhancer annotations

We used enhancer annotations from the Roadmap Epigenomics Project 15-state
ChromHMM segmentation (Roadmap Epigenomics Consortium et al. 2015). We treated
the Roadmap enhancer states (EnhG and Enh) as enhancer regions and focused on
three epigenomes spanning the differentiation trajectory of this
iPSC-to-cardiomyocyte study:

| Epigenome | Tissue | Stage |
|---|---|---|
| E020 | iPS-20b | starting state |
| E013 | hESC-derived CD56+ mesoderm | intermediate state |
| E095 | left ventricle | terminal state |

Each tested variant is treated as a single base and intersected against the
enhancer intervals of each epigenome, giving one indicator per variant per
epigenome over all
`r format_integer(background_variant_count)` tested variants.

## Section 1: FASH versus Strober

```{r section1, fig.cap="Figure 1. Enhancer enrichment for current FASH, Strober linear, and Strober quadratic/nonlinear discoveries, using all discovered variants (left) and one lead variant per gene (right). Enhancer annotations are the Roadmap 15-state ChromHMM enhancer states for E020 (iPSC), E013 (mesoderm), and E095 (left ventricle). Fold enrichment is the overlap rate in the variant set divided by the overlap rate among all tested variants; the dashed line marks no enrichment. Bars are leave-one-autosome-out jackknife 95% intervals."}
plot_method_comparison()
```

With one lead variant per gene, FASH's `r set_size("current_lead")` lead
variants are enriched at the intermediate and terminal states:
`r describe_fold("current_lead", e013)` at the E013 mesoderm enhancers and
`r describe_fold("current_lead", e095)` at the E095 left-ventricle enhancers.
The starting state gives no signal,
`r describe_fold("current_lead", e020)`. Both Strober analyses point the same
way at E013 and E095, with intervals that overlap FASH's.

Using every discovered variant the FASH estimates are smaller:
`r describe_fold("current_all", e095)` at E095, and
`r describe_fold("current_all", e013)` at E013, which is not distinguishable
from the background. The Strober sets carry somewhat larger estimates in this
view. The three discovery sets differ in size
and membership and each has its own gene composition, so these differences are
not a formal comparison between methods.

## Section 2: FASH-specific discoveries

An rsID is counted as FASH-only if it appears in neither Strober analysis,
regardless of which gene it was paired with; this is the conservative reading,
since an enhancer annotation is a property of a variant rather than of a
gene-variant pair. For the lead column, FASH's lowest-lfdr variant is chosen
for each gene first and the exclusion is applied afterwards, so a gene whose
best variant Strober also found drops out rather than falling back to a weaker
one. That leaves `r set_size("current_only_all")` of FASH's
`r set_size("current_all")` discovered variants, and
`r set_size("current_only_lead")` of its `r set_size("current_lead")` lead
variants.

```{r section2, fig.cap="Figure 2. Enhancer enrichment for FASH discoveries at variants that neither Strober analysis identified (blue), against all FASH discoveries as a reference (grey), using all discovered variants (left) and one lead variant per gene (right). Enhancer annotations are the Roadmap 15-state ChromHMM enhancer states for E020 (iPSC), E013 (mesoderm), and E095 (left ventricle). Fold enrichment is the overlap rate in the variant set divided by the overlap rate among all tested variants; the dashed line marks no enrichment. Bars are leave-one-autosome-out jackknife 95% intervals."}
plot_fash_only()
```

The lead-per-gene enrichment survives the exclusion largely intact:
`r describe_fold("current_only_lead", e013)` at E013 and
`r describe_fold("current_only_lead", e095)` at E095, against
`r fold_value("current_lead", e013)`-fold and
`r fold_value("current_lead", e095)`-fold for the unfiltered FASH lead set.
Using every FASH-only variant the estimates are weaker, with intervals that
reach or cross 1: `r describe_fold("current_only_all", e013)` at E013 and
`r describe_fold("current_only_all", e095)` at E095.

Taken together, FASH discoveries — including lead variants that neither Strober
analysis identified — show enrichment in enhancer annotations relevant to this
differentiation, which offers orthogonal support for their regulatory
relevance. The signal is clearest for the mesoderm and left-ventricle states
and under lead-per-gene selection.

## Methods

**Background.** All `r format_integer(background_variant_count)` variants
tested in this study that lie on an autosome and participate in at least one
tested gene-variant pair. No external reference panel enters, so the background
is exactly the cis-eQTL testing universe from which the discoveries were drawn.

**Fold enrichment.** For a variant set and an enhancer annotation,

$$\text{fold enrichment} = \frac{\text{proportion of selected variants overlapping the enhancer}}{\text{proportion of all tested variants overlapping the enhancer}}$$

The background excludes the set being tested, so a set is never part of its own
comparator. No matched controls are used: the purpose here is a simple
exploratory check, and every background variant already comes from the same
testing universe.

**Discovery sets.** FASH uses cumulative lfdr at FDR 0.05 after the BF
correction; both Strober analyses use their reported eFDR 0.05, with
`non_linear_dynamic_eqtls_5_pc.txt` labelled quadratic/nonlinear throughout.
"One lead variant per gene" keeps, for each discovered gene, the variant with
the lowest lfdr (FASH) or the smallest p-value (Strober). This reduces the
repeated contribution of multiple variants tagging the same signal; the
all-variant view reports the complete discovery set.

**Uncertainty.** Uncertainty was assessed using a leave-one-autosome-out
jackknife to account approximately for local dependence among variants. All
intervals shown are 95% jackknife intervals.

## Reproducibility

The figures on this page are computed at render time by
[`enrichment_api.R`](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/internal/fash_strober_enhancer_comparison/enrichment_api.R),
through exactly the calls below:

```{r api-demo, eval=FALSE}
source(
  "code/revision_simulations/internal/fash_strober_enhancer_comparison/enrichment_api.R"
)

# 1. Background: every variant tested in this study.
background <- load_background()

# 2. Variant sets: the FASH and Strober discovery sets, all and lead-per-gene.
sets <- load_variant_sets()

# 3. The three Roadmap enhancer annotations.
annotations <- load_annotation_groups("custom", groups = c(
  "Roadmap E020 iPS-20b: Enhancer",
  "Roadmap E013 hESC-derived CD56+ mesoderm: Enhancer",
  "Roadmap E095 left ventricle: Enhancer"
))

# 4. Fold enrichment against the tested-variant background, with
# 5. leave-one-autosome-out jackknife 95% intervals.
compute_enrichment(sets, background, annotations, controls = "background")
```

`compute_enrichment()` takes any named list of variant vectors, so substituting
your own sets needs no other change:

```{r api-own-sets, eval=FALSE}
compute_enrichment(
  list(my_hits = readLines("my_variants.txt"), fash_only = sets$current_only_lead),
  background, annotations, controls = "background"
)
```

Rebuilding the inputs from source — the annotation intervals and the tested
variant universe, then the discovery sets:

```{bash rebuild, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/variant_annotation_enrichment/download_variant_annotations.R
Rscript --vanilla \
  code/revision_simulations/internal/variant_annotation_enrichment/run_variant_annotation_enrichment.R
Rscript --vanilla \
  code/revision_simulations/internal/fash_strober_enhancer_comparison/run_fash_strober_enhancer_comparison.R
```

Intermediate data stays under `output/`, which is not tracked in git; the
repository carries code, not data. Rebuilding this page:

```{bash reproduction-command, eval=FALSE}
Rscript --vanilla -e \
  'workflowr::wflow_build("analysis/revision_enhancer_enrichment.rmd")'
```

## Reference

Roadmap Epigenomics Consortium, Kundaje, A., Meuleman, W., Ernst, J.,
Bilenky, M., Yen, A., et al. (2015). Integrative analysis of 111 reference
human epigenomes. *Nature* **518**, 317–330.
<https://doi.org/10.1038/nature14248>
