---
title: "Revision Simulation R1: real-genotype eQTLs with random B-spline effects"
author: "Ziang Zhang"
date: "2026-08-22"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9,
  fig.height = 6.5,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source("code/revision_simulations/shared/simulation_functions.R")
source("code/revision_simulations/shared/real_genotype_one_per_gene.R")
source("code/revision_simulations/r1_random_bspline/reporting.R")
```

<style>
@media (max-width: 767px) {
  html,
  body {
    max-width: 100%;
    overflow-x: hidden;
  }
  .main-container table {
    display: block;
    width: 100%;
    max-width: 100%;
    overflow-x: auto;
    -webkit-overflow-scrolling: touch;
  }
  .main-container .MathJax_Display {
    max-width: 100%;
    overflow-x: auto;
    overflow-y: hidden;
    -webkit-overflow-scrolling: touch;
  }
}
</style>

## Overview

This real-genotype simulation uses random cubic B-spline dynamic effects as a
non-IWP data-generating mechanism. It is the broad, smoothly varying baseline;
the separate [spiky transient analysis](revision_combined_spiky_genotype_simulation.html)
examines sharply localized effects.

### Result provenance for the 0.1.43 refresh

The FASH-IWP1 and FASH-linear rows were recomputed on the same five fixed
simulation seeds with `fashr` 0.1.43 at RemoteSha
`bf223df75da6e41ae48607a56b4cd12d7c3b24e7`. The targeted result family
completed at `r current_fash_manifest$generated_at`. The direct-interaction
permutation tests were not rerun: their rows come from the unchanged formal
historical cache. Before combining the rows, the report verifies every
retained SHA-256 digest and requires exact selected-pair/genotype agreement
plus numerical agreement of the simulated inputs to `1e-12`.

```{r result-provenance-table}
knitr::kable(
  result_provenance_table,
  align = "l",
  caption = paste(
    "Table R1.1: Computation source and validation status for each result",
    "component in the refreshed page."
  )
)
```

For donor $i$, variant $j$, and time $t$, expression is generated as

$$
Y_{ijt} = \alpha_{jt} + G_{ij}\beta_j(t) +
X_i^T\gamma_{jt} + \epsilon_{ijt},
\qquad \epsilon_{ijt} \sim N(0, 1).
$$

The genotype matrix is taken from the real YRI data used in the manuscript.
For each simulation seed, one tested cis variant is sampled uniformly from
each of the `6,362` tested genes, and its observed dosage vector is retained
for the same `19` donors. The resulting design has one locus-level unit per
gene and an empirical MAF distribution; a variant tested for multiple genes
may therefore appear more than once. R1, R2, R3A, and R3B use the exact same
sampled gene-variant pairs and genotype matrix for a given seed.

The five columns of $X$ remain simulated independently from $N(0,1)$ and are
standardized across donors. Independently for each covariate, gene, and time,
$\gamma_{kjt}\sim N(0,0.5^2)$. The current simulation sets $\alpha_{jt}=0$;
each fitted regression still includes an intercept. Effect trajectories,
covariates, covariate effects, and expression noise remain simulated.

```{r r1-simulate-genotypes-and-covariates, eval=FALSE}
seed <- 12345
component_seeds <- revision_component_seeds(seed)

genotype_cache <- readRDS(
  "output/revision_simulations/shared/real_genotype_one_per_gene_J6362_pilot5/genotype_samples.rds"
)
genotype_sample <- genotype_cache$samples[[as.character(seed)]]
G <- genotype_sample$G
covariates <- simulate_covariate_matrix(
  n_donors = 19,
  n_covariates = 5,
  seed = component_seeds[["covariates"]]
)
```

### Real-genotype design and provenance

```{r genotype-provenance-table}
knitr::kable(
  genotype_provenance_table,
  align = "l",
  caption = "Table R1.2: Provenance and selection rule for the real-genotype simulation design."
)
```

```{r genotype-sampling-table}
knitr::kable(
  genotype_sampling_table,
  align = "l",
  caption = paste(
    "Table R1.3: Per-seed genotype sampling summary. Repeated assignments",
    "refer to the same tested variant selected for more than one gene."
  )
)
```

Each simulated dataset has `19` donors, `16` time points, `6,362` gene-level
units, and five PC-like covariates. The classes are fixed at:

- `r format_integer(dynamic_count)` dynamic eQTLs with random cubic B-spline effects;
- `r format_integer(constant_count)` non-dynamic eQTLs with constant effects drawn from $N(0,1)$;
- `r format_integer(zero_count)` zero-effect units.

For a dynamic variant,

$$
\beta_j(t) = \mu_j + h_j(t),
\qquad
\mu_j \sim N(0,1),
\qquad
h_j(t) = \sum_{k=1}^{6} b_{jk} B_k(t),
\qquad b_{jk} \stackrel{\mathrm{iid}}{\sim} N(0,1).
$$

Only the time-varying deviation $h_j(t)$ is mean-centered over the observed
times and scaled to have maximum absolute value `2`. The independent
$\mu_j$ is the genetic main effect, so the final trajectory is not constrained
to be centered at zero. With exact integer class counts, the true dynamic-null
proportion is $\pi_0 = 5090/6362 =$ `r format_decimal(true_pi0, 4)`.

```{r r1-simulate-effects-and-expression, eval=FALSE}
class_probs <- c(
  dynamic_bspline = 0.20,
  constant = 0.40,
  zero = 0.40
)
effect_sim <- simulate_variant_effect_curves(
  n_variants = 6362,
  time_grid = 0:15,
  class_probs = class_probs,
  dynamic_amplitude = 2,
  bspline_df = 6,
  bspline_coefficient_sd = 1,
  constant_sd = 1,
  dynamic_main_effect_sd = 1,
  exact_class_counts = TRUE,
  seed = component_seeds[["functional_truth"]],
  scenario = "r1_real_genotype_one_per_gene_random_bspline_main_effect_dynamic_eqtl"
)
effect_sim <- reassign_effect_simulation_by_maf(
  effect_sim = effect_sim,
  maf = genotype_sample$variant_info$observed_maf,
  class_probs = class_probs,
  seed = component_seeds[["classes"]],
  n_strata = 10
)
expression_sim <- simulate_eqtl_expression_from_genotypes(
  G = G,
  beta_matrix = effect_sim$beta_matrix,
  time_grid = 0:15,
  covariates = covariates,
  expression_noise_sd = 1,
  covariate_effect_sd = 0.5,
  intercept_sd = 0,
  seed = component_seeds[["expression"]]
)
```

At each time point, the genetic effect and its standard error are estimated
from `Y ~ G + PC1 + PC2 + PC3 + PC4 + PC5`. The real-data-like t-statistic
standard-error correction uses

$$
\mathrm{df} = 19 - 7 = 12,
$$

corresponding to the intercept, genotype, and five covariates. All methods use
the same per-time estimates and corrected standard errors.

```{r r1-estimate-and-fit-fash, eval=FALSE}
eqtl_summary <- estimate_eqtl_summaries_from_genotypes(
  G = G,
  expression = expression_sim$expression,
  covariates = covariates,
  apply_t_se_correction = TRUE
)
datasets <- make_fash_datasets_from_eqtl_summary(
  beta_hat = eqtl_summary$beta_hat,
  se = eqtl_summary$se,
  true_beta = effect_sim$beta_matrix,
  time_grid = 0:15,
  unit_info = effect_sim$unit_info,
  scenario = "r1_real_genotype_one_per_gene_random_bspline_main_effect_dynamic_eqtl"
)

common_sd_grid <- default_revision_grid()
common_pred_step <- 1
common_penalty <- 10

fash_iwp1_raw <- fashr::fash(
  Y = "y",
  smooth_var = "x",
  S = "sd",
  data_list = datasets,
  order = 1,
  grid = common_sd_grid,
  num_basis = 20,
  pred_step = common_pred_step,
  penalty = common_penalty,
  num_cores = 4,
  verbose = FALSE
)
fash_iwp1_bf <- fashr::BF_update(fash_iwp1_raw, plot = FALSE)

fash_linear_raw <- fit_linear_mixture_fash(
  datasets = datasets,
  grid = common_sd_grid,
  pred_step = common_pred_step,
  penalty = common_penalty
)
fash_linear_bf <- BF_update_linear_mixture_fash(fash_linear_raw)
```

The reported curves and tables average five independently simulated datasets
with seeds `12345`, `22345`, `32345`, `42345`, and `52345`. The full
seed-12345 object is used only to display representative trajectories.

Truth labels are reassigned within empirical-MAF deciles so that dynamic,
constant, and zero-effect units have closely matched MAF distributions.

```{r truth-maf-balance-table}
knitr::kable(
  truth_maf_balance_table,
  align = "l",
  caption = "Table R1.4: Empirical MAF balance across truth classes for each simulation seed."
)
```

## Example simulated eQTLs

The panels show three zero-effect variants, three constant eQTLs, and three
random B-spline dynamic eQTLs. Points are per-time genetic effect estimates,
bars show two corrected standard errors, and red lines are the true effects.

```{r example-trajectories, fig.width=13, fig.height=11, fig.cap="Figure R1.1: Representative zero-effect, constant-effect, and random B-spline dynamic eQTLs from seed 12345. Points are estimated genetic effects, vertical bars span two corrected standard errors, and red curves are the true effects."}
plot_genotype_eqtl_examples(
  out,
  classes = c("zero", "constant", "dynamic_bspline"),
  n_per_class = 3,
  seed = 1,
  true_curve_n = 500
)
```

## Comparison 1: IWP versus linear FASH

This comparison aligns the exact null, scale grid, `pred_step`, and
null-favoring penalty. `FASH-IWP1` uses the first-order IWP alternative. For
`FASH-linear`, let $h=1$ and $z(t)=\{t-\min(t)\}/h$. Its prior is

$$
\beta_j(t)=a_j+\theta_j z(t),
\qquad
\theta_j\sim\pi_0\delta_0+
\sum_{k=1}^{K}\pi_k N(0,\tau_k^2).
$$

The zero-grid component is the same unrestricted constant null used by IWP1.
For each positive component, $\tau_k$ is the prior SD of the linear departure
at $\min(t)+h$. The complete grid is one exact zero plus 51 positive values
from $\exp(-5)$ to `1`; all weights are estimated jointly with `penalty = 10`.
IWP1 uses the same numerical grid for its one-step predictive SD. This matches
the one-step marginal scale while leaving the linear and IWP covariance shapes
different.

The BF update does not retain the raw penalized non-null weights. Following
`fashr::BF_update()`, it re-fits an unpenalized full-grid mixture, normalizes
the positive-grid weights to define the conditional alternative, collapses
that alternative into a Bayes factor, and then updates the total null weight.
The table below summarizes both the number of active positive-grid components
and their conditional RMS pred-step SD.

### Estimated dynamic-null proportion

The table reports the fitted prior weight on each method's constant component.
The true finite-simulation dynamic-null proportion is
$\pi_0 = 5090/6362 =$ `r format_decimal(true_pi0, 4)`.

```{r pi0-table}
knitr::kable(
  pi0_table,
  align = "l",
  caption = paste(
    "Table R1.5: Estimated dynamic-null proportions and linear-mixture prior",
    "diagnostics across five random B-spline simulation replicates."
  )
)
```

### Results at alpha 0.05

```{r fash-alpha-table}
knitr::kable(
  fash_table,
  align = "l",
  caption = paste(
    "Table R1.6: Monte Carlo power and empirical FDR at alpha 0.05 for IWP",
    "and linear FASH under random B-spline effects."
  )
)
```

### Power as a function of alpha

```{r fash-power, fig.cap="Figure R1.2: Monte Carlo power across nominal alpha for IWP and linear FASH under random B-spline effects. Curves identify the raw and BF-updated methods shown in the in-figure legend; uncertainty summarizes five simulation seeds."}
plot_mc_alpha_curves(
  fash_alpha,
  metric = "power",
  title = "IWP versus linear FASH: Monte Carlo power",
  subtitle = "Five seeds; real YRI genotypes; N = 19, J = 6362 genes",
  style_profile = "combined"
)
```

### Empirical FDR as a function of alpha

```{r fash-fdr, fig.cap="Figure R1.3: Monte Carlo empirical FDR across nominal alpha for IWP and linear FASH under random B-spline effects. Curves identify methods in the in-figure legend, and the diagonal reference marks equality between empirical FDR and nominal alpha."}
plot_mc_alpha_curves(
  fash_alpha,
  metric = "fdr",
  title = "IWP versus linear FASH: Monte Carlo empirical FDR",
  subtitle = "Five seeds; real YRI genotypes; N = 19, J = 6362 genes",
  legend_position = "topleft",
  style_profile = "combined"
)
```

At $\alpha = 0.05$, IWP-Raw and IWP-BF have mean power
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-IWP1-Raw", "mean_power"))`
and
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-IWP1-BF", "mean_power"))`,
compared with
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-linear-Raw", "mean_power"))`
and
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-linear-BF", "mean_power"))`
for the corresponding linear fits. Their empirical FDR estimates are
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-IWP1-Raw", "mean_fdr"))`
and
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-IWP1-BF", "mean_fdr"))`
for IWP-Raw and IWP-BF, and
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-linear-Raw", "mean_fdr"))`
and
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-linear-BF", "mean_fdr"))`
for linear-Raw and linear-BF. These values are paired across the same five
simulated datasets; they are pilot empirical-FDR summaries rather than a claim
of general calibration.

## Comparison 2: IWP FASH versus direct interaction tests

The direct baselines fit individual-level expression using linear or
linear-plus-quadratic genotype-by-time interactions and use permutation-based
eFDR. To make this comparison favorable to the direct methods, their eFDR
calculation is supplied with the true $\pi_0 = 5090/6362$ and uses `B = 100`
donor-level permutations per replicate.

Only the IWP FASH rows in this section were recomputed with `fashr` 0.1.43.
The direct linear and quadratic rows are read unchanged from the historical
formal cache after the report confirms that its simulated genotype,
covariates, truth, and expression agree with the retained 0.1.43 payload.

```{r r1-direct-interaction-efdr, eval=FALSE}
out <- add_direct_interaction_efdr_results_to_genotype_output(
  out = out,
  n_permutations = 100,
  alpha = 0.05,
  seed = seed + 10000L,
  lambda = 0.5,
  pi0_method = "conservative",
  true_pi0 = 5090 / 6362,
  include_true_pi0 = TRUE,
  permute_covariates_with_expression = TRUE,
  num_cores = 4,
  overwrite = TRUE,
  verbose = FALSE
)
```

### Results at alpha 0.05

```{r direct-alpha-table}
knitr::kable(
  direct_table,
  align = "l",
  caption = paste(
    "Table R1.7: Monte Carlo power and empirical FDR at alpha 0.05 for IWP",
    "FASH and direct interaction tests under random B-spline effects."
  )
)
```

### Power as a function of alpha

```{r direct-power, fig.cap="Figure R1.4: Monte Carlo power across nominal alpha for IWP FASH and direct linear or quadratic interaction tests. The in-figure legend identifies methods; direct-test eFDR uses the true pi0 and 100 donor-level permutations."}
plot_mc_alpha_curves(
  direct_alpha,
  metric = "power",
  title = "IWP FASH versus direct interaction: Monte Carlo power",
  subtitle = "Five seeds; direct eFDR uses true pi0 = 0.8001 and B = 100 permutations",
  style_profile = "combined"
)
```

### Empirical FDR as a function of alpha

```{r direct-fdr, fig.cap="Figure R1.5: Monte Carlo empirical FDR across nominal alpha for IWP FASH and direct interaction tests. The in-figure legend identifies methods, and the diagonal reference marks equality between empirical FDR and nominal alpha."}
plot_mc_alpha_curves(
  direct_alpha,
  metric = "fdr",
  title = "IWP FASH versus direct interaction: Monte Carlo empirical FDR",
  subtitle = "Five seeds; direct eFDR uses true pi0 = 0.8001 and B = 100 permutations",
  legend_position = "bottomright",
  style_profile = "combined"
)
```

At $\alpha = 0.05$, IWP-BF has mean power
`r format_decimal(get_mc_metric(direct_alpha_005, "FASH-IWP1-BF", "mean_power"))`,
compared with
`r format_decimal(get_mc_metric(direct_alpha_005, "Direct-linear-LRT-eFDR-true-pi0", "mean_power"))`
for the direct linear test and
`r format_decimal(get_mc_metric(direct_alpha_005, "Direct-quadratic-LRT-eFDR-true-pi0", "mean_power"))`
for the direct quadratic test. IWP-BF has the highest power while maintaining
mean empirical FDR
`r format_decimal(get_mc_metric(direct_alpha_005, "FASH-IWP1-BF", "mean_fdr"))`.
The direct tests are more conservative but less powerful, even with the true
null proportion supplied.

## Interpretation

For broad random B-spline effects, the tables and curves above compare IWP and
linear covariance shapes after matching their exact null, one-step SD grid,
and null-favoring penalty. R1 and R2 are exactly paired on their sampled
gene-variant units and real genotype matrices for every seed; only their
simulated effect geometries differ. The separate spiky transient page applies
the same comparison to sharply localized one-, two-, and three-peak effects.

## Reproducibility

The
[R1 simulation driver](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/r1_random_bspline/run_random_bspline_mc_replication.R)
uses the
[shared simulation functions](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/shared/simulation_functions.R).
Real-genotype selection, dosage extraction, validation, and MAF-balanced truth
assignment are implemented in the
[shared real-genotype helper](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/shared/real_genotype_one_per_gene.R),
and the versioned input cache is created by the
[shared cache builder](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/shared/build_real_genotype_one_per_gene_cache.R).
The page-specific cache validation and formatting are in
[R1 reporting code](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/r1_random_bspline/reporting.R).
The shared
[0.1.43 reporting overlay](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/r1_r2_fashr0143/reporting_overlay.R)
validates the completed targeted cache and combines its FASH rows with only
the direct rows from the historical cache. The following command regenerates
the full historical-style Monte Carlo cache, including direct tests. It is
shown for reproducibility but is intentionally not evaluated during workflowr
builds; the displayed FASH rows read the separate targeted 0.1.43 cache.

The full Monte Carlo cache can be regenerated with:

```{bash regenerate-r1-cache, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/shared/build_real_genotype_one_per_gene_cache.R

Rscript --vanilla \
  code/revision_simulations/r1_random_bspline/run_random_bspline_mc_replication.R \
  --J 6362 \
  --n-donors 19 \
  --n-covariates 5 \
  --noise-sd 1 \
  --dynamic-main-effect-sd 1 \
  --efdr-permutations 100 \
  --seed-list 12345,22345,32345,42345,52345 \
  --genotype-cache output/revision_simulations/shared/real_genotype_one_per_gene_J6362_pilot5/genotype_samples.rds \
  --output-id r1_real_genotype_one_per_gene_J6362_random_bspline_main_effect_linear_mixture_predstep1_penalty10_pilot5 \
  --num-cores 4
```

After the formal R1-R3 caches have been generated, exact pairing of every
sampled gene-variant unit and genotype matrix is validated with:

```{bash validate-r1-r2-r3-pairing, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/shared/validate_r1_r2_real_genotype_pairing.R \
  --r1-output-id r1_real_genotype_one_per_gene_J6362_random_bspline_main_effect_linear_mixture_predstep1_penalty10_pilot5 \
  --r2-output-id r2_real_genotype_one_per_gene_J6362_timed_cosine_one_two_three_peak_main_effect_linear_mixture_predstep1_penalty10_pilot5 \
  --r3-output-id r3_real_genotype_one_per_gene_J6362_matched_functional_relative_clearance_main_effect_pilot5
```

The report requires a successful
`summary/real_genotype_pairing_validation.csv` artifact and will stop if the
R1/R2/R3 genotype designs are not exactly paired.

The recomputed FASH result family is stored under:

```text
output/revision_simulations/mc/r1_r2_fashr0143/
```

The unchanged historical full fit and direct-interaction results are stored
under:

```text
output/revision_simulations/mc/r1_real_genotype_one_per_gene_J6362_random_bspline_main_effect_linear_mixture_predstep1_penalty10_pilot5/
```
