---
title: "Revision Real-data Analysis R7: matched FASH-linear comparison"
author: "Ziang Zhang"
date: "2026-08-21"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9.6,
  fig.height = 7.2,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(file.path(
  "code",
  "revision_simulations",
  "internal",
  "matched_fash_linear_real_data",
  "reporting.R"
))
```

<style>
.matched-banner {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border: 1px solid #d9a441;
  border-left: 6px solid #d9a441;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.matched-result {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.matched-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
  margin-bottom: 1.2em;
}
.matched-table-scroll caption,
p.caption {
  color: #243447;
  font-weight: 650;
  text-align: left;
  padding: 0.35em 0 0.55em;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
.MathJax_Display,
pre {
  max-width: 100%;
  overflow-x: auto;
}
</style>

<div class="matched-banner">
<strong>Revision analysis R7.</strong> This page reports a fresh full-data
matched FASH-linear rerun with `fashr` 0.1.43.
</div>

## Analysis definition

Both FASH fits use the same 1,009,173 time-specific-PC effect estimates,
standard errors, exact constant null, 52-point one-day predictive-SD grid,
`betaprec=0`, `pred_step=1`, `penalty=10`, corrected BF adjustment, and
cumulative-lfdr FDR 0.05.

For FASH-linear,

\[
\beta_j(t)=a_j+b_j(t-t_{j,\min}),\qquad
b_j\sim\sum_{k=1}^{51}\pi_k
N\!\left(0,\left\{s_k/\mathrm{pred\_step}\right\}^2\right),
\]

where $s_k=\operatorname{SD}(b_j\,\mathrm{pred\_step})$ is
`sd_linear_step`. The exact-null component is $s_0=0$. Its 51 positive
values are numerically identical to the IWP1 `psd` grid.

The exact fitting code from the full-data runner is shown below. The linear
likelihood does not use IWP basis coefficients, so `betaprec` is retained as a
matched setting rather than passed to the linear-mixture fitting helper.

```{r fash-linear-fitting-code, eval=FALSE}
# Match the predictive-SD grid and fitting options to the IWP1 reference fit.
matched_grid <- as.numeric(current_raw$psd_grid)
matched_pred_step <- as.numeric(current_raw$settings$pred_step)
matched_penalty <- as.integer(current_raw$settings$penalty)
matched_betaprec <- as.numeric(current_raw$settings$betaprec)
matched_likelihood <- as.character(current_raw$settings$likelihood)

# Fit the exact-null plus 51-component linear-slope mixture.
linear_raw <- fit_linear_mixture_fash_from_stats(
  statistics = statistics,
  grid = matched_grid,
  pred_step = matched_pred_step,
  penalty = matched_penalty,
  statistic_time_span = 15,
  n_time = 16
)

# Record the matched likelihood settings in the fitted object.
linear_raw$settings$betaprec <- matched_betaprec
linear_raw$settings$likelihood <- matched_likelihood

# Apply the same corrected Bayes-factor adjustment used for FASH-IWP1.
linear_bf <- BF_update_linear_mixture_fash(linear_raw)
```

```{r matched-settings-table, results='asis'}
settings_to_report <- matched_settings_table[, c(
  "setting", "fash_iwp1", "fash_linear", "alignment"
)]
names(settings_to_report) <- c(
  "Setting", "FASH-IWP1", "FASH-linear", "Alignment"
)
render_scrollable_table_matched(
  settings_to_report,
  caption = "Table 1. Matched data, prior-scale, fitting, and discovery settings.",
  minimum_width = "1040px"
)
```

The scientific selection rules used below are:

```{r discovery-rules, eval=FALSE}
alpha <- 0.05

iwp1_bf_indices <- select_cumulative_lfdr_calls(
  fash_iwp1_bf$lfdr,
  alpha
)
linear_bf_indices <- select_cumulative_lfdr_calls(
  fash_linear_bf$lfdr,
  alpha
)
strober_quadratic_discoveries <- subset(strober_quadratic, eFDR <= alpha)
strober_linear_discoveries <- subset(strober_linear, eFDR <= alpha)
```

## Discovery counts

<div class="matched-result">
After BF adjustment, FASH-IWP1 identifies
`r format_integer_matched(iwp_bf_counts$pair_count)` pairs,
`r format_integer_matched(iwp_bf_counts$gene_count)` genes, and
`r format_integer_matched(iwp_bf_counts$variant_count)` unique variants.
FASH-linear identifies
`r format_integer_matched(linear_bf_counts$pair_count)` pairs,
`r format_integer_matched(linear_bf_counts$gene_count)` genes, and
`r format_integer_matched(linear_bf_counts$variant_count)` unique variants.
</div>

```{r discovery-counts-table, results='asis'}
render_scrollable_table_matched(
  discovery_counts_display,
  caption = "Table 2. Raw and BF-adjusted cumulative-lfdr FDR-0.05 discovery counts.",
  digits = 6L,
  minimum_width = "820px"
)
```

The directional coverage is the fraction of each IWP discovery set also found
by FASH-linear.

```{r iwp1-linear-proportion-table, results='asis'}
render_scrollable_table_matched(
  iwp1_linear_proportion_display,
  caption = "Table 3. BF-adjusted FASH-IWP1 discoveries covered by BF-adjusted FASH-linear at the three reporting levels.",
  minimum_width = "860px"
)
```

```{r iwp2-linear-proportion-table, results='asis'}
render_scrollable_table_matched(
  iwp2_linear_proportion_display,
  caption = "Table 4. BF-adjusted FASH-IWP2 discoveries covered by BF-adjusted FASH-linear at the three reporting levels.",
  minimum_width = "860px"
)
```

FASH-linear covers 92.6% of the FASH-IWP1 pair discoveries and 47.7% of the
FASH-IWP2 pair discoveries.

## Four-method overlap

The following Venn diagrams extend the real-data comparison by adding
BF-adjusted FASH-linear to Strober quadratic, Strober linear, and BF-adjusted
FASH-IWP1. Strober discoveries use eFDR at most 0.05.

```{r venn-pairs, fig.width=9.8, fig.height=7.4, fig.cap="Figure 1. Gene-variant-pair overlap among Strober quadratic, Strober linear, BF-adjusted FASH-IWP1, and BF-adjusted FASH-linear. FASH sets use cumulative-lfdr FDR 0.05; Strober sets use eFDR 0.05."}
pair_sets <- list(
  `Strober quad.` =
    venn_sets$`Gene-variant pairs`$`Strober quadratic`,
  `Strober linear` =
    venn_sets$`Gene-variant pairs`$`Strober linear`,
  `FASH-linear` =
    venn_sets$`Gene-variant pairs`$`FASH-linear BF`,
  `FASH-IWP1` =
    venn_sets$`Gene-variant pairs`$`FASH-IWP1 BF`
)

ggVennDiagram::ggVennDiagram(
  pair_sets,
  label = "count",
  label_alpha = 0,
  label_size = 3.7,
  set_size = 3.8,
  edge_size = 0.8
) +
  ggplot2::scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  ggplot2::theme(
    legend.position = "right",
    plot.margin = ggplot2::margin(20, 40, 20, 40)
  )
```

```{r venn-genes, fig.width=9.8, fig.height=7.4, fig.cap="Figure 2. Gene-level overlap among Strober quadratic, Strober linear, BF-adjusted FASH-IWP1, and BF-adjusted FASH-linear. A gene enters a set when at least one of its tested variants is discovered by that method."}
gene_sets <- list(
  `Strober quad.` = venn_sets$Genes$`Strober quadratic`,
  `Strober linear` = venn_sets$Genes$`Strober linear`,
  `FASH-linear` = venn_sets$Genes$`FASH-linear BF`,
  `FASH-IWP1` = venn_sets$Genes$`FASH-IWP1 BF`
)

ggVennDiagram::ggVennDiagram(
  gene_sets,
  label = "count",
  label_alpha = 0,
  label_size = 3.7,
  set_size = 3.8,
  edge_size = 0.8
) +
  ggplot2::scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  ggplot2::theme(
    legend.position = "right",
    plot.margin = ggplot2::margin(20, 40, 20, 40)
  )
```

```{r venn-variants, fig.width=9.8, fig.height=7.4, fig.cap="Figure 3. Unique-variant overlap among Strober quadratic, Strober linear, BF-adjusted FASH-IWP1, and BF-adjusted FASH-linear. Each rsID is counted once per method, regardless of the number of associated genes."}
variant_sets <- list(
  `Strober quad.` = venn_sets$Variants$`Strober quadratic`,
  `Strober linear` = venn_sets$Variants$`Strober linear`,
  `FASH-linear` = venn_sets$Variants$`FASH-linear BF`,
  `FASH-IWP1` = venn_sets$Variants$`FASH-IWP1 BF`
)

ggVennDiagram::ggVennDiagram(
  variant_sets,
  label = "count",
  label_alpha = 0,
  label_size = 3.7,
  set_size = 3.8,
  edge_size = 0.8
) +
  ggplot2::scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  ggplot2::theme(
    legend.position = "right",
    plot.margin = ggplot2::margin(20, 40, 20, 40)
  )
```

The BF-adjusted IWP1--linear intersections are
`r format_integer_matched(iwp_linear_overlap$Intersection[iwp_linear_overlap$Unit == "Gene-variant pairs"])`
pairs,
`r format_integer_matched(iwp_linear_overlap$Intersection[iwp_linear_overlap$Unit == "Genes"])`
genes, and
`r format_integer_matched(iwp_linear_overlap$Intersection[iwp_linear_overlap$Unit == "Variants"])`
unique variants. These are real-data set comparisons, not known-truth power
comparisons.

## Reproducibility and validation

The page reads the versioned cache and does not refit the million-pair model.

```{bash reproduction-command, eval=FALSE}
OMP_NUM_THREADS=1 \
OPENBLAS_NUM_THREADS=1 \
MKL_NUM_THREADS=1 \
VECLIB_MAXIMUM_THREADS=1 \
BLIS_NUM_THREADS=1 \
Rscript --vanilla \
  code/revision_simulations/internal/matched_fash_linear_real_data/\
run_matched_fash_linear_real_data.R

Rscript --vanilla \
  code/revision_simulations/internal/matched_fash_linear_real_data/\
build_iwp_overlap_proportion_cache.R

Rscript --vanilla -e \
  'workflowr::wflow_build("analysis/revision_internal_matched_fash_linear_real_data.rmd", local = TRUE)'
```

```{r software-table, results='asis'}
render_scrollable_table_matched(
  package_display,
  caption = "Table 5. Software provenance for the fresh FASH-linear rerun.",
  minimum_width = "1120px"
)
```

```{r validation-table, results='asis'}
render_scrollable_table_matched(
  validation,
  caption = "Table 6. Numerical, provenance, set-construction, and resource validations.",
  minimum_width = "760px"
)
```

```{r runtime-table, results='asis'}
render_scrollable_table_matched(
  runtime_summary,
  caption = "Table 7. Single-process runtime for fitting and cache construction.",
  digits = 2L,
  minimum_width = "620px"
)
```
