---
title: "Revision Simulation R1: genotype-level eQTLs with random B-spline effects"
author: "Ziang Zhang"
date: "2026-07-30"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9,
  fig.height = 6.5,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source("code/revision_simulations/shared/simulation_functions.R")
source("code/revision_simulations/r1_random_bspline/reporting.R")
```

## Overview

This genotype-level simulation uses random cubic B-spline dynamic effects as a
non-IWP data-generating mechanism. It is the broad, smoothly varying baseline;
the separate [spiky transient analysis](revision_combined_spiky_genotype_simulation.html)
examines sharply localized effects.

For donor $i$, variant $j$, and time $t$, expression is generated as

$$
Y_{ijt} = \alpha_{jt} + G_{ij}\beta_j(t) +
X_i^T\gamma_{jt} + \epsilon_{ijt},
\qquad \epsilon_{ijt} \sim N(0, 1).
$$

For each variant, the minor-allele frequency is drawn from
$p_j\sim\mathrm{Uniform}(0.1,0.5)$ and
$G_{ij}\sim\mathrm{Binomial}(2,p_j)$; monomorphic draws are repeated.
The five columns of $X$ are generated independently from $N(0,1)$ and
standardized across donors. Independently for each covariate, variant, and
time, $\gamma_{kjt}\sim N(0,0.5^2)$. The current simulation sets
$\alpha_{jt}=0$; each fitted regression still includes an intercept. Genotype,
covariate, covariate-effect, and noise draws are mutually independent.

```{r r1-simulate-genotypes-and-covariates, eval=FALSE}
seed <- 12345
set.seed(seed)

genotype_sim <- simulate_genotype_matrix(
  n_donors = 19,
  n_variants = 1000,
  maf_range = c(0.1, 0.5)
)
covariates <- simulate_covariate_matrix(
  n_donors = 19,
  n_covariates = 5
)
```

Each simulated dataset has `19` donors, `16` time points, `1000` variants, and
five PC-like covariates. The variant classes are fixed at:

- `200` dynamic eQTLs with random cubic B-spline effects;
- `400` non-dynamic eQTLs with constant effects drawn from $N(0,1)$;
- `400` zero-effect variants.

For a dynamic variant,

$$
\beta_j(t) = \mu_j + h_j(t),
\qquad
\mu_j \sim N(0,1),
\qquad
h_j(t) = \sum_{k=1}^{6} b_{jk} B_k(t),
\qquad b_{jk} \stackrel{\mathrm{iid}}{\sim} N(0,1).
$$

Only the time-varying deviation $h_j(t)$ is mean-centered over the observed
times and scaled to have maximum absolute value `2`. The independent
$\mu_j$ is the genetic main effect, so the final trajectory is not constrained
to be centered at zero. Thus, the true dynamic-null proportion is
$\pi_0 = 0.8$.

```{r r1-simulate-effects-and-expression, eval=FALSE}
class_probs <- c(
  dynamic_bspline = 0.20,
  constant = 0.40,
  zero = 0.40
)
effect_sim <- simulate_variant_effect_curves(
  n_variants = 1000,
  time_grid = 0:15,
  class_probs = class_probs,
  dynamic_amplitude = 2,
  bspline_df = 6,
  bspline_coefficient_sd = 1,
  constant_sd = 1,
  dynamic_main_effect_sd = 1,
  exact_class_counts = TRUE,
  seed = revision_component_seeds(seed)[["functional_truth"]],
  scenario = "genotype_random_bspline_main_effect_dynamic_eqtl"
)
expression_sim <- simulate_eqtl_expression_from_genotypes(
  G = genotype_sim$G,
  beta_matrix = effect_sim$beta_matrix,
  time_grid = 0:15,
  covariates = covariates,
  expression_noise_sd = 1,
  covariate_effect_sd = 0.5,
  intercept_sd = 0
)
```

At each time point, the genetic effect and its standard error are estimated
from `Y ~ G + PC1 + PC2 + PC3 + PC4 + PC5`. The real-data-like t-statistic
standard-error correction uses

$$
\mathrm{df} = 19 - 7 = 12,
$$

corresponding to the intercept, genotype, and five covariates. All methods use
the same per-time estimates and corrected standard errors.

```{r r1-estimate-and-fit-fash, eval=FALSE}
eqtl_summary <- estimate_eqtl_summaries_from_genotypes(
  G = genotype_sim$G,
  expression = expression_sim$expression,
  covariates = covariates,
  apply_t_se_correction = TRUE
)
datasets <- make_fash_datasets_from_eqtl_summary(
  beta_hat = eqtl_summary$beta_hat,
  se = eqtl_summary$se,
  true_beta = effect_sim$beta_matrix,
  time_grid = 0:15,
  unit_info = effect_sim$unit_info,
  scenario = "genotype_random_bspline_main_effect_dynamic_eqtl"
)

fash_iwp1_raw <- fashr::fash(
  Y = "y",
  smooth_var = "x",
  S = "sd",
  data_list = datasets,
  order = 1,
  grid = default_revision_grid(),
  num_basis = 20,
  penalty = 10,
  num_cores = 4,
  verbose = FALSE
)
fash_iwp1_bf <- fashr::BF_update(fash_iwp1_raw, plot = FALSE)

linear_sigma_grid <- exp(seq(log(0.05), log(5), length.out = 25))
fash_linear_raw <- fit_simplified_fash(
  datasets = datasets,
  estimate_sigma = TRUE,
  sigma_beta_grid = linear_sigma_grid,
  scale_time = TRUE
)
fash_linear_bf <- BF_update_simplified_fash(fash_linear_raw)
```

The reported curves and tables average five independently simulated datasets
with seeds `12345`, `22345`, `32345`, `42345`, and `52345`. The full
seed-12345 object is used only to display representative trajectories.

## Example simulated eQTLs

The panels show three zero-effect variants, three constant eQTLs, and three
random B-spline dynamic eQTLs. Points are per-time genetic effect estimates,
bars show two corrected standard errors, and red lines are the true effects.

```{r example-trajectories, fig.width=13, fig.height=11, fig.cap="Figure R1.1: Representative zero-effect, constant-effect, and random B-spline dynamic eQTLs from seed 12345. Points are estimated genetic effects, vertical bars span two corrected standard errors, and red curves are the true effects."}
plot_genotype_eqtl_examples(
  out,
  classes = c("zero", "constant", "dynamic_bspline"),
  n_per_class = 3,
  seed = 1,
  true_curve_n = 500
)
```

## Comparison 1: IWP versus linear FASH

This comparison changes only the alternative prior. `FASH-IWP1` uses the
first-order IWP alternative. `FASH-linear` uses a constant-versus-linear model
on time scaled to $[0,1]$: its alternative adds a slope
$b\sim N(0,\sigma_\beta^2)$ to the unrestricted intercept. Within each seed,
the fit profiles $\pi_0$ separately at each of 25 log-spaced candidate values
of $\sigma_\beta$ from `0.05` to `5`, then selects the pair with the largest
marginal likelihood. The BF adjustment updates the mixture weight while
holding the selected $\sigma_\beta$ fixed. Across the five seeds, the selected
slope SD has mean
`r format_decimal(linear_sigma_summary$mean_selected_sigma)` and observed range
[`r format_decimal(linear_sigma_summary$min_selected_sigma)`,
`r format_decimal(linear_sigma_summary$max_selected_sigma)`].

### Estimated dynamic-null proportion

The table reports the fitted prior weight on each method's constant component.
The true dynamic-null proportion is $\pi_0 = 0.8$.

```{r pi0-table}
knitr::kable(
  pi0_table,
  align = "l",
  caption = paste(
    "Table R1.1: Estimated dynamic-null proportions and selected linear slab",
    "scales across five random B-spline simulation replicates."
  )
)
```

### Results at alpha 0.05

```{r fash-alpha-table}
knitr::kable(
  fash_table,
  align = "l",
  caption = paste(
    "Table R1.2: Monte Carlo power and empirical FDR at alpha 0.05 for IWP",
    "and linear FASH under random B-spline effects."
  )
)
```

### Power as a function of alpha

```{r fash-power, fig.cap="Figure R1.2: Monte Carlo power across nominal alpha for IWP and linear FASH under random B-spline effects. Curves identify the raw and BF-updated methods shown in the in-figure legend; uncertainty summarizes five simulation seeds."}
plot_mc_alpha_curves(
  fash_alpha,
  metric = "power",
  title = "IWP versus linear FASH: Monte Carlo power",
  subtitle = "Five seeds; random cubic B-spline effects; N = 19, J = 1000",
  style_profile = "combined"
)
```

### Empirical FDR as a function of alpha

```{r fash-fdr, fig.cap="Figure R1.3: Monte Carlo empirical FDR across nominal alpha for IWP and linear FASH under random B-spline effects. Curves identify methods in the in-figure legend, and the diagonal reference marks equality between empirical FDR and nominal alpha."}
plot_mc_alpha_curves(
  fash_alpha,
  metric = "fdr",
  title = "IWP versus linear FASH: Monte Carlo empirical FDR",
  subtitle = "Five seeds; random cubic B-spline effects; N = 19, J = 1000",
  legend_position = "topleft",
  style_profile = "combined"
)
```

At $\alpha = 0.05$, IWP-Raw and IWP-BF have mean power
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-IWP1-Raw", "mean_power"))`
and
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-IWP1-BF", "mean_power"))`,
compared with
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-linear-Raw", "mean_power"))`
and
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-linear-BF", "mean_power"))`
for the corresponding linear fits. Their empirical FDR estimates are
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-IWP1-Raw", "mean_fdr"))`
and
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-IWP1-BF", "mean_fdr"))`.
The IWP alternative therefore provides substantially higher power for these
nonlinear broad curves. BF correction raises the estimated IWP null proportion
and produces the expected more conservative fit.

## Comparison 2: IWP FASH versus direct interaction tests

The direct baselines fit individual-level expression using linear or
linear-plus-quadratic genotype-by-time interactions and use permutation-based
eFDR. To make this comparison favorable to the direct methods, their eFDR
calculation is supplied with the true $\pi_0 = 0.8$ and uses `B = 100`
donor-level permutations per replicate.

```{r r1-direct-interaction-efdr, eval=FALSE}
out <- add_direct_interaction_efdr_results_to_genotype_output(
  out = out,
  n_permutations = 100,
  alpha = 0.05,
  seed = seed + 10000L,
  lambda = 0.5,
  pi0_method = "conservative",
  true_pi0 = 0.8,
  include_true_pi0 = TRUE,
  permute_covariates_with_expression = TRUE,
  num_cores = 4,
  overwrite = TRUE,
  verbose = FALSE
)
```

### Results at alpha 0.05

```{r direct-alpha-table}
knitr::kable(
  direct_table,
  align = "l",
  caption = paste(
    "Table R1.3: Monte Carlo power and empirical FDR at alpha 0.05 for IWP",
    "FASH and direct interaction tests under random B-spline effects."
  )
)
```

### Power as a function of alpha

```{r direct-power, fig.cap="Figure R1.4: Monte Carlo power across nominal alpha for IWP FASH and direct linear or quadratic interaction tests. The in-figure legend identifies methods; direct-test eFDR uses the true pi0 and 100 donor-level permutations."}
plot_mc_alpha_curves(
  direct_alpha,
  metric = "power",
  title = "IWP FASH versus direct interaction: Monte Carlo power",
  subtitle = "Five seeds; direct eFDR uses true pi0 = 0.8 and B = 100 permutations",
  style_profile = "combined"
)
```

### Empirical FDR as a function of alpha

```{r direct-fdr, fig.cap="Figure R1.5: Monte Carlo empirical FDR across nominal alpha for IWP FASH and direct interaction tests. The in-figure legend identifies methods, and the diagonal reference marks equality between empirical FDR and nominal alpha."}
plot_mc_alpha_curves(
  direct_alpha,
  metric = "fdr",
  title = "IWP FASH versus direct interaction: Monte Carlo empirical FDR",
  subtitle = "Five seeds; direct eFDR uses true pi0 = 0.8 and B = 100 permutations",
  legend_position = "bottomright",
  style_profile = "combined"
)
```

At $\alpha = 0.05$, IWP-BF has mean power
`r format_decimal(get_mc_metric(direct_alpha_005, "FASH-IWP1-BF", "mean_power"))`,
compared with
`r format_decimal(get_mc_metric(direct_alpha_005, "Direct-linear-LRT-eFDR-true-pi0", "mean_power"))`
for the direct linear test and
`r format_decimal(get_mc_metric(direct_alpha_005, "Direct-quadratic-LRT-eFDR-true-pi0", "mean_power"))`
for the direct quadratic test. IWP-BF has the highest power while maintaining
mean empirical FDR
`r format_decimal(get_mc_metric(direct_alpha_005, "FASH-IWP1-BF", "mean_fdr"))`.
The direct tests are more conservative but less powerful, even with the true
null proportion supplied.

## Interpretation

For broad random B-spline effects, IWP FASH is substantially more powerful than
linear FASH and both direct interaction baselines. The BF-updated IWP fit
retains high power and conservative empirical FDR across this five-seed pilot.
The separate spiky transient page evaluates whether this ordering persists for
sharply localized one-, two-, and three-peak effects.

## Reproducibility

The
[R1 simulation driver](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/r1_random_bspline/run_random_bspline_mc_replication.R)
uses the
[shared simulation functions](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/shared/simulation_functions.R).
The page-specific cache validation and formatting are in
[R1 reporting code](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/r1_random_bspline/reporting.R).
The following command regenerates the full Monte Carlo cache. It is shown for
reproducibility but is intentionally not evaluated during workflowr builds;
the table and figure chunks above read the saved cache.

The full Monte Carlo cache can be regenerated with:

```{bash regenerate-r1-cache, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/r1_random_bspline/run_random_bspline_mc_replication.R \
  --J 1000 \
  --n-donors 19 \
  --n-covariates 5 \
  --noise-sd 1 \
  --dynamic-main-effect-sd 1 \
  --efdr-permutations 100 \
  --seed-list 12345,22345,32345,42345,52345 \
  --output-id r1_random_bspline_main_effect_profile_sigma_pilot5 \
  --num-cores 4
```

Cached replicate metrics, summaries, and figures are stored under:

```text
output/revision_simulations/mc/r1_random_bspline_main_effect_profile_sigma_pilot5/
```
