---
title: "Revision Simulation R3: functional testing under broad and transient effects"
author: "Ziang Zhang"
date: "2026-07-30"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 10,
  fig.height = 8,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source("code/revision_simulations/shared/simulation_functions.R")
source("code/revision_simulations/r3_functional_testing/reporting.R")
```

## Overview

This simulation evaluates FASH functional testing after global dynamic-eQTL
discovery. It uses two non-IWP truth mechanisms matched on sample size,
variant classes, genetic main effects, functional labels, and analysis.

For donor $i$, variant $j$, and time $t$,

$$
Y_{ijt} = \alpha_{jt} + G_{ij}\beta_j(t) +
X_i^T\gamma_{jt} + \epsilon_{ijt},
\qquad \epsilon_{ijt}\sim N(0,1).
$$

For each variant, the minor-allele frequency is drawn from
$p_j\sim\mathrm{Uniform}(0.1,0.5)$ and
$G_{ij}\sim\mathrm{Binomial}(2,p_j)$; monomorphic draws are repeated.
The five columns of $X$ are generated independently from $N(0,1)$ and
standardized across donors. Independently for each covariate, variant, and
time, $\gamma_{kjt}\sim N(0,0.5^2)$. The current simulation sets
$\alpha_{jt}=0$; each fitted regression still includes an intercept. Genotype,
covariate, covariate-effect, and noise draws are mutually independent.

```{r r3-simulate-genotypes-and-covariates, eval=FALSE}
seed <- 12345
component_seeds <- revision_component_seeds(seed)

genotype_sim <- simulate_genotype_matrix(
  n_donors = 19,
  n_variants = 1000,
  maf_range = c(0.1, 0.5),
  seed = component_seeds[["genotype"]]
)
covariates <- simulate_covariate_matrix(
  n_donors = 19,
  n_covariates = 5,
  seed = component_seeds[["covariates"]]
)
```

Each replicate contains `19` donors, `16` time points, five PC-like
covariates, `200` dynamic eQTLs, `400` constant eQTLs, and `400` zero-effect
variants. For every dynamic candidate,

$$
\beta_j(t)=\mu_j+h_j(t),
\qquad \mu_j\sim N(0,1).
$$

The dynamic eQTLs are balanced as evenly as possible across six preassigned
truth cells: early, middle, or late crossed with switch or non-switch. Each
candidate draws a new $\mu_j$ from $N(0,1)$; rejection sampling retains it only
when the final trajectory $\mu_j+h_j(t)$ satisfies its assigned cell.

### R3A: broad random B-spline effects

The deviation $h_j(t)$ is a cubic B-spline with six iid $N(0,1)$ coefficients.
It is centered over the 16 observed times and scaled to maximum absolute value
`2` before adding $\mu_j$.

### R3B: compact raised-cosine effects

The deviation $h_j(t)$ contains one, two, or three compact raised-cosine peaks
with half-width `1.5`. Peak locations vary within the early, middle, and late
windows, secondary amplitudes are random, and the deviation is scaled to
centered RMS `0.9` before adding $\mu_j$.

### Truth definitions

Early, middle, and late use the same max-contrast functionals as the
manuscript. The target contrast must be at least `0.1`, and the two non-target
contrasts must be at most `-0.1`.

For the switch threshold $c=0.25$,

$$
T_{\mathrm{switch}}\{\beta\}
=
\min\left[
\max_t\{\beta(t):\beta(t)>0\},
\max_t\{|\beta(t)|:\beta(t)<0\}
\right]-c.
$$

A switch truth must have $T_{\mathrm{switch}}\geq 0.1$; equivalently, it has
both a positive and a negative effect of magnitude at least `0.35`. A
non-switch truth remains the same sign over the full dense grid. Writing
$R_j=\max_t\beta_j(t)-\min_t\beta_j(t)$, its clearance from zero must satisfy

$$
\min_t|\beta_j(t)|\geq\max(0.1,\,0.1R_j).
$$

Thus, a larger-amplitude non-switch curve must remain proportionally farther
from zero. These margins make the simulation labels unambiguous.

```{r r3-simulate-functional-truth, eval=FALSE}
truth_mechanism <- "random_bspline" # Use "raised_cosine" for R3B.
scenario <- paste0(
  if (truth_mechanism == "random_bspline") "r3a_" else "r3b_",
  "matched_functional_",
  truth_mechanism,
  "_relative_clearance_main_effect"
)
time_grid <- 0:15
evaluation_grid <- seq(0, 15, by = 0.1)
class_probs <- c(
  dynamic_bspline = 0.20,
  constant = 0.40,
  zero = 0.40
)

effect_sim <- simulate_matched_functional_effect_set(
  n_variants = 1000,
  truth_mechanism = truth_mechanism,
  time_grid = time_grid,
  evaluation_grid = evaluation_grid,
  class_probs = class_probs,
  dynamic_main_effect_sd = 1,
  constant_sd = 1,
  bspline_amplitude = 2,
  bspline_df = 6,
  bspline_coefficient_sd = 1,
  cosine_width_half = 1.5,
  cosine_spike_counts = 1:3,
  cosine_relative_amplitude_range = c(0.35, 0.75),
  cosine_target_centered_rms = 0.9,
  switch_threshold = 0.25,
  location_truth_margin = 0.10,
  switch_truth_margin = 0.10,
  non_switch_min_abs = 0.10,
  non_switch_min_range_fraction = 0.10,
  exact_class_counts = TRUE,
  seed = seed,
  class_seed = component_seeds[["classes"]],
  constant_seed = component_seeds[["constant_effects"]],
  shape_seed = component_seeds[["functional_truth"]],
  scenario = scenario
)
expression_sim <- simulate_eqtl_expression_from_genotypes(
  G = genotype_sim$G,
  beta_matrix = effect_sim$beta_matrix,
  time_grid = time_grid,
  covariates = covariates,
  expression_noise_sd = 1,
  covariate_effect_sd = 0.5,
  intercept_sd = 0,
  seed = component_seeds[["expression"]]
)
```

At each observed time, the genetic effect and standard error are estimated
from `Y ~ G + PC1 + PC2 + PC3 + PC4 + PC5`. The t-statistic correction uses
`df = 19 - 7 = 12`; FASH is fit to these estimates and corrected standard
errors.

## Example truth curves

Two examples are shown for each of the six functional truth cells. Points and
bars are the per-time estimates and two corrected standard errors used by
FASH; orange lines are the continuous truths. Shading marks the target timing
window.

### R3A examples

```{r r3a-examples, fig.width=11, fig.height=23, fig.cap="Figure R3.1: Representative broad random B-spline truth curves for each functional target and timing cell. Points and bars show estimated effects and two corrected standard errors, orange lines are continuous truths, and shading marks the target timing window."}
plot_truth_examples(
  example_curves$random_bspline,
  mechanism_labels[["random_bspline"]]
)
```

### R3B examples

```{r r3b-examples, fig.width=11, fig.height=23, fig.cap="Figure R3.2: Representative compact raised-cosine truth curves for each functional target and timing cell. Points and bars show estimated effects and two corrected standard errors, orange lines are continuous truths, and shading marks the target timing window."}
plot_truth_examples(
  example_curves$raised_cosine,
  mechanism_labels[["raised_cosine"]]
)
```

## Inference procedure

At each nominal level $\alpha$, the same procedure is applied to
`FASH-IWP1-Raw` and `FASH-IWP1-BF`:

1. Apply global dynamic-eQTL FDR control at level $\alpha$.
2. For each dynamically selected variant, evaluate the early, middle, late,
   and switch functionals using posterior samples.
3. Rank functional lfsr values and call variants whose cumulative FSR is at
   most $\alpha$.

Power is the fraction of true positive functionals that are called. Empirical
FSR is the fraction of functional calls whose known true functional is
non-positive. The figures report five-seed Monte Carlo means and 95% t-based
intervals across the full alpha grid.

```{r r3-fit-and-test-functionals, eval=FALSE}
out <- run_genotype_level_dynamic_eqtl_simulation(
  G = genotype_sim$G,
  time_grid = time_grid,
  covariates = covariates,
  class_probs = class_probs,
  maf_range = c(0.1, 0.5),
  expression_noise_sd = 1,
  covariate_effect_sd = 0.5,
  intercept_sd = 0,
  dynamic_main_effect_sd = 1,
  alpha = 0.05,
  seed = seed,
  num_cores = 4,
  num_basis = 20,
  scenario = scenario,
  save_outputs = FALSE,
  effect_sim = effect_sim,
  expression_sim = expression_sim
)

true_functionals <- effect_sim$true_functionals[
  out$unit_info$variant_id,
  c("early", "middle", "late", "switch"),
  drop = FALSE
]
true_dynamic <- out$unit_info$effect_class == "dynamic_bspline"

raw_results <- evaluate_fash_functional_testing(
  fit = out$fash_fits$fash_iwp1_raw,
  true_functionals = true_functionals,
  evaluation_grid = evaluation_grid,
  alpha_grid = seq(0.005, 0.20, by = 0.005),
  method = "FASH-IWP1-Raw",
  switch_threshold = 0.25,
  true_dynamic = true_dynamic,
  num_cores = 4,
  seed = component_seeds[["functional_posterior"]]
)
bf_results <- evaluate_fash_functional_testing(
  fit = out$fash_fits$fash_iwp1_bf,
  true_functionals = true_functionals,
  evaluation_grid = evaluation_grid,
  alpha_grid = seq(0.005, 0.20, by = 0.005),
  method = "FASH-IWP1-BF",
  switch_threshold = 0.25,
  true_dynamic = true_dynamic,
  num_cores = 4,
  seed = component_seeds[["functional_posterior"]] + 100L
)
```

## R3A: broad random B-spline results

### Results at alpha 0.05

```{r r3a-table}
knitr::kable(
  r3a_table,
  align = "l",
  caption = paste(
    "Table R3.1: Functional-testing power and empirical false sign rate at",
    "alpha 0.05 under broad random B-spline truth."
  )
)
```

### Power as a function of alpha

```{r r3a-power, fig.height=9, fig.cap="Figure R3.3: Monte Carlo functional-testing power under broad random B-spline truth. Panels identify early, middle, late, and switch targets; curves identify raw and BF-updated FASH in the in-figure legend, with 95% intervals across five seeds."}
plot_mc_functional_curve_grid(
  mc_curve = r3a_alpha,
  metric = "power",
  title = "Functional-testing power: broad random B-spline truth"
)
```

### Empirical FSR as a function of alpha

```{r r3a-fsr, fig.height=9, fig.cap="Figure R3.4: Monte Carlo empirical false sign rate under broad random B-spline truth. Panels identify functional targets, curves identify raw and BF-updated FASH, and the diagonal reference denotes nominal alpha."}
plot_mc_functional_curve_grid(
  mc_curve = r3a_alpha,
  metric = "empirical_fsr",
  title = "Empirical FSR: broad random B-spline truth"
)
```

## R3B: compact raised-cosine results

### Results at alpha 0.05

```{r r3b-table}
knitr::kable(
  r3b_table,
  align = "l",
  caption = paste(
    "Table R3.2: Functional-testing power and empirical false sign rate at",
    "alpha 0.05 under compact raised-cosine truth."
  )
)
```

### Power as a function of alpha

```{r r3b-power, fig.height=9, fig.cap="Figure R3.5: Monte Carlo functional-testing power under compact raised-cosine truth. Panels identify early, middle, late, and switch targets; curves identify raw and BF-updated FASH in the in-figure legend, with 95% intervals across five seeds."}
plot_mc_functional_curve_grid(
  mc_curve = r3b_alpha,
  metric = "power",
  title = "Functional-testing power: compact raised-cosine truth"
)
```

### Empirical FSR as a function of alpha

```{r r3b-fsr, fig.height=9, fig.cap="Figure R3.6: Monte Carlo empirical false sign rate under compact raised-cosine truth. Panels identify functional targets, curves identify raw and BF-updated FASH, and the diagonal reference denotes nominal alpha."}
plot_mc_functional_curve_grid(
  mc_curve = r3b_alpha,
  metric = "empirical_fsr",
  title = "Empirical FSR: compact raised-cosine truth"
)
```

## Dynamic-null proportion

The first-stage dynamic-null proportion is $\pi_0=0.8$. The table reports the
fitted constant-component weights for the raw and BF-adjusted IWP fits.

```{r pi0-table}
knitr::kable(
  pi0_table,
  align = "l",
  caption = paste(
    "Table R3.3: Estimated first-stage dynamic-null proportions for raw and",
    "BF-updated IWP FASH under both functional truth mechanisms."
  )
)
```

## Interpretation

For R3A, BF-updated FASH has mean empirical FSR
`r format_decimal(max(r3a_alpha_005$mean_empirical_fsr[
  r3a_alpha_005$method == "FASH-IWP1-BF"
]), 3)` or lower for all four targets at $\alpha=0.05$. Its switch power is
`r format_decimal(metric_at(
  r3a_alpha_005, "switch", "FASH-IWP1-BF", "mean_power"
), 3)`.

For the more localized R3B trajectories, BF-updated FASH makes few timing
calls and no false timing calls in these five replicates. Its switch power is
`r format_decimal(metric_at(
  r3b_alpha_005, "switch", "FASH-IWP1-BF", "mean_power"
), 3)`, but its mean switch FSR is
`r format_decimal(metric_at(
  r3b_alpha_005, "switch", "FASH-IWP1-BF", "mean_empirical_fsr"
), 3)` at nominal $\alpha=0.05$. In this five-seed pilot, the scale-relative
clearance condition separates switch and non-switch truths in both mechanisms
without removing the compact transient behavior. These results support
functional testing for the stated margin-separated designs; they do not imply
that BF adjustment alone guarantees FSR control for every functional
hypothesis or every truth geometry.

## Reproducibility

The
[R3 simulation driver](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/r3_functional_testing/run_matched_functional_testing_mc_replication.R)
uses the
[shared simulation functions](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/shared/simulation_functions.R).
The page-specific cache validation and formatting are in
[R3 reporting code](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/r3_functional_testing/reporting.R).
The following command regenerates the paired five-seed cache. It is shown for
reproducibility but is intentionally not evaluated during workflowr builds;
the table and figure chunks above read the saved cache.

The paired five-seed cache can be regenerated with:

```{bash regenerate-r3-cache, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/r3_functional_testing/run_matched_functional_testing_mc_replication.R \
  --J 1000 \
  --n-donors 19 \
  --n-covariates 5 \
  --noise-sd 1 \
  --dynamic-main-effect-sd 1 \
  --truth-mechanisms random_bspline,raised_cosine \
  --seed-list 12345,22345,32345,42345,52345 \
  --output-id r3_matched_functional_relative_clearance_main_effect_pilot5 \
  --num-cores 4
```

Cached replicate metrics, call-level diagnostics, examples, summaries, and
figures are stored under:

```text
output/revision_simulations/mc/r3_matched_functional_relative_clearance_main_effect_pilot5/
```
