---
title: "Revision R4: unit-specific null correlations"
author: "Ziang Zhang"
date: "2026-08-24"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 10,
  fig.height = 5.5,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source("code/revision_simulations/internal/covariance_estimation/donor_null_permutation_helpers.R")
source("code/revision_simulations/r4_correlated_errors/unit_specific_residual_permutation_helpers.R")
source("code/revision_simulations/r4_correlated_errors/reporting.R")
```

## Question and estimand

For a single gene--variant unit $j$, we estimate the null correlation of its
16 t-adjusted z-scores,

$$
\operatorname{Corr}\{\widehat z_j(t),\widehat z_j(s)
\mid G_j, X, \text{donor coverage};\ \beta_j=0\}.
$$

This is intentionally a **per-unit** diagnostic. It does not pool units to
produce a common real-data covariance matrix, and it is not an estimate of a
population-average correlation. The six units below were selected uniformly at
random from the BF-adjusted FASH fit with a fixed seed. Whether a selected unit
is a genuine biological null is not the identifying assumption: the
randomization removes its fitted genotype effect before generating the null
draws.

## Signal-stripped residual randomization

At every time point, we fit the original model
`expression ~ 1 + PC1 + ... + PC5 + genotype`. For each permutation, we retain
the fitted intercept-and-PC component, remove the fitted genotype component,
and add a donor-permuted full-model residual. One donor map is used for all 16
time points of a draw, so each donor's residual trajectory remains intact.
Maps are restricted to donors with the same observed-time pattern, preserving
the original missingness structure. The null response is then refit against the
original genotype and covariates.

For each refit, its raw regression SE is converted by the same t-to-normal
adjustment used for the FASH input, and
$\widehat z_j(t)=\widehat\beta_j(t)/\widehat{\mathrm{SE}}_{\mathrm{t-adjusted},j}(t)$.
Each unit has 400 independently sampled donor maps; the empirical correlation
matrix is the correlation across its 400 resulting null z-score vectors. The
procedure conditions on observed $G_j$ and $X$, but not on an unobserved true
genetic effect.

```{r selected-units, results='asis'}
knitr::kable(
  r4_selected_units[, c(
    "unit", "pair_key", "source_bf_lfdr", "maximum_observed_beta_difference"
  )],
  caption = paste(
    "Table 1. Six randomly selected units. The final column verifies that",
    "the raw-data refit reproduces the saved FASH beta estimates."
  ),
  align = "llll"
)
```

## Example unit-specific null correlation matrices

The following two panels show the first two units in the fixed random sample.
They are examples rather than representative or averaged correlation patterns.

```{r unit-null-correlation-heatmaps, fig.width=11, fig.height=5.3, fig.cap="Figure 1. Estimated 16 by 16 t-adjusted null z-score correlation matrices from 400 signal-stripped residual-trajectory permutations for two randomly selected units. Each unit has its own donor maps and correlation estimate."}
plot_r4_unit_null_heatmaps(r4_analysis$unit_results[1:2])
```

## Estimated variograms

For each unit, the standardized semivariogram at lag $h$ is

$$
\widehat\gamma_j(h) = 1 -
\frac{1}{16-h}\sum_{t=1}^{16-h}\widehat R_j(t,t+h).
$$

Under a lag-independent correlation of zero it is one. The curves display
within-unit variation in the null dependence rather than a fitted common
time-series model. The shaded intervals are 95% joint-draw bootstrap percentile
intervals: each bootstrap replication resamples whole 16-time z-score
draws, so it retains their cross-time structure. They quantify Monte Carlo
uncertainty from the 400 permutations, not between-unit uncertainty.

```{r unit-null-variograms, fig.width=10, fig.height=5.5, fig.cap="Figure 2. Unit-specific null variograms estimated from 400 residual-permutation draws per randomly selected unit. Shaded bands are 95% joint-draw bootstrap percentile intervals; the dashed line is the value under zero correlation."}
plot_r4_unit_null_variograms(r4_analysis$unit_results)
```

```{r variogram-summary, results='asis'}
knitr::kable(
  r4_variogram_summary,
  caption = "Table 2. Short- and long-lag summaries of the unit-specific null correlations.",
  align = "llrr"
)
```

## Full-model residual variograms

For a complementary residual-based calculation, we refit the original
time-specific FASH analysis model for every gene and time,
`Y ~ 1 + PC1 + ... + PC5 + G`, and retain its full-model cell-line residual
$\widehat e(t)$. Thus both PC and fitted genotype components are removed.
For each unit, we calculate `cor(e_hat, use = "pairwise.complete.obs")` across
the 16 time columns and derive the same variogram. Because the observed
missingness patterns differ by time, each correlation entry uses 13--19 shared
donors. These are observed full-model residual correlations, not null z-score
correlations and not a residual-permutation estimate.

```{r full-model-residual-variograms, fig.width=10, fig.height=5.5, fig.cap="Figure 3. Variograms derived from correlations of the full-model cell-line residuals for the same six units. At each time, residuals are from Y ~ 1 + PC1 + ... + PC5 + G; each correlation uses pairwise-complete donors. The dashed line is the value under zero correlation."}
plot_r4_full_model_residual_variograms(r4_residual_analysis$unit_results)
```

## Interpretation and scope

The heatmaps and variograms answer a narrow question: conditional on the
observed design for a particular unit, what temporal dependence appears in
effect estimates when the fitted genotype signal is removed and residual donor
trajectories are randomized? They are useful for checking that a single common
correlation matrix would hide meaningful unit-to-unit heterogeneity.

They do not establish a population covariance model, calibrate FASH discovery
FDR, or imply that every selected observed unit is biologically null. A formal
simulation or calibration analysis would require a separate target estimand and
repeated datasets.

## Reproducibility

The cache was generated from the retained BF-adjusted FASH fit and the original
YRI dosage, expression, and time-specific PC inputs. It records input hashes,
the six selected pair keys, all 2,400 donor maps, and the 2,400 t-adjusted
null z-score draws. Recreate it from the workflowr root with:

```{bash, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/r4_correlated_errors/run_unit_specific_residual_permutation.R \
  --n-units 6 \
  --n-permutations 400 \
  --selection-seed 20260824 \
  --permutation-seed 20260825 \
  --output-id r4_unit_specific_residual_permutation_z

Rscript --vanilla \
  code/revision_simulations/r4_correlated_errors/compute_full_model_residual_correlation.R
```
