---
title: "Internal note: correlated-observation FASH model"
author: "Ziang Zhang"
date: "2026-08-13"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

<style>
.model-banner {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border: 1px solid #d9a441;
  border-left: 6px solid #d9a441;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.model-answer {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.model-caution {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #d55e00;
  border-radius: 4px;
  background: #fff4ec;
  color: #59331e;
}
pre,
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: hidden;
}
</style>

<div class="model-banner">
<strong>Internal mathematical note.</strong> This page is intentionally
unlinked from the workflowr homepage and navigation. It documents a software
capability; it does not change the manuscript model, the primary real-data
analysis, or the response letter.
</div>

## Short answer

<div class="model-answer">
<strong>Yes.</strong> The conditionally independent Gaussian likelihood in
the FASH manuscript is a diagonal special case of the Gaussian likelihood
already implemented in <code>fashr</code>. For each unit \(j\), the software
can use an arbitrary user-supplied symmetric positive-definite precision
matrix \(\mathbf{\Omega}_j\), allowing the coordinates of
\(\hat{\boldsymbol\beta}_j\) to be correlated conditional on the entire effect
function \(\beta_j\).
</div>

The key distinction is **within-unit** versus **between-unit** dependence.
The `Omega` interface models dependence among the \(R_j\) effect estimates
within unit \(j\). The empirical-Bayes objective still combines one marginal
likelihood per unit and does not accept a joint covariance matrix coupling
different units.

## The manuscript model as a special case

Following the manuscript, let

$$
\hat{\boldsymbol\beta}_j
\mathrel{\mathop:}=
\left(
  \hat\beta_j(t_{j1}),\ldots,\hat\beta_j(t_{jR_j})
\right)^T,
\qquad
\boldsymbol\beta_j
\mathrel{\mathop:}=
\left(
  \beta_j(t_{j1}),\ldots,\beta_j(t_{jR_j})
\right)^T.
$$

The manuscript assumes

$$
\hat\beta_j(t_{jr})\mid\beta_j,s_{jr}
\overset{\mathrm{ind}}{\sim}
N\!\left(\beta_j(t_{jr}),s_{jr}^2\right),
\qquad r\in[R_j].
$$

Equivalently, with

$$
\mathbf D_j
\mathrel{\mathop:}=
\operatorname{diag}(s_{j1},\ldots,s_{jR_j}),
\qquad
\mathbf S_j\mathrel{\mathop:}=\mathbf D_j^2,
$$

the vector likelihood is

$$
\hat{\boldsymbol\beta}_j\mid\beta_j,\mathbf S_j
\sim
N_{R_j}\!\left(\boldsymbol\beta_j,\mathbf S_j\right).
$$

Thus conditional independence is exactly the restriction that the
conditional covariance matrix is diagonal.

## General within-unit correlated observation model

Let \(\mathbf\Omega_j\in\mathbb R^{R_j\times R_j}\) be a symmetric
positive-definite precision matrix, and let
\(\mathbf\Sigma_j\mathrel{\mathop:}=\mathbf\Omega_j^{-1}\). The general
Gaussian
observation model supported by `fashr` is

$$
\boxed{
\hat{\boldsymbol\beta}_j\mid\beta_j,\mathbf\Omega_j
\sim
N_{R_j}\!\left(
  \boldsymbol\beta_j+\boldsymbol o_j,
  \mathbf\Omega_j^{-1}
\right)
}
\qquad j\in[J],
$$

The bold matrix \(\mathbf\Omega_j\) is distinct from the manuscript's
unbold \(\Omega\), which denotes the domain of the condition variable \(t\).
Here \(\boldsymbol o_j\) is an optional known offset. For the summary-effect
setting in the manuscript, \(\boldsymbol o_j=\boldsymbol 0\). The conditional
log-density is

$$
\log p(\hat{\boldsymbol\beta}_j\mid\beta_j,\mathbf\Omega_j)
=
-\frac{1}{2}\left\{
  (\hat{\boldsymbol\beta}_j-\boldsymbol o_j-\boldsymbol\beta_j)^T
  \mathbf\Omega_j
  (\hat{\boldsymbol\beta}_j-\boldsymbol o_j-\boldsymbol\beta_j)
  -\log|\mathbf\Omega_j|
  +R_j\log(2\pi)
\right\}.
$$

Consequently,

$$
\operatorname{Cov}\!\left(
  \hat{\boldsymbol\beta}_j\mid\beta_j,\mathbf\Omega_j
\right)
=\mathbf\Sigma_j,
$$

whose off-diagonal entries encode conditional covariance among settings of
\(t\). The manuscript likelihood is recovered by taking

$$
\mathbf\Omega_j=\mathbf S_j^{-1}
=\operatorname{diag}\!\left(
  s_{j1}^{-2},\ldots,s_{jR_j}^{-2}
\right).
$$

### Preserving the reported marginal standard errors

Suppose \(\mathbf P_j\) is the desired conditional correlation matrix and the
reported marginal standard errors remain
\(\boldsymbol s_j=(s_{j1},\ldots,s_{jR_j})^T\). Then the appropriate covariance
and precision matrices are

$$
\mathbf\Sigma_j=\mathbf D_j\mathbf P_j\mathbf D_j,
\qquad
\mathbf\Omega_j
=\mathbf D_j^{-1}\mathbf P_j^{-1}\mathbf D_j^{-1}.
$$

Passing only \(\mathbf P_j^{-1}\) would implicitly replace all marginal
standard errors by one and would therefore specify a different likelihood.

## Full FASH hierarchy under correlated observations

The observation model changes, but the FASH prior remains the same. In the
manuscript's notation,

$$
\begin{aligned}
\beta_j
&\overset{\mathrm{iid}}{\sim} g_\beta,
\\
g_\beta
&=\sum_{k=0}^K\pi_k\,
L\text{-GP}(\beta;\sigma_k),
\qquad
\pi_k\geq0,
\quad
\sum_{k=0}^K\pi_k=1,
\\
\hat{\boldsymbol\beta}_j\mid\beta_j,\mathbf\Omega_j
&\sim
N_{R_j}\!\left(
  \boldsymbol\beta_j+\boldsymbol o_j,
  \mathbf\Omega_j^{-1}
\right).
\end{aligned}
$$

Here \(\sigma_0=0\) represents no deviation from the baseline space
\(S_0=\operatorname{Null}\{L\}\), exactly as in the manuscript. Conditional on
the functions and the supplied precision matrices, `fashr` forms the
unit-level factorization

$$
p(\hat{\boldsymbol\beta}\mid
  \boldsymbol\beta,\{\mathbf\Omega_j\}_{j=1}^J)
=
\prod_{j=1}^J
p(\hat{\boldsymbol\beta}_j\mid\beta_j,\mathbf\Omega_j).
$$

If units are dependent in the data-generating process, the resulting sum of
unit-level log likelihoods is a composite likelihood; `Omega` does not model
cross-unit dependence.

## Marginal likelihood and empirical-Bayes fit

Let the \(k\)th \(L\)-GP component induce

$$
\boldsymbol\beta_j\mid k
\sim N_{R_j}(\boldsymbol0,\mathbf C_{jk}),
$$

where \(\mathbf C_{jk}\) is the component covariance evaluated at
\((t_{j1},\ldots,t_{jR_j})\). If all units share the same observation grid,
this reduces to the manuscript's notation \(\mathbf C_k\). Define the
offset-adjusted data

$$
\boldsymbol y_j
\mathrel{\mathop:}=
\hat{\boldsymbol\beta}_j-\boldsymbol o_j.
$$

Gaussian conjugacy gives the component marginal likelihood

$$
\begin{aligned}
\ell_{jk}
&=
\int
p(\boldsymbol y_j\mid\boldsymbol\beta_j,\mathbf\Sigma_j)
p_k(\boldsymbol\beta_j)\,d\boldsymbol\beta_j
\\
&=
d\mathcal N\!\left(
  \boldsymbol y_j;
  \boldsymbol0,
  \mathbf C_{jk}+\mathbf\Sigma_j
\right).
\end{aligned}
$$

Therefore the empirical-Bayes objective has exactly the same form as in the
paper,

$$
\hat{\boldsymbol\pi}
=\operatorname*{argmax}_{\boldsymbol\pi}
\sum_{j=1}^J
\log\left\{
  \sum_{k=0}^K\pi_k\ell_{jk}
\right\},
$$

but each \(\ell_{jk}\) now uses the full
\(\mathbf\Sigma_j=\mathbf\Omega_j^{-1}\), not the diagonal
\(\mathbf S_j\).

## Posterior under the general model

The posterior remains a mixture of \(K+1\) GPs,

$$
p(\beta_j\mid\hat{\boldsymbol\beta}_j,
  \mathbf\Omega_j,\hat{\boldsymbol\pi})
=
\sum_{k=0}^K
\tilde\pi_{jk}\,
p_k(\beta_j\mid\hat{\boldsymbol\beta}_j,\mathbf\Omega_j),
$$

with unchanged posterior mixture weights

$$
\tilde\pi_{jk}
=
\frac{\hat\pi_k\ell_{jk}}
{\sum_{k'=0}^K\hat\pi_{k'}\ell_{jk'}}.
$$

For a proper Gaussian component, the posterior of the function values at the
observed settings is

$$
\boldsymbol\beta_j\mid\boldsymbol y_j,k
\sim
N_{R_j}(\boldsymbol m_{jk},\mathbf V_{jk}),
$$

where

$$
\begin{aligned}
\boldsymbol m_{jk}
&=
\mathbf C_{jk}
(\mathbf C_{jk}+\mathbf\Sigma_j)^{-1}
\boldsymbol y_j,
\\
\mathbf V_{jk}
&=
\mathbf C_{jk}
-\mathbf C_{jk}
(\mathbf C_{jk}+\mathbf\Sigma_j)^{-1}
\mathbf C_{jk}.
\end{aligned}
$$

For an arbitrary \(t\in\Omega\), let

$$
\boldsymbol c_{jk}(t)
\mathrel{\mathop:}=
\operatorname{Cov}_k\!\left(
  \boldsymbol\beta_j,\beta_j(t)
\right).
$$

Then

$$
\begin{aligned}
\mathbb E_k\!\left[\beta_j(t)\mid\boldsymbol y_j\right]
&=
\boldsymbol c_{jk}(t)^T
(\mathbf C_{jk}+\mathbf\Sigma_j)^{-1}
\boldsymbol y_j,
\\
\operatorname{Cov}_k\!\left(
  \beta_j(s),\beta_j(t)\mid\boldsymbol y_j
\right)
&=
C_k(s,t)
-\boldsymbol c_{jk}(s)^T
(\mathbf C_{jk}+\mathbf\Sigma_j)^{-1}
\boldsymbol c_{jk}(t).
\end{aligned}
$$

Thus smoothing, lfdr, lfsr, and other posterior functionals retain their FASH
definitions; only the unit-level Gaussian likelihood is generalized.

## Mapping the model to `fashr`

In `fashr` 0.1.42:

1. `fash(..., likelihood = "gaussian", Omega = ...)` selects the dependent
   Gaussian likelihood when `S` is `NULL`.
2. `Omega` may be one precision matrix shared by all units or a list containing
   one precision matrix per unit.
3. The likelihood and posterior-sampling paths both use the supplied precision
   matrix.
4. If both `S` and `Omega` are supplied, the package warns and uses the
   diagonal `S` path. They are alternative parameterizations, not additive
   sources of uncertainty.
5. The package checks dimensions but does not currently guarantee symmetry or
   positive definiteness. These conditions must be validated before fitting.
6. `fashr` consumes \(\mathbf\Omega_j\); it does not estimate the observation
   correlation matrix from the effect estimates.

```{r omega-interface, eval=FALSE}
correlation <- estimated_error_correlation

omega <- lapply(datasets, function(dataset) {
  standard_errors <- dataset$SE
  scale_matrix <- diag(standard_errors)
  covariance <- scale_matrix %*% correlation %*% scale_matrix
  precision <- solve(covariance)

  # Validate the user-supplied precision before calling fashr.
  chol(precision)
  precision
})

correlated_fit <- fashr::fash(
  Y = "beta",
  smooth_var = "time",
  data_list = datasets,
  S = NULL,
  Omega = omega,
  likelihood = "gaussian",
  num_basis = 20,
  order = 1,
  betaprec = 0,
  pred_step = 1,
  penalty = 10,
  grid = fine_grid
)
```

<div class="model-caution">
<strong>Software capability is not the same as the fitted model in the
manuscript.</strong> The retained dynamic-eQTL analysis in
<code>code/01_fash.R</code> calls <code>fash()</code> with
<code>S = "SE"</code>, after the marginal \(t\)-to-normal standard-error
adjustment. It therefore uses the diagonal conditional-independence model. A
full-<code>Omega</code> fit is available as a sensitivity or alternative model,
but it requires a defensible externally supplied conditional error covariance.
</div>

## Source provenance and existing validation

This note was checked against the following local sources:

- manuscript source `revision-local/fash.tex`, especially the problem setup,
  FASH likelihood, adaptive \(L\)-GP prior, posterior mixture, and likelihood
  matrix;
- `fashr` source checkout at commit
  `3c1a9dee015f24a176571acbc133fabe48ae1009` and package version 0.1.42;
- `R/00_setup.R`, `R/01_likelihood.R`, `R/03_sampling.R`, `R/04_fash.R`,
  `src/Gaussian_dep.cpp`, and `src/Gaussian_dep_fixed.cpp` in that checkout;
- the retained exact audit in
  `code/revision_simulations/internal/omega_precision_sanity_check/`.

The retained audit compared the dependent-Gaussian path with independent
closed-form multivariate-normal calculations. All 12 exact checks passed, and
the maximum absolute likelihood-matrix error was
\(3.55\times10^{-15}\). That audit establishes the implementation contract for
valid precision matrices; it does not establish how \(\mathbf\Omega_j\) should
be estimated in a particular scientific application.

For the existing real-data sensitivity that constructs
\(\mathbf\Omega_j=(\mathbf D_j\mathbf P\mathbf D_j)^{-1}\), see
[Internal sensitivity analysis: correlated FASH likelihoods](revision_internal_correlated_likelihood_sensitivity.html).
