---
title: "Revision Simulation R2: real-genotype eQTLs with spiky transient effects"
author: "Ziang Zhang"
date: "2026-08-22"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9,
  fig.height = 6.5,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source("code/revision_simulations/shared/simulation_functions.R")
source("code/revision_simulations/shared/real_genotype_one_per_gene.R")
source("code/revision_simulations/r2_spiky_transient/reporting.R")
```

<style>
@media (max-width: 767px) {
  html,
  body {
    max-width: 100%;
    overflow-x: hidden;
  }
  .main-container table {
    display: block;
    width: 100%;
    max-width: 100%;
    overflow-x: auto;
    -webkit-overflow-scrolling: touch;
  }
  .main-container .MathJax_Display {
    max-width: 100%;
    overflow-x: auto;
    overflow-y: hidden;
    -webkit-overflow-scrolling: touch;
  }
}
</style>

## Overview

This real-genotype simulation evaluates dynamic-eQTL detection when the true
effects are continuous but sharply localized and are not generated from the
IWP prior used by FASH. It uses the same individual-level pipeline as the
[broad random B-spline simulation](revision_genotype_level_simulation.html).

### Result provenance for the 0.1.43 refresh

The FASH-IWP1 and FASH-linear rows were recomputed on the same five fixed
simulation seeds with `fashr` 0.1.43 at RemoteSha
`bf223df75da6e41ae48607a56b4cd12d7c3b24e7`. The targeted result family
completed at `r current_fash_manifest$generated_at`. The direct-interaction
permutation tests were not rerun: their rows come from the unchanged formal
historical cache. Before combining the rows, the report verifies every
retained SHA-256 digest and requires exact selected-pair/genotype agreement
plus numerical agreement of the simulated inputs to `1e-12`.

```{r result-provenance-table}
knitr::kable(
  result_provenance_table,
  align = "l",
  caption = paste(
    "Table R2.1: Computation source and validation status for each result",
    "component in the refreshed page."
  )
)
```

For donor $i$, variant $j$, and time $t$,

$$
Y_{ijt} = \alpha_{jt} + G_{ij}\beta_j(t) +
X_i^T\gamma_{jt} + \epsilon_{ijt},
\qquad \epsilon_{ijt} \sim N(0,1).
$$

The genotype matrix is taken from the real YRI data used in the manuscript.
For each simulation seed, one tested cis variant is sampled uniformly from
each of the `6,362` tested genes, and its observed dosage vector is retained
for the same `19` donors. The resulting design has one locus-level unit per
gene and an empirical MAF distribution; a variant tested for multiple genes
may therefore appear more than once. R1, R2, R3A, and R3B use the exact same
sampled gene-variant pairs and genotype matrix for a given seed.

The five columns of $X$ remain simulated independently from $N(0,1)$ and are
standardized across donors. Independently for each covariate, gene, and time,
$\gamma_{kjt}\sim N(0,0.5^2)$. The current simulation sets $\alpha_{jt}=0$;
each fitted regression still includes an intercept. Effect trajectories,
covariates, covariate effects, and expression noise remain simulated.

```{r r2-simulate-genotypes-and-covariates, eval=FALSE}
seed <- 12345
component_seeds <- revision_component_seeds(seed)

genotype_cache <- readRDS(
  "output/revision_simulations/shared/real_genotype_one_per_gene_J6362_pilot5/genotype_samples.rds"
)
genotype_sample <- genotype_cache$samples[[as.character(seed)]]
G <- genotype_sample$G
covariates <- simulate_covariate_matrix(
  n_donors = 19,
  n_covariates = 5,
  seed = component_seeds[["covariates"]]
)
```

### Real-genotype design and provenance

```{r genotype-provenance-table}
knitr::kable(
  genotype_provenance_table,
  align = "l",
  caption = "Table R2.2: Provenance and selection rule for the real-genotype simulation design."
)
```

```{r genotype-sampling-table}
knitr::kable(
  genotype_sampling_table,
  align = "l",
  caption = paste(
    "Table R2.3: Per-seed genotype sampling summary. Repeated assignments",
    "refer to the same tested variant selected for more than one gene."
  )
)
```

Each replicate has `19` donors, `16` time points, `6,362` gene-level units,
and five PC-like covariates. There are `r format_integer(dynamic_count)` dynamic
eQTLs, `r format_integer(constant_count)` constant eQTLs with effects drawn
from $N(0,1)$, and `r format_integer(zero_count)` zero-effect units. Therefore,
the true dynamic-null proportion is $\pi_0=5090/6362=$
`r format_decimal(true_pi0, 4)`.

For a dynamic eQTL,

$$
\beta_j(t) = \mu_j + h_j(t),
\qquad \mu_j \sim N(0,1),
$$

where $h_j(t)$ is a sum of one, two, or three compact raised-cosine peaks,

$$
h(t;c,w)=
\frac{1+\cos\{\pi(t-c)/w\}}{2}
\mathbf{1}(|t-c|\leq w),
\qquad w=1.5.
$$

The largest peak is placed at a random location in an early, middle, or late
window. Additional peaks, when present, occur in different windows and have
random amplitudes between `35%` and `75%` of the primary peak. Multi-peak
curves include both same-sign and alternating-sign patterns. The deviation
$h_j$ is not mean-centered; it is scaled to centered RMS `0.9`, and the
independent genetic main effect $\mu_j$ remains as the background effect
outside the compact supports.

```{r r2-simulate-effects-and-expression, eval=FALSE}
class_probs <- c(
  dynamic_bspline = 0.20,
  constant = 0.40,
  zero = 0.40
)
shape_cell_probs <- c(
  k1__spiky__single = 1 / 3,
  `k2__spiky__same-sign` = 1 / 6,
  `k2__spiky__alternating-sign` = 1 / 6,
  `k3__spiky__same-sign` = 1 / 6,
  `k3__spiky__alternating-sign` = 1 / 6
)
evaluation_grid <- seq(0, 15, by = 0.1)

effect_sim <- simulate_raised_cosine_multipeak_effect_set(
  n_variants = 6362,
  time_grid = 0:15,
  evaluation_grid = evaluation_grid,
  class_probs = class_probs,
  width_levels = c(spiky = 1.5),
  spike_counts = 1:3,
  shape_cell_probs = shape_cell_probs,
  primary_time_groups = c("early", "middle", "late"),
  relative_amplitude_range = c(0.35, 0.75),
  center_by_observed_mean = FALSE,
  validate_functional_labels = FALSE,
  switch_threshold = 0.25,
  target_centered_rms = 0.9,
  baseline_sd = 1,
  constant_sd = 1,
  dynamic_baseline_sd = 1,
  exact_class_counts = TRUE,
  seed = seed,
  class_seed = component_seeds[["classes"]],
  constant_seed = component_seeds[["constant_effects"]],
  shape_seed = component_seeds[["functional_truth"]],
  scenario = "r2_real_genotype_one_per_gene_timed_cosine_one_two_three_peak_main_effect_dynamic_eqtl"
)
effect_sim <- reassign_effect_simulation_by_maf(
  effect_sim = effect_sim,
  maf = genotype_sample$variant_info$observed_maf,
  class_probs = class_probs,
  seed = component_seeds[["classes"]],
  n_strata = 10
)
expression_sim <- simulate_eqtl_expression_from_genotypes(
  G = G,
  beta_matrix = effect_sim$beta_matrix,
  time_grid = 0:15,
  covariates = covariates,
  expression_noise_sd = 1,
  covariate_effect_sd = 0.5,
  intercept_sd = 0,
  seed = component_seeds[["expression"]]
)
```

The `r format_integer(dynamic_count)` dynamic units are exactly balanced at
`424` per one-, two-, and three-peak group. Alternating peak signs describe geometry only; this page tests
global dynamic eQTL status and does not equate an alternating pattern with the
manuscript's switch functional. Functional testing is evaluated separately.

### Dynamic shape and timing allocation

```{r allocation-table}
knitr::kable(
  allocation_table,
  align = "l",
  caption = paste(
    "Table R2.4: Allocation of dynamic eQTLs across peak counts, sign",
    "patterns, and primary timing windows in the spiky simulation."
  )
)
```

At each time, effects and standard errors are estimated with
`Y ~ G + PC1 + PC2 + PC3 + PC4 + PC5`. The t-statistic correction uses
`df = 19 - 7 = 12`. FASH receives only these per-time estimates and corrected
standard errors. The direct interaction tests use the full individual-level
data.

```{r r2-estimate-and-fit-fash, eval=FALSE}
eqtl_summary <- estimate_eqtl_summaries_from_genotypes(
  G = G,
  expression = expression_sim$expression,
  covariates = covariates,
  apply_t_se_correction = TRUE
)
datasets <- make_fash_datasets_from_eqtl_summary(
  beta_hat = eqtl_summary$beta_hat,
  se = eqtl_summary$se,
  true_beta = effect_sim$beta_matrix,
  time_grid = 0:15,
  unit_info = effect_sim$unit_info,
  scenario = "r2_real_genotype_one_per_gene_timed_cosine_one_two_three_peak_main_effect_dynamic_eqtl"
)

common_sd_grid <- default_revision_grid()
common_pred_step <- 1
common_penalty <- 10

fash_iwp1_raw <- fashr::fash(
  Y = "y",
  smooth_var = "x",
  S = "sd",
  data_list = datasets,
  order = 1,
  grid = common_sd_grid,
  num_basis = 20,
  pred_step = common_pred_step,
  penalty = common_penalty,
  num_cores = 4,
  verbose = FALSE
)
fash_iwp1_bf <- fashr::BF_update(fash_iwp1_raw, plot = FALSE)

fash_linear_raw <- fit_linear_mixture_fash(
  datasets = datasets,
  grid = common_sd_grid,
  pred_step = common_pred_step,
  penalty = common_penalty
)
fash_linear_bf <- BF_update_linear_mixture_fash(fash_linear_raw)
```

All tables and curves below average five independently simulated datasets with
seeds `12345`, `22345`, `32345`, `42345`, and `52345`. The seed-12345 full fit
is used only for representative trajectories.

Truth labels are reassigned within empirical-MAF deciles so that dynamic,
constant, and zero-effect units have closely matched MAF distributions.

```{r truth-maf-balance-table}
knitr::kable(
  truth_maf_balance_table,
  align = "l",
  caption = "Table R2.5: Empirical MAF balance across truth classes for each simulation seed."
)
```

## Example simulated eQTLs

The first figure shows zero-effect and constant dynamic-null examples. Points
are per-time effect estimates, bars show two corrected standard errors, and
the colored line is the true effect.

```{r null-examples, fig.width=10, fig.height=5.2, fig.cap="Figure R2.1: Representative zero-effect and constant-effect dynamic-null eQTLs from seed 12345. Points are estimated genetic effects, vertical bars span two corrected standard errors, and colored curves are the true effects."}
plot_genotype_eqtl_examples(
  out,
  classes = c("zero", "constant"),
  n_per_class = 3,
  seed = 1,
  true_curve_n = 500
)
```

The second figure shows one representative from each available
shape-by-primary-window cell. Truth curves are drawn on a dense time grid so
that compact peaks between the 16 observed time points remain visible.

```{r dynamic-examples, fig.width=12, fig.height=18, fig.cap="Figure R2.2: Representative compact dynamic eQTLs across peak-count, sign-pattern, and primary-window cells. Points and error bars show the 16 estimated effects and two corrected standard errors; colored dense curves show the continuous truths."}
plot_cosine_peak_cell_examples(
  out,
  seed = 1,
  include_time_groups = TRUE
)
```

## Comparison 1: IWP versus linear FASH

This comparison aligns the exact null, scale grid, `pred_step`, and
null-favoring penalty. `FASH-IWP1` uses the first-order IWP alternative. For
`FASH-linear`, let $h=1$ and $z(t)=\{t-\min(t)\}/h$. Its prior is

$$
\beta_j(t)=a_j+\theta_j z(t),
\qquad
\theta_j\sim\pi_0\delta_0+
\sum_{k=1}^{K}\pi_k N(0,\tau_k^2).
$$

The zero-grid component is the same unrestricted constant null used by IWP1.
For each positive component, $\tau_k$ is the prior SD of the linear departure
at $\min(t)+h$. The complete grid is one exact zero plus 51 positive values
from $\exp(-5)$ to `1`; all weights are estimated jointly with `penalty = 10`.
IWP1 uses the same numerical grid for its one-step predictive SD. This matches
the one-step marginal scale while leaving the linear and IWP covariance shapes
different.

In the BF-adjusted object, `fashr::BF_update()` does not simply retain the raw
penalized non-null weights. It re-fits an unpenalized full-grid mixture,
normalizes the positive-grid weights to define the conditional alternative,
collapses that alternative into a Bayes factor, and then updates the total null
weight. The table below summarizes both the number of active positive-grid
components and their conditional RMS pred-step SD across the five seeds.

### Estimated dynamic-null proportion

```{r pi0-table}
knitr::kable(
  pi0_table,
  align = "l",
  caption = paste(
    "Table R2.6: Estimated dynamic-null proportions and linear-mixture prior",
    "diagnostics across five spiky-effect simulation replicates."
  )
)
```

### Results at alpha 0.05

```{r fash-table}
knitr::kable(
  fash_table,
  align = "l",
  caption = paste(
    "Table R2.7: Monte Carlo power and empirical FDR at alpha 0.05 for IWP",
    "and linear FASH under compact spiky effects."
  )
)
```

### Power as a function of alpha

```{r fash-power, fig.cap="Figure R2.3: Monte Carlo power across nominal alpha for IWP and linear FASH under compact one-, two-, and three-peak effects. Curves identify raw and BF-updated methods in the in-figure legend; uncertainty summarizes five seeds."}
plot_mc_alpha_curves(
  fash_alpha,
  metric = "power",
  title = "IWP versus linear FASH: Monte Carlo power",
  subtitle = "Five seeds; real YRI genotypes; N = 19, J = 6362 genes",
  style_profile = "combined"
)
```

### Empirical FDR as a function of alpha

```{r fash-fdr, fig.cap="Figure R2.4: Monte Carlo empirical FDR across nominal alpha for IWP and linear FASH under compact spiky effects. Curves identify methods in the in-figure legend, and the diagonal reference marks nominal calibration."}
plot_mc_alpha_curves(
  fash_alpha,
  metric = "fdr",
  title = "IWP versus linear FASH: Monte Carlo empirical FDR",
  subtitle = "Five seeds; real YRI genotypes; N = 19, J = 6362 genes",
  legend_position = "topleft",
  style_profile = "combined"
)
```

At $\alpha=0.05$, IWP-BF has mean power
`r format_decimal(metric_at(fash_alpha_005, "FASH-IWP1-BF", "mean_power"))`,
compared with
`r format_decimal(metric_at(fash_alpha_005, "FASH-linear-BF", "mean_power"))`
for linear-BF. Their empirical FDR estimates are
`r format_decimal(metric_at(fash_alpha_005, "FASH-IWP1-BF", "mean_fdr"))`
and
`r format_decimal(metric_at(fash_alpha_005, "FASH-linear-BF", "mean_fdr"))`.

## Comparison 2: IWP FASH versus direct interaction tests

The direct linear LRT tests a linear genotype-by-time interaction. The direct
quadratic LRT jointly tests linear and quadratic interaction terms. Their
permutation eFDR calculations use `B=100` donor-level permutations and are
supplied with the true $\pi_0=5090/6362$, deliberately removing null-proportion
estimation uncertainty from the direct baselines.

Only the IWP FASH rows in this section were recomputed with `fashr` 0.1.43.
The direct linear and quadratic rows are read unchanged from the historical
formal cache after the report confirms that its simulated genotype,
covariates, truth, and expression agree with the retained 0.1.43 payload.

```{r r2-direct-interaction-efdr, eval=FALSE}
out <- add_direct_interaction_efdr_results_to_genotype_output(
  out = out,
  n_permutations = 100,
  alpha = 0.05,
  seed = component_seeds[["permutations"]],
  lambda = 0.5,
  pi0_method = "conservative",
  true_pi0 = 5090 / 6362,
  include_true_pi0 = TRUE,
  permute_covariates_with_expression = TRUE,
  num_cores = 4,
  overwrite = TRUE,
  verbose = FALSE
)
```

### Results at alpha 0.05

```{r direct-table}
knitr::kable(
  direct_table,
  align = "l",
  caption = paste(
    "Table R2.8: Monte Carlo power and empirical FDR at alpha 0.05 for IWP",
    "FASH and direct interaction tests under compact spiky effects."
  )
)
```

### Power as a function of alpha

```{r direct-power, fig.cap="Figure R2.5: Monte Carlo power across nominal alpha for IWP FASH and direct linear or quadratic interaction tests under compact spiky effects. The in-figure legend identifies methods; direct-test eFDR uses the true pi0 and 100 permutations."}
plot_mc_alpha_curves(
  direct_alpha,
  metric = "power",
  title = "IWP FASH versus direct interaction: Monte Carlo power",
  subtitle = "Five seeds; direct eFDR uses true pi0 = 0.8001 and B = 100",
  style_profile = "combined"
)
```

### Empirical FDR as a function of alpha

```{r direct-fdr, fig.cap="Figure R2.6: Monte Carlo empirical FDR across nominal alpha for IWP FASH and direct interaction tests under compact spiky effects. The in-figure legend identifies methods, and the diagonal reference marks nominal calibration."}
plot_mc_alpha_curves(
  direct_alpha,
  metric = "fdr",
  title = "IWP FASH versus direct interaction: Monte Carlo empirical FDR",
  subtitle = "Five seeds; direct eFDR uses true pi0 = 0.8001 and B = 100",
  legend_position = "bottomleft",
  style_profile = "combined"
)
```

At $\alpha=0.05$, mean power is
`r format_decimal(metric_at(direct_alpha_005, "FASH-IWP1-BF", "mean_power"))`
for IWP-BF,
`r format_decimal(metric_at(direct_alpha_005, "Direct-linear-LRT-eFDR-true-pi0", "mean_power"))`
for direct linear, and
`r format_decimal(metric_at(direct_alpha_005, "Direct-quadratic-LRT-eFDR-true-pi0", "mean_power"))`
for direct quadratic. The corresponding empirical FDR estimates are
`r format_decimal(metric_at(direct_alpha_005, "FASH-IWP1-BF", "mean_fdr"))`,
`r format_decimal(metric_at(direct_alpha_005, "Direct-linear-LRT-eFDR-true-pi0", "mean_fdr"))`,
and
`r format_decimal(metric_at(direct_alpha_005, "Direct-quadratic-LRT-eFDR-true-pi0", "mean_fdr"))`.

## Power by number of peaks

The global discovery threshold is fitted once across all `6,362` gene-level units.
Power is then summarized within the true one-, two-, and three-peak dynamic
subgroups; no subgroup-specific threshold is estimated.

### Results at alpha 0.05

```{r peak-table}
knitr::kable(
  peak_table,
  align = "l",
  caption = paste(
    "Table R2.9: Monte Carlo power at alpha 0.05 stratified by the true",
    "number of compact peaks; all calls use the same global threshold."
  )
)
```

### Power as a function of alpha

```{r peak-power, fig.width=10, fig.height=6, fig.cap="Figure R2.7: Monte Carlo discovery power across nominal alpha, stratified by one-, two-, and three-peak truth. Panels denote peak count, curves identify methods in the in-figure legend, and the FDR threshold is fitted globally before subgrouping."}
plot_mc_shape_power_grid(
  peak_alpha[peak_alpha$method %in% peak_methods, ],
  shape_order = c("1-peak", "2-peak", "3-peak"),
  title = "Power by number of compact peaks",
  subtitle = "Five seeds; global FDR threshold applied before subgrouping",
  style_profile = "combined",
  width = 2600
)
```

## Interpretation

The regenerated tables and curves provide the paired comparison of IWP and
linear covariance shapes after matching their exact null, one-step SD grid,
and null-favoring penalty. The peak-stratified analysis reports how each method
behaves as the trajectory contains one, two, or three localized features. The
direct methods are given the true null proportion, whereas FASH estimates its
mixture weights from the summary statistics. R1 and R2 are exactly paired on
their sampled gene-variant units and real genotype matrices for every seed;
only their simulated effect geometries differ.

## Reproducibility

The
[R2 simulation driver](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/r2_spiky_transient/run_spiky_transient_mc_replication.R)
uses the
[shared simulation functions](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/shared/simulation_functions.R).
Real-genotype selection, dosage extraction, validation, and MAF-balanced truth
assignment are implemented in the
[shared real-genotype helper](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/shared/real_genotype_one_per_gene.R),
and the versioned input cache is created by the
[shared cache builder](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/shared/build_real_genotype_one_per_gene_cache.R).
The page-specific cache validation and formatting are in
[R2 reporting code](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/r2_spiky_transient/reporting.R).
The shared
[0.1.43 reporting overlay](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/r1_r2_fashr0143/reporting_overlay.R)
validates the completed targeted cache and combines its FASH rows with only
the direct rows from the historical cache. The following command regenerates
the full historical-style five-seed cache, including direct tests. It is shown
for reproducibility but is intentionally not evaluated during workflowr
builds; the displayed FASH rows read the separate targeted 0.1.43 cache.

The full five-seed cache can be regenerated with:

```{bash regenerate-r2-cache, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/shared/build_real_genotype_one_per_gene_cache.R

Rscript --vanilla \
  code/revision_simulations/r2_spiky_transient/run_spiky_transient_mc_replication.R \
  --J 6362 \
  --n-donors 19 \
  --n-covariates 5 \
  --noise-sd 1 \
  --dynamic-main-effect-sd 1 \
  --width-half 1.5 \
  --target-centered-rms 0.9 \
  --efdr-permutations 100 \
  --seed-list 12345,22345,32345,42345,52345 \
  --genotype-cache output/revision_simulations/shared/real_genotype_one_per_gene_J6362_pilot5/genotype_samples.rds \
  --output-id r2_real_genotype_one_per_gene_J6362_timed_cosine_one_two_three_peak_main_effect_linear_mixture_predstep1_penalty10_pilot5 \
  --num-cores 4
```

After the formal R1-R3 caches have been generated, exact pairing of every
sampled gene-variant unit and genotype matrix is validated with:

```{bash validate-r1-r2-r3-pairing, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/shared/validate_r1_r2_real_genotype_pairing.R \
  --r1-output-id r1_real_genotype_one_per_gene_J6362_random_bspline_main_effect_linear_mixture_predstep1_penalty10_pilot5 \
  --r2-output-id r2_real_genotype_one_per_gene_J6362_timed_cosine_one_two_three_peak_main_effect_linear_mixture_predstep1_penalty10_pilot5 \
  --r3-output-id r3_real_genotype_one_per_gene_J6362_matched_functional_relative_clearance_main_effect_pilot5
```

The report requires a successful
`summary/real_genotype_pairing_validation.csv` artifact and will stop if the
R1/R2/R3 genotype designs are not exactly paired.

The recomputed FASH result family is stored under:

```text
output/revision_simulations/mc/r1_r2_fashr0143/
```

The unchanged historical full fit and direct-interaction results are stored
under:

```text
output/revision_simulations/mc/r2_real_genotype_one_per_gene_J6362_timed_cosine_one_two_three_peak_main_effect_linear_mixture_predstep1_penalty10_pilot5/
```
