---
title: "Internal sensitivity analysis: original Middle window"
author: "Ziang Zhang"
date: "2026-08-20"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 8.0,
  fig.height = 7.2,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)

cache_dir <- file.path(
  "output",
  "revision_simulations",
  "internal",
  "evaluation_grid_mc_all_category_baseline_discoveries_middle_4_11"
)

configuration <- readRDS(file.path(cache_dir, "configuration.rds"))
population <- read.csv(
  file.path(cache_dir, "baseline_discovery_population.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
population_summary <- read.csv(
  file.path(cache_dir, "population_summary.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
pair_lfsr <- read.csv(
  file.path(cache_dir, "pair_lfsr_by_setting.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
comparison_summary <- read.csv(
  file.path(cache_dir, "comparison_summary.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
runtime <- read.csv(
  file.path(cache_dir, "runtime.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
cache_validation <- read.csv(
  file.path(cache_dir, "validation.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

category_order <- c("early", "middle", "late", "switch")
category_labels <- c(
  early = "Early",
  middle = "Middle",
  late = "Late",
  switch = "Switch"
)
expected_pair_counts <- c(
  early = 124L,
  middle = 24L,
  late = 20L,
  switch = 984L
)
expected_gene_counts <- c(
  early = 8L,
  middle = 5L,
  late = 12L,
  switch = 250L
)
expected_setting_ids <- c(
  "grid_0p15_M3000",
  "grid_0p10_M3000",
  "grid_0p05_M3000",
  "grid_0p10_M2000",
  "grid_0p10_M5000"
)
expected_comparison_ids <- c(
  "grid_0p10_vs_0p15_M3000",
  "grid_0p10_vs_0p05_M3000",
  "M2000_vs_M3000_grid_0p10",
  "M3000_vs_M5000_grid_0p10"
)

if (
  configuration$seed != 20260820L ||
  configuration$num_cores != 4L ||
  configuration$alpha != 0.05 ||
  configuration$switch_threshold != 0.25 ||
  configuration$category_pair_count != sum(expected_pair_counts) ||
  configuration$unique_pair_count != 1145L ||
  configuration$middle_definition != "closed_4_11" ||
  !identical(configuration$category_order, category_order) ||
  !identical(
    unname(configuration$expected_pair_counts),
    unname(expected_pair_counts)
  ) ||
  !identical(
    unname(configuration$expected_gene_counts),
    unname(expected_gene_counts)
  ) ||
  !setequal(pair_lfsr$setting_id, expected_setting_ids) ||
  !identical(
    unique(comparison_summary$comparison_id),
    expected_comparison_ids
  ) ||
  nrow(population) != sum(expected_pair_counts) ||
  nrow(pair_lfsr) != sum(expected_pair_counts) *
    length(expected_setting_ids) ||
  nrow(comparison_summary) != length(category_order) *
    length(expected_comparison_ids) ||
  anyDuplicated(population[c("category", "pair_id")]) ||
  anyDuplicated(pair_lfsr[c("category", "pair_id", "setting_id")]) ||
  any(!is.finite(pair_lfsr$lfsr)) ||
  any(pair_lfsr$lfsr < 0 | pair_lfsr$lfsr > 1) ||
  !all(cache_validation$passed)
) {
  stop("The all-category numerical-sensitivity cache failed page validation.")
}

population_summary <- population_summary[
  match(category_order, population_summary$category),
  ,
  drop = FALSE
]
comparison_summary <- comparison_summary[
  order(
    match(comparison_summary$comparison_id, expected_comparison_ids),
    match(comparison_summary$category, category_order)
  ),
  ,
  drop = FALSE
]
```

<style type="text/css">
.sensitivity-key-result {
  border-left: 5px solid #2c7fb8;
  background: #eef6fb;
  padding: 0.9em 1.1em;
  margin: 1em 0 1.4em 0;
}

table > caption,
p.caption {
  color: #1f2937 !important;
  font-size: 1.10em !important;
  font-weight: 700 !important;
  line-height: 1.35;
  text-align: left !important;
  margin-top: 0.7em;
  margin-bottom: 0.7em;
}

table > caption {
  caption-side: top;
}

.table-scroll {
  width: 100%;
  overflow-x: auto;
  margin-bottom: 1em;
}

.table-scroll table {
  min-width: 1040px;
}
</style>

# Question and conclusion

This temporary comparison changes only the Middle functional back to the
original window $4\leq t\leq11$. It asks whether evaluation-grid step 0.10 and
3,000 posterior draws are sufficiently stable for the baseline-discovery pairs
in Early, original Middle, Late, and Switch.

<div class="sensitivity-key-result">
The original Middle functional is <strong>sensitive to the evaluation
grid</strong>. Its mean absolute LFSR change is 0.0135 for grid 0.10 versus 0.15
and 0.0396 for grid 0.10 versus 0.05, compared with 0.0005 and 0.0003 under the
open $3<t<12$ definition. Monte Carlo sensitivity remains small: the original
Middle mean absolute change is 0.0017 for 3,000 versus 5,000 draws. Thus, the
instability is specific to how the original Middle boundary is represented on
the evaluation grid, rather than to posterior Monte Carlo size.
</div>

This is a **reported-pair sensitivity analysis**: the baseline discovery
population is held fixed. It does not rerank all 9,205 candidate pairs under
each numerical setting and therefore does not claim identical full discovery
sets. A full-universe server run would be the appropriate follow-up only if
this targeted diagnostic showed material instability.

# Estimands and fixed analysis population

For posterior trajectory draw $b^{(m)}(t)$, the four statistics are

$$
\begin{aligned}
T_{\mathrm{early}}^{(m)}
&= \sup_{t\leq3}|b^{(m)}(t)|-\sup_{t>3}|b^{(m)}(t)|,\\
T_{\mathrm{middle}}^{(m)}
&= \sup_{4\leq t\leq11}|b^{(m)}(t)|
   -\sup_{t<4\;\mathrm{or}\;t>11}|b^{(m)}(t)|,\\
T_{\mathrm{late}}^{(m)}
&= \sup_{t\geq12}|b^{(m)}(t)|-\sup_{t<12}|b^{(m)}(t)|,\\
T_{\mathrm{switch}}^{(m)}
&= \min\!\left\{\sup_t b^{(m)}(t),-\inf_t b^{(m)}(t)\right\}-0.25.
\end{aligned}
$$

These are continuous-time functionals. Each numerical setting approximates
the suprema and infima by extrema on its stated evaluation grid.

For each category $c$, the estimated local false sign rate is

$$
\widehat{\mathrm{LFSR}}_c
=\frac{1}{M}\sum_{m=1}^{M}I\!\left\{T_c^{(m)}\leq0\right\}.
$$

All four categories use the saved manuscript discoveries with cumulative FSR
at most 0.05. In particular, the original Middle population contains 24 pairs
from 5 genes. The same category-specific pairs are retained in every
comparison.

```{r analysis-population-table}
# Display the exact fixed population used in every sensitivity comparison.
population_table <- data.frame(
  Category = unname(category_labels[population_summary$category]),
  Definition = c(
    "Largest absolute effect at t <= 3",
    "Largest absolute effect at 4 <= t <= 11",
    "Largest absolute effect at t >= 12",
    "Positive and negative excursions both exceed 0.25"
  ),
  `Baseline pairs` = population_summary$n_pairs,
  `Baseline genes` = population_summary$n_genes,
  check.names = FALSE
)

knitr::kable(
  population_table,
  align = c("l", "l", "r", "r"),
  caption = paste(
    "Table 1. Fixed baseline-discovery population for the original-Middle",
    "conditional sensitivity analysis."
  )
)
```

# Numerical construction

Grid sensitivity is evaluated at steps 0.15, 0.10, and 0.05 with the same
3,000 posterior draws. The 0.15 and 0.10 grids are exact row subsets of the
0.05 grid.

For Monte Carlo sensitivity, `predict.fash()` returns draws grouped by sampled
prior PSD component. Therefore, taking its unmodified first columns would not
produce a representative posterior prefix. For each fitted pair, we generate
5,000 draws once, apply a fixed pair-specific random column permutation, and
use the first 2,000, 3,000, and 5,000 permuted columns. This construction makes
the Monte Carlo comparisons nested while holding the posterior sample stream
fixed. Seven pairs occur in more than one category, so 1,152 category-pair rows
require 1,145 unique posterior-sampling calls.

```{r numerical-design-code, eval=FALSE}
# Reproducible construction used by the cache-building script for each pair.
set.seed(20260820L + pair_index)
posterior_samples <- predict(
  fash_fit1_update,
  index = pair_index,
  smooth_var = seq(0, 15, by = 0.05),
  only.samples = TRUE,
  M = 5000L
)

# predict.fash groups columns by PSD component, so randomize column order
# deterministically before taking nested Monte Carlo prefixes.
set.seed(20260820L + 1000000L + pair_index)
posterior_samples <- posterior_samples[
  ,
  sample.int(ncol(posterior_samples)),
  drop = FALSE
]

samples_M2000 <- posterior_samples[, seq_len(2000L), drop = FALSE]
samples_M3000 <- posterior_samples[, seq_len(3000L), drop = FALSE]
samples_M5000 <- posterior_samples[, seq_len(5000L), drop = FALSE]
```

```{r comparison-summary-table, results='asis'}
# Report continuous agreement before the auxiliary individual-LFSR cutoff.
summary_table <- data.frame(
  Comparison = comparison_summary$comparison,
  Category = unname(category_labels[comparison_summary$category]),
  Pairs = comparison_summary$n_pairs,
  `Mean |change|` = sprintf(
    "%.4f",
    comparison_summary$mean_absolute_change
  ),
  `90th percentile |change|` = sprintf(
    "%.4f",
    comparison_summary$q90_absolute_change
  ),
  `Maximum |change|` = sprintf(
    "%.4f",
    comparison_summary$maximum_absolute_change
  ),
  Spearman = sprintf("%.3f", comparison_summary$spearman_correlation),
  `Within 0.01` = paste0(
    sprintf("%.1f", 100 * comparison_summary$fraction_within_0p01),
    "%"
  ),
  `LFSR-0.05 agreement` = paste0(
    sprintf(
      "%.1f",
      100 * comparison_summary$lfsr_0p05_classification_agreement
    ),
    "%"
  ),
  check.names = FALSE
)

summary_table_html <- knitr::kable(
  summary_table,
  format = "html",
  align = c("l", "l", "r", rep("r", 6)),
  caption = paste(
    "Table 2. Paired LFSR agreement for every baseline-discovery pair.",
    "The last column is an auxiliary individual-LFSR diagnostic, not a",
    "recomputed cumulative-FSR discovery classification."
  )
)
knitr::asis_output(paste0(
  '<div class="table-scroll">',
  summary_table_html,
  '</div>'
))
```

The continuous metrics show that the original Middle grid effect is not merely
an isolated 0.05 crossing. For grid 0.10 versus 0.05, only 12.5% of the 24
Middle pairs are within 0.01 and the maximum absolute change is 0.076.

# Four paired LFSR comparisons

The plotting code below explicitly joins the same category-pair rows between
the requested horizontal- and vertical-axis settings. Every panel uses the
same 0--0.16 scale. This is slightly wider than the open-Middle page so that
the original-Middle LFSR of 0.152 is not clipped. The solid diagonal is
equality, dotted diagonals mark an absolute difference of 0.01, and orange
points cross the auxiliary individual-LFSR value 0.05.

```{r scatter-construction}
library(ggplot2)

build_comparison_data <- function(comparison_id) {
  comparison_spec <- configuration$comparisons[
    configuration$comparisons$comparison_id == comparison_id,
    ,
    drop = FALSE
  ]
  if (nrow(comparison_spec) != 1L) {
    stop("The requested comparison is not uniquely specified.")
  }

  x_rows <- pair_lfsr[
    pair_lfsr$setting_id == comparison_spec$x_setting_id,
    c("category", "pair_id", "gene_symbol", "lfsr"),
    drop = FALSE
  ]
  y_rows <- pair_lfsr[
    pair_lfsr$setting_id == comparison_spec$y_setting_id,
    c("category", "pair_id", "lfsr"),
    drop = FALSE
  ]
  names(x_rows)[names(x_rows) == "lfsr"] <- "x_lfsr"
  names(y_rows)[names(y_rows) == "lfsr"] <- "y_lfsr"

  plot_data <- merge(
    x_rows,
    y_rows,
    by = c("category", "pair_id"),
    all = FALSE,
    sort = FALSE
  )
  expected_rows <- sum(configuration$expected_pair_counts)
  if (nrow(plot_data) != expected_rows ||
      anyDuplicated(plot_data[c("category", "pair_id")])) {
    stop("The paired plotting population is incomplete.")
  }

  plot_data$category_panel <- factor(
    plot_data$category,
    levels = category_order,
    labels = unname(category_labels[category_order])
  )
  plot_data$cutoff_status <- ifelse(
    (plot_data$x_lfsr <= 0.05) == (plot_data$y_lfsr <= 0.05),
    "Same side of 0.05",
    "Crosses 0.05"
  )
  plot_data
}

build_comparison_plot <- function(comparison_id, x_label, y_label) {
  plot_data <- build_comparison_data(comparison_id)

  ggplot(plot_data, aes(x = x_lfsr, y = y_lfsr)) +
    geom_abline(
      intercept = c(-0.01, 0.01),
      slope = 1,
      color = "grey65",
      linewidth = 0.45,
      linetype = "dotted"
    ) +
    geom_abline(
      intercept = 0,
      slope = 1,
      color = "grey20",
      linewidth = 0.55
    ) +
    geom_vline(
      xintercept = 0.05,
      color = "grey72",
      linewidth = 0.4,
      linetype = "dashed"
    ) +
    geom_hline(
      yintercept = 0.05,
      color = "grey72",
      linewidth = 0.4,
      linetype = "dashed"
    ) +
    geom_point(
      aes(color = cutoff_status),
      alpha = 0.62,
      size = 1.35
    ) +
    facet_wrap(~category_panel, ncol = 2) +
    scale_color_manual(
      values = c(
        "Same side of 0.05" = "#2c7fb8",
        "Crosses 0.05" = "#d95f02"
      )
    ) +
    scale_x_continuous(
      limits = c(0, 0.16),
      breaks = c(0, 0.025, 0.05, 0.075, 0.10, 0.125, 0.15),
      labels = scales::label_number(accuracy = 0.001),
      expand = expansion(mult = c(0.01, 0.02))
    ) +
    scale_y_continuous(
      limits = c(0, 0.16),
      breaks = c(0, 0.025, 0.05, 0.075, 0.10, 0.125, 0.15),
      labels = scales::label_number(accuracy = 0.001),
      expand = expansion(mult = c(0.01, 0.02))
    ) +
    coord_equal() +
    labs(
      x = x_label,
      y = y_label,
      color = NULL
    ) +
    theme_bw(base_size = 11.5) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey92", linewidth = 0.35),
      strip.background = element_rect(fill = "grey94", color = "grey70"),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom",
      legend.margin = margin(t = -2),
      panel.spacing = grid::unit(0.8, "lines")
    )
}
```

```{r grid-0p10-vs-0p15, fig.cap="Figure 1. Grid-step sensitivity with the original Middle window: 0.10 versus 0.15 at 3,000 posterior draws. Panels contain all baseline-discovery pairs for each category; the discovery population is held fixed."}
build_comparison_plot(
  comparison_id = "grid_0p10_vs_0p15_M3000",
  x_label = "LFSR at grid step 0.10",
  y_label = "LFSR at grid step 0.15"
)
```

Early, Late, and Switch remain close to equality, but original Middle does not.
Its mean absolute change is 0.0135, its maximum change is 0.0327, and only 25%
of its pairs are within 0.01.

```{r grid-0p10-vs-0p05, fig.cap="Figure 2. Grid-step sensitivity with the original Middle window: 0.10 versus 0.05 at 3,000 posterior draws. Panels contain all baseline-discovery pairs for each category; the discovery population is held fixed."}
build_comparison_plot(
  comparison_id = "grid_0p10_vs_0p05_M3000",
  x_label = "LFSR at grid step 0.10",
  y_label = "LFSR at grid step 0.05"
)
```

The original-Middle separation becomes larger on the finer grid: mean absolute
change 0.0396, 90th percentile 0.0650, maximum 0.0760, and only 12.5% of pairs
within 0.01. The other three category panels are unchanged from the
open-Middle analysis.

```{r M2000-vs-M3000, fig.cap="Figure 3. Posterior Monte Carlo sensitivity with the original Middle window: 2,000 versus 3,000 nested draws at grid step 0.10. Panels contain all baseline-discovery pairs for each category; columns were deterministically permuted before taking prefixes."}
build_comparison_plot(
  comparison_id = "M2000_vs_M3000_grid_0p10",
  x_label = "LFSR with 2,000 posterior draws",
  y_label = "LFSR with 3,000 posterior draws"
)
```

The original-Middle mean absolute change is 0.0025, all 24 pairwise changes are
within 0.01, and the signed mean change is close to zero. This is consistent
with ordinary Monte Carlo fluctuation.

```{r M3000-vs-M5000, fig.cap="Figure 4. Posterior Monte Carlo sensitivity with the original Middle window: 3,000 versus 5,000 nested draws at grid step 0.10. Panels contain all baseline-discovery pairs for each category; columns were deterministically permuted before taking prefixes."}
build_comparison_plot(
  comparison_id = "M3000_vs_M5000_grid_0p10",
  x_label = "LFSR with 3,000 posterior draws",
  y_label = "LFSR with 5,000 posterior draws"
)
```

The original-Middle mean absolute change decreases to 0.0017 and every pair is
within 0.01. Thus, 3,000 posterior draws remain adequate; increasing Monte
Carlo size does not resolve the original functional's grid sensitivity.

# Final numerical choices

- Functional definitions: Early $t\leq3$; Middle $4\leq t\leq11$; Late $t\geq12$;
  Switch threshold 0.25.
- Evaluation grid under comparison: `seq(0, 15, by = 0.10)`.
- Posterior Monte Carlo size: 3,000 draws.
- Sensitivity population: all baseline-discovery pairs within each category,
  held fixed across settings.
- Randomization: pair-specific seed 20260820 plus fitted-pair index, followed
  by a pair-specific deterministic column permutation for nested prefixes.
- Computation: one R job with four workers; the production posterior-sampling
  stage took `r sprintf("%.1f", runtime$elapsed_seconds[runtime$stage == "posterior_sampling_and_evaluation"])`
  seconds.
