---
title: "Internal analysis: predictive-SD mixture FASH-linear real-data ablation"
author: "Ziang Zhang"
date: "2026-08-11"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 11,
  fig.height = 7,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(file.path(
  "code",
  "revision_simulations",
  "internal",
  "fash_linear_real_data_ablation",
  "reporting.R"
))
```

<style>
.ablation-banner {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border: 1px solid #d9a441;
  border-left: 6px solid #d9a441;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.ablation-answer {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #b2182b;
  border-radius: 4px;
  background: #fff1f0;
  color: #5a1f25;
}
.ablation-note {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #2c7fb8;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.ablation-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
.MathJax_Display,
pre {
  max-width: 100%;
  overflow-x: auto;
}
</style>

::: {.ablation-banner}
<strong>Internal exploratory ablation.</strong> This page is intentionally
absent from the workflowr homepage, manuscript, and response letter. It uses
the primary time-specific-PC effect estimates and compares the retained IWP1
FASH fit with the predictive-SD finite-mixture FASH-linear comparator. The
separate FASH-CL analysis is explicitly deferred; this page does not update or
reinterpret that comparison.
:::

## Executive answer

::: {.ablation-answer}
<strong>The more comparable mixture prior does not reverse the main real-data
count comparison, but it does change discovery membership.</strong> At
cumulative-lfdr FDR 0.05 after BF adjustment, IWP1 FASH
identifies `r format_integer(current_bf_count$pair_count)` pairs,
`r format_integer(current_bf_count$gene_count)` genes, and
`r format_integer(current_bf_count$variant_count)` unique variants. The
predictive-SD mixture FASH-linear comparator identifies
`r format_integer(linear_bf_count$pair_count)` pairs,
`r format_integer(linear_bf_count$gene_count)` genes, and
`r format_integer(linear_bf_count$variant_count)` variants.
:::

Relative to the historical profiled single-slab version, the new linear
mixture has `r format_integer(abs(bf_version_comparison$pair_count_difference))`
fewer BF-adjusted calls
(`r format_percent(abs(bf_version_comparison$pair_count_difference) / bf_version_comparison$profile_pair_count)`),
and the all-pair lfdr rankings remain highly correlated
(Spearman `r format_decimal(bf_version_comparison$spearman_lfdr, 3L)`). In
contrast, its raw count is higher by
`r format_integer(raw_version_comparison$pair_count_difference)` pairs
(`r format_percent(raw_version_comparison$pair_count_difference / raw_version_comparison$profile_pair_count)`).
The prior migration is therefore modest after BF adjustment but material before
it.

IWP1 FASH identifies
`r format_integer(variant_overlap$current_only_count)` unique variants that
FASH-linear misses, so flexibility produces a substantial distinct discovery
set. However, FASH-linear identifies
`r format_integer(variant_overlap$linear_only_count)` variants that IWP1 FASH
misses. These real-data counts and overlaps do not reveal which calls are true
and cannot establish a power ordering between the covariance models.

## A strongest FASH-linear-only example

The example is selected deterministically rather than for visual appearance:
among BF-adjusted FASH-linear discoveries that are not called by IWP1 FASH,
choose the pair with the smallest FASH-linear lfdr; break ties by the largest
IWP1 lfdr and then pair key. This selects
`r example_summary$gene_symbol` / `r example_summary$variant_id`.

```{r linear-only-example, fig.width=8.4, fig.height=5.4, fig.cap="A strongest predictive-SD mixture FASH-linear-only example. Points and error bars show the time-specific effect estimates and plus or minus two adjusted standard errors. Curves and ribbons show posterior means and 95% intervals under BF-adjusted IWP1 FASH and FASH-linear."}
plot_linear_only_example()
```

For this pair, the BF-adjusted lfdr is
`r format_decimal(example_summary$linear_lfdr, 5L)` under FASH-linear and
`r format_decimal(example_summary$current_lfdr, 3L)` under IWP1. The fitted
curves visualize how the rank-one linear and IWP covariance geometries pool the
same observations differently. This deterministic example is diagnostic of
model behavior; it is not external evidence that either call is true.

## Discordant trajectory gallery

To avoid drawing a conclusion from one selected pair, the gallery below uses
a symmetric deterministic rule: the three strongest BF-adjusted
FASH-linear-only calls and the three strongest BF-adjusted IWP1-only calls,
requiring a different gene for each panel. Each panel reports both FASH lfdr
values and the linear/nonlinear Strober p-values.

```{r discordant-example-gallery, fig.width=12.4, fig.height=8.6, fig.cap="Deterministic discordant examples. The top row contains strong FASH-linear-only calls; the bottom row contains strong IWP1-only calls. Points and error bars are time-specific estimates and plus or minus two adjusted standard errors; curves and ribbons are posterior means and 95% intervals."}
plot_example_gallery()
```

The panels show that discordance is bidirectional. The two covariance models
can concentrate and shrink the same time-specific estimates differently even
after their scale grids, exact-null component, penalty, and BF procedure are
aligned. These examples are diagnostics of model behavior, not independent
validation of either discovery set.

## Interactive trajectory explorer

The companion local Shiny app searches all 1,009,173 tested pairs by HGNC
symbol, Ensembl ID, or rsID. After selecting a pair, checkboxes independently
show the observed estimates, IWP1 FASH, FASH-linear, and posterior intervals.
The plot displays raw and BF-adjusted lfdr for both FASH fits and both Strober
p-values; the adjacent table also displays Strober eFDR and discovery calls.

[Open the FASH trajectory explorer](http://127.0.0.1:7421/) after starting the
local app below.

Run from the workflowr project root:

```{r explorer-command, eval=FALSE}
Rscript --vanilla \
  apps/fash_trajectory_explorer/build_explorer_cache.R

Rscript --vanilla -e \
  'shiny::runApp("apps/fash_trajectory_explorer", host = "127.0.0.1", port = 7421, launch.browser = TRUE)'
```

The compact search index is already retained in the project output directory,
so the first command is needed only when an underlying fit or Strober table
changes. App startup loads the large BF-adjusted IWP1 fit once; selected
posterior trajectories are then computed on demand and cached in memory.

## Matched model definition

Both analyses use the same 1,009,173 time-specific-PC gene-variant effect
estimates, t-adjusted standard errors, time grid, tested pairs, BF adjustment
procedure, and cumulative-lfdr FDR threshold. For pair $j$, the linear model
is

\[
\beta_j(t) = a_j + d_j\frac{t-t_{j,\min}}{h}, \qquad h=1\text{ day},
\]

where the intercept $a_j$ is unrestricted under every component. The null is
the exact point mass $d_j=0$, so $H_0$ is a constant trajectory rather than
a zero trajectory. Under the alternative,

\[
d_j \sim \sum_{k=1}^{51}\pi_k N(0,s_k^2),
\]

and the complete 52-point grid is $s_0=0$ plus the same 51 positive one-day
predictive-SD values used for IWP1. Both fits use the same exact-zero component,
null-favoring penalty of 10, and `fashr` BF-update semantics. This alignment is
on predictive scale, not covariance geometry: FASH-linear induces a rank-one
linear covariance across time, whereas IWP1 induces its integrated-Wiener
covariance.

```{r model-settings, eval=FALSE}
grid <- default_revision_grid()
linear_fit_raw <- fit_linear_mixture_fash_from_stats(
  statistics = statistics,
  grid = grid,
  pred_step = 1,
  penalty = 10,
  statistic_time_span = 15,
  n_time = 16
)
linear_fit_bf <- BF_update_linear_mixture_fash(linear_fit_raw)
alpha <- 0.05
```

The retained sufficient statistics use endpoint-scaled time
$(t-t_{j,\min})/15$. The runner multiplies every one-day SD by 15 before the
rank-one likelihood calculation, exactly bridging the reported one-day grid to
that cached coefficient scale.

The BF update follows `fashr` exactly. It performs an unpenalized full-grid
mixture refit, drops the fitted null component and renormalizes the positive
weights to form the conditional alternative, collapses the likelihood to null
versus that alternative, computes each Bayes factor, estimates the
BF-controlled total null weight, and rebuilds prior and posterior weights. Thus
the BF-adjusted positive weights are based on this conditional-alternative
refit, not the raw penalized fit's conditional weights.

## Discovery counts

```{r discovery-counts, results='asis'}
render_scrollable_table(
  discovery_counts,
  minimum_width = "820px"
)
```

The predictive-SD mixture has
`r format_integer(linear_raw_count$pair_count)` raw calls versus
`r format_integer(current_raw_count$pair_count)` for IWP1. After BF adjustment,
the corresponding counts are `r format_integer(linear_bf_count$pair_count)`
and `r format_integer(current_bf_count$pair_count)`. Count differences do not
identify which model has higher power because the truth is unknown in these
data.

## Learned predictive-SD mixture weights

```{r prior-weights, fig.width=10.5, fig.height=7.4, fig.cap="Learned exact-null and conditional-alternative weights. The upper panels show the fitted exact-null weight for each method; the lower panels show weights on the shared positive one-day predictive-SD grid after conditioning on the alternative. Points mark active positive components."}
plot_prior_weights()
```

```{r prior-summary, results='asis'}
render_scrollable_table(
  prior_summary,
  digits = 6L,
  minimum_width = "640px"
)
```

The linear mixture uses eight active positive components in both fits. Its
null weight changes from `r format_decimal(linear_raw_prior$null_weight, 4L)`
to `r format_decimal(linear_bf_prior$null_weight, 4L)` after BF adjustment;
the conditional-alternative RMS one-day SD is
`r format_decimal(linear_raw_prior$alternative_rms_predstep_sd, 4L)` raw and
`r format_decimal(linear_bf_prior$alternative_rms_predstep_sd, 4L)` after BF
adjustment. Sharing a grid does not force the two covariance models to learn
the same weights.

```{r active-prior-weights, results='asis'}
render_scrollable_table(
  active_prior_weights,
  digits = 6L,
  minimum_width = "760px"
)
```

## Historical profile-to-mixture migration

The profiled single-Gaussian slab is shown here only as the historical version
being replaced; it is not the active FASH-linear comparator.

```{r historical-linear-version-comparison, results='asis'}
render_scrollable_table(
  linear_version_comparison,
  digits = 6L,
  minimum_width = "1500px"
)
```

For raw lfdr, the mixture retains every historical profiled-slab discovery and
adds `r format_integer(raw_version_comparison$mixture_only_pair_count)` pairs;
the set Jaccard index is
`r format_decimal(raw_version_comparison$pair_jaccard, 3L)`. After BF
adjustment, the total changes from
`r format_integer(bf_version_comparison$profile_pair_count)` to
`r format_integer(bf_version_comparison$mixture_pair_count)`, but membership is
not identical:
`r format_integer(bf_version_comparison$profile_only_pair_count)` pairs are
historical-profile-only and
`r format_integer(bf_version_comparison$mixture_only_pair_count)` are
mixture-only (Jaccard
`r format_decimal(bf_version_comparison$pair_jaccard, 3L)`). The BF-adjusted
lfdr Spearman correlation is
`r format_decimal(bf_version_comparison$spearman_lfdr, 3L)`.

## BF-adjusted discovery overlap

```{r overlap-table, results='asis'}
render_scrollable_table(
  overlap_summary,
  digits = 4L,
  minimum_width = "980px"
)
```

```{r overlap-venns, fig.width=12, fig.height=4.4, fig.cap="BF-adjusted cumulative-lfdr FDR-0.05 overlap between IWP1 FASH and predictive-SD mixture FASH-linear at the gene-variant-pair, gene, and unique-variant levels."}
plot_overlap_venns()
```

The pair-level intersection is
`r format_integer(pair_overlap$intersection_count)`.
IWP1 FASH contributes
`r format_integer(pair_overlap$current_only_count)`
exclusive pairs, whereas FASH-linear contributes
`r format_integer(pair_overlap$linear_only_count)`.

## Global lfdr comparison

The left panel contains all tested pairs. The right panel selects one pair per
gene using only the minimum current-FASH BF-adjusted lfdr, then evaluates the
same pair under FASH-linear.

```{r lfdr-scatter, fig.width=11.5, fig.height=6.1, fig.cap="BF-adjusted lfdr comparison for all tested pairs and one current-FASH-selected top pair per gene. Colors indicate cumulative-lfdr FDR-0.05 discovery membership."}
plot_lfdr_scatter()
```

The all-pair Spearman correlation is
`r format_decimal(lfdr_summary$spearman_correlation[lfdr_summary$selection == "All tested gene-variant pairs"], 3L)`;
for current-FASH-selected top pairs it is
`r format_decimal(lfdr_summary$spearman_correlation[lfdr_summary$selection == "Current-FASH top pair per gene"], 3L)`.

## Interpretation for manuscript strategy

::: {.ablation-note}
The matched mixture makes FASH-linear substantially more comparable to IWP1 in
effect-size prior, exact-null, penalty, and BF semantics. The BF-adjusted result
is broadly stable relative to the historical profile version: the linear count
falls by `r format_integer(abs(bf_version_comparison$pair_count_difference))`
(`r format_percent(abs(bf_version_comparison$pair_count_difference) / bf_version_comparison$profile_pair_count)`),
and lfdr rankings remain highly correlated. The raw mixture count, however, is
materially higher, and neither the historical nor IWP1 discovery memberships
are identical to the new mixture membership.
:::

The remaining scientific contrast is primarily covariance geometry, but the
models also learn their mixture weights separately from the data. This
real-data analysis has no known truth, so it cannot establish relative power or
validate either model's exclusive discoveries. Orthogonal evidence such as
replication, functional enrichment, or prespecified shape diagnostics would be
needed for that claim. The FASH-CL comparison remains deferred to its separate
page and is outside this update.

## Reproducibility

Routine page builds read the retained cache and do not refit the million-pair
comparator.

```{r reproduction-command, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/fash_linear_real_data_ablation/\
run_fash_linear_real_data_ablation.R

Rscript --vanilla \
  apps/fash_trajectory_explorer/build_explorer_cache.R

Rscript --vanilla \
  code/revision_simulations/internal/fash_linear_real_data_ablation/\
build_example_gallery.R

Rscript --vanilla -e \
  'workflowr::wflow_build("analysis/revision_internal_fash_linear_real_data_ablation.rmd", local = TRUE)'
```

```{r runtime-table, results='asis'}
render_scrollable_table(runtime_summary, digits = 1L, minimum_width = "760px")
```

The reporting loader requires a completed versioned run, the retained cache and
compact-fit MD5 values, exact grid settings, and all 11 unchanged-IWP checks
before exposing any page object.

```{r unchanged-iwp-validation, results='asis'}
render_scrollable_table(
  unchanged_iwp_validation,
  minimum_width = "620px"
)
```

```{r fit-provenance-table, results='asis'}
render_scrollable_table(fit_provenance, minimum_width = "1280px")
```

```{r provenance-table, results='asis'}
render_scrollable_table(input_provenance, minimum_width = "1420px")
```
