---
title: "Revision Real-data Analysis R6: enhancer enrichment of FASH dynamic eQTLs"
author: "Ziang Zhang"
date: "2026-08-26"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 2
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 8.6,
  fig.height = 6.4,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(file.path(
  "code",
  "revision_simulations",
  "internal",
  "fash_strober_enhancer_comparison",
  "reporting.R"
))
```

<style>
html,
body {
  max-width: 100%;
  overflow-x: hidden;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
pre {
  max-width: 100%;
  overflow-x: auto;
}
</style>

## Question

Do dynamic eQTLs discovered by FASH show orthogonal evidence of regulatory
function? We tested whether one lead variant per discovered gene preferentially
overlaps enhancer annotations from biologically relevant Roadmap epigenomes,
using the Roadmap 15-state ChromHMM enhancer definition of states 6, 7 and 12
described by He and Wang (2017).

Section 1 compares the FASH and Strober discovery sets. Section 2 examines the
FASH lead variants that neither Strober analysis identified.

The analysis is a descriptive, exploratory check, in six steps:

1. take the published FASH and Strober discovery sets;
2. keep one lead variant per discovered gene;
3. define Roadmap enhancers as ChromHMM states 6, 7 and 12;
4. record whether each variant overlaps each epigenome's enhancer annotation;
5. compare the overlap proportion with the remaining tested variants;
6. estimate uncertainty with a leave-one-autosome-out jackknife.

## Roadmap enhancer annotations

We used the Roadmap Epigenomics 15-state ChromHMM annotations (Roadmap
Epigenomics Consortium et al. 2015) and defined enhancers as states `6_EnhG`,
`7_Enh` and `12_EnhBiv` (He and Wang 2017). Each of the
`r format_integer(background_variant_count)` tested variants was assigned an
overlap indicator for each of the thirteen epigenomes below.

```{r panel, echo=FALSE}
panel_table()
```

## Section 1: FASH versus Strober

```{r section1, fig.cap="Figure 1. Enhancer enrichment for the FASH, Strober linear, and Strober quadratic/nonlinear lead-per-gene discovery sets across thirteen Roadmap epigenomes. Points are fold enrichment, bars are pointwise leave-one-autosome-out jackknife 95% intervals, and the dashed line marks no enrichment. One cell falls below the ten-variant overlap floor and is absent: E065 (Strober linear)."}
plot_method_comparison()
```

FASH's `r set_size("current_lead")` lead variants show modest enrichment across
several of the annotations: `r count_above_one("current_lead")` Roadmap
annotations had pointwise jackknife intervals entirely above 1
(`r labels_above_one("current_lead")`), spanning
`r range_above_one("current_lead")`. Both Strober lead sets point in broadly
similar directions, with wider intervals that likely reflect in part their
smaller discovery sets (`r set_size("linear_lead")` and
`r set_size("quadratic_lead")` lead variants). The pattern is spread across
pluripotent, germ-layer and cardiac annotations, so it supports general
regulatory enrichment rather than stage or tissue specificity. The three sets
differ in size, membership and gene composition, and no formal statistical
comparison between them is made here.

## Section 2: FASH-specific discoveries

An rsID is counted as FASH-only if neither Strober analysis identified it,
regardless of gene pairing. Lead selection happens first and the exclusion is
applied afterwards, leaving `r set_size("current_only_lead")` of FASH's
`r set_size("current_lead")` lead variants. This is rsID-level exclusivity, not
LD-independent locus-level novelty.

```{r section2, fig.cap="Figure 2. Enhancer enrichment for FASH lead variants at rsIDs that neither Strober analysis identified (blue), against all FASH lead variants (grey). Same annotations, estimator and jackknife as Figure 1."}
plot_fash_only()
```

The enrichment estimates were similar after excluding rsIDs identified by either
Strober analysis: `r count_above_one("current_only_lead")` annotations had
pointwise intervals entirely above 1, spanning
`r range_above_one("current_only_lead")` against
`r range_above_one("current_lead")` for the unfiltered lead set. This gives
orthogonal support for the regulatory relevance of the additional FASH
discoveries as a set, rather than validating individual variants.

## Methods

**Discovery sets and lead variants.** FASH uses cumulative lfdr at FDR 0.05
after the BF correction; both Strober analyses use their reported eFDR 0.05,
with `non_linear_dynamic_eqtls_5_pc.txt` labelled quadratic/nonlinear. For each
discovered gene we selected its strongest discovered variant — lowest lfdr
(FASH) or smallest p-value (Strober) — and then evaluated the resulting unique
lead variants, so a duplicated rsID counts once. A variant is FASH-only if
neither Strober analysis reported that rsID.

**Enhancer annotations.** Roadmap Epigenomics 15-state ChromHMM segmentations
for the thirteen epigenomes above, with enhancers defined as states `6_EnhG`,
`7_Enh` and `12_EnhBiv` (Roadmap Epigenomics Consortium et al. 2015; He and
Wang 2017). Results were qualitatively similar using the broader
enhancer-related definition (`EnhG`, `Enh`, `EnhBiv` and `BivFlnk`) implemented
in the companion code for Strober et al. (2019)
(`r sensitivity_summary()`).

**Fold enrichment.** The proportion of selected variants overlapping the
annotation divided by the same proportion among the remaining tested variants,
which excludes the selected set from its own comparator. No matched controls are
used. A pair with fewer than `r minimum_overlap` overlapping selected variants is
reported as not estimable.

**Uncertainty.** Uncertainty was assessed using a leave-one-autosome-out
jackknife to account approximately for local dependence among variants. The
intervals are pointwise, and the related Roadmap annotations should not be
interpreted as independent tests; no multiplicity correction is applied.

## Reproducibility

`code/revision_simulations/internal/fash_strober_enhancer_comparison/run_r6_roadmap_panel.R`
builds the annotation indicators and every estimate this page displays, reusing
the epigenome panel and segmentation reader from
`roadmap_enhancer_exploration/` and the estimator from `enrichment_api.R`:

```{r workflow, eval=FALSE}
source(
  "code/revision_simulations/internal/fash_strober_enhancer_comparison/enrichment_api.R"
)
source(
  "code/revision_simulations/internal/roadmap_enhancer_exploration/roadmap_enhancer_exploration_helpers.R"
)

background <- load_background()                       # tested variants
sets <- load_variant_sets()[c(                        # lead per gene
  "current_lead", "linear_lead", "quadratic_lead", "current_only_lead"
)]

annotations <- build_enhancer_annotation_matrix(       # Roadmap indicators
  background[, c("variant_id", "chromosome", "position")],
  definitions = list(EpiCompare = c("6_EnhG", "7_Enh", "12_EnhBiv"))
)

compute_enrichment(                                   # fold enrichment and
  sets, background, annotations, controls = "background"  # chromosome jackknife
)
```

Rebuilding the inputs and then this page's cache:

```{bash rebuild, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/variant_annotation_enrichment/download_variant_annotations.R
Rscript --vanilla \
  code/revision_simulations/internal/variant_annotation_enrichment/run_variant_annotation_enrichment.R
Rscript --vanilla \
  code/revision_simulations/internal/fash_strober_enhancer_comparison/run_fash_strober_enhancer_comparison.R
Rscript --vanilla \
  code/revision_simulations/internal/roadmap_enhancer_exploration/run_roadmap_enhancer_exploration.R
Rscript --vanilla \
  code/revision_simulations/internal/fash_strober_enhancer_comparison/run_r6_roadmap_panel.R
```

The run script also checks that the state-6/7/12 indicators contain the
state-6/7 indicators and are contained in the state-6/7/11/12 indicators, on
every epigenome. Intermediate data stays under `output/`, which is not tracked in
git. Rebuilding this page:

```{bash reproduction-command, eval=FALSE}
Rscript --vanilla -e \
  'workflowr::wflow_build("analysis/revision_enhancer_enrichment.rmd")'
```

## References

He, Y. and Wang, T. (2017). EpiCompare: an online tool to define and explore
genomic regions with tissue or cell type-specific epigenomic features.
*Bioinformatics* **33**(20), 3268–3275.
<https://doi.org/10.1093/bioinformatics/btx371>

Roadmap Epigenomics Consortium, Kundaje, A., Meuleman, W., Ernst, J.,
Bilenky, M., Yen, A., et al. (2015). Integrative analysis of 111 reference
human epigenomes. *Nature* **518**, 317–330.
<https://doi.org/10.1038/nature14248>

Strober, B., Elorbany, R., Rhodes, K., Krishnan, N., Tayeb, K., Battle, A., and
Gilad, Y. (2019). Dynamic genetic regulation of gene expression during cellular
differentiation. *Science* **364**, 1287–1290.
<https://doi.org/10.1126/science.aaw0040>
