---
title: "Simulation 2: Real-genotype simulation"
author: "Ziang Zhang"
date: "2026-09-18"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9.4,
  fig.height = 5.8,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = TRUE
)
project_root <- if (
  file.exists("code/revision_simulations/shared/simulation_functions.R")
) "." else ".."
experiment <- file.path(
  project_root,
  "code/revision_simulations/internal/r3_two_stage_functional_testing_pilot10"
)
source(file.path(experiment, "pilot10_helpers.R"))
source(file.path(experiment, "reporting.R"))
comparison_experiment <- file.path(
  project_root, "code/revision_simulations/internal/r3_stage1_method_comparison"
)
source(file.path(comparison_experiment, "comparison_helpers.R"))
source(file.path(comparison_experiment, "reporting.R"))
report <- r3c_load_report(project_root)
stage1_summary <- report$tables$stage1_mc_summary
stage1_alpha005 <- report$tables$stage1_alpha005_summary
stage2_summary <- report$tables$functional_mc_summary
truth_example_pool <- r3c_load_truth_examples(project_root, report)
peak_power <- r3c_peak_power(project_root, report)
location_power <- r3c_location_power(report)
peak_difference <- function(peaks) {
  x <- peak_power$summary
  x <- x[x$peak_count == peaks & abs(x$alpha - 0.05) < 1e-12, ]
  delta <- x$mean_power[x$method == "Direct-quadratic-LRT-eFDR-true-pi0"] -
    x$mean_power[x$method == "FASH-IWP1-BF"]
  stopifnot(length(delta) == 1L)
  sprintf("%+.1f", 100 * delta)
}
mechanism_labels <- report$baseline$config$contract$mechanism_labels

stage1_value <- function(mechanism, field, digits = 3L, method = "FASH-IWP1-BF") {
  row <- stage1_alpha005[
    stage1_alpha005$truth_mechanism == mechanism &
    stage1_alpha005$method == method,
    ,
    drop = FALSE
  ]
  stopifnot(nrow(row) == 1L)
  formatC(row[[field]], format = "f", digits = digits)
}
```

<style>
p.caption {
  font-weight: 600;
  color: #333;
}
.r3-two-stage-table {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.r3-two-stage-table table {
  width: 100%;
  margin-bottom: 1em;
}
.r3-two-stage-table caption {
  font-weight: 600;
  color: #333;
}
.main-container p code {
  white-space: normal;
  overflow-wrap: anywhere;
  word-break: break-word;
}
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: hidden;
}
</style>

## Overview

This simulation evaluates dynamic eQTL screening followed by temporal
classification. It uses ten Monte Carlo seeds and two truth mechanisms: broad
random B-spline trajectories and compact raised-cosine trajectories. The
analysis proceeds in two stages:

1. Compare BF-adjusted FASH-IWP1, BF-adjusted FASH-linear, and direct linear
   and quadratic interaction tests for dynamic discovery across nominal FDR
   levels from `0.005` through `0.20`, using all `6,362` units.
2. At the fixed threshold $q_{\mathrm{dyn}}=0.05$, classify the FASH-IWP1
   dynamic discoveries as Early, Middle, Late, or Switch, separately for each
   functional, using $q_{\mathrm{func}}=0.05$.

For donor $i$, unit $j$, and observed time $t$,

$$
Y_{ijt}=\alpha_{jt}+G_{ij}\beta_j(t)+X_i^T\gamma_{jt}+\epsilon_{ijt},
\qquad \epsilon_{ijt}\sim N(0,1).
$$

For each seed, one tested cis variant is sampled uniformly for each of the
`6,362` tested genes, and its YRI dosage vector is retained for the same `19`
donors. Each replicate has `16` time points and five standardized covariates.
Per-time regressions include an intercept, genotype, and five covariates;
their t-adjusted standard errors use 12 residual degrees of freedom.
The dynamic, constant, and zero-effect counts are exactly
`1,272/2,545/2,545`, with truth classes balanced across empirical-MAF deciles.

The dynamic effects have Early/Middle/Late probabilities `0.29/0.42/0.29`.
Switch status is balanced within each location, giving six exact cell counts
`185/184/267/267/185/184`. The final trajectory, including its random genetic
main effect, must satisfy the assigned location and Switch margins. Early is
$t\leq3$, Middle is $3<t<12$, and Late is $t\geq12$ on the dense 0.1-day grid;
Switch uses threshold 0.25. The two mechanisms retain the same truth contract:

- Broad effects are cubic B-splines with six independent standard-normal
  coefficients, centered over the observed times and scaled to maximum absolute
  deviation 2.
- Compact effects contain one, two, or three raised-cosine peaks with half-width
  1.5 and are scaled to centered RMS 0.9.

### Examples of true effect curves

Figure 1 shows six examples from seed `12345`: broad B-spline effects
classified as Early, Middle, or Late (A-C), and transient raised-cosine effects
with one, two, or three generating peaks (D-F). Each panel uses the first
eligible unit in the retained illustration pool, ordered by original unit
index, without filtering on discoveries. The missing two-peak illustration
is recovered with the original seed and frozen input generator, with truth,
estimates, and standard errors checked against the retained examples.
Black points are estimated effects,
gray bars span $\pm 2$ reported standard errors, and orange curves show the
true $\beta_j(t)$, including the genetic main effect.

```{r truth-examples, fig.width=10, fig.height=6.2, fig.cap="Figure 1. Six representative simulated trajectories. A-C: broad B-spline effects with Early, Middle, and Late truth, respectively. D-F: transient raised-cosine effects with one, two, and three generating peaks, respectively. Each panel shows one unit: black points are estimated effects, gray bars span +/-2 reported standard errors, and the orange curve is the true effect. Panels use separate effect scales."}
# Select one stored example per category, without using method scores or calls.
example_members <- unique(truth_example_pool[, c("truth_mechanism", "target",
  "unit_index", "pair_key", "true_functional", "peak_count")])
example_members <- example_members[order(example_members$unit_index), ]
panel_labels <- c("A. Broad: Early", "B. Broad: Middle", "C. Broad: Late",
                  "D. Transient: one peak", "E. Transient: two peaks",
                  "F. Transient: three peaks")
selected_examples <- do.call(rbind, lapply(seq_along(panel_labels), function(i) {
  eligible <- if (i <= 3L) {
    example_members$truth_mechanism == "random_bspline" &
      example_members$target == c("early", "middle", "late")[[i]] &
      example_members$true_functional > 0
  } else {
    example_members$truth_mechanism == "raised_cosine" &
      !is.na(example_members$peak_count) & example_members$peak_count == i - 3L
  }
  member <- head(example_members[eligible, c("truth_mechanism", "unit_index",
    "pair_key", "peak_count")], 1L)
  stopifnot(nrow(member) == 1L)
  member$panel <- panel_labels[[i]]
  member
}))
# Check stored peak counts against the validated original generating labels.
peak_labels <- readRDS(file.path(project_root,
  "output/revision_simulations/internal/r3_peak_stratified_power_20260918",
  "peak_labels.rds"))$records[["12345"]]$units
transient_examples <- selected_examples[4:6, ]
stopifnot(identical(transient_examples$peak_count, 1:3),
  identical(transient_examples$peak_count,
    peak_labels$peak_count[match(transient_examples$pair_key, peak_labels$pair_key)]))
example_curves <- merge(unique(truth_example_pool[, c("truth_mechanism",
  "unit_index", "pair_key", "time", "true_effect")]), selected_examples,
  by = c("truth_mechanism", "unit_index", "pair_key"), sort = FALSE)
example_observed <- merge(attr(truth_example_pool, "observed"), selected_examples,
  by = c("truth_mechanism", "unit_index", "pair_key"), sort = FALSE)
example_curves$panel <- factor(example_curves$panel, levels = panel_labels)
example_observed$panel <- factor(example_observed$panel, levels = panel_labels)
stopifnot(!anyNA(example_curves$panel), !anyNA(example_observed$panel),
  all(table(example_curves$panel) == 151L),
  all(table(example_observed$panel) == 16L))

# Use the appendixB convention: estimated effects with +/-2 SE and orange truth.
ggplot2::ggplot(example_observed, ggplot2::aes(time, estimate)) +
  ggplot2::geom_errorbar(ggplot2::aes(ymin = estimate - 2 * se,
    ymax = estimate + 2 * se), width = 0, color = "grey55", linewidth = 0.35) +
  ggplot2::geom_point(size = 1.15, color = "black") +
  ggplot2::geom_line(data = example_curves, ggplot2::aes(time, true_effect),
    color = "#D55E00", linewidth = 0.8) +
  ggplot2::facet_wrap(~ panel, ncol = 3, scales = "free_y") +
  ggplot2::scale_x_continuous(breaks = c(0, 5, 10, 15)) +
  ggplot2::labs(x = "Time (days)", y = "Effect") +
  ggplot2::theme_bw(base_size = 11) +
  ggplot2::theme(panel.grid.minor = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(size = 10),
    axis.text = ggplot2::element_text(size = 9))
```

## Stage 1: dynamic discovery across alpha

FASH-IWP1 and FASH-linear share the unrestricted constant null, the one-day
predictive-SD grid (zero plus 51 positive values from $\exp(-5)$ to 1), and
`penalty = 10`. FASH-linear uses a finite mixture of Gaussian linear slopes.
Both fits use the pinned BF update, which retains the conditional alternative
mixture and updates its total weight against the constant null.

The direct linear LRT tests the genotype-by-time term; the quadratic LRT
jointly tests the linear and quadratic interaction terms. Their eFDR uses
100 donor-level permutations, with expression trajectories and covariates
permuted together relative to genotype. The
direct tests are supplied with the true $\pi_0=5090/6362$; FASH estimates its
null weight. All methods use the same simulated data. FASH receives per-time
regression summaries, while the direct tests receive individual-level data.

Let $D_m(\alpha)$ be the discoveries for method $m$. FASH uses cumulative mean
lfdr, and direct tests use monotone permutation eFDR q-values. For each replicate,

$$
\mathrm{Power}_m(\alpha)=\frac{|D_m(\alpha)\cap H_1|}{|H_1|},
\qquad
\mathrm{FDP}_m(\alpha)=
\frac{|D_m(\alpha)\cap H_0|}{\max\{|D_m(\alpha)|,1\}}.
$$

The curves show the ten-seed means of power and FDP (empirical FDR). Ribbons
are pointwise replicate minima and maxima. Each panel is one truth mechanism;
the vertical line marks $\alpha=0.05$, and the FDR diagonal marks nominal
calibration. Both FASH methods also retain Raw results in the cache for later
Raw-versus-BF comparisons.

```{r stage1-selection, eval=FALSE}
# FASH: rank all-unit lfdr values and select by their cumulative mean.
lfdr_order <- order(lfdr, unit_index)
fdr <- cumsum(lfdr[lfdr_order]) / seq_along(lfdr_order)
fash_calls <- lfdr_order[fdr <= alpha]

# Direct LRTs: apply the same donor permutations to both interaction degrees.
permutation_null <- compute_direct_interaction_permutation_null(
  out = individual_level_data,
  n_permutations = 100L,
  interaction_degrees = c(1L, 2L),
  seed = seed + 10000L,
  permute_covariates_with_expression = TRUE,
  num_cores = num_cores
)
direct_qvalues <- empirical_fdr_qvalues(
  observed_pvalues = observed_lrt$pvalue,
  null_pvalues = as.vector(permutation_null$null_pvalues[[degree_label]]),
  pi0_method = "fixed",
  fixed_pi0 = 5090 / 6362
)$qvalue
direct_calls <- which(direct_qvalues <= alpha)
```

### Power as a function of alpha

```{r stage1-power, fig.width=10.4, fig.height=5.2, fig.cap="Figure 2. Dynamic discovery power for BF-adjusted FASH-IWP1, BF-adjusted FASH-linear, and direct linear and quadratic interaction tests. Panels show broad random B-spline and compact raised-cosine truth; curves are ten-seed means and ribbons are pointwise replicate minima and maxima. Direct-test eFDR uses the true null proportion and 100 donor permutations."}
# Select the four main-comparison methods; Raw FASH results remain in the cache.
power_curves <- stage1_summary[
  stage1_summary$method %in% report$contract$display_methods,
  c("truth_mechanism", "method", "alpha", "mean_power", "power_min", "power_max")
]
power_curves <- power_curves[order(
  power_curves$truth_mechanism,
  match(power_curves$method, report$contract$display_methods),
  power_curves$alpha
), ]
r3c_plot_stage1(power_curves, metric = "power")
```

### Compact-truth power by number of peaks

We split the compact-truth power comparison by the number of generating
raised-cosine peaks. For each seed and group, power is the fraction of that
group's true dynamic units selected using the original global method scores
and thresholds. The three strata exhaust the dynamic units; curves average
these within-group powers across the same ten seeds. Each replicate contains
426, 424, and 422 true dynamic units with one, two, and three peaks, respectively.

```{r stage1-power-by-peaks, fig.width=11, fig.height=4.8, fig.cap="Figure 3. Dynamic discovery power under compact raised-cosine truth, stratified by one, two, or three generating peaks. All four methods retain their original global discovery thresholds. Each curve is the ten-seed mean of within-group power; ribbons show pointwise replicate minima and maxima. The vertical line marks alpha = 0.05."}
# Each stratum's denominator is all true dynamic units with that peak count.
# The reporting gate verifies count-weighted powers reproduce Figure 2B per seed.
peak_curves <- peak_power$summary[
  peak_power$summary$method %in% report$contract$display_methods,
  c("peak_count", "method", "alpha", "mean_power", "power_min", "power_max")
]
method_labels <- setNames(c("FASH-IWP1 (BF)", "FASH-linear (BF)",
  "Direct linear", "Direct quadratic"), report$contract$display_methods)
peak_curves$method_label <- factor(method_labels[peak_curves$method],
  levels = unname(method_labels))
peak_curves$peak_group <- factor(peak_curves$peak_count, levels = 1:3,
  labels = c("1 peak\n426 units per seed", "2 peaks\n424 units per seed",
             "3 peaks\n422 units per seed"))
stopifnot(nrow(peak_curves) == 480L, !anyNA(peak_curves$method_label),
          !anyNA(peak_curves$peak_group))
method_colors <- setNames(c("#0072B2", "#D55E00", "#009E73", "#CC79A7"),
  unname(method_labels))
method_lines <- setNames(c("solid", "longdash", "dotted", "dotdash"),
  unname(method_labels))
ggplot2::ggplot(peak_curves, ggplot2::aes(alpha, mean_power,
    color = method_label, fill = method_label, linetype = method_label)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = power_min, ymax = power_max),
    alpha = 0.09, color = NA, show.legend = FALSE) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::geom_vline(xintercept = 0.05, color = "grey50",
    linetype = "dashed", linewidth = 0.4) +
  ggplot2::facet_wrap(~peak_group, nrow = 1) +
  ggplot2::scale_color_manual(values = method_colors, drop = FALSE) +
  ggplot2::scale_fill_manual(values = method_colors, drop = FALSE) +
  ggplot2::scale_linetype_manual(values = method_lines, drop = FALSE) +
  ggplot2::scale_x_continuous(breaks = c(0.005, 0.05, 0.10, 0.15, 0.20)) +
  ggplot2::coord_cartesian(ylim = c(0, 1)) +
  ggplot2::labs(x = "Nominal Stage-1 FDR level", y = "Mean within-group dynamic power",
    color = NULL, fill = NULL, linetype = NULL) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(legend.position = "bottom", panel.grid.minor = ggplot2::element_blank(),
    strip.background = ggplot2::element_rect(fill = "grey95")) +
  ggplot2::guides(color = ggplot2::guide_legend(nrow = 2, byrow = TRUE))
```

At $\alpha=0.05$, direct quadratic minus BF-adjusted FASH-IWP1 mean power is
`r peak_difference(1L)`, `r peak_difference(2L)`, and `r peak_difference(3L)`
percentage points for one, two, and three peaks, respectively. Figure 5 reports
the corresponding overall empirical FDR comparison.

### Power by temporal category under both truth mechanisms

We split both the broad and compact truth comparisons by true Early, Middle,
and Late category, with 369, 534, and 369 dynamic units per seed under each
mechanism. Each group includes
both Switch and non-Switch units. Power uses all true dynamic units in that
category as its denominator and the original global discovery thresholds.
This summarizes Stage-1 discovery within true temporal categories; Stage-2
conditional classification is evaluated separately below.

```{r stage1-power-by-location, fig.width=11, fig.height=7.4, fig.cap="Figure 4. Dynamic discovery power stratified by true Early, Middle, and Late category. Rows show broad random B-spline (top) and compact raised-cosine (bottom) truth. Each stratum includes both Switch and non-Switch units. Curves are ten-seed mean within-category powers using the original global discovery thresholds; ribbons show pointwise replicate minima and maxima. The vertical line marks alpha = 0.05."}
# Each denominator is all true dynamic units with a positive temporal functional.
# The reporting gate verifies count-weighted powers reproduce both panels of Figure 2 per seed.
location_curves <- location_power$summary[
  location_power$summary$method %in% report$contract$display_methods,
  c("truth_mechanism", "target", "method", "alpha", "mean_power", "power_min", "power_max")
]
method_labels <- setNames(c("FASH-IWP1 (BF)", "FASH-linear (BF)",
  "Direct linear", "Direct quadratic"), report$contract$display_methods)
location_curves$method_label <- factor(method_labels[location_curves$method],
  levels = unname(method_labels))
location_curves$location_group <- factor(location_curves$target,
  levels = c("early", "middle", "late"),
  labels = c("Early\n369 units per seed", "Middle\n534 units per seed",
             "Late\n369 units per seed"))
location_curves$mechanism <- factor(location_curves$truth_mechanism,
  levels = c("random_bspline", "raised_cosine"),
  labels = c("Broad random B-spline", "Compact raised cosine"))
stopifnot(nrow(location_curves) == 960L, !anyNA(location_curves$mechanism), !anyNA(location_curves$method_label),
          !anyNA(location_curves$location_group))
method_colors <- setNames(c("#0072B2", "#D55E00", "#009E73", "#CC79A7"),
  unname(method_labels))
method_lines <- setNames(c("solid", "longdash", "dotted", "dotdash"),
  unname(method_labels))
ggplot2::ggplot(location_curves, ggplot2::aes(alpha, mean_power,
    color = method_label, fill = method_label, linetype = method_label)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = power_min, ymax = power_max),
    alpha = 0.09, color = NA, show.legend = FALSE) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::geom_vline(xintercept = 0.05, color = "grey50",
    linetype = "dashed", linewidth = 0.4) +
  ggplot2::facet_grid(mechanism ~ location_group) +
  ggplot2::scale_color_manual(values = method_colors, drop = FALSE) +
  ggplot2::scale_fill_manual(values = method_colors, drop = FALSE) +
  ggplot2::scale_linetype_manual(values = method_lines, drop = FALSE) +
  ggplot2::scale_x_continuous(breaks = c(0.005, 0.05, 0.10, 0.15, 0.20)) +
  ggplot2::coord_cartesian(ylim = c(0, 1)) +
  ggplot2::labs(x = "Nominal Stage-1 FDR level", y = "Mean within-group dynamic power",
    color = NULL, fill = NULL, linetype = NULL) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(legend.position = "bottom", panel.grid.minor = ggplot2::element_blank(),
    strip.background = ggplot2::element_rect(fill = "grey95")) +
  ggplot2::guides(color = ggplot2::guide_legend(nrow = 2, byrow = TRUE))
```

### Empirical FDR as a function of alpha

```{r stage1-fdr, fig.width=10.4, fig.height=5.2, fig.cap="Figure 5. Empirical dynamic FDR for the same four methods and paired datasets. Curves average replicate-specific false-discovery proportions across ten seeds; ribbons are pointwise replicate minima and maxima. The diagonal marks nominal FDR, and the vertical line marks alpha = 0.05."}
# Use the same truth, method order, and alpha grid as the power comparison.
fdr_curves <- stage1_summary[
  stage1_summary$method %in% report$contract$display_methods,
  c("truth_mechanism", "method", "alpha", "mean_empirical_fdr",
    "empirical_fdr_min", "empirical_fdr_max")
]
fdr_curves <- fdr_curves[order(
  fdr_curves$truth_mechanism,
  match(fdr_curves$method, report$contract$display_methods),
  fdr_curves$alpha
), ]
r3c_plot_stage1(fdr_curves, metric = "fdr")
```

At $\alpha=0.05$, mean power for BF-adjusted FASH-IWP1 and direct quadratic
is `r stage1_value("random_bspline", "mean_power")` and
`r stage1_value("random_bspline", "mean_power", method = "Direct-quadratic-LRT-eFDR-true-pi0")`
under broad truth, and `r stage1_value("raised_cosine", "mean_power")` and
`r stage1_value("raised_cosine", "mean_power", method = "Direct-quadratic-LRT-eFDR-true-pi0")`
under compact truth. Under compact truth, their mean empirical FDR values are
`r stage1_value("raised_cosine", "mean_empirical_fdr")` and
`r stage1_value("raised_cosine", "mean_empirical_fdr", method = "Direct-quadratic-LRT-eFDR-true-pi0")`,
respectively; power should therefore be read together with empirical FDR.

## Stage 2: conditional functional classification at 0.05

For the remainder of the analysis, $D(0.05)$ denotes the BF-adjusted
FASH-IWP1 discoveries at 0.05. This candidate set remains fixed for Stage 2.
The fitted prior remains the full-data prior. For each candidate and each
functional, 3,000 posterior draws are evaluated on the dense time grid. The
candidate lfsr values are then ranked separately for Early, Middle, Late, and
Switch, and calls are made while their cumulative FSR is at most 0.05. The four
classifications are evaluated separately and need not be mutually exclusive.

For target $k$, let $S_k\subseteq D(0.05)$ be its final calls and let
$T_k\{\beta_j\}>0$ denote true membership. We report

$$
\mathrm{FSR}_k=
\frac{\sum_{j\in S_k}\mathbf{1}[T_k\{\beta_j\}\leq0]}
{\max(|S_k|,1)},
$$

and conditional classification power

$$
\mathrm{Power}_{k\mid\mathrm{Stage\ 1}}=
\frac{\sum_{j\in S_k}\mathbf{1}[T_k\{\beta_j\}>0]}
{\sum_{j\in D(0.05)}\mathbf{1}[T_k\{\beta_j\}>0]}.
$$

The tables average each replicate-specific rate and call count across the ten
seeds.

```{r stage2-selection, eval=FALSE}
dynamic_candidates <- r3ts_stage1(
  fdr_table,
  n_units = 6362L,
  q = 0.05
)
lfsr_by_target <- compute_functional_lfsr(
  fit = fit_iwp1_bf,
  functionals = temporal_functionals,
  indices = dynamic_candidates,
  smooth_var = seq(0, 15, by = 0.1),
  num_cores = num_cores,
  seed = functional_posterior_seed
)
stage2_calls <- r3ts_stage2(
  dynamic_candidates,
  lfsr_by_target,
  n_units = 6362L,
  q = 0.05
)
```

```{r stage2-broad-table, results='asis'}
# Select the truth mechanism and retain only the three displayed estimands.
broad_table <- stage2_summary[
  stage2_summary$truth_mechanism == "random_bspline",
  c("truth_mechanism", "target", "mean_empirical_fsr",
    "mean_conditional_classification_power", "mean_calls")
]
broad_table <- broad_table[
  match(c("early", "middle", "late", "switch"), broad_table$target),
]
r3ts10_show_table(
  r3ts10_format_stage2_table(broad_table, "random_bspline"),
  paste(
    "Table 1. Ten-seed mean Stage-2 results under broad random B-spline truth.",
    "Classification power is conditional on true target members retained by",
    "Stage 1 at q = 0.05."
  )
)
```

```{r stage2-compact-table, results='asis'}
# Select the truth mechanism and retain only the three displayed estimands.
compact_table <- stage2_summary[
  stage2_summary$truth_mechanism == "raised_cosine",
  c("truth_mechanism", "target", "mean_empirical_fsr",
    "mean_conditional_classification_power", "mean_calls")
]
compact_table <- compact_table[
  match(c("early", "middle", "late", "switch"), compact_table$target),
]
r3ts10_show_table(
  r3ts10_format_stage2_table(compact_table, "raised_cosine"),
  paste(
    "Table 2. Ten-seed mean Stage-2 results under compact raised-cosine truth.",
    "Classification power is conditional on true target members retained by",
    "Stage 1 at q = 0.05."
  )
)
```

Stage-1 screening and Stage-2 classification answer distinct questions. The
curves quantify which true dynamic effects reach the candidate set; the tables
then quantify functional classification among true target members in that set.
The two nominal 0.05 thresholds are operating rules, and the ten-seed empirical
rates describe performance under the two retained truth mechanisms.

## Reproducibility

The additional comparison is implemented in
`code/revision_simulations/internal/r3_stage1_method_comparison/` and reuses
the completed ten-seed simulation inputs and IWP1 results. Its input-pairing checks
validate genotype, truth, regression summaries, individual-level expression,
and covariates before fitting the additional methods. The original two-stage
runner remains in
`code/revision_simulations/internal/r3_two_stage_functional_testing_pilot10/`.
All FASH fits use `fashr` 0.1.43 at Git SHA
`bf223df75da6e41ae48607a56b4cd12d7c3b24e7`.

The compact comparison cache retains all-unit scores, Raw/BF prior weights,
per-seed curves, and direct-test diagnostics. Full Raw/BF fit objects and
permutation null matrices remain in server checkpoints. Stage-2 tables are
copied unchanged from the validated two-stage cache. This page reads only
completed, hash-validated caches. To run the additional methods on Midway3:

```bash
sbatch /project/mstephens/ziangzhang/fash/workspace/slurm/run_r3_stage1_method_comparison.sbatch
```

```{r cache-identity, results='asis', echo=FALSE}
cat("Comparison cache ID: `", report$contract$result_id, "`.\n", sep = "")
```
