---
title: "Internal analysis: enhancer enrichment by FASH temporal class"
author: "Ziang Zhang"
date: "2026-08-25"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 2
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9.5,
  fig.height = 4.1,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(file.path(
  "code",
  "revision_simulations",
  "internal",
  "fash_temporal_class_enhancer_enrichment",
  "reporting.R"
))
```

<style>
html,
body {
  max-width: 100%;
  overflow-x: hidden;
}
.temporal-class-internal-banner {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border: 1px solid #d9a441;
  border-left: 6px solid #d9a441;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.temporal-class-definition {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.temporal-class-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.temporal-class-table-scroll table {
  margin-bottom: 1em;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
pre,
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
}
</style>

<div class="temporal-class-internal-banner">
<strong>Internal analysis page.</strong> This page is not linked from the
workflowr homepage, manuscript, or response letter.
</div>

## Question

The R6 page shows that FASH's dynamic-eQTL discoveries fall inside Roadmap
enhancers of this differentiation more often than the tested variants do. This
page splits those same discoveries by *when* the effect peaks — early, middle,
late — and asks whether the enrichment concentrates in a specific enhancer
category.

The naive prediction is stage matching: early variants in E020 (iPSC)
enhancers, middle in E013 (mesoderm), late in E095 (left ventricle). The answer
below is that stage matching does **not** hold, but the temporal classes do
differ sharply in which regulatory compartment they occupy.

Everything here is descriptive and exploratory. Nothing is refit: the discovery
set, the timing classification, and the enrichment machinery all come from
existing caches and from `enrichment_api.R` unchanged.

## Temporal classes

The manuscript classifies a discovered pair by where the peak of
$|\beta(t)|$ falls in the 15-day time course, using the Strober-style
functionals

$$
\text{early: } \sup_{t\le 3}|\beta(t)| - \sup_{t>3}|\beta(t)| > 0,\qquad
\text{middle: } \sup_{3<t<12}|\beta(t)| - \sup_{t\le 3 \text{ or } t\ge 12}|\beta(t)| > 0,
$$
$$
\text{late: } \sup_{t\ge 12}|\beta(t)| - \sup_{t<12}|\beta(t)| > 0.
$$

`testing_functional()` returns, for each functional, the posterior probability
that it is not positive (`lfsr`), so $1 - \texttt{lfsr}$ is the posterior
probability that the peak lies in that window. These are already cached for the
FASH-IWP1 fit in `output/dynamic_eQTL_real/classify_dyn_eQTLs_*.RData`.

<div class="temporal-class-definition">
<ul>
<li><strong>Discovery set:</strong> the R6 set — all
`r format_integer(nrow(labelled_pairs))` gene-variant pairs
(`r format_integer(length(discovered_variants))` unique variants) called by
BF-adjusted FASH-IWP1 at cumulative-lfdr FDR
`r format_decimal(analysis_alpha, 2L)`.</li>
<li><strong>Class:</strong> each pair is assigned to the window with the
largest posterior probability. Early $[0,3]$, Middle $(3,12)$, and Late
$[12,15]$ partition the evaluation grid. Their positive-window probabilities
sum to the posterior non-null probability rather than one because an exact-null
trajectory belongs to no timing class; the three cached classifications were
sampled separately, so their empirical sum also contains Monte Carlo error
(median row sum `r format_decimal(median(probability_row_sums))`).</li>
<li><strong>All variants:</strong> every unique rsID in the class.</li>
<li><strong>One lead variant per gene:</strong> the lowest-lfdr pair of each
gene <em>within</em> the class, so a gene whose variants split across classes
contributes one lead to each.</li>
</ul>
</div>

```{r class-summary-table, results='asis'}
render_internal_table(
  class_summary_display,
  align = "lrrrrrr",
  minimum_width = "980px"
)
```

The last two columns are why the assignment is by rank rather than by a
significance threshold. Applying the manuscript's own
`cfsr` $\le$ `r format_decimal(analysis_alpha, 2L)` cutoff leaves only
`r format_integer(confident_call_counts$pairs[1])` early,
`r format_integer(confident_call_counts$pairs[2])` middle, and
`r format_integer(confident_call_counts$pairs[3])` late pairs, from
`r format_integer(confident_call_counts$genes[1])`,
`r format_integer(confident_call_counts$genes[2])`, and
`r format_integer(confident_call_counts$genes[3])` genes respectively. Those
sets are at or below the `r MINIMUM_OVERLAP`-overlap floor an enrichment
estimate needs, so a strict reading of the classification cannot answer this
question at all. The
ranked assignment uses every discovery and buys the power at the cost of
calling many pairs whose timing is uncertain; the confident-subset sensitivity
below reruns the analysis with a probability floor.

The class sizes are very unequal: the late class holds
`r class_count("late", "pairs")` of the `r format_integer(nrow(labelled_pairs))`
pairs. Read every comparison below with that in mind — the late class is close
to the whole FASH discovery set, so its estimates reproduce the R6 numbers
almost by construction, whereas the early class rests on
`r class_count("early", "genes")` genes.

## Enrichment against the tested background

Same comparator, definition, and interval as the R6 page: fold enrichment is
the overlap rate in the set divided by the overlap rate among all
`r format_integer(background_variant_count)` tested autosomal variants, with the
set excluded from its own comparator, and the bars are 22-block
leave-one-autosome-out jackknife 95% intervals.

```{r background-enrichment, fig.cap="Figure 1. Enhancer enrichment of the early, middle, and late FASH discovery classes relative to all tested variants, using every variant in the class (left) and one lead variant per gene (right). Annotations are the Roadmap 15-state ChromHMM enhancer states of the three epigenomes bracketing the differentiation. The dashed line marks no enrichment; bars are leave-one-autosome-out jackknife 95% intervals. Cells whose overlap fell below the floor of 10 variants are not plotted."}
plot_background_enrichment()
```

```{r background-enrichment-table, results='asis'}
render_internal_table(
  enrichment_display_table(background_enrichment),
  align = "llrrr",
  minimum_width = "980px"
)
```

Stage matching fails at the first row. The early class is **depleted**, not
enriched, at the iPSC enhancers it was supposed to match:
`r describe_fold("early_all", e020)` at E020, and equally depleted at E013,
`r describe_fold("early_all", e013)`. Its lead set is too small to estimate at
E020 and E013 — `r describe_fold("early_lead", e020)`.

The late class carries the R6 signal: `r describe_fold("late_all", e013)` at
E013 and `r describe_fold("late_all", e095)` at E095 using every variant, and
`r describe_fold("late_lead", e013)` and `r describe_fold("late_lead", e095)`
under lead selection. These sit close to the R6 estimates for the whole FASH
lead set (1.47 at E013, 1.26 at E095), which is expected given that
the late class is two thirds of the discoveries.

The middle class points at the mesoderm state under lead selection,
`r describe_fold("middle_lead", e013)`, which is the largest estimate in its
row and the one place where the stage-matched prediction is directionally
supported — but the interval covers 1 and this is one contrast among eighteen.

## Is any of it specific to the class?

Figure 1 compares each class with the whole testing universe, so a class
inherits whatever enrichment FASH discoveries have in general. Replacing the
comparator with the other `r format_integer(length(discovered_variants))`
FASH-discovered variants asks the sharper question: given that a variant was
discovered, does its timing class predict which enhancers it hits?

```{r within-discovery-enrichment, fig.cap="Figure 2. The same three classes, now measured against the other FASH-discovered variants rather than against the full testing universe. A value above 1 means the class hits that enhancer state more often than the rest of FASH's discoveries do."}
plot_within_discovery_enrichment()
```

```{r within-discovery-table, results='asis'}
render_internal_table(
  enrichment_display_table(within_discovery_enrichment),
  align = "llrrr",
  minimum_width = "980px"
)
```

Timing does carry information, in one direction. Against its fellow
discoveries the early class is depleted at all three states —
`r describe_fold("early_all", e020, within_discovery_enrichment)` at E020,
`r describe_fold("early_all", e013, within_discovery_enrichment)` at E013,
`r describe_fold("early_all", e095, within_discovery_enrichment)` at E095 —
and the late class is enriched,
`r describe_fold("late_all", e013, within_discovery_enrichment)` at E013 and
`r describe_fold("late_all", e095, within_discovery_enrichment)` at E095. The
middle class is indistinguishable from the rest of the discoveries.

So the answer to "is there a specific enhancer category" is: not a
*stage-specific* one. What separates the classes is enhancers versus
everything else, and the separation runs late-enriched, early-depleted, in all
three epigenomes at once.

## Where the early class sits instead

An enhancer depletion is only interpretable next to the compartments the same
variants do occupy. The panel below adds the cell-type-agnostic ENCODE cCRE
enhancer-like annotation and the promoter/TSS annotations to the same
all-variant comparison against the tested background.

```{r regulatory-context, fig.height=4.6, fig.cap="Figure 3. Regulatory context for the three classes, all-variant view, against the tested background. Left: enhancer annotations. Right: promoter and active-TSS annotations. The early class is displaced from enhancers towards promoters and active TSS; the late class shows the opposite tilt."}
plot_regulatory_context()
```

```{r regulatory-context-table, results='asis'}
render_internal_table(
  enrichment_display_table(context_enrichment, context_annotations),
  align = "llrrrrrrrrr",
  minimum_width = "1580px"
)
```

The early class is promoter-shifted: `r describe_fold("early_all", "Roadmap E020 iPS-20b: Active TSS", context_enrichment)`
at the E020 active-TSS state, `r describe_fold("early_all", "Roadmap E013 hESC-derived CD56+ mesoderm: Active TSS", context_enrichment)`
at the E013 active-TSS state, and
`r describe_fold("early_all", "GENCODE promoter +/-2 kb", context_enrichment)`
at GENCODE promoters, against
`r fold_value("early_all", "ENCODE cCRE enhancer-like", context_enrichment)`-fold
at cCRE enhancer-like elements. The late class shows the mirror pattern, with
its enrichment on the enhancer side.

That is a coherent reading of the depletion rather than a contradiction of it:
variants whose effect is strongest in the first three days sit
disproportionately at promoters and active TSS of the starting epigenome, and
the variants that drive the enhancer enrichment R6 reports are the ones whose
effect peaks late.

## Sensitivities

### Matched controls

The classes differ in allele frequency and TSS distance, and enhancers are
themselves TSS-proximal, so the background comparison is confounded in
principle. Repeating Figure 1 against matched controls — five comparable
variants per selected variant on the MAF, TSS-distance, variant-density and
tested-gene strata, pooled over 100 seeds, the sampler R6 retains but does not
display — leaves the pattern intact.

```{r matched-enrichment, fig.cap="Figure 4. Figure 1 recomputed against matched control variants instead of the raw tested background."}
plot_matched_enrichment()
```

```{r matched-table, results='asis'}
render_internal_table(
  enrichment_display_table(matched_enrichment),
  align = "llrrr",
  minimum_width = "980px"
)
```

The early depletion survives matching:
`r describe_fold("early_all", e013, matched_enrichment)` at E013 and
`r describe_fold("early_all", e020, matched_enrichment)` at E020, against
`r fold_value("early_all", e013)`-fold and `r fold_value("early_all", e020)`-fold
unmatched. It is not an artefact of the early class sitting closer to a TSS.

### Confident class calls only

Restricting to pairs whose winning window probability is at least
`r format_decimal(confident_probability_floor, 1L)` keeps
`r format_integer(length(confident_variant_sets$early_all))` early,
`r format_integer(length(confident_variant_sets$middle_all))` middle, and
`r format_integer(length(confident_variant_sets$late_all))` late variants.

```{r confident-table, results='asis'}
render_internal_table(
  enrichment_display_table(confident_enrichment),
  align = "llrrr",
  minimum_width = "980px"
)
```

Most early cells fall below the overlap floor at this stringency, so the early
depletion cannot be re-checked here. Among what survives, the stage-matched
reading looks slightly better than it does in the full assignment: the middle
lead set reaches `r describe_fold("middle_lead", e013, confident_enrichment)`
at the mesoderm state, and the late lead set
`r describe_fold("late_lead", e095, confident_enrichment)` at the
left-ventricle state. Both are single post-hoc contrasts on small sets with
wide intervals and neither is corrected for the number of comparisons on this
page; they are worth a follow-up, not a claim.

## Readout

1. **No stage matching.** Early variants are not enriched in iPSC enhancers;
   they are significantly depleted there, and their point estimate is below 1
   at every enhancer annotation on this page.
2. **The R6 enhancer enrichment is a late-class phenomenon.** The late class is
   enriched at E013 and E095 both against the tested background and against the
   other FASH discoveries; the middle class is indistinguishable from the rest
   of the discoveries.
3. **The classes differ by compartment, not by cell type.** Early discoveries
   are promoter- and TSS-shifted, late discoveries enhancer-shifted, and this
   survives matched controls.
4. **The strict classification is underpowered for this question.** At
   `cfsr` $\le$ `r format_decimal(analysis_alpha, 2L)` the three classes hold
   `r format_integer(confident_call_counts$pairs[1])`,
   `r format_integer(confident_call_counts$pairs[2])`, and
   `r format_integer(confident_call_counts$pairs[3])` pairs, which no
   annotation-overlap analysis can use.

## Caveats

- The class assignment is a ranking of three posterior probabilities, not a
  significance call. Half the discoveries have a winning probability below
  `r format_decimal(stats::median(labelled_pairs$probability))`.
- Class sizes are unequal by a factor of five, and the effective number of
  independent units is the gene count
  (`r class_count("early", "genes")` / `r class_count("middle", "genes")` /
  `r class_count("late", "genes")`), not the variant count. The chromosome
  jackknife absorbs within-chromosome dependence only approximately.
- No multiplicity correction is applied anywhere on this page, and the
  annotation contrasts are correlated (the three epigenomes share much of their
  enhancer catalogue).
- Enrichment is orthogonal support for regulatory relevance. It does not
  validate individual discoveries or their assigned timing.

## Reproduction

```{r reproduction-commands, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/fash_temporal_class_enhancer_enrichment/test_temporal_class_helpers.R

Rscript --vanilla \
  code/revision_simulations/internal/fash_temporal_class_enhancer_enrichment/test_reporting.R

Rscript --vanilla -e \
  'workflowr::wflow_build("analysis/revision_internal_fash_temporal_class_enhancer_enrichment.rmd")'
```

Page-specific source files:

- `code/revision_simulations/internal/fash_temporal_class_enhancer_enrichment/temporal_class_helpers.R`
- `code/revision_simulations/internal/fash_temporal_class_enhancer_enrichment/test_temporal_class_helpers.R`
- `code/revision_simulations/internal/fash_temporal_class_enhancer_enrichment/reporting.R`
- `code/revision_simulations/internal/fash_temporal_class_enhancer_enrichment/test_reporting.R`

Cached inputs, neither of them produced here:

- `output/revision_simulations/internal/fash_strober_enhancer_comparison_fashr0143/discovery_pair_tables.rds`
  (R6 discovery sets)
- `output/dynamic_eQTL_real/classify_dyn_eQTLs_{early,middle,late}.RData`
  (`testing_functional()` results reproducible from `code/02_dyn_lfsr.R`)
- `output/revision_simulations/internal/variant_annotation_enrichment/`
  (annotation matrix and matching covariates)
