---
title: "Internal analysis: tested variants per gene by FASH discovery status"
author: "Ziang Zhang"
date: "2026-08-19"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 8.8,
  fig.height = 6.8,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(file.path(
  "code",
  "revision_simulations",
  "internal",
  "fash_discovery_variant_count",
  "reporting.R"
))
```

<style>
.variant-count-internal-banner {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border: 1px solid #d9a441;
  border-left: 6px solid #d9a441;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.variant-count-definition {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.variant-count-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.variant-count-table-scroll table {
  margin-bottom: 1em;
}
.variant-count-path {
  white-space: normal;
  overflow-wrap: anywhere;
  word-break: break-word;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
pre,
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
}
</style>

<div class="variant-count-internal-banner">
<strong>Internal analysis page.</strong> This page is not linked from the
workflowr homepage, manuscript, or response letter.
</div>

## Analysis definition

<div class="variant-count-definition">
<ul>
<li><strong>Tested universe:</strong> `r format_integer(expected_pair_count)`
gene-variant pairs from `r format_integer(expected_gene_count)` genes.</li>
<li><strong>FASH result:</strong> BF-adjusted FASH-IWP1 with cumulative-lfdr
FDR `r format_decimal(analysis_alpha, 2L)`.</li>
<li><strong>FASH-discovered gene:</strong> a gene with at least one called
gene-variant pair.</li>
<li><strong>Not-discovered gene:</strong> a tested gene with no called
gene-variant pair.</li>
<li><strong>Gene-level value:</strong> the number of tested variants associated
with the gene in the full FASH analysis.</li>
</ul>
</div>

The fixed FASH result contains
`r format_integer(length(discovered_pair_indices))` called pairs from
`r format_integer(length(discovered_gene_ids))` genes. The two comparison
groups contain `r format_integer(group_summary$Genes[2])` FASH-discovered genes
and `r format_integer(group_summary$Genes[1])` not-discovered genes.

## Distribution

Both panels apply `log(1 + x)` to each gene-level variant count. The upper
panel shows boxplots and reports medians on the original count scale. The lower
panel overlays the two histograms after normalizing each group to density.

```{r variant-count-boxplot, fig.height=10.5, fig.cap="Boxplot and overlapping density-normalized histogram of log(1 + number of tested variants per gene) by BF-adjusted FASH-IWP1 discovery status."}
plot_variant_count_distribution_panel()
```

## Group summary

All values in the table below are computed on the original variant-count
scale.

```{r group-summary-table, results='asis'}
render_internal_table(
  group_summary_display,
  align = "lrrrrr",
  minimum_width = "780px"
)
```

## One-sided non-parametric comparison

The analysis uses a one-sided Wilcoxon rank-sum test on the original variant
counts, with the FASH-discovered group supplied as the first sample and the
not-discovered group supplied as the second sample.

Test settings:

- `alternative = "greater"`
- `exact = FALSE`
- `correct = TRUE`

```{r wilcoxon-test-table, results='asis'}
render_internal_table(
  wilcoxon_test_table,
  align = "lllllrr",
  minimum_width = "1420px"
)
```

## Input provenance

```{r input-provenance-table, results='asis'}
render_internal_table(
  input_provenance_table,
  align = "llrll",
  minimum_width = "1320px"
)
```

Page-specific source files:

- <span class="variant-count-path">`code/revision_simulations/internal/fash_discovery_variant_count/reporting.R`</span>
- <span class="variant-count-path">`code/revision_simulations/internal/fash_discovery_variant_count/test_reporting.R`</span>

Reproduction commands from the workflowr project root:

```{r reproduction-commands, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/fash_discovery_variant_count/test_reporting.R

Rscript --vanilla -e \
  'workflowr::wflow_build("analysis/revision_internal_fash_discovery_variant_count.rmd")'
```
