---
title: "Internal analysis: functional annotation enrichment of FASH variants"
author: "Ziang Zhang"
date: "2026-08-07"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9.5,
  fig.height = 7.2,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(
  "code/revision_simulations/internal/variant_annotation_enrichment/reporting.R"
)
```

<style>
.variant-enrichment-internal-banner {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border: 1px solid #d9a441;
  border-left: 6px solid #d9a441;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.variant-enrichment-key-result {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.variant-enrichment-caution {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #cc7a00;
  border-radius: 4px;
  background: #fff8ec;
  color: #563a0c;
}
.variant-enrichment-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.variant-enrichment-table-scroll table {
  margin-bottom: 1em;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
pre {
  max-width: 100%;
  overflow-x: auto;
}
</style>

<div class="variant-enrichment-internal-banner">
<strong>Internal exploratory page.</strong> This page is intentionally absent
from the workflowr homepage, navigation, manuscript, and response letter. It
reads completed retained caches and does not modify any existing FASH result.
</div>

## Question and provisional answer

This analysis asks two related questions:

1. Are **all BF-adjusted FASH dynamic-eQTL variant discoveries** enriched in
   regulatory annotations relative to comparable tested variants?
2. After removing variants that are identical to, or tagged by, a Strober
   discovery, do the remaining **FASH-specific discoveries** retain evidence
   of enrichment?

The first question is the more stable primary analysis. The second is more
diagnostic: method-specific labels can change when nearby variants tag the
same association signal.

<div class="variant-enrichment-key-result">
<strong>Provisional readout.</strong>
`r primary_signal_sentence`
`r specific_signal_sentence`
`r specific_displayed_signal_sentence`
`r comparison_signal_sentence`
The exact-ID definition leaves
`r format_integer(lookup_summary("Exact FASH-specific variants"))` FASH-specific
variants; the within-study LD-aware definition leaves
`r format_integer(lookup_summary("LD-aware FASH-specific variants"))`.
</div>

The result is evidence about **functional plausibility**, not proof that an
individual variant is causal and not an independent validation of FASH's FDR.

## Variant sets

The unit throughout is a unique rsID. FASH calls use cumulative lfdr FDR at
0.05 after the paper's BF update. Strober linear and nonlinear calls use their
reported eFDR at 0.05. All controls come from variants tested by all three
methods and present in the study VCF.

```{r discovery-summary, results='asis'}
render_scrollable_table(
  discovery_summary,
  align = "lr",
  minimum_width = "620px"
)
```

The primary sets contain `r format_integer(lookup_summary("FASH unique variants"))`
FASH variants, `r format_integer(lookup_summary("Strober linear unique variants"))`
Strober-linear variants, and
`r format_integer(lookup_summary("Strober nonlinear unique variants"))`
Strober-nonlinear variants. The union of the two Strober sets has
`r format_integer(lookup_summary("Strober union unique variants"))` variants.

## Primary enrichment result

Each discovered variant is matched to five non-discovered tested variants on
chromosome, cohort MAF, minimum distance to a tested target-gene TSS, local
tested-variant density within +/-1 Mb, and the number of tested genes. Matching
is repeated for 100 fixed seeds. Points pool the matched controls across seeds;
intervals are delete-one-autosome jackknife 95% intervals. Filled points have
BH q below 0.05 within the displayed analysis panel.
Categories with fewer than 10 overlaps in either the discovery set or pooled
controls are retained as missing estimates rather than extrapolated from very
sparse counts.

```{r primary-enrichment, fig.width=9.6, fig.height=8.6, fig.cap="Figure 1. Functional-annotation enrichment for FASH, Strober linear, and Strober nonlinear variant discoveries at their reported 0.05 thresholds. Enrichment is relative to matched variants from the common tested universe. Error bars are delete-one-chromosome 95% intervals."}
plot_primary_enrichment()
```

The plot includes gene architecture from
[GENCODE v19](https://www.gencodegenes.org/human/release_19.html), the
cell-type-agnostic hg19 ENCODE cCRE registry
[ENCFF788SJC](https://www.encodeproject.org/files/ENCFF788SJC/), and Roadmap
15-state ChromHMM annotations for iPS-20b (E020), hESC-derived CD56+ mesoderm
(E013), and left ventricle (E095). These tissues provide an iPSC-centered
comparison and two differentiation-relevant references; they are not exact
matches to every sampled time point.

## Are FASH-only discoveries enriched?

An rsID-only comparison classifies
`r format_integer(lookup_summary("Exact FASH-specific variants"))` variants as
FASH-specific. This can overstate specificity when FASH and Strober select
different variants in the same LD block. The sensitivity analysis therefore
also removes a FASH variant when a Strober-union variant lies within 1 Mb and
has dosage r<sup>2</sup> &ge; 0.8 in the 19 study donors.

```{r specific-enrichment, fig.width=9.6, fig.height=8.6, fig.cap="Figure 2. Functional-annotation enrichment for exact-ID-specific and LD-aware-specific FASH and Strober variants. The LD-aware definition uses dosage correlation in the 19 study donors and is therefore an approximate sensitivity analysis."}
plot_specific_enrichment()
```

<div class="variant-enrichment-caution">
<strong>Interpretation boundary.</strong> With only 19 donors, sample dosage
correlations are noisy. The LD-aware panel is useful for checking whether an
exact-ID result is obviously driven by neighboring tags, but it is not a
high-resolution fine-mapping analysis. A reference-panel LD analysis would be
a sensible follow-up if this result becomes manuscript-facing.
</div>

The table below gives the FASH estimates for intronic overlap and the seven
core regulatory categories. Intronic overlap is included because it is the
strongest positive FASH-specific result, although it is not equivalent to
active regulatory annotation.

```{r key-result-table, results='asis'}
render_scrollable_table(
  key_result_table,
  digits = 4L,
  align = "rlrrrrr",
  minimum_width = "1050px"
)
```

## Unequal set size and repeated-locus sensitivities

Two additional analyses probe design choices that can make one method appear
more enriched than another:

- **Equal top-K:** retain the top
  `r format_integer(lookup_summary("Equal top-K"))` unique variants for each
  method, where K is the size of the smallest primary discovery set.
- **One lead variant per discovered gene:** retain the lowest-lfdr FASH variant
  or lowest-p-value Strober variant per discovered gene.

```{r design-sensitivities, fig.width=9.6, fig.height=8.4, fig.cap="Figure 3. Equal-top-K and one-lead-variant-per-gene sensitivity analyses for core regulatory annotations. These panels test whether unequal discovery counts or repeated nearby discoveries dominate the primary comparison."}
plot_design_sensitivities()
```

These are sensitivity estimands, not replacements for the reported-threshold
analysis. In particular, top-K changes the significance criterion, while one
lead variant per gene changes both locus weighting and the number of variants.

## Matching diagnostics

The standardized mean differences below compare each discovery set with its
matched controls for every seed. Dotted lines mark +/-0.1 as a visual reference,
not a formal acceptance threshold.

```{r matching-balance-plot, fig.width=10.2, fig.height=7.2, fig.cap="Figure 4. Matching balance over 100 deterministic seeds. Boxes summarize standardized mean differences between discovered variants and their matched controls."}
plot_matching_balance()
```

<details>
<summary><strong>Maximum absolute standardized differences</strong></summary>

```{r matching-balance-table, results='asis'}
render_scrollable_table(
  matching_balance_summary,
  digits = 4L,
  align = "lllr",
  minimum_width = "880px"
)
```

</details>

Matching starts with the full coarsened stratum. If no control is available,
it drops local-density bin, then target-distance bin, and finally chooses from
the nearest standardized covariate profiles within the same chromosome and
MAF bin. Chromosome and MAF bin are never relaxed.

<details>
<summary><strong>Matching relaxation audit</strong></summary>

```{r relaxation-table, results='asis'}
render_scrollable_table(
  relaxation_summary,
  digits = 4L,
  align = "llrrr",
  minimum_width = "820px"
)
```

</details>

## Annotation provenance

All coordinates use GRCh37/hg19; no lift-over is performed. The downloaded
files and observed checksums are retained in the analysis cache.

```{r source-table, results='asis'}
render_scrollable_table(
  source_table,
  digits = 4L,
  align = "llllrr",
  minimum_width = "1180px"
)
```

The Roadmap state labels follow the official
[15-state core-mark model](https://egg2.wustl.edu/roadmap/web_portal/chr_state_learning.html)
and are collapsed into active TSS, enhancer, transcribed, bivalent/poised,
Polycomb-repressed, heterochromatin/repeats, and quiescent categories. The full
set of collapsed states is retained in the cache even though the main figures
focus on regulatory categories.

## Limitations and next decision

- Functional overlap is not causal fine-mapping; linked variants can share an
  annotation signal.
- The control design addresses measured chromosome/MAF/distance/density/gene-
  multiplicity differences, but unmeasured discovery-process differences can
  remain.
- ENCODE cCREs are cell-type agnostic, and the three Roadmap epigenomes are
  biologically motivated references rather than exact time-resolved matches.
- Panel-level q-values are descriptive because annotation categories overlap
  and the discovery sets were generated from the same tested universe.

In this run, all-FASH and LD-aware FASH-specific variants agree in showing no
clear positive enrichment in the core promoter/cCRE/iPSC regulatory
annotations. The mild intronic result is too indirect to support a claim that
FASH uniquely recovers more functional variants, especially because the
Strober nonlinear sets show the clearer enhancer pattern. The current evidence
therefore argues for keeping this analysis internal. If the question remains
important, the next useful step is a deliberately chosen time-resolved
iPSC/cardiomyocyte regulatory atlas plus reference-panel LD, not promotion of
the present result to the manuscript.

## Reproduction

The page reads only retained caches. Recreate them from the fixed FASH fit,
Strober tables, study VCF, and versioned annotations with:

```{r reproduction-command, eval=FALSE}
Rscript --vanilla code/revision_simulations/internal/variant_annotation_enrichment/run_variant_annotation_enrichment.R
```

The cache was generated at `r configuration$generated_at`. The full command,
input checksums, session information, per-seed balance summaries, enrichment
tables, and LD classifications are retained under
`output/revision_simulations/internal/variant_annotation_enrichment/`.
