---
title: "Internal sensitivity analysis: all four functional categories"
author: "Ziang Zhang"
date: "2026-08-20"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 8.0,
  fig.height = 7.2,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)

cache_dir <- file.path(
  "output",
  "revision_simulations",
  "internal",
  "evaluation_grid_mc_all_category_baseline_discoveries_open_middle"
)

configuration <- readRDS(file.path(cache_dir, "configuration.rds"))
population <- read.csv(
  file.path(cache_dir, "baseline_discovery_population.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
population_summary <- read.csv(
  file.path(cache_dir, "population_summary.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
pair_lfsr <- read.csv(
  file.path(cache_dir, "pair_lfsr_by_setting.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
comparison_summary <- read.csv(
  file.path(cache_dir, "comparison_summary.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
runtime <- read.csv(
  file.path(cache_dir, "runtime.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
cache_validation <- read.csv(
  file.path(cache_dir, "validation.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

category_order <- c("early", "middle", "late", "switch")
category_labels <- c(
  early = "Early",
  middle = "Middle",
  late = "Late",
  switch = "Switch"
)
expected_pair_counts <- c(
  early = 124L,
  middle = 60L,
  late = 20L,
  switch = 984L
)
expected_gene_counts <- c(
  early = 8L,
  middle = 15L,
  late = 12L,
  switch = 250L
)
expected_setting_ids <- c(
  "grid_0p15_M3000",
  "grid_0p10_M3000",
  "grid_0p05_M3000",
  "grid_0p10_M2000",
  "grid_0p10_M5000"
)
expected_comparison_ids <- c(
  "grid_0p10_vs_0p15_M3000",
  "grid_0p10_vs_0p05_M3000",
  "M2000_vs_M3000_grid_0p10",
  "M3000_vs_M5000_grid_0p10"
)

if (
  configuration$seed != 20260820L ||
  configuration$num_cores != 4L ||
  configuration$alpha != 0.05 ||
  configuration$switch_threshold != 0.25 ||
  configuration$category_pair_count != sum(expected_pair_counts) ||
  configuration$unique_pair_count != 1178L ||
  !identical(configuration$category_order, category_order) ||
  !identical(
    unname(configuration$expected_pair_counts),
    unname(expected_pair_counts)
  ) ||
  !identical(
    unname(configuration$expected_gene_counts),
    unname(expected_gene_counts)
  ) ||
  !setequal(pair_lfsr$setting_id, expected_setting_ids) ||
  !identical(
    unique(comparison_summary$comparison_id),
    expected_comparison_ids
  ) ||
  nrow(population) != sum(expected_pair_counts) ||
  nrow(pair_lfsr) != sum(expected_pair_counts) *
    length(expected_setting_ids) ||
  nrow(comparison_summary) != length(category_order) *
    length(expected_comparison_ids) ||
  anyDuplicated(population[c("category", "pair_id")]) ||
  anyDuplicated(pair_lfsr[c("category", "pair_id", "setting_id")]) ||
  any(!is.finite(pair_lfsr$lfsr)) ||
  any(pair_lfsr$lfsr < 0 | pair_lfsr$lfsr > 1) ||
  !all(cache_validation$passed)
) {
  stop("The all-category numerical-sensitivity cache failed page validation.")
}

population_summary <- population_summary[
  match(category_order, population_summary$category),
  ,
  drop = FALSE
]
comparison_summary <- comparison_summary[
  order(
    match(comparison_summary$comparison_id, expected_comparison_ids),
    match(comparison_summary$category, category_order)
  ),
  ,
  drop = FALSE
]
```

<style type="text/css">
.sensitivity-key-result {
  border-left: 5px solid #2c7fb8;
  background: #eef6fb;
  padding: 0.9em 1.1em;
  margin: 1em 0 1.4em 0;
}

table > caption,
p.caption {
  color: #1f2937 !important;
  font-size: 1.10em !important;
  font-weight: 700 !important;
  line-height: 1.35;
  text-align: left !important;
  margin-top: 0.7em;
  margin-bottom: 0.7em;
}

table > caption {
  caption-side: top;
}

.table-scroll {
  width: 100%;
  overflow-x: auto;
  margin-bottom: 1em;
}

.table-scroll table {
  min-width: 1040px;
}
</style>

# Question and conclusion

This page asks whether the current numerical choices---evaluation-grid step
0.10 and 3,000 posterior draws---are sufficiently stable for the pairs already
selected in the four illustrative functional categories: Early, Middle, Late,
and Switch.

<div class="sensitivity-key-result">
The answer is <strong>yes for the conditional robustness of the reported
pairs</strong>. Across categories, grid comparisons have mean absolute LFSR
changes of 0.0003--0.0017. The 3,000-versus-5,000-draw comparison has mean
absolute changes of 0.0014--0.0021, smaller in every category than the
corresponding 2,000-versus-3,000 change. These results support retaining
<strong>grid step 0.10 and 3,000 posterior draws</strong> for the illustrative
functional analysis.
</div>

This is a **reported-pair sensitivity analysis**: the baseline discovery
population is held fixed. It does not rerank all 9,205 candidate pairs under
each numerical setting and therefore does not claim identical full discovery
sets. A full-universe server run would be the appropriate follow-up only if
this targeted diagnostic showed material instability.

# Estimands and fixed analysis population

For posterior trajectory draw $b^{(m)}(t)$, the four statistics are

$$
\begin{aligned}
T_{\mathrm{early}}^{(m)}
&= \sup_{t\leq3}|b^{(m)}(t)|-\sup_{t>3}|b^{(m)}(t)|,\\
T_{\mathrm{middle}}^{(m)}
&= \sup_{3<t<12}|b^{(m)}(t)|
   -\sup_{t\leq3\;\mathrm{or}\;t\geq12}|b^{(m)}(t)|,\\
T_{\mathrm{late}}^{(m)}
&= \sup_{t\geq12}|b^{(m)}(t)|-\sup_{t<12}|b^{(m)}(t)|,\\
T_{\mathrm{switch}}^{(m)}
&= \min\!\left\{\sup_t b^{(m)}(t),-\inf_t b^{(m)}(t)\right\}-0.25.
\end{aligned}
$$

These are continuous-time functionals. Each numerical setting approximates
the suprema and infima by extrema on its stated evaluation grid.

For each category $c$, the estimated local false sign rate is

$$
\widehat{\mathrm{LFSR}}_c
=\frac{1}{M}\sum_{m=1}^{M}I\!\left\{T_c^{(m)}\leq0\right\}.
$$

Early, Late, and Switch pairs are the saved manuscript discoveries with
cumulative FSR at most 0.05. Middle uses the less boundary-ambiguous open
window $3<t<12$ and its 60 baseline discoveries at grid 0.10 and $M=3{,}000$.
The same category-specific pairs are retained in every comparison.

```{r analysis-population-table}
# Display the exact fixed population used in every sensitivity comparison.
population_table <- data.frame(
  Category = unname(category_labels[population_summary$category]),
  Definition = c(
    "Largest absolute effect at t <= 3",
    "Largest absolute effect at 3 < t < 12",
    "Largest absolute effect at t >= 12",
    "Positive and negative excursions both exceed 0.25"
  ),
  `Baseline pairs` = population_summary$n_pairs,
  `Baseline genes` = population_summary$n_genes,
  check.names = FALSE
)

knitr::kable(
  population_table,
  align = c("l", "l", "r", "r"),
  caption = paste(
    "Table 1. Fixed baseline-discovery population for the four-category",
    "conditional sensitivity analysis."
  )
)
```

# Numerical construction

Grid sensitivity is evaluated at steps 0.15, 0.10, and 0.05 with the same
3,000 posterior draws. The 0.15 and 0.10 grids are exact row subsets of the
0.05 grid.

For Monte Carlo sensitivity, `predict.fash()` returns draws grouped by sampled
prior PSD component. Therefore, taking its unmodified first columns would not
produce a representative posterior prefix. For each fitted pair, we generate
5,000 draws once, apply a fixed pair-specific random column permutation, and
use the first 2,000, 3,000, and 5,000 permuted columns. This construction makes
the Monte Carlo comparisons nested while holding the posterior sample stream
fixed. Ten pairs occur in more than one category, so 1,188 category-pair rows
require 1,178 unique posterior-sampling calls.

```{r numerical-design-code, eval=FALSE}
# Reproducible construction used by the cache-building script for each pair.
set.seed(20260820L + pair_index)
posterior_samples <- predict(
  fash_fit1_update,
  index = pair_index,
  smooth_var = seq(0, 15, by = 0.05),
  only.samples = TRUE,
  M = 5000L
)

# predict.fash groups columns by PSD component, so randomize column order
# deterministically before taking nested Monte Carlo prefixes.
set.seed(20260820L + 1000000L + pair_index)
posterior_samples <- posterior_samples[
  ,
  sample.int(ncol(posterior_samples)),
  drop = FALSE
]

samples_M2000 <- posterior_samples[, seq_len(2000L), drop = FALSE]
samples_M3000 <- posterior_samples[, seq_len(3000L), drop = FALSE]
samples_M5000 <- posterior_samples[, seq_len(5000L), drop = FALSE]
```

```{r comparison-summary-table, results='asis'}
# Report continuous agreement before the auxiliary individual-LFSR cutoff.
summary_table <- data.frame(
  Comparison = comparison_summary$comparison,
  Category = unname(category_labels[comparison_summary$category]),
  Pairs = comparison_summary$n_pairs,
  `Mean |change|` = sprintf(
    "%.4f",
    comparison_summary$mean_absolute_change
  ),
  `90th percentile |change|` = sprintf(
    "%.4f",
    comparison_summary$q90_absolute_change
  ),
  `Maximum |change|` = sprintf(
    "%.4f",
    comparison_summary$maximum_absolute_change
  ),
  Spearman = sprintf("%.3f", comparison_summary$spearman_correlation),
  `Within 0.01` = paste0(
    sprintf("%.1f", 100 * comparison_summary$fraction_within_0p01),
    "%"
  ),
  `LFSR-0.05 agreement` = paste0(
    sprintf(
      "%.1f",
      100 * comparison_summary$lfsr_0p05_classification_agreement
    ),
    "%"
  ),
  check.names = FALSE
)

summary_table_html <- knitr::kable(
  summary_table,
  format = "html",
  align = c("l", "l", "r", rep("r", 6)),
  caption = paste(
    "Table 2. Paired LFSR agreement for every baseline-discovery pair.",
    "The last column is an auxiliary individual-LFSR diagnostic, not a",
    "recomputed cumulative-FSR discovery classification."
  )
)
knitr::asis_output(paste0(
  '<div class="table-scroll">',
  summary_table_html,
  '</div>'
))
```

The continuous metrics are more informative than isolated 0.05 crossings.
For example, Late has only 20 pairs, several near 0.05, so one or two small
changes can noticeably reduce the last-column percentage even when every LFSR
change is below 0.01.

# Four paired LFSR comparisons

The plotting code below explicitly joins the same category-pair rows between
the requested horizontal- and vertical-axis settings. Every panel uses the
same 0--0.13 scale, which contains all plotted LFSRs. The solid diagonal is
equality, dotted diagonals mark an absolute difference of 0.01, and orange
points cross the auxiliary individual-LFSR value 0.05.

```{r scatter-construction}
library(ggplot2)

build_comparison_data <- function(comparison_id) {
  comparison_spec <- configuration$comparisons[
    configuration$comparisons$comparison_id == comparison_id,
    ,
    drop = FALSE
  ]
  if (nrow(comparison_spec) != 1L) {
    stop("The requested comparison is not uniquely specified.")
  }

  x_rows <- pair_lfsr[
    pair_lfsr$setting_id == comparison_spec$x_setting_id,
    c("category", "pair_id", "gene_symbol", "lfsr"),
    drop = FALSE
  ]
  y_rows <- pair_lfsr[
    pair_lfsr$setting_id == comparison_spec$y_setting_id,
    c("category", "pair_id", "lfsr"),
    drop = FALSE
  ]
  names(x_rows)[names(x_rows) == "lfsr"] <- "x_lfsr"
  names(y_rows)[names(y_rows) == "lfsr"] <- "y_lfsr"

  plot_data <- merge(
    x_rows,
    y_rows,
    by = c("category", "pair_id"),
    all = FALSE,
    sort = FALSE
  )
  expected_rows <- sum(configuration$expected_pair_counts)
  if (nrow(plot_data) != expected_rows ||
      anyDuplicated(plot_data[c("category", "pair_id")])) {
    stop("The paired plotting population is incomplete.")
  }

  plot_data$category_panel <- factor(
    plot_data$category,
    levels = category_order,
    labels = unname(category_labels[category_order])
  )
  plot_data$cutoff_status <- ifelse(
    (plot_data$x_lfsr <= 0.05) == (plot_data$y_lfsr <= 0.05),
    "Same side of 0.05",
    "Crosses 0.05"
  )
  plot_data
}

build_comparison_plot <- function(comparison_id, x_label, y_label) {
  plot_data <- build_comparison_data(comparison_id)

  ggplot(plot_data, aes(x = x_lfsr, y = y_lfsr)) +
    geom_abline(
      intercept = c(-0.01, 0.01),
      slope = 1,
      color = "grey65",
      linewidth = 0.45,
      linetype = "dotted"
    ) +
    geom_abline(
      intercept = 0,
      slope = 1,
      color = "grey20",
      linewidth = 0.55
    ) +
    geom_vline(
      xintercept = 0.05,
      color = "grey72",
      linewidth = 0.4,
      linetype = "dashed"
    ) +
    geom_hline(
      yintercept = 0.05,
      color = "grey72",
      linewidth = 0.4,
      linetype = "dashed"
    ) +
    geom_point(
      aes(color = cutoff_status),
      alpha = 0.62,
      size = 1.35
    ) +
    facet_wrap(~category_panel, ncol = 2) +
    scale_color_manual(
      values = c(
        "Same side of 0.05" = "#2c7fb8",
        "Crosses 0.05" = "#d95f02"
      )
    ) +
    scale_x_continuous(
      limits = c(0, 0.13),
      breaks = c(0, 0.025, 0.05, 0.075, 0.10, 0.125),
      labels = scales::label_number(accuracy = 0.001),
      expand = expansion(mult = c(0.01, 0.02))
    ) +
    scale_y_continuous(
      limits = c(0, 0.13),
      breaks = c(0, 0.025, 0.05, 0.075, 0.10, 0.125),
      labels = scales::label_number(accuracy = 0.001),
      expand = expansion(mult = c(0.01, 0.02))
    ) +
    coord_equal() +
    labs(
      x = x_label,
      y = y_label,
      color = NULL
    ) +
    theme_bw(base_size = 11.5) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey92", linewidth = 0.35),
      strip.background = element_rect(fill = "grey94", color = "grey70"),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom",
      legend.margin = margin(t = -2),
      panel.spacing = grid::unit(0.8, "lines")
    )
}
```

```{r grid-0p10-vs-0p15, fig.cap="Figure 1. Grid-step sensitivity: 0.10 versus 0.15 at 3,000 posterior draws. Panels contain all baseline-discovery pairs for each category; the discovery population is held fixed."}
build_comparison_plot(
  comparison_id = "grid_0p10_vs_0p15_M3000",
  x_label = "LFSR at grid step 0.10",
  y_label = "LFSR at grid step 0.15"
)
```

The points remain close to equality: category-specific mean absolute changes
range from 0.0005 to 0.0017. The largest change is 0.0110 for one Switch pair;
99.9% of Switch pairs and all Early, Middle, and Late pairs are within 0.01.

```{r grid-0p10-vs-0p05, fig.cap="Figure 2. Grid-step sensitivity: 0.10 versus 0.05 at 3,000 posterior draws. Panels contain all baseline-discovery pairs for each category; the discovery population is held fixed."}
build_comparison_plot(
  comparison_id = "grid_0p10_vs_0p05_M3000",
  x_label = "LFSR at grid step 0.10",
  y_label = "LFSR at grid step 0.05"
)
```

The finer-grid comparison is at least as stable. Mean absolute changes range
from 0.0003 to 0.0017, and every pair is within 0.01. This supports treating
grid step 0.10 as adequate for these reported pairs.

```{r M2000-vs-M3000, fig.cap="Figure 3. Posterior Monte Carlo sensitivity: 2,000 versus 3,000 nested draws at grid step 0.10. Panels contain all baseline-discovery pairs for each category; columns were deterministically permuted before taking prefixes."}
build_comparison_plot(
  comparison_id = "M2000_vs_M3000_grid_0p10",
  x_label = "LFSR with 2,000 posterior draws",
  y_label = "LFSR with 3,000 posterior draws"
)
```

Mean absolute changes range from 0.0021 to 0.0027. Signed mean changes are all
within 0.0003 of zero, consistent with Monte Carlo fluctuation rather than a
systematic shift.

```{r M3000-vs-M5000, fig.cap="Figure 4. Posterior Monte Carlo sensitivity: 3,000 versus 5,000 nested draws at grid step 0.10. Panels contain all baseline-discovery pairs for each category; columns were deterministically permuted before taking prefixes."}
build_comparison_plot(
  comparison_id = "M3000_vs_M5000_grid_0p10",
  x_label = "LFSR with 3,000 posterior draws",
  y_label = "LFSR with 5,000 posterior draws"
)
```

Mean absolute changes decrease to 0.0014--0.0021, with at least 99.2% of pairs
within 0.01 in every category. The smaller changes after 3,000 draws provide
the direct saturation check requested here and support retaining $M=3{,}000$.

# Final numerical choices

- Functional definitions: Early $t\leq3$; Middle $3<t<12$; Late $t\geq12$;
  Switch threshold 0.25.
- Evaluation grid: `seq(0, 15, by = 0.10)`.
- Posterior Monte Carlo size: 3,000 draws.
- Sensitivity population: all baseline-discovery pairs within each category,
  held fixed across settings.
- Randomization: pair-specific seed 20260820 plus fitted-pair index, followed
  by a pair-specific deterministic column permutation for nested prefixes.
- Computation: one R job with four workers; the production posterior-sampling
  stage took `r sprintf("%.1f", runtime$elapsed_seconds[runtime$stage == "posterior_sampling_and_evaluation"])`
  seconds.
