---
title: "Revision Real-data Analysis R6: enhancer enrichment of FASH dynamic eQTLs"
author: "Ziang Zhang"
date: "2026-08-21"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 2
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 11,
  fig.height = 8.4,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(file.path(
  "code",
  "revision_simulations",
  "internal",
  "fash_strober_enhancer_comparison",
  "reporting.R"
))
```

<style>
html,
body {
  max-width: 100%;
  overflow-x: hidden;
}
.fash-strober-key-result {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.fash-strober-caution {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #cc7a00;
  border-radius: 4px;
  background: #fff8ec;
  color: #563a0c;
}
.fash-strober-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.fash-strober-table-scroll table {
  margin-bottom: 1em;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
pre {
  max-width: 100%;
  overflow-x: auto;
}
</style>

## The question

Do the gene-variant pairs that FASH discovers, and Strober's linear and
quadratic/nonlinear analyses do not, carry independent regulatory evidence?

| Section | What is being tested |
|---|---|
| 1 | FASH versus Strober linear versus Strober quadratic/nonlinear |
| 2 | FASH discoveries at variants Strober never discovered |

Both figures show the same two selection strategies side by side: **all
discovered variants** and **one lead variant per gene**. FASH uses cumulative
lfdr at FDR 0.05 after the BF correction; Strober uses its reported eFDR 0.05.
The existing Strober `non_linear_dynamic_eqtls_5_pc.txt` result is labelled
quadratic/nonlinear throughout.

**FASH-only, variant level:** an rsID is not FASH-only if it appears in *any*
Strober linear or quadratic discovery, regardless of which gene it was paired
with. This is the conservative reading of "new evidence", because an enhancer
annotation is a property of a variant rather than of a pair. For the lead
column, the per-gene lowest-lfdr variant is chosen **first** and the exclusion
applied afterwards, so every variant shown is genuinely FASH's top call for its
gene; a gene whose best variant Strober also found drops out rather than falling
back to a weaker one.

## Discovery-set construction

```{r discovery-table, results='asis'}
render_scrollable_table(
  discovery_table,
  caption = "Table 1. Discovery sets used in the enhancer-enrichment analysis.",
  align = "rlllrrr",
  minimum_width = "1120px"
)
```

Of the
`r format_integer(strict_exclusivity_audit$original_variant_count[1])`
variants FASH discovers,
`r format_integer(strict_exclusivity_audit$excluded_variant_count[1])`
were also discovered by Strober linear or quadratic/nonlinear and are removed,
leaving `r format_integer(strict_exclusivity_audit$fash_only_variant_count[1])`
that only FASH found.

<details>
<summary><strong>Annotation coverage for all eight sets</strong></summary>

```{r coverage-table, results='asis'}
render_scrollable_table(
  coverage_table,
  caption = "Table 2. Annotation coverage for all eight discovery sets.",
  align = "llllrrr",
  minimum_width = "1240px"
)
```

</details>

## Section 1: FASH versus Strober

```{r section1, fig.width=11.5, fig.height=8.8, fig.cap="Figure 1. Enhancer enrichment for current FASH, Strober linear, and Strober quadratic/nonlinear. Intervals are delete-one-autosome 95% intervals; filled points mark nominal p below 0.05 without multiplicity correction. Chevrons mark intervals extending beyond the plotted range; exact bounds are returned by compute_enrichment()."}
plot_method_comparison("Current FASH")
```

Under lead-per-gene selection the ordering reverses at the
E013 mesoderm enhancer, where FASH holds the strongest estimate of the three
methods: `r describe_estimate("Custom regulatory", "current_lead", e013)`,
against `r format_decimal(enhancer_lookup("Custom regulatory", "linear_lead", e013)$enrichment, 2L)`-fold
for Strober linear and
`r format_decimal(enhancer_lookup("Custom regulatory", "quadratic_lead", e013)$enrichment, 2L)`-fold
for Strober quadratic.

## Section 2: variants Strober never discovered

The grey series repeats the unfiltered FASH set from Section 1 so the effect of
the exclusion is directly visible.

```{r section2, fig.width=11.5, fig.height=8.8, fig.cap="Figure 2. Enhancer enrichment for FASH discoveries at variants absent from both Strober analyses, against the unfiltered FASH reference. Chevrons mark intervals extending beyond the plotted range; exact bounds are returned by compute_enrichment()."}
plot_fash_only()
```

**All variants** (`r set_size("current_only_all")` variants): every FASH
discovery at a variant Strober never found. Here
`r describe_significant("current_only_all")`.

**One lead variant per gene** (`r set_size("current_only_lead")` variants):
FASH's lowest-lfdr variant for each gene, keeping only those Strober never
found. A gene whose best variant Strober also discovered drops out rather than
falling back to a weaker one, so
`r format_integer(strict_exclusivity_audit$fash_only_lead_gene_count[1])` of
FASH's `r format_integer(lead_gene_count)` genes survive. Here the enriched
annotations are `r describe_significant("current_only_lead")`.

<div class="fash-strober-caution">
<strong>Exploratory, uncorrected.</strong> Fold enrichment, 95% intervals, and
nominal p-values are reported for
`r format_integer(n_enhancer_tests)` estimable enhancer estimates across
`r length(expected_sets)` variant sets and two annotation systems, with no
multiplicity adjustment; the p-values are descriptive. The two annotation
systems evaluate the same discoveries and are not independent replications.
Each set has its own matched controls and the sets differ in size and
membership, so a larger estimate for one method is not a formal test that it
outperforms another - in particular, Strober's higher all-variant enrichment
must not be read that way.
</div>

<div class="fash-strober-key-result">
<strong>Readout.</strong> FASH discoveries at variants Strober never found are
enriched for mesoderm enhancers when one lead variant per gene is taken, and
indistinguishable from the null when every discovered variant is counted. E013
is the hESC-derived CD56+ mesoderm enhancer set, the transitional state of this
iPSC-to-cardiomyocyte differentiation, between the E020 starting state and the
E095 terminal state. That observation was made after seeing the results and is
not a pre-specified hypothesis.
</div>

## Methods

Enough detail to reimplement the analysis from scratch. The
[How the enrichment is computed](#how-the-enrichment-is-computed) section below
shows the same thing as four lines of code.

### Annotation provenance

```{r provenance-table, results='asis'}
render_scrollable_table(
  provenance_table,
  caption = "Table 3. Provenance of the enhancer annotations.",
  align = "llllllll",
  minimum_width = "1560px",
  escape = FALSE
)
```

### How the custom annotation matrix was built

1. **Variant universe.** All variants tested in the study that lie on an
   autosome, have a finite cohort MAF, participate in at least one tested
   gene-variant pair, and have a computable local variant density:
   `r format_integer(745867)` variants. Everything downstream is defined on this
   set, so no external reference panel enters.
2. **Sources.** The five files in the table above, each checksummed on download.
3. **Interval construction.**
   - *GENCODE v19*: `gene` and `exon` records only. Gene body is the `gene`
     interval; exons are reduced to merge overlapping transcripts; **introns are
     `setdiff(gene body, exon)`**; promoters are TSS &plusmn; 2 kb; **intergenic
     is the complement of gene body**, not an interval set.
   - *ENCODE cCRE*: the class is decoded from the RGB field of the BED file, not
     from a name column - `255,0,0` promoter-like, `255,205,0` **enhancer-like**,
     `0,176,240` CTCF-only. An unrecognised RGB is a hard error.
   - *Roadmap*: the 15 ChromHMM states are collapsed to 7 categories. **Enhancer
     is states 6 and 7** (genic enhancers and enhancers). States 1-2 are Active
     TSS, 3-5 Transcribed, 8-9 Heterochromatin/repeats, 10-12 Bivalent/poised,
     13-14 Polycomb-repressed, 15 Quiescent. An unrecognised state is a hard
     error.
4. **Coordinate conversion.** BED files are 0-based half-open and GTF is 1-based
   closed, so BED starts are incremented by one before use.
5. **Overlap.** Each variant is treated as a single-base interval and
   intersected against every annotation with `GenomicRanges::findOverlaps`,
   giving a `r format_integer(745867)` &times; 29 logical matrix. Four of those
   29 columns are the pre-specified custom enhancers.

baselineLD v2.2 is used as published: its rows are the 1000G Phase 3 **EUR** SNP
list, which is why only
`r format_integer(coverage_by_set$covered_variant_count[coverage_by_set$annotation_system == "baselineLD v2.2" & coverage_by_set$discovery_set == "current_all"][1] * 0 + 557405)`
of the `r format_integer(745867)` tested variants have a row. The enhancer
*definitions* in both systems are ancestry-free genomic intervals; only
baselineLD's variant coverage is ancestry-dependent.

### Matched controls

For each discovery set, every selected variant is matched to
`r configuration$controls_per_variant` non-selected tested variants drawn from
the same stratum. Strata are the interaction of chromosome with decile bins of
cohort MAF, log10 minimum target-gene TSS distance, log1p local tested-variant
count within 1 Mb, and a coarsened tested-gene count (1, 2, 3-4, 5+). Matching
is repeated over `r configuration$matching_seed_count` fixed seeds beginning at
`r format_integer(min(configuration$matching_seeds))`, and control counts are
pooled across seeds. For the baselineLD system both the selected set and the
control pool are first restricted to baselineLD-covered variants, so numerator
and denominator stay on the same footing.

### Enrichment and interval

With $a$ selected variants overlapping an annotation out of $n_{\text{sel}}$,
and $b$ control overlaps out of $n_{\text{ctrl}}$ pooled across seeds:

$$\hat{\theta} = \log_2\!\left[\frac{a/n_{\text{sel}}}{b/n_{\text{ctrl}}}\right]$$

Uncertainty is a delete-one-chromosome block jackknife. For each autosome $c$,
recompute $\hat{\theta}_{(c)}$ with every variant on $c$ removed from both the
selected and the control counts, then

$$\mathrm{se} = \sqrt{\frac{m-1}{m}\sum_{c}\left(\hat{\theta}_{(c)} - \bar{\theta}\right)^2},
\qquad m = 22$$

The reported interval is $\hat{\theta} \pm 1.96\,\mathrm{se}$ on the log2 scale,
displayed as folds; the nominal p-value is $2\Phi(-|\hat{\theta}/\mathrm{se}|)$.
Chromosome blocks absorb within-chromosome correlation, which is why these
p-values are orders of magnitude more conservative than a Fisher test that
assumes independent variants. An estimate is reported only when both the
selected and the control overlap reach 10.

<div class="fash-strober-caution">
<strong>Known limitations of the interval.</strong> Only 22 jackknife blocks, so
the standard error is itself noisy and the normal quantile is optimistic in the
tails. Chromosomes differ greatly in size while the standard delete-one formula
assumes equal blocks. The Monte Carlo variability of the
`r configuration$matching_seed_count` matching seeds is not propagated; controls
are pooled and treated as fixed. The interval is symmetric in log2 and therefore
asymmetric in fold space.
</div>

## How the enrichment is computed

Stripped of bookkeeping, the analysis is three objects and one function. The
module [`enrichment_api.R`](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/internal/fash_strober_enhancer_comparison/enrichment_api.R)
is written to make that visible:

```{r api-demo, eval=FALSE}
source("code/revision_simulations/internal/fash_strober_enhancer_comparison/enrichment_api.R")

# 1. ANNOTATIONS - which variants belong to which group
annotations <- load_annotation_groups("custom")   # or "baselineld"

# 2. INPUT - a variant set to test, and the background of all tested variants
background <- load_background()                   # 745,867 tested variants
sets       <- load_variant_sets()                 # the eight published sets

# 3. OUTPUT - one enrichment score per set per annotation group
compute_enrichment(sets["current_only_lead"], background, annotations)
```

`compute_enrichment()` takes **any** named list of variant vectors, so swapping
in your own sets needs no other change:

```{r api-own-sets, eval=FALSE}
my_sets <- list(
  my_hits     = readLines("my_variants.txt"),
  fash_only   = sets$current_only_lead,
  strober_lin = sets$linear_lead
)
compute_enrichment(my_sets, background, annotations)
```

### The two control choices

The same function answers the question two ways, and the difference is the
whole reason the matching machinery exists:

```{r api-controls, eval=FALSE}
# Plain over-representation: rate in the set / rate in the background.
compute_enrichment(sets["current_all"], background, annotations,
                   controls = "background")

# Matched: each selected variant is compared against comparable variants drawn
# from the background, so the comparison is not confounded.
compute_enrichment(sets["current_all"], background, annotations,
                   controls = "matched")
```

Discovered variants are systematically higher-MAF and closer to a TSS than the
average tested variant, and enhancers are themselves TSS-proximal, so the plain
comparison attributes some purely positional signal to biology. For FASH's
all-variant set at the E095 enhancer the plain fold is
`r format_decimal(enhancer_lookup("Custom regulatory", "current_all", "Roadmap E095 left ventricle: Enhancer")$unmatched_enrichment, 2L)`,
while the matched fold is
`r format_decimal(enhancer_lookup("Custom regulatory", "current_all", "Roadmap E095 left ventricle: Enhancer")$enrichment, 2L)`
at nominal chromosome-jackknife p =
`r format(signif(enhancer_lookup("Custom regulatory", "current_all", "Roadmap E095 left ventricle: Enhancer")$p_value, 2L), scientific = FALSE)`.
Every headline number on this page uses `controls = "matched"`.

### Available annotation groups

The pre-specified enhancer panel is the default, but both systems carry more:

```{r api-groups, eval=FALSE}
list_annotation_groups("custom")      # 29 groups: GENCODE, ENCODE cCRE, Roadmap
list_annotation_groups("baselineld")  # 83 binary baselineLD v2.2 annotations

load_annotation_groups("custom", groups = "GENCODE promoter +/-2 kb")
```

### Checking the interface against this page

`enrichment_api.R` is a readable front end to the same audited matching engine
the analysis uses, so it can drift. This check recomputes all eight published
sets through `compute_enrichment()` and diffs them against the estimates shown
above:

```{bash api-verify, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/fash_strober_enhancer_comparison/enrichment_api.R \
  --verify
```

### Rebuilding the annotation matrix from source

```{bash rebuild-annotations, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/variant_annotation_enrichment/download_variant_annotations.R
Rscript --vanilla \
  code/revision_simulations/internal/variant_annotation_enrichment/run_variant_annotation_enrichment.R
```

The first downloads and checksums the five annotation sources; the second
rebuilds the variant universe and the 29-column annotation matrix from them.
Intermediate data stays under `output/`, which is not tracked in git - the
repository carries code, not data.

<details>
<summary><strong>Retained input provenance</strong></summary>

```{r input-provenance-table, results='asis'}
render_scrollable_table(
  input_provenance,
  caption = "Table 4. Provenance of every retained R6 analysis input.",
  align = "lrlll",
  minimum_width = "1360px"
)
```

</details>

## Reproducibility

Routine page builds read the retained fashr 0.1.43 cache at:

`output/revision_simulations/internal/fash_strober_enhancer_comparison_fashr0143/`

The earlier cache remains unchanged for historical comparison. The retained
configuration records the corrected BF-updated fit SHA-256, exact package
version and RemoteSha, matching seeds, control ratio, and input checksums.

```{bash reproduction-command, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/fash_strober_enhancer_comparison/test_fash_strober_enhancer_helpers.R

Rscript --vanilla \
  code/revision_simulations/internal/fash_strober_enhancer_comparison/run_fash_strober_enhancer_comparison.R

Rscript --vanilla \
  code/revision_simulations/internal/fash_strober_enhancer_comparison/enrichment_api.R \
  --verify

Rscript --vanilla -e \
  'workflowr::wflow_build("analysis/revision_enhancer_enrichment.rmd")'
```
