---
title: "Revision Real-data Sensitivity R4: temporal correlation"
author: "Ziang Zhang"
date: "2026-08-05"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 10,
  fig.height = 6.5,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source("code/revision_simulations/r4_correlated_errors/reporting.R")
```

<style>
.r4-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.r4-table-scroll table {
  margin-bottom: 1em;
}
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: hidden;
}
</style>

## Question

The FASH-IWP1 likelihood treats the 16 effect estimates within a trajectory as
conditionally independent given its true effect function. The real data reuse
the same donors at multiple times, so this assumption is not exact. This page
asks three increasingly targeted questions:

1. What full within-trajectory dependence patterns are visible among the 500
   most null-like gene-variant pairs?
2. Does FASH-IWP1 remain calibrated when the complete estimated 16 by 16
   patterns are inserted into the R1 data-generating mechanism?
3. How do power, empirical FDR, and the estimated dynamic-null proportion
   change as a local lag-1 correlation varies from `-0.3` to `0.3`?

Only FASH-IWP1 before and after BF adjustment is reported. The purpose is to
test whether the fitted method used in the main analysis is sensitive to
moderate violations of its conditional-independence working model.

## Real-data dependence pattern

### Null-like one-pair-per-gene subset

The BF-adjusted FASH fit contains more than one variant for many genes. To
avoid allowing a gene with many variants to dominate the diagnostic, we first
retain the variant with the **highest BF-adjusted lfdr within each gene** and
then take the 500 representatives with the highest lfdr overall. The selected
lfdr values range from
`r format_decimal(min(real_analysis$selected_units$lfdr))` to
`r format_decimal(max(real_analysis$selected_units$lfdr))`.

For FASH-IWP1, a high lfdr supports a constant effect trajectory, not
necessarily a zero effect. Two complementary full-matrix estimators are
therefore considered.

### Direct centered estimator

For unit $j$, let $\widehat\beta_j(t)$ and $s_j(t)$ be the effect estimate and
the t-adjusted standard error supplied to FASH. We estimate its unknown
constant effect by inverse-variance weighting,

$$
\widehat\mu_j =
\frac{\sum_t \widehat\beta_j(t)/s_j^2(t)}
{\sum_t 1/s_j^2(t)},
\qquad
z_j(t)=\frac{\widehat\beta_j(t)-\widehat\mu_j}{s_j(t)},
$$

and compute the empirical correlation matrix across the 500 rows of $z_j(t)$.
This is a descriptive correlation of **centered** residual trajectories.
Estimating and removing $\widehat\mu_j$ itself induces negative dependence, so
this matrix should not be interpreted as an uncentered error covariance.

### Pairwise-difference estimator

An alternative removes the unknown constant without explicitly estimating it.
For each time pair $(t,s)$, define

$$
a_j=s_j^2(t)+s_j^2(s),\qquad
c_j=\frac{2s_j(t)s_j(s)}{a_j},\qquad
q_j=\frac{\{\widehat\beta_j(t)-\widehat\beta_j(s)\}^2}{a_j}.
$$

Under a constant truth and conditionally Gaussian errors with correlation
$R_{ts}$, $E(1-q_j)=c_jR_{ts}$. We therefore use the through-origin
method-of-moments estimate

$$
\widehat R_{ts}=
\frac{\sum_j c_j(1-q_j)}{\sum_j c_j^2}.
$$

Both symmetric estimates are projected to the nearest positive-definite
correlation matrix before simulation.

```{r real-correlation-heatmaps, fig.width=10, fig.height=9.2, fig.cap="Figure R4.1: Real-data within-trajectory correlation diagnostics from 500 null-like one-pair-per-gene representatives. Panels distinguish the direct-centered and pairwise-difference estimators before and after nearest-positive-definite projection; color encodes correlation."}
plot_r4_real_data_correlation_heatmaps(real_analysis)
```

```{r projection-diagnostics-table, results='asis'}
render_scrollable_table(
  projection_table,
  caption = paste(
    "Table R4.1: Positive-definite projection diagnostics for the two",
    "real-data 16 by 16 correlation estimates."
  ),
  align = "lrrrrrr",
  minimum_width = "980px"
)
```

The direct matrix is already positive definite. The pairwise matrix has a
slightly negative minimum eigenvalue, but nearest-PD projection changes any
entry by at most
`r format_decimal(real_analysis$projection_diagnostics$maximum_absolute_change[real_analysis$projection_diagnostics$matrix_id == "pairwise_adjusted"], 6)`.
Thus the projection enforces a valid generating matrix without materially
changing its pattern.

### Why a matched independence benchmark is necessary

The pairwise moment equation is valid before selecting unusually flat
trajectories. Ranking by a dynamic-null lfdr is itself a selection on
within-trajectory variation. To expose this ascertainment effect, we generated
independent standard-normal errors with the observed SE patterns and compared:

- the fixed top-500 SE patterns without re-selection; and
- the 500 flattest trajectories selected from all 6,362 one-per-gene SE
  patterns.

```{r real-lag1-benchmark-table, results='asis'}
render_scrollable_table(
  benchmark_lag1_table,
  caption = paste(
    "Table R4.2: Observed lag-1 dependence and matched independent-error",
    "benchmarks for the direct-centered and pairwise-difference estimators."
  ),
  align = "lrrr",
  minimum_width = "1080px"
)
```

For the direct estimator, observed lag-1 correlation is
`r format_interval(
  get_real_lag_metric("Direct centered", "mean_correlation_observed"),
  get_real_lag_metric("Direct centered", "correlation_ci_lower"),
  get_real_lag_metric("Direct centered", "correlation_ci_upper")
)`, compared with
`r format_interval(
  get_benchmark_metric("Direct centered", "Fixed selected SE patterns", "mean_correlation"),
  get_benchmark_metric("Direct centered", "Fixed selected SE patterns", "correlation_ci_lower"),
  get_benchmark_metric("Direct centered", "Fixed selected SE patterns", "correlation_ci_upper")
)` under independent errors with the same centering and SE patterns. The
additional short-lag negative dependence is therefore modest.

The raw pairwise estimate is strongly positive,
`r format_interval(
  get_real_lag_metric("Pairwise difference", "mean_correlation_observed"),
  get_real_lag_metric("Pairwise difference", "correlation_ci_lower"),
  get_real_lag_metric("Pairwise difference", "correlation_ci_upper")
)`, but the flatness-selected independence benchmark is already
`r format_interval(
  get_benchmark_metric("Pairwise difference", "Flatness-selected gene representatives", "mean_correlation"),
  get_benchmark_metric("Pairwise difference", "Flatness-selected gene representatives", "correlation_ci_lower"),
  get_benchmark_metric("Pairwise difference", "Flatness-selected gene representatives", "correlation_ci_upper")
)`. Consequently, the pairwise matrix is retained as a deliberately
conservative full-matrix stress pattern, not presented as an unbiased estimate
of the unselected residual correlation.

The conclusion is not driven by the t-to-normal SE adjustment. Inverting that
adjustment gives the following lag-1 sensitivity:

```{r se-scale-sensitivity-table, results='asis'}
render_scrollable_table(
  se_sensitivity_table,
  caption = paste(
    "Table R4.3: Sensitivity of the real-data lag-1 diagnostics to the",
    "t-to-normal standard-error scaling."
  ),
  align = "llrr",
  minimum_width = "700px"
)
```

### Variograms with 95% intervals

For a unit-diagonal correlation matrix, the standardized semivariogram at lag
$h$ is

$$
\gamma(h)=1-\frac{1}{16-h}\sum_{t=1}^{16-h}R_{t,t+h}.
$$

The blue bands below are percentile intervals from 1,000 gene bootstraps. The
gray and orange bands are 1,000 matched independent-error replications.

```{r real-data-variograms, fig.width=11, fig.height=5.2, fig.cap="Figure R4.2: Lag-specific real-data correlation and semivariogram diagnostics. Blue curves and bands are observed estimates with 95% gene-bootstrap intervals; grey and orange curves and bands are the two matched independent-error benchmarks identified in the in-figure legend."}
plot_r4_real_data_variograms(
  real_analysis$bootstrap_lags,
  real_analysis$independence_benchmark
)
```

The direct centered diagnostic shows its largest departure at lag 1 and moves
toward its matched independence benchmark at longer lags, although the pattern
is not exactly lag-1-only. The pairwise diagnostic is dominated by a broad
positive common mode that is nearly reproduced by flatness selection under
independence. These observations motivate testing both complete matrices first;
the subsequent one-parameter sweep uses lag 1 as a parsimonious local
sensitivity direction rather than as a fitted time-series model.

## Paired R1 simulation with the complete matrices

### Data-generating mechanism

The full-matrix experiment reuses R1 without changing its scientific setting:
19 donors, 16 times, 1,000 variants, five PC-like covariates, and exact class
counts of 200 random B-spline dynamic effects, 400 constant effects, and 400
zero effects. Thus the true dynamic-null proportion is $\pi_0=0.8$.

At every time, `Y ~ G + PC1 + ... + PC5` is fitted and the genotype standard
error receives the same t-to-normal correction used in R1 and the real-data
analysis. Within each seed, the independent, direct-full, and pairwise-full
conditions share the exact same genotype, covariates, truth, nuisance effects,
and Gaussian innovations. Only the Cholesky transform of the expression errors
changes.

```{r full-matrix-r1-setting, eval=FALSE}
correlation_matrices <- readRDS(file.path(
  "output",
  "revision_simulations",
  "real_data",
  "r4_null_like_top500_full_correlations",
  "simulation_correlation_matrices.rds"
))

out <- run_genotype_level_bspline_eqtl_simulation(
  n_donors = 19,
  n_variants = 1000,
  time_grid = 0:15,
  n_covariates = 5,
  class_probs = c(
    dynamic_bspline = 0.20,
    constant = 0.40,
    zero = 0.40
  ),
  expression_noise_sd = 1,
  dynamic_main_effect_sd = 1,
  seed = seed,
  num_basis = 20,
  expression_error_correlation = correlation,
  save_outputs = FALSE
)
```

### Target-versus-realized dependence

```{r full-matrix-target-realized, fig.width=10, fig.height=9.2, fig.cap="Figure R4.3: Target and realized dependence in the paired R1 complete-matrix simulations. Panels distinguish independent, direct-centered, and pairwise-difference conditions and the relevant error scale; color encodes correlation."}
plot_r4_full_matrix_target_realized(
  full_configuration,
  full_matrix_summary
)
```

```{r full-matrix-lag1-table, results='asis'}
render_scrollable_table(
  full_lag1_table,
  caption = paste(
    "Table R4.4: Target and realized mean lag-1 correlations in the paired",
    "complete-matrix R1 simulations."
  ),
  align = "lrrrr",
  minimum_width = "1120px"
)
```

The realized expression-error and truth-known beta-hat error correlations track
both complete targets closely. Centering the dynamic-null beta-hat trajectories
again shifts the correlations, which is why the target matrix should not be
compared directly with a post-centering diagnostic without reproducing the same
operation.

### Power, empirical FDR, and estimated pi0

```{r full-matrix-metrics-figure, fig.width=12, fig.height=5.2, fig.cap="Figure R4.4: Power, empirical FDR, and estimated pi0 for paired R1 simulations using independent errors or either complete real-data-derived correlation matrix. Colors identify conditions and symbols identify raw versus BF-updated FASH in the in-figure legend; interval bars summarize five paired seeds."}
plot_r4_full_matrix_metrics(
  full_condition_alpha,
  full_condition_pi0,
  true_pi0 = 0.8,
  nominal_alpha = 0.05
)
```

```{r full-matrix-results-table, results='asis'}
render_scrollable_table(
  full_results_table,
  caption = paste(
    "Table R4.5: Monte Carlo performance and paired changes for raw and",
    "BF-updated IWP FASH under the complete correlation matrices."
  ),
  align = "llrrrrrr",
  minimum_width = "1450px"
)
```

For the primary BF-adjusted IWP1 fit, the direct full matrix changes power from
`r format_decimal(get_full_metric("Independent", "FASH-IWP1-BF", "mean_power"))`
to
`r format_decimal(get_full_metric("Direct centered full matrix", "FASH-IWP1-BF", "mean_power"))`
and empirical FDR from
`r format_decimal(get_full_metric("Independent", "FASH-IWP1-BF", "mean_fdr"))`
to
`r format_decimal(get_full_metric("Direct centered full matrix", "FASH-IWP1-BF", "mean_fdr"))`.
The paired changes are
`r format_signed_interval(
  get_full_metric("Direct centered full matrix", "FASH-IWP1-BF", "mean_delta_power"),
  get_full_metric("Direct centered full matrix", "FASH-IWP1-BF", "power_delta_ci_lower"),
  get_full_metric("Direct centered full matrix", "FASH-IWP1-BF", "power_delta_ci_upper")
)` for power and
`r format_signed_interval(
  get_full_metric("Direct centered full matrix", "FASH-IWP1-BF", "mean_delta_fdr"),
  get_full_metric("Direct centered full matrix", "FASH-IWP1-BF", "fdr_delta_ci_lower"),
  get_full_metric("Direct centered full matrix", "FASH-IWP1-BF", "fdr_delta_ci_upper")
)` for FDR. Estimated pi0 is
`r format_decimal(get_full_metric("Direct centered full matrix", "FASH-IWP1-BF", "mean_estimated_pi0"))`,
compared with
`r format_decimal(get_full_metric("Independent", "FASH-IWP1-BF", "mean_estimated_pi0"))`
under independence.

The selection-sensitive pairwise stress pattern is also conservative: its
BF-adjusted power is
`r format_decimal(get_full_metric("Pairwise-difference full matrix", "FASH-IWP1-BF", "mean_power"))`,
its mean realized FDP is
`r format_decimal(get_full_metric("Pairwise-difference full matrix", "FASH-IWP1-BF", "mean_fdr"))`,
and its estimated pi0 is
`r format_decimal(get_full_metric("Pairwise-difference full matrix", "FASH-IWP1-BF", "mean_estimated_pi0"))`.
Zero false discoveries in five replicates should be read as conservative
behavior in this focused experiment, not as proof that the population FDR is
exactly zero.

Thus neither complete matrix produces FDR inflation. The empirically relevant
direct pattern yields essentially unchanged power and pi0, while the much
stronger common-mode pairwise stress pattern is slightly less powerful and more
conservative.

## Sensitivity across lag-1 correlation strengths

The complete-matrix results address the observed structures directly. To make
the robustness boundary interpretable, we next vary only the first
off-diagonal correlation,

$$
R_{tt}=1,\qquad R_{t,t+1}=R_{t+1,t}=\rho,\qquad
R_{ts}=0\ \text{for }|t-s|>1,
$$

over $\rho\in\{-0.3,-0.2,\ldots,0.3\}$. This is a lag-1-only sensitivity model,
not AR(1). The same five seeds and common innovations are used at every rho.

```{r lag1-sweep-setting, eval=FALSE}
rho_grid <- seq(-0.3, 0.3, by = 0.1)
correlation_matrices <- lapply(
  rho_grid,
  function(rho) make_lag1_correlation(16, rho)
)
```

```{r lag1-sweep-power-fdr, fig.width=11, fig.height=5.2, fig.cap="Figure R4.5: Power and empirical FDR across the lag-1-only correlation sweep. Curves identify raw and BF-updated IWP FASH in the in-figure legend; the vertical reference marks the approximate real-data direct-centered estimate and the horizontal FDR reference marks 0.05."}
plot_r4_correlation_sweep(
  sweep_alpha_005,
  empirical_rho = -0.09,
  nominal_alpha = 0.05,
  methods = c("FASH-IWP1-Raw", "FASH-IWP1-BF")
)
```

```{r lag1-sweep-pi0, fig.width=10, fig.height=5.2, fig.cap="Figure R4.6: Estimated dynamic-null proportion across the lag-1-only correlation sweep. Curves identify raw and BF-updated IWP FASH, the vertical reference marks the approximate real-data direct-centered estimate, and the horizontal reference marks the true pi0 of 0.8."}
plot_r4_correlation_sweep_pi0(
  sweep_pi0,
  empirical_rho = -0.09,
  true_pi0 = 0.8
)
```

```{r lag1-sweep-primary-table, results='asis'}
render_scrollable_table(
  sweep_primary_table,
  caption = paste(
    "Table R4.6: BF-updated IWP FASH performance and paired changes across",
    "the lag-1-only correlation sensitivity grid."
  ),
  align = "rrrrrrr",
  minimum_width = "1400px"
)
```

Negative adjacent correlation is conservative in this setting. Near the
real-data direct estimate, `rho = -0.1`, BF-adjusted IWP1 has power
`r format_decimal(get_sweep_metric(-0.1, "mean_power"))`, empirical FDR
`r format_decimal(get_sweep_metric(-0.1, "mean_fdr"))`, and estimated pi0
`r format_decimal(get_sweep_pi0_metric(-0.1))`.

At mild positive correlation, `rho = +0.1`, power is
`r format_decimal(get_sweep_metric(0.1, "mean_power"))`, the paired power change
is
`r format_signed_interval(
  get_sweep_paired_metric(0.1, "mean_delta_power"),
  get_sweep_paired_metric(0.1, "power_delta_ci_lower"),
  get_sweep_paired_metric(0.1, "power_delta_ci_upper")
)`, empirical FDR is
`r format_decimal(get_sweep_metric(0.1, "mean_fdr"))`, and estimated pi0 remains
`r format_decimal(get_sweep_pi0_metric(0.1))`.

Inflation becomes visible only as positive correlation grows. At `rho = +0.2`,
BF-adjusted empirical FDR is
`r format_decimal(get_sweep_metric(0.2, "mean_fdr"))`; at `rho = +0.3`, it is
`r format_decimal(get_sweep_metric(0.3, "mean_fdr"))`, with paired change
`r format_signed_interval(
  get_sweep_paired_metric(0.3, "mean_delta_fdr"),
  get_sweep_paired_metric(0.3, "fdr_delta_ci_lower"),
  get_sweep_paired_metric(0.3, "fdr_delta_ci_upper")
)`. BF adjustment substantially attenuates this failure mode: at `rho = +0.3`,
Raw IWP1 FDR is
`r format_decimal(get_sweep_metric(0.3, "mean_fdr", "FASH-IWP1-Raw"))`,
versus
`r format_decimal(get_sweep_metric(0.3, "mean_fdr", "FASH-IWP1-BF"))`
after BF adjustment.

The pi0 diagnostic tells the same story more sharply. At `rho = +0.3`, the Raw
estimate falls to
`r format_decimal(get_sweep_pi0_metric(0.3, method = "FASH-IWP1-Raw"))`, while
the BF-adjusted estimate remains
`r format_decimal(get_sweep_pi0_metric(0.3, method = "FASH-IWP1-BF"))`, close
to the true value 0.8.

## Interpretation and limitations

Together, the diagnostics support a local robustness claim:

- the centered real-data pattern shows modest additional negative short-lag
  dependence beyond the matched centering benchmark;
- inserting its complete 16 by 16 matrix into R1 does not inflate FDR or
  materially change power or estimated pi0;
- an intentionally strong, broad positive full-matrix stress pattern is also
  conservative; and
- in the controlled lag-1 sweep, mild negative or positive correlation has a
  small effect after BF adjustment, whereas moderate positive correlation
  eventually lowers power and inflates FDR.

Several boundaries matter. The top-lfdr subset is selected for flatness, so the
pairwise moment estimator is not unbiased after selection; its matched
benchmark is shown explicitly. The direct matrix is a post-centering
description rather than an identified raw covariance. Monte Carlo intervals
use only five paired seeds and are therefore imprecise, especially for realized
FDP. Finally, these experiments vary within-trajectory Gaussian correlation;
they do not establish robustness to arbitrary dependence, non-Gaussian errors,
or dependence across genes and variants.

## Reproducibility

The implementation is in the
[R4 correlation directory](https://github.com/stephenslab/fashr-paper/tree/master/code/revision_simulations/r4_correlated_errors).
Routine workflowr builds read the completed versioned caches and do not rerun
the real-data bootstrap or Monte Carlo fits.

The real-data matrices can be regenerated with:

```{bash regenerate-real-data-correlations, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/r4_correlated_errors/estimate_real_data_error_correlations.R \
  --top-n 500 \
  --bootstrap-reps 1000 \
  --benchmark-reps 1000 \
  --output-id r4_null_like_top500_full_correlations
```

The paired complete-matrix experiment can be regenerated with:

```{bash regenerate-full-matrix-r4, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/r4_correlated_errors/run_full_empirical_correlation_mc.R \
  --J 1000 \
  --seed-list 12345,22345,32345,42345,52345 \
  --output-id r4_full_empirical_correlations_top500_pilot5 \
  --num-cores 4
```

The lag-1 sweep with pi0 summaries can be regenerated with:

```{bash regenerate-lag1-pi0-sweep, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/r4_correlated_errors/run_lag1_correlation_sweep.R \
  --J 1000 \
  --rho-list=-0.3,-0.2,-0.1,0,0.1,0.2,0.3 \
  --seed-list 12345,22345,32345,42345,52345 \
  --output-id r4_lag1_correlation_sweep_m0p3_to_p0p3_pi0_pilot5 \
  --num-cores 4
```

The versioned caches are:

```text
output/revision_simulations/real_data/
  r4_null_like_top500_full_correlations/

output/revision_simulations/mc/
  r4_full_empirical_correlations_top500_pilot5/
  r4_lag1_correlation_sweep_m0p3_to_p0p3_pi0_pilot5/
```
