---
title: "Internal R3 full calibration: constrained B-spline truth scenarios"
author: "Ziang Zhang"
date: "2026-08-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
project_root <- if (file.exists("code/revision_simulations/shared/simulation_functions.R")) {
  "."
} else if (file.exists("../code/revision_simulations/shared/simulation_functions.R")) {
  ".."
} else {
  stop("Could not locate the workflowr project root.")
}
output_root <- file.path(project_root, "output", "revision_simulations", "internal")
seed_list <- c(12345L, 22345L, 32345L, 42345L, 52345L)
target_order <- c("early", "middle", "late", "switch")
designs <- list(
  constrained_global = list(
    prefix = "r3a_constrained_global_bspline_open_middle_J6362",
    truth_mechanism = "constrained_global_bspline",
    label = "Constrained global random cubic B-spline"
  ),
  targeted_local = list(
    prefix = "r3a_targeted_broad_bspline_open_middle_J6362",
    truth_mechanism = "targeted_broad_bspline",
    label = "Targeted local broad cubic B-spline"
  )
)

read_completed_design <- function(design) {
  paths <- file.path(
    output_root, paste0(design$prefix, "_seed", seed_list), "pilot_result.rds"
  )
  if (any(!file.exists(paths))) {
    stop("The five-seed J=6362 output is incomplete for ", design$label, ".")
  }
  results <- lapply(paths, readRDS)
  configurations <- lapply(results, `[[`, "configuration")
  reference <- configurations[[1L]]
  comparable_truth_generation_settings <- function(settings) {
    if (!is.list(settings)) return(settings)
    if (!is.null(names(settings))) {
      settings <- settings[!grepl("(^seed$|_seed$)", names(settings))]
    }
    lapply(settings, comparable_truth_generation_settings)
  }
  if (!identical(reference$J, 6362L) ||
      !identical(reference$truth_settings$truth_mechanism, design$truth_mechanism) ||
      !identical(reference$middle_window, c(3, 12)) ||
      !identical(reference$middle_boundary, "open") ||
      !identical(reference$truth_generation_settings$middle_window, c(3, 12)) ||
      !identical(reference$truth_generation_settings$middle_boundary, "open")) {
    stop("The recorded configuration does not match the intended full scenario.")
  }
  if (!all(vapply(configurations, function(configuration) {
    identical(configuration$J, reference$J) &&
      identical(configuration$truth_settings, reference$truth_settings) &&
      identical(
        comparable_truth_generation_settings(
          configuration$truth_generation_settings
        ),
        comparable_truth_generation_settings(reference$truth_generation_settings)
      ) &&
      identical(configuration$middle_window, reference$middle_window) &&
      identical(configuration$middle_boundary, reference$middle_boundary) &&
      identical(configuration$middle_definition, reference$middle_definition) &&
      identical(configuration$package_provenance, reference$package_provenance)
  }, logical(1)))) {
    stop("Seed-level configurations are not comparable for ", design$label, ".")
  }
  alpha <- do.call(rbind, Map(function(result, seed) {
    rows <- result$functional_alpha
    rows$seed <- seed
    rows
  }, results, seed_list))
  alpha$alpha <- as.numeric(alpha$alpha)
  list(configuration = reference, alpha = alpha)
}

summarize_alpha <- function(alpha, label) {
  rows <- alpha[
    alpha$method == "FASH-IWP1-BF" & abs(alpha$alpha - 0.05) < 1e-12,
    , drop = FALSE
  ]
  output <- do.call(rbind, lapply(split(rows, rows$target), function(group) {
    discoveries <- sum(group$n_discoveries)
    false_discoveries <- sum(group$false_discoveries)
    data.frame(
      design = label, target = group$target[[1L]],
      mean_power = mean(group$power),
      mean_empirical_fsr = mean(group$empirical_fsr),
      maximum_seed_empirical_fsr = max(group$empirical_fsr),
      pooled_discoveries = discoveries,
      pooled_false_discoveries = false_discoveries,
      pooled_empirical_fsr = if (discoveries == 0L) 0 else {
        false_discoveries / discoveries
      }
    )
  }))
  output[match(target_order, output$target), ]
}

summarize_gate <- function(alpha, label) {
  rows <- alpha[
    alpha$method == "FASH-IWP1-BF" &
      alpha$alpha >= 0.05 & alpha$alpha <= 0.20,
    , drop = FALSE
  ]
  curve <- do.call(rbind, lapply(
    split(rows, list(rows$target, rows$alpha)),
    function(group) data.frame(
      target = group$target[[1L]],
      alpha = group$alpha[[1L]],
      mean_empirical_fsr = mean(group$empirical_fsr),
      mean_fsr_excess = mean(group$empirical_fsr) - group$alpha[[1L]]
    )
  ))
  maximum <- curve[which.max(curve$mean_fsr_excess), , drop = FALSE]
  data.frame(
    design = label,
    maximum_mean_fsr_excess = maximum$mean_fsr_excess,
    target_at_maximum = maximum$target,
    alpha_at_maximum = maximum$alpha,
    calibration_gate_passed = maximum$mean_fsr_excess <= 0.03
  )
}

full_results <- lapply(designs, read_completed_design)
alpha005 <- do.call(rbind, Map(
  function(design, result) summarize_alpha(result$alpha, design$label),
  designs, full_results
))
gate_summary <- do.call(rbind, Map(
  function(design, result) summarize_gate(result$alpha, design$label),
  designs, full_results
))
global_gate <- gate_summary[gate_summary$design == designs$constrained_global$label, ]
local_gate <- gate_summary[gate_summary$design == designs$targeted_local$label, ]
global_configuration <- full_results$constrained_global$configuration
local_configuration <- full_results$targeted_local$configuration
```

## Scope and decision rule

This internal page records two predefined, full real-genotype R3 truth
scenarios. Each has five cached genotype seeds and J = 6362 gene--variant
pairs per seed, with 19 donors, 16 observed times, five covariates, and the
pinned fashr 0.1.43 revision bf223df75da6e41ae48607a56b4cd12d7c3b24e7.

All truth acceptance, saved truth functionals, and posterior testing on this
page use the same open Middle estimand, `3 < t < 12`. The earlier outputs with
prefixes lacking `open_middle` used legacy closed-Middle truth generation and
are explicitly superseded; they are not used anywhere on this page.

For FASH-IWP1-BF, the prespecified calibration gate ranges over Early, Middle,
Late, Switch, and alpha from 0.05 to 0.20. It passes only if the largest mean
seed-level empirical FSR minus alpha is no greater than 0.03. Power is
descriptive; seed is the Monte Carlo replication.

## Truth mechanisms

The constrained-global scenario samples global random cubic B-splines and
accepts curves only when target contrast is at least 0.60 and the
target-to-outside absolute peak ratio is at least 1.40. It retains broad
global B-spline geometry. Its centered RMS is 0.90.

The targeted-local scenario was defined before the global confirmation result.
It uses a broad local cubic B-spline with the same contrast and ratio
thresholds. Non-switch curves have a same-signed baseline of at least 0.75
of the primary amplitude and background at most 0.05. This makes temporal
labels cleaner, but is a more local and less conservative misspecification
stress test.

## Results at alpha 0.05

```{r alpha005-table, echo=FALSE}
display <- alpha005
display$target <- tools::toTitleCase(display$target)
for (column in c(
  "mean_power", "mean_empirical_fsr", "maximum_seed_empirical_fsr",
  "pooled_empirical_fsr"
)) display[[column]] <- formatC(display[[column]], format = "f", digits = 3)
names(display) <- c(
  "Design", "Target", "Mean power", "Mean seed-level FSR",
  "Maximum seed FSR", "Pooled calls", "Pooled false calls", "Pooled FSR"
)
knitr::kable(
  display, align = "l",
  caption = paste(
    "Table 1. Five-seed J=6362 FASH-IWP1-BF results at alpha 0.05.",
    "Pooled quantities are descriptive; the seed is the replication."
  )
)
```

## Full-grid calibration decision

```{r gate-table, echo=FALSE}
display <- gate_summary
display$maximum_mean_fsr_excess <- formatC(
  display$maximum_mean_fsr_excess, format = "f", digits = 4
)
display$alpha_at_maximum <- formatC(
  display$alpha_at_maximum, format = "f", digits = 3
)
names(display) <- c(
  "Design", "Maximum mean FSR excess", "Target at maximum",
  "Alpha at maximum", "Gate passed"
)
knitr::kable(
  display, align = "l",
  caption = paste(
    "Table 2. Prespecified FASH-IWP1-BF calibration gate.",
    "Passing requires maximum mean empirical FSR minus alpha no greater than 0.03."
  )
)
```

```{r decision-text, results="asis", echo=FALSE}
cat(
  "The constrained-global B-spline has maximum excess ",
  sprintf("%.4f", global_gate$maximum_mean_fsr_excess),
  " at ", tools::toTitleCase(global_gate$target_at_maximum),
  " and alpha ", sprintf("%.3f", global_gate$alpha_at_maximum), ". ",
  if (global_gate$calibration_gate_passed) {
    "It passes the numerical gate."
  } else {
    "It does not qualify as the primary calibration scenario."
  },
  "\n\n",
  "The targeted local broad B-spline has maximum excess ",
  sprintf("%.4f", local_gate$maximum_mean_fsr_excess),
  " at ", tools::toTitleCase(local_gate$target_at_maximum),
  " and alpha ", sprintf("%.3f", local_gate$alpha_at_maximum), ". ",
  if (local_gate$calibration_gate_passed) {
    "It passes the prespecified numerical gate."
  } else {
    "It does not pass the prespecified numerical gate."
  },
  sep = ""
)
```

The local mechanism must retain its stated limitation: it evaluates a sharply
specified temporal-label scenario, not a replacement for the global B-spline
misspecification stress test. The strict single-lobe R3B raised-cosine design
remains a low-power label-clean sensitivity rather than a primary scenario.

## Reproduction

The seed-level runner records both the requested Middle definition and the
actual truth-generator settings, recomputes the saved truth functionals, and
stops if they differ. The page independently requires all five seed-level
configurations to agree on these fields before displaying results. Jobs were
submitted with
`code/revision_simulations/internal/pilots/run_r3_truth_specification_pilot.sbatch`
using the two `open_middle_J6362` prefixes shown in the setup chunk.
