---
title: "Internal exploration: mashr-style null-correlation screens"
author: "Ziang Zhang"
date: "2026-08-07"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 10,
  fig.height = 5.8,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(
  paste0(
    "code/revision_simulations/internal/covariance_estimation/",
    "mashr_mean_z_reporting.R"
  )
)
```

<style>
.mean-z-banner {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border: 1px solid #d9a441;
  border-left: 6px solid #d9a441;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.mean-z-result {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.mean-z-caution {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #d55e00;
  border-radius: 4px;
  background: #fff4ec;
  color: #59331e;
}
.mean-z-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.mean-z-table-scroll table {
  margin-bottom: 1em;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
pre {
  max-width: 100%;
  overflow-x: auto;
}
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: hidden;
}
</style>

<div class="mean-z-banner">
<strong>Internal exploratory analysis.</strong> This page is intentionally
unlinked. It does not modify the reviewer-facing R4 page, manuscript, response
letter, or any existing R4 cache.
</div>

## Question and answer

After drawing one variant per gene before inspecting the outcomes, how do the
following three screens compare?

Screen 1 is the standard mashr-style maximum-z rule

$$
\max_t |z_j(t)| < 2
$$

and Screen 2 adds the proposed mean-Z rule

$$
\max_t |z_j(t)| < 2
\quad\text{and}\quad
|Z_{\mathrm{mean},j}| < 2,
\qquad
Z_{\mathrm{mean},j}=\sqrt{16}\,\overline z_j?
$$

Screen 3 instead starts from Screen 1 and removes pairs called significant by
a mashr model fitted to all thinned pairs:

$$
\max_t |z_j(t)| < 2
\quad\text{and}\quad
\operatorname{lfdr}_{\mathrm{mashr},j}>0.05.
$$

Here the pair-level mashr lfdr is the posterior probability assigned to the
exact all-zero effect component. Thus Screen 3 removes a pair only when mashr
assigns posterior null probability at most 0.05.

For each retained set, the empirical time correlation is estimated exactly as

$$
\widehat C=\operatorname{cor}(Z_{\mathcal N}),
$$

where rows are thinned gene--variant units and columns are the 16 time points.
No trajectory is centered within a unit.

<div class="mean-z-result">
<strong>Result.</strong> Screen 1 retained `r format_integer(n_maximum)` of
`r format_integer(n_candidates)` candidates. Screen 3 retained all
`r format_integer(n_mashr)` of them: mashr called
`r format_integer(mashr_mean_z_analysis$mashr_screen_summary$n_pair_lfdr_at_or_below_threshold)`
pairs significant among the complete thinned set, but none of those pairs had
passed Screen 1. The
minimum mashr pair lfdr inside Screen 1 was
`r format_decimal(mashr_mean_z_analysis$mashr_screen_summary$minimum_screen1_pair_lfdr, 3)`.
Consequently, Screen 3 has exactly the same empirical correlation matrix,
lag-correlation curve, and variogram as Screen 1.

By contrast, Screen 2 retained `r format_integer(n_combined)`, removing
`r format_integer(n_removed_by_mean_z)` units, or
`r format_percent(fraction_removed_from_maximum)` of Screen 1. Its mean lag-1
correlation is `r format_decimal(combined_lag[1], 3)`, compared with
`r format_decimal(maximum_lag[1], 3)` under Screens 1 and 3; at lag 15 the
corresponding values are `r format_decimal(combined_lag[15], 3)` and
`r format_decimal(maximum_lag[15], 3)`.
</div>

The Screen-3 negative result matters: with this threshold and model, a mashr
significance filter does not repair the constant-effect contamination that can
remain after Screen 1. Screen 2 changes the estimate, but that still is not
evidence that its estimate is automatically closer to the residual
correlation. It selects directly on a common mode of the same vector whose
correlation is being estimated.

## Design

The analysis starts from the complete BF-adjusted real-data FASH object with
`r format_integer(mashr_mean_z_configuration$n_fash_pairs)` tested
gene--variant pairs from `r format_integer(mashr_mean_z_configuration$n_genes)`
genes. Using seed `r mashr_mean_z_configuration$thinning_seed`, it draws one
variant uniformly within each gene before extracting beta estimates, adjusted
standard errors, z scores, lfdr values, or any other outcome-derived quantity.
All three screens are then applied to exactly the same
`r format_integer(n_candidates)` thinned candidates.

The second statistic uses the user-specified definition

$$
Z_{\mathrm{mean},j}=\sqrt{R}\,\overline z_j=4\overline z_j,
\qquad R=16.
$$

Thus $|Z_{\mathrm{mean},j}|<2$ is equivalent to
$|\overline z_j|<0.5$. The $\sqrt R$ scaling is calibrated when the 16
coordinates are independent with unit variance. Because cross-time correlation
is precisely the object under study, this page treats $Z_{\mathrm{mean}}$ as an
independence-scaled sensitivity screen, not as a calibrated test statistic.

For Screen 3, mashr is fitted once to the complete set of
`r format_integer(n_candidates)` thinned pairs. The null correlation is
estimated internally using
`mashr::estimate_null_correlation_simple()` with its standard threshold 2.
This function selects the Screen-1 rows and computes

$$
\widehat C_{\mathrm{mashr}}
=\operatorname{cov2cor}\!\left\{
  \operatorname{cov}\left(Z_{\mathcal N_1}\right)
\right\}
=\operatorname{cor}\left(Z_{\mathcal N_1}\right),
$$

where $\mathcal N_1$ is the Screen-1 set. Therefore mashr's estimated null
correlation and the Screen-1 empirical correlation are the same estimator,
not two independent estimates; their maximum numerical difference here is
`r format(mashr_mean_z_configuration$mashr_null_correlation_maximum_difference, scientific = TRUE, digits = 3)`.
The full mashr fit then uses the canonical identity, singleton, equal-effects,
and simple-heterogeneity covariance families with a null-biased prior. Screen
3 is applied only after this full-data fit.

```{r selection-table, results='asis'}
render_scrollable_table(
  selection_table,
  align = "lrrrrr",
  minimum_width = "1160px",
  digits = 3L
)
```

```{r mashr-screen-summary-table, results='asis'}
render_scrollable_table(
  mashr_screen_summary_table,
  align = "lr",
  minimum_width = "720px",
  digits = 6L
)
```

The small time-column means in this table do not prove that individual units
have zero constant effects. Positive and negative unit-specific intercepts can
cancel in each column while still contributing a positive common covariance
across time.

## Empirical correlation matrices

```{r correlation-heatmaps, fig.width=15, fig.height=5.7, fig.cap="Empirical z-score correlations under the three screens. All panels use the same off-diagonal scale; the unit diagonal is omitted to make the off-diagonal structure visible. Screens 1 and 3 are exactly identical."}
plot_mashr_mean_z_heatmaps()
```

For the specific day-0/day-1 pair, the correlation is
`r format_decimal(maximum_correlation[1, 2], 3)` under both Screens 1 and 3 and
`r format_decimal(combined_correlation[1, 2], 3)` after adding the mean-Z
screen. Screen 3 makes no entrywise change. Across all 120 off-diagonal
entries, Screen 2 preserves much of the ordering but shifts almost every entry
downward relative to Screens 1 and 3.

```{r matrix-comparison-table, results='asis'}
render_scrollable_table(
  matrix_comparison_table,
  align = "llrrrrr",
  minimum_width = "1640px",
  digits = 4L
)
```

## Lag correlation and variogram

Intervals are pointwise percentile 95% intervals from 1,000 gene-level
bootstrap resamples within each retained set.

```{r lag-correlation, fig.width=9.5, fig.height=5.8, fig.cap="Lag-averaged empirical correlations with pointwise gene-bootstrap 95% intervals."}
plot_lag_correlation_comparison()
```

```{r variogram, fig.width=9.5, fig.height=5.8, fig.cap="Standardized semivariograms, defined as one minus the lag-averaged correlation, with pointwise gene-bootstrap 95% intervals."}
plot_variogram_comparison()
```

```{r key-lag-table, results='asis'}
render_scrollable_table(
  key_lag_table,
  align = "llrlrl",
  minimum_width = "1160px",
  digits = 3L
)
```

Screens 1 and 3 have the same persistent positive component that decays from
lag 1 to lag 15. The mean-Z restriction in Screen 2 weakens lag 1 and pushes
the longer-lag average below zero. Therefore the apparent disappearance of
long-range positive dependence is inseparable from the Screen-2 selection
rule; the mashr significance filter in Screen 3 provides no change at all.

## Why the mean-Z screen can induce negative correlation

<div class="mean-z-caution">
The stricter rule is not an outcome-neutral refinement. It conditions on the
sum of the 16 coordinates. Even if a zero-effect vector originally had
independent coordinates, forcing its mean toward zero makes a positive value at
one time point more likely to be offset by negative values elsewhere.
</div>

The limiting calculation makes the mechanism transparent. If
$z\sim N(0,I_R)$ and one conditions on the exact constraint
$\mathbf 1^Tz=0$, then

$$
\operatorname{Cov}(z\mid\mathbf 1^Tz=0)
=I_R-\frac{1}{R}\mathbf 1\mathbf 1^T.
$$

After rescaling to a correlation matrix, every off-diagonal entry is
$-1/(R-1)$, which is approximately $-0.067$ for $R=16$. The actual screen is a
band rather than an exact constraint, and the real coordinates are not
independent, so this calculation is not a prediction of the observed matrix.
It does show why negative long-lag entries can arise mechanically after
screening on $|Z_{\mathrm{mean}}|$.

Consequently, the comparison identifies a genuine sensitivity but does not by
itself identify the true residual correlation. The maximum-only set may retain
unit-specific constant effects; the combined set more aggressively removes
that common direction but can bias the correlation downward through
conditioning.

## Audit of the motivating bad example

The visually suspicious Screen-1 example identified in the first gallery is
not removed by Screen 3. Its estimated effects are positive at every time
point, with an inverse-variance-weighted constant effect near 1, yet its
individual z scores all remain below 2. Mashr assigns it pair lfdr
`r format_decimal(mashr_mean_z_analysis$bad_example_audit$mashr_pair_lfdr, 3)`,
well above the 0.05 removal threshold. Screen 2 removes it because its
$Z_{\mathrm{mean}}$ is `r format_decimal(mashr_mean_z_analysis$bad_example_audit$mean_z_score, 3)`.

```{r bad-example-table, results='asis'}
render_scrollable_table(
  bad_example_table,
  align = "lllrrrrrlll",
  minimum_width = "1760px",
  digits = 4L
)
```

This demonstrates why the Screen-3 filter is not equivalent to controlling the
common direction. The marginal z scores are deliberately bounded by Screen 1,
and under the fitted mashr mixture the accumulated evidence for this smooth,
moderate trajectory is not strong enough to make its posterior null
probability at most 0.05.

## Selected-unit audit {.tabset .tabset-pills}

The three tabs show fixed random samples of 25 retained units. The first sample is
drawn from the full maximum-z set; `r sum(get_gallery(maximum_filter_id)$metadata$passes_maximum_z_and_mean_z)`
of its 25 units also pass the added mean-Z rule and the other
`r sum(!get_gallery(maximum_filter_id)$metadata$passes_maximum_z_and_mean_z)`
do not. The second sample is drawn independently from the combined set. The
third sample is drawn independently from Screen 3. Screens 1 and 3 contain the
same pairs, but their displayed samples use different fixed seeds, so their
gallery panels need not match. Blue, orange, or green lines are z-score
trajectories, gray lines mark zero and $\pm2$, and the green dashed line is the
within-trajectory mean z. Full gene and variant identifiers are listed beneath
each plot.

### Screen 1: max|z(t)| < 2

```{r maximum-z-gallery, fig.width=13, fig.height=13, fig.cap="Twenty-five fixed-random units selected under the maximum-z rule."}
plot_gallery(maximum_filter_id)
```

```{r maximum-z-gallery-table, results='asis'}
render_scrollable_table(
  gallery_table(maximum_filter_id),
  align = "rllrrrrll",
  minimum_width = "1540px",
  digits = 4L
)
```

### Screen 2: max|z(t)| < 2 and |Z_mean| < 2

```{r combined-gallery, fig.width=13, fig.height=13, fig.cap="Twenty-five fixed-random units selected after adding the mean-Z rule."}
plot_gallery(combined_filter_id)
```

```{r combined-gallery-table, results='asis'}
render_scrollable_table(
  gallery_table(combined_filter_id),
  align = "rllrrrrll",
  minimum_width = "1540px",
  digits = 4L
)
```

### Screen 3: max|z(t)| < 2 and mashr pair lfdr > 0.05

```{r mashr-gallery, fig.width=13, fig.height=13, fig.cap="Twenty-five fixed-random units selected after removing mashr discoveries from Screen 1. The retained set is exactly the Screen-1 set in this fit."}
plot_gallery(mashr_filter_id)
```

```{r mashr-gallery-table, results='asis'}
render_scrollable_table(
  gallery_table(mashr_filter_id),
  align = "rllrrrrll",
  minimum_width = "1540px",
  digits = 4L
)
```

## Interpretation and limitations

1. Letting mashr estimate its own null correlation does not provide an
   independent validation of Screen 1: the standard estimator is exactly the
   Screen-1 empirical z-score correlation.
2. At pair lfdr 0.05, Screen 3 is redundant after Screen 1. It removes zero of
   the `r format_integer(n_maximum)` Screen-1 pairs and leaves the suspicious
   constant-effect example in the set.
3. The proposed mean-Z criterion materially changes the estimated correlation,
   despite using the same one-variant-per-gene thinning. It is the only tested
   screen here that removes the motivating example.
4. The Screen-2 matrix has a similar time-pattern ordering but a much lower
   baseline. That is consistent with removing a broad common component, but
   the rule also truncates that direction and can create negative dependence.
5. None of the selected sets is known to contain only exact
   zero-marginal-effect pairs. Screens 1 and 3 can retain constant-effect
   heterogeneity, while Screen 2 can have selection-induced downward bias.
6. These are empirical correlations of adjusted z scores, not beta-scale
   covariance matrices and not identified gene-specific error covariances.

The defensible conclusion is therefore twofold. First, the proposed mashr-lfdr
filter does not improve the maximum-z set in this dataset. Second, the estimated
residual-correlation level is sensitive to adding a mean-Z restriction, but
that restriction has its own selection effect. This page should remain
internal until the estimand and selection argument are settled.

## Reproduction

The cache contains the exact candidate z matrix, fitted beta estimates,
adjusted standard errors, selected indices, all three complete empirical
correlation matrices, the fitted mashr object and posterior weights, bootstrap
intervals, gallery membership, input-file provenance, and CSV audit tables.

```{r reproduction-command, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/covariance_estimation/run_mashr_mean_z_null_correlation.R

workflowr::wflow_build(
  "analysis/revision_internal_mashr_mean_z_null_correlation.rmd",
  local = TRUE,
  verbose = TRUE
)
```
