---
title: "Internal R1 diagnostic: residual normality and t-adjusted SEs"
author: "Ziang Zhang"
date: "2026-08-20"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 10.5,
  fig.height = 5.8,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(file.path(
  "code",
  "revision_simulations",
  "internal",
  "r1_normality_assumption",
  "reporting.R"
))
```

<style>
.normality-banner {
  margin: 1.2em 0 1.5em;
  padding: 0.9em 1.1em;
  border: 1px solid #d9a441;
  border-left: 6px solid #d9a441;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.normality-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.main-container .figure {
  max-width: 980px;
  margin: 1.3em auto;
}
.main-container .figure img {
  display: block;
  width: 100%;
  max-width: 100%;
  height: auto;
}
.MathJax_Display,
pre {
  max-width: 100%;
  overflow-x: auto;
}
</style>

::: {.normality-banner}
<strong>Two complementary checks.</strong> Part 1 is a descriptive real-data
residual diagnostic, stratified by day. Part 2 is the seed-12345 R1
known-truth consequence analysis and is the main justification for the
t-statistic-based SE adjustment.
:::

## 1. Real-data residual normality check

We use the fixed one-variant-per-gene subset (6,362 genes) and fit the original
regression separately at Day 1, Day 3, and Day 9,

$$
Y_{gti}=\alpha_{gt}+\beta_{gt}G_{gi}
+\sum_{k=1}^{5}\gamma_{gtk}\mathrm{PC}_{tik}+e_{gti}.
$$

Each regression has 19 donors, seven fitted coefficients, and residual df 12.
The conventional standardized residual is

$$
r_{gti}=\frac{e_{gti}}{s_{gt}\sqrt{1-h_{gtii}}},
$$

the quantity returned by `rstandard()` for an ordinary linear model. We
compare it with the standard Normal as in a traditional regression diagnostic.
Days are not pooled; within each day, the plot summarizes the 19 residuals
from each of the 6,362 gene-specific regressions.

```{r real-data-residual-distributions, fig.width=11, fig.height=8.6, out.width='97%', fig.cap="Figure 1. Traditional standardized-residual diagnostics for the fixed one-variant-per-gene real-data subset, shown separately for Day 1, Day 3, and Day 9. The upper row contains standard-Normal QQ plots. The lower row contains density-scaled histograms with the standard-Normal density overlaid in red. Each day contains 120,878 residuals from 6,362 regressions with 19 donors and residual df 12."}
# Retain the conventional standardized residual from every valid gene-specific
# regression, and set the requested Day 1, Day 3, Day 9 panel order explicitly.
residual_plot_data <- standardized_residuals[
  standardized_residuals$day %in% c(1L, 3L, 9L),
  c("day", "day_label", "standardized_residual"),
  drop = FALSE
]
residual_plot_data$day_label <- factor(
  residual_plot_data$day_label,
  levels = c("Day 1", "Day 3", "Day 9")
)

# For each day, compare empirical standardized-residual quantiles with the
# corresponding standard-Normal quantiles on one shared probability grid.
qq_probabilities <- c(
  seq(0.001, 0.010, by = 0.001),
  seq(0.015, 0.985, by = 0.005),
  seq(0.990, 0.999, by = 0.001)
)
residual_qq_plot_data <- do.call(rbind, lapply(
  levels(residual_plot_data$day_label),
  function(current_day) {
    current_residuals <- residual_plot_data$standardized_residual[
      residual_plot_data$day_label == current_day
    ]
    data.frame(
      day_label = current_day,
      normal_reference_quantile = stats::qnorm(qq_probabilities),
      empirical_quantile = as.numeric(stats::quantile(
        current_residuals,
        probs = qq_probabilities,
        names = FALSE,
        type = 8
      )),
      stringsAsFactors = FALSE
    )
  }
))
residual_qq_plot_data$day_label <- factor(
  residual_qq_plot_data$day_label,
  levels = levels(residual_plot_data$day_label)
)

residual_qq_plot <- ggplot2::ggplot(
  residual_qq_plot_data,
  ggplot2::aes(
    x = normal_reference_quantile,
    y = empirical_quantile
  )
) +
  ggplot2::geom_abline(
    intercept = 0,
    slope = 1,
    color = "grey45",
    linewidth = 0.7,
    linetype = 2
  ) +
  ggplot2::geom_point(
    color = "#2166ac",
    size = 0.9,
    alpha = 0.75
  ) +
  ggplot2::facet_wrap(ggplot2::vars(day_label), nrow = 1) +
  ggplot2::labs(
    title = "Standard-Normal QQ comparison",
    x = "Standard-Normal quantile",
    y = "Empirical standardized-residual quantile"
  ) +
  normality_theme()

# Overlay the standard-Normal density on density-scaled histograms of the same
# day-specific standardized residuals shown in the QQ plots.
normal_reference_density <- data.frame(
  standardized_residual = seq(-3.6, 3.6, length.out = 1001L)
)
normal_reference_density$density <- stats::dnorm(
  normal_reference_density$standardized_residual
)

residual_histogram_plot <- ggplot2::ggplot(
  residual_plot_data,
  ggplot2::aes(x = standardized_residual)
) +
  ggplot2::geom_histogram(
    ggplot2::aes(y = ggplot2::after_stat(density)),
    binwidth = 0.08,
    boundary = 0,
    fill = "#92c5de",
    color = "white",
    linewidth = 0.15
  ) +
  ggplot2::geom_line(
    data = normal_reference_density,
    ggplot2::aes(x = standardized_residual, y = density),
    inherit.aes = FALSE,
    color = "#b2182b",
    linewidth = 0.9
  ) +
  ggplot2::facet_wrap(ggplot2::vars(day_label), nrow = 1) +
  ggplot2::coord_cartesian(xlim = c(-3.6, 3.6)) +
  ggplot2::labs(
    title = "Histogram with standard-Normal density",
    x = "Standardized residual",
    y = "Density"
  ) +
  normality_theme()

residual_qq_plot / residual_histogram_plot +
  patchwork::plot_layout(heights = c(1.05, 0.95))
```

```{r residual-summary-table}
# Report the exact day-specific denominators, distribution summaries, and
# two-sided exceedance rates at standard-Normal 5%, 1%, and 0.1% cutoffs.
tail_rate_005 <- residual_tail_summary[
  residual_tail_summary$nominal_two_sided_rate == 0.05,
  c("day", "observed_two_sided_rate"),
  drop = FALSE
]
tail_rate_001 <- residual_tail_summary[
  residual_tail_summary$nominal_two_sided_rate == 0.01,
  c("day", "observed_two_sided_rate"),
  drop = FALSE
]
tail_rate_0001 <- residual_tail_summary[
  residual_tail_summary$nominal_two_sided_rate == 0.001,
  c("day", "observed_two_sided_rate"),
  drop = FALSE
]

residual_summary_display <- data.frame(
  Day = residual_day_summary$day_label,
  Residuals = format_integer(residual_day_summary$n_residuals),
  SD = format_decimal(residual_day_summary$sd, 3L),
  Skewness = format_decimal(residual_day_summary$skewness, 3L),
  `Excess kurtosis` = format_decimal(
    residual_day_summary$excess_kurtosis,
    3L
  ),
  `Outside +/-1.96` = format_percent(
    tail_rate_005$observed_two_sided_rate[
      match(residual_day_summary$day, tail_rate_005$day)
    ],
    2L
  ),
  `Outside +/-2.58` = format_percent(
    tail_rate_001$observed_two_sided_rate[
      match(residual_day_summary$day, tail_rate_001$day)
    ],
    2L
  ),
  `Outside +/-3.29` = format_percent(
    tail_rate_0001$observed_two_sided_rate[
      match(residual_day_summary$day, tail_rate_0001$day)
    ],
    3L
  ),
  check.names = FALSE
)
render_scrollable_table(
  residual_summary_display,
  caption = paste(
    "Day-stratified standardized-residual summaries.",
    "Normal-reference tail rates are 5%, 1%, and 0.1%."
  ),
  align = "lrrrrrrr",
  minimum_width = "900px"
)
```

The deviations are modest: the day-specific SDs are
`r paste(format_decimal(residual_day_summary$sd, 3L), collapse = ", ")`,
and the absolute skewness is at most
`r format_decimal(max(abs(residual_day_summary$skewness)), 3L)`.
The extreme tails are somewhat lighter than the standard Normal, especially
at Day 1. This is a descriptive diagnostic rather than a formal pooled test:
residuals are dependent within each regression, and the same donors recur
across genes.

## 2. Consequence of omitting the t-statistic SE adjustment

The same Gaussian R1 data, IWP1 prior grid, penalty, and cumulative-lfdr rule
are used in both SE arms. Only the supplied SE changes. Raw and BF denote the
original FASH fit and its BF-updated dynamic-null weight, respectively.

```{r realized-fdp-curves, fig.width=10.5, fig.height=5.8, out.width='95%', fig.cap="Figure 2. Seed-12345 realized FDP before and after the t-statistic-based adjustment of the regression SE. Colors distinguish Raw and BF-updated FASH. Grey lines show equality with the nominal cumulative-lfdr level. These curves are realized FDP from one simulated dataset, not Monte Carlo empirical FDR."}
# Retain the Gaussian R1 arm and order the two SE scales and FASH stages.
fdp_curve_data <- performance_alpha_curve[
  performance_alpha_curve$error_distribution == "Gaussian",
  c("se_scale", "fit_stage", "alpha", "realized_fdp"),
  drop = FALSE
]
fdp_curve_data$se_label <- factor(
  fdp_curve_data$se_scale,
  levels = c("raw regression SE", "t-adjusted SE"),
  labels = c("Raw regression SE", "t-adjusted SE")
)
fdp_curve_data$fit_stage <- factor(
  fdp_curve_data$fit_stage,
  levels = c("Raw", "BF")
)

# Draw the equality reference separately within each SE panel.
nominal_fdp_reference <- expand.grid(
  alpha = c(0, 0.20),
  se_label = levels(fdp_curve_data$se_label),
  stringsAsFactors = FALSE
)
nominal_fdp_reference$realized_fdp <- nominal_fdp_reference$alpha
nominal_fdp_reference$se_label <- factor(
  nominal_fdp_reference$se_label,
  levels = levels(fdp_curve_data$se_label)
)

ggplot2::ggplot(
  fdp_curve_data,
  ggplot2::aes(
    x = alpha,
    y = realized_fdp,
    color = fit_stage
  )
) +
  ggplot2::geom_line(
    data = nominal_fdp_reference,
    ggplot2::aes(x = alpha, y = realized_fdp, group = se_label),
    inherit.aes = FALSE,
    color = "grey45",
    linewidth = 0.7,
    linetype = 3
  ) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::facet_wrap(ggplot2::vars(se_label), nrow = 1) +
  ggplot2::scale_color_manual(
    values = c("Raw" = "#d95f02", "BF" = "#1b9e77")
  ) +
  ggplot2::scale_x_continuous(
    breaks = seq(0, 0.20, by = 0.05),
    labels = scales::label_percent(accuracy = 1)
  ) +
  ggplot2::scale_y_continuous(
    labels = scales::label_percent(accuracy = 1)
  ) +
  ggplot2::labs(
    title = "Realized FDP before and after t adjustment",
    subtitle = "R1 seed 12345; grey lines mark the nominal level",
    x = "Nominal cumulative-lfdr level",
    y = "Realized FDP",
    color = "FASH stage"
  ) +
  normality_theme()
```

```{r performance-alpha005-table}
# Select the four Gaussian R1 results at cumulative-lfdr level 0.05.
gaussian_performance_alpha005 <- performance_alpha005[
  performance_alpha005$error_distribution == "Gaussian" &
    abs(performance_alpha005$alpha - 0.05) < 1e-12,
  ,
  drop = FALSE
]
gaussian_performance_alpha005 <- gaussian_performance_alpha005[order(
  match(
    gaussian_performance_alpha005$se_scale,
    c("raw regression SE", "t-adjusted SE")
  ),
  match(gaussian_performance_alpha005$fit_stage, c("Raw", "BF"))
), , drop = FALSE]

# Format counts and realized FDP for a compact collaborator-facing table.
performance_display <- data.frame(
  `SE scale` = gaussian_performance_alpha005$se_scale,
  `FASH stage` = gaussian_performance_alpha005$fit_stage,
  Discoveries = format_integer(
    gaussian_performance_alpha005$n_discoveries
  ),
  `False discoveries` = format_integer(
    gaussian_performance_alpha005$false_discoveries
  ),
  `Realized FDP` = format_percent(
    gaussian_performance_alpha005$realized_fdp,
    1L
  ),
  check.names = FALSE
)
render_scrollable_table(
  performance_display,
  caption = "Seed-12345 results at cumulative-lfdr level 0.05.",
  align = "llrrr",
  minimum_width = "680px"
)
```

At level 0.05, t adjustment reduces realized FDP from
`r format_percent(gaussian_performance_alpha005$realized_fdp[
  gaussian_performance_alpha005$se_scale == "raw regression SE" &
    gaussian_performance_alpha005$fit_stage == "Raw"
], 1L)` to
`r format_percent(gaussian_performance_alpha005$realized_fdp[
  gaussian_performance_alpha005$se_scale == "t-adjusted SE" &
    gaussian_performance_alpha005$fit_stage == "Raw"
], 1L)` for Raw FASH, and from
`r format_percent(gaussian_performance_alpha005$realized_fdp[
  gaussian_performance_alpha005$se_scale == "raw regression SE" &
    gaussian_performance_alpha005$fit_stage == "BF"
], 1L)` to
`r format_percent(gaussian_performance_alpha005$realized_fdp[
  gaussian_performance_alpha005$se_scale == "t-adjusted SE" &
    gaussian_performance_alpha005$fit_stage == "BF"
], 1L)` after BF updating. This practical difference is the main result:
with only 19 donors, the finite-sample t-statistic uncertainty cannot be
ignored when supplying a Normal-likelihood method with effect estimates and
standard errors.

This page reads versioned caches; routine builds do not rerun the regressions
or FASH.

```{bash test-and-build-r1-normality-page, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/r1_normality_assumption/test_real_data_residual_normality_helpers.R

Rscript --vanilla \
  code/revision_simulations/internal/r1_normality_assumption/test_reporting.R

Rscript --vanilla -e \
  'workflowr::wflow_build("analysis/revision_internal_r1_normality_assumption.rmd")'
```
