---
title: "Internal exploratory analysis: residual covariance estimation among FASH-null trajectories"
author: "Ziang Zhang"
date: "2026-08-06"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 10,
  fig.height = 6,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(
  "code/revision_simulations/internal/covariance_estimation/pairwise_reporting.R"
)
source(
  paste0(
    "code/revision_simulations/internal/covariance_estimation/",
    "permuted_variogram_covariance_reporting.R"
  )
)
```

<style>
.internal-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.internal-table-scroll table {
  margin-bottom: 1em;
}
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: hidden;
}
</style>

## Status and question

This is an **internal exploratory analysis**. It is not linked from the
reviewer-facing R4 page, the manuscript, or the response letter. The existing
R4 page and its simulation caches are unchanged.

The previous empirical-covariance experiment showed why near-zero z scores are
not sufficient to identify measurement-error correlation: many selected
trajectories have a persistent nonzero genetic effect. Here we focus on a
different estimator that cancels an unknown constant effect algebraically,
without estimating or subtracting a separate intercept for each unit.

The final section adds a stronger donor-permutation experiment. It uses a
matched independent-time permutation as the variogram reference, allowing the
lag covariance and a full empirical covariance matrix to be estimated without
interpreting centering-induced negative correlation as residual dependence.

The target is the common residual correlation matrix $C$ in

$$
\widehat\beta_j \mid \beta_j
\sim N\!\left(\beta_j,
D_j C D_j\right),
\qquad
D_j=\operatorname{diag}(s_{j1},\ldots,s_{j16}).
$$

Because the standardized residuals have unit marginal variances, $C$ is both
their covariance and correlation matrix. The beta-hat-scale covariance
$D_j C D_j$ differs across units because their standard errors differ.

## Pairwise-difference estimator

Under the FASH dynamic null, unit $j$ has an unknown constant effect,

$$
\beta_j(t)=\mu_j.
$$

For time points $t$ and $s$, define

$$
A_{j,ts}=s_{jt}^2+s_{js}^2,
\qquad
x_{j,ts}=\frac{2s_{jt}s_{js}}{A_{j,ts}},
\qquad
q_{j,ts}=\frac{
  \{\widehat\beta_j(t)-\widehat\beta_j(s)\}^2
}{A_{j,ts}}.
$$

If the selected unit is a true FASH null, the unknown $\mu_j$ cancels and

$$
E\!\left(1-q_{j,ts}\mid s_{jt},s_{js}\right)
=x_{j,ts}C_{ts}.
$$

We therefore estimate each off-diagonal entry by OLS through the origin,

$$
\widehat C_{ts}
=
\frac{\sum_j x_{j,ts}(1-q_{j,ts})}
     {\sum_j x_{j,ts}^2},
\qquad
\widehat C_{tt}=1.
$$

```{r pairwise-estimator-setting, eval=FALSE}
variance_sum <- se[, time_t]^2 + se[, time_s]^2
x <- 2 * se[, time_t] * se[, time_s] / variance_sum
q <- (beta_hat[, time_t] - beta_hat[, time_s])^2 / variance_sum

raw_correlation[time_t, time_s] <- sum(x * (1 - q)) / sum(x^2)
raw_correlation[time_s, time_t] <- raw_correlation[time_t, time_s]
diag(raw_correlation) <- 1
```

No unit-specific intercept is estimated, no trajectory is centered, and no
positive-definite projection is applied. If a selected unit is dynamically
non-null, its squared true difference enters $q_{j,ts}$ and moves the estimate
downward. Thus the estimator equals residual correlation in expectation for an
externally fixed set of true constant-effect nulls, but is a lower estimate
when dynamic alternatives contaminate that fixed set.

There is a separate selection issue: the fitted lfdr is computed from the same
16 observations. Requiring a high observed lfdr can favor trajectories that
look unusually flat because of their realized noise. Therefore, high lfdr is
an approximate null selector, not an oracle label; the analysis below examines
whether the estimate is sensitive to the chosen lfdr cutoff.

## LFDR-threshold selection

Within each gene, we first retain the variant with the highest BF-adjusted FASH
lfdr, using the pair key to break exact ties. We then use all gene
representatives satisfying the strict rules

$$
\operatorname{lfdr}>0.97,
\qquad
\operatorname{lfdr}>0.96,
\qquad
\operatorname{lfdr}>0.95.
$$

The three sets are nested and contain
`r paste(format(selection_counts$n_selected, big.mark = ","), collapse = ", ")`
distinct genes, respectively, out of
`r format(configuration$n_gene_representatives, big.mark = ",")`
one-per-gene representatives.

```{r threshold-summary-table, results='asis'}
render_scrollable_table(
  threshold_summary_table,
  align = "lrlrrrrrr",
  minimum_width = "1550px"
)
```

## Raw empirical matrices

```{r pairwise-raw-matrices, fig.width=14, fig.height=5.6}
plot_pairwise_raw_matrices()
```

All three matrices show a broad positive common mode rather than dependence
concentrated near the first off-diagonal. Mean off-diagonal correlation ranges
only from
`r format_decimal(off_diagonal_range[1])` to
`r format_decimal(off_diagonal_range[2])`
as the selected set grows from
`r format(selection_counts$n_selected[1], big.mark = ",")` to
`r format(selection_counts$n_selected[3], big.mark = ",")` genes.

The raw matrices have one negative eigenvalue each, with minimum eigenvalues
`r paste(format_decimal(matrix_diagnostics$minimum_eigenvalue, 4), collapse = ", ")`.
This is allowed in the present descriptive analysis: these are the unmodified
pairwise moment estimates requested here. They cannot be used directly as
Gaussian simulation covariance matrices without a later regularization or
projection step.

## Variogram

For lag $h$, the plotted standardized semivariogram is

$$
\widehat\gamma(h)
=\frac{1}{|\mathcal P_h|}
\sum_{(t,s)\in\mathcal P_h}\{1-\widehat C_{ts}\},
\qquad
\mathcal P_h=\{(t,s):|t-s|=h\}.
$$

The bands are percentile 95% intervals from 1,000 gene-bootstrap
replications, performed separately within each threshold-selected set.

```{r pairwise-variograms, fig.width=9.5, fig.height=6.2}
plot_pairwise_variograms()
```

The variogram does not increase with lag. Instead, mean correlation rises from
`r format_decimal(lag1_range[1])`--`r format_decimal(lag1_range[2])`
at lag 1 to
`r format_decimal(lag15_range[1])`--`r format_decimal(lag15_range[2])`
at lag 15, so the semivariogram declines. This is not the usual signature of a
short-range temporal error process. It is more consistent with a broad common
mode, time-specific heterogeneity, or ascertainment for flat trajectories than
with a lag-1-only dependence pattern.

The bootstrap bands quantify gene-sampling uncertainty conditional on the
selection rule. They do not quantify bias from using the observed lfdr to
select the same trajectories.

## Sensitivity to the LFDR threshold

```{r pairwise-threshold-sensitivity, fig.width=8.8, fig.height=5.8}
plot_pairwise_threshold_sensitivity()
```

The lag-1 estimates change by only
`r format_decimal(diff(lag1_range))`
across the three thresholds; the mean off-diagonal estimates change by only
`r format_decimal(diff(off_diagonal_range))`.
Comparing the two endpoint matrices, the mean absolute off-diagonal difference
is
`r format_decimal(extreme_matrix_difference[["Mean absolute off-diagonal difference"]])`
and the maximum entrywise difference is
`r format_decimal(extreme_matrix_difference[["Maximum absolute entry difference"]])`.

```{r matrix-difference-table, results='asis'}
render_scrollable_table(
  matrix_difference_table,
  align = "lrrr",
  minimum_width = "920px",
  digits = 4
)
```

Thus the empirical magnitude and broad matrix pattern are not materially
sensitive to choosing 0.95, 0.96, or 0.97. This stability is useful, but it
does not by itself distinguish genuine residual common-mode correlation from
selection-induced flatness because all three sets are defined by the same
trajectory-based lfdr.

## Internal interpretation

This experiment supports three conclusions.

1. **The algebraic cancellation works for the FASH null.** Unlike
   within-trajectory centering, the estimator removes an arbitrary constant
   effect without estimating it and does not mechanically impose a sum-to-zero
   constraint.
2. **The result is stable across the requested lfdr thresholds.** Reducing the
   cutoff from 0.97 to 0.95 nearly quadruples the selected sample size while
   changing the average correlation only slightly.
3. **The estimated pattern is not a mild lag-1 process.** It is a strong,
   nearly global positive common mode, with correlation tending to be larger
   rather than smaller at long lags. The threshold sweep therefore does not
   reproduce the mild negative lag-1 pattern suggested by the centered
   diagnostic.

For this observed-data selection experiment alone, the strongest defensible
summary is:

> Among one-per-gene trajectories with BF-adjusted FASH lfdr above 0.95--0.97,
> the unprojected pairwise-difference estimator yields a strong positive common
> mode that is stable to the lfdr cutoff. If these selected trajectories were
> an externally fixed set of true constant-effect nulls, this would estimate
> residual correlation directly. Because the same trajectories are selected
> using their observed lfdr, residual selection bias remains unresolved.

## Donor-permutation resolution

The observed-data analysis cannot separate residual covariance from selection
for flat trajectories. We therefore fix the selected gene--variant pairs
before generating a genotype-null dataset and compare two residual-block
randomizations:

1. **Synchronized permutation.** One donor residual-block mapping is shared by
   all selected units and all 16 time points. This breaks the
   genotype--expression association while preserving the observed cross-time
   donor pairing.
2. **Independent-time benchmark.** A separate donor residual-row permutation
   is used at every time point. This preserves each time's marginal residual
   distribution but removes cross-time pairing.

Both designs use the original genotypes, the time-specific PC1--PC5 designs,
the fixed one-variant-per-gene selections, and 100 deterministic
permutations. The primary analysis uses the 4,025 units with strict
BF-adjusted FASH `lfdr > 0.95` and the t-adjusted SE used by FASH.

### From variogram to covariance

For estimator $m$, let $\widehat\gamma^{\mathrm{sync}}_{m,ts}$ and
$\widehat\gamma^{\mathrm{ind}}_{m,ts}$ be the covariance-scale
semivariograms under the synchronized and independent-time designs. The
matched covariance estimate is

$$
\widehat K^{(m)}_{ts}
=
\widehat\gamma^{\mathrm{ind}}_{m,ts}
-
\widehat\gamma^{\mathrm{sync}}_{m,ts}.
$$

The independent-time direct zero-null covariance has mean off-diagonal
`r format_decimal(permuted_independent_off_diagonal_mean, 4)` and range
[`r format_decimal(permuted_independent_off_diagonal_range[1], 4)`,
`r format_decimal(permuted_independent_off_diagonal_range[2], 4)`], confirming
that it is an effective zero-cross-time-covariance reference. For the full
matrices, the diagonal is supplied by the synchronized direct zero-null
marginal variances because every variogram has a zero diagonal.

```{r permuted-variogram-covariance, fig.width=9.5, fig.height=8.2}
plot_permuted_variogram_covariance()
```

The synchronized variograms increase toward their method-matched independence
benchmarks. Subtracting them from those benchmarks shows positive covariance
that decreases with lag rather than covariance that increases with lag.

```{r permuted-primary-lag-table, results='asis'}
render_scrollable_table(
  permuted_primary_lag_table,
  align = "llll",
  minimum_width = "1150px"
)
```

The pairwise OLS and direct zero-null estimates agree numerically. Across all
120 off-diagonal entries, their matrix correlation is
`r format_decimal(permuted_method_agreement_table[1, 2], 4)` and their RMSE is
`r format_decimal(permuted_method_agreement_table[1, 3], 4)`. Their mean
covariance difference is only
`r format_decimal(permuted_method_agreement_table[1, 4], 4)`.

The within-unit centered estimate agrees closely in shape, with matrix
correlation
`r format_decimal(permuted_method_agreement_table[2, 2], 4)`, but is not
numerically identical: its off-diagonal RMSE is
`r format_decimal(permuted_method_agreement_table[2, 3], 4)` and its mean
covariance is lower by
`r format_decimal(abs(permuted_method_agreement_table[2, 4]), 4)`. The gap
grows to about 0.04 at the longest lag. This residual discrepancy is expected
because the implemented centering removes an inverse-variance weighted
constant on the beta-hat scale and then divides by unequal SEs; centering is
therefore not exactly invariant on the final standardized scale.

```{r permuted-method-agreement-table, results='asis'}
render_scrollable_table(
  permuted_method_agreement_table,
  align = "lrrrr",
  minimum_width = "1150px",
  digits = 4
)
```

Thus all three estimators support the same qualitative lag pattern and a
non-negligible positive covariance. Pairwise OLS and direct zero-null provide
the strongest quantitative agreement; the centered estimator is a mildly
conservative diagnostic, especially at long lags.

### Empirical covariance matrix

```{r permuted-covariance-matrix-comparison, fig.width=14, fig.height=5.3}
plot_permuted_covariance_matrices()
```

```{r permuted-matrix-diagnostics-table, results='asis'}
render_scrollable_table(
  permuted_matrix_diagnostics_table,
  align = "lrrrrrrr",
  minimum_width = "1450px",
  digits = 4
)
```

The covariance is not negligible on the standardized beta-hat scale. For the
primary direct zero-null estimate, the mean diagonal is
`r format_decimal(permuted_direct_diagnostics$mean_diagonal, 3)`, the mean
off-diagonal covariance is
`r format_decimal(permuted_direct_diagnostics$mean_off_diagonal, 3)`, and the
off-diagonal entries range from
`r format_decimal(permuted_direct_diagnostics$minimum_off_diagonal, 3)` to
`r format_decimal(permuted_direct_diagnostics$maximum_off_diagonal, 3)`.
The mean matrix is positive definite, with minimum eigenvalue
`r format_decimal(permuted_direct_diagnostics$minimum_eigenvalue, 3)`, and its
leading eigenvalue explains
`r format_decimal(100 * permuted_direct_diagnostics$leading_eigenvalue_fraction, 1)`%
of the total variance.

```{r permuted-direct-covariance-matrix, fig.width=9.4, fig.height=8.4}
plot_permuted_direct_covariance_matrix()
```

The matrix is not well described by a stationary AR(1) model. It has a clear
near-diagonal component, but neighboring late-time estimates are often around
0.45--0.56, whereas many early-versus-late entries are around 0.10--0.20.
There is therefore both persistent positive covariance and substantial
time-specific heterogeneity.

The numeric matrix is retained below for exact inspection.

```{r permuted-direct-covariance-table, results='asis'}
render_scrollable_table(
  permuted_primary_matrix_table,
  align = paste0("l", paste(rep("r", 16), collapse = "")),
  minimum_width = "1900px",
  digits = 3
)
```

### Selection-threshold sensitivity

```{r permuted-threshold-sensitivity-table, results='asis'}
render_scrollable_table(
  permuted_threshold_sensitivity_table,
  align = "lrrrr",
  minimum_width = "980px",
  digits = 3
)
```

The direct zero-null covariance profile is stable across the three fixed
selection thresholds. At lag 1 it ranges from approximately 0.40 to 0.43;
lag-9 and lag-15 estimates are essentially unchanged while their intervals
widen as the selected set becomes smaller.

The strongest internal conclusion from the permutation experiment is:

> After method-matched independence calibration, pairwise OLS and direct
> zero-null estimation recover nearly identical empirical covariance matrices.
> Within-unit centering recovers the same shape with a modest downward
> long-lag bias caused by unequal SEs. The residual covariance is not
> negligible: it averages about 0.28 off diagonal, is about 0.43 at lag 1, and
> remains about 0.13 at lag 15. Its nonstationary 16-time structure is richer
> than a lag-1-only model.

## Limitations

- The FASH lfdr is a posterior null probability, not an observed true-null
  label; every threshold-selected set can contain dynamic alternatives.
- Selection on the same trajectory can favor small pairwise differences even
  when the underlying unit is a true FASH null.
- The estimator uses the t-adjusted SE stored in the fitted FASH object; its
  moment interpretation treats those SEs as the relevant marginal error
  scales.
- The gene bootstrap conditions on the fitted lfdr and selected variant per
  gene. It does not repeat the full regression and FASH fitting pipeline.
- The observed-data pairwise matrices are intentionally unprojected and each
  has a small negative eigenvalue. Their lag summary is descriptive and is not
  guaranteed to be a valid stationary variogram. In contrast, all three mean
  covariance matrices from the matched permutation analysis are positive
  definite.
- The matched permutation covariance uses 100 randomization replicates. The
  long-lag intervals are consequently wider than the short-lag intervals.
- The independent-time benchmark removes cross-time donor pairing by design;
  it is a calibration device rather than a proposed biological data-generating
  model.
- The within-unit centering comparison is only approximately invariant because
  the SEs differ across times and units. Pairwise OLS and direct zero-null are
  the primary quantitative estimators in the permuted null.
- The permutation identifies the covariance of the time-specific regression
  estimates under the fixed selected units and observed residual blocks. It
  does not separately attribute that covariance to particular biological or
  technical sources.
- This page does not change or rerun the existing R4 sensitivity analysis.

## Reproducibility

Routine builds read the completed cache at
`output/revision_simulations/internal/pairwise_lfdr_threshold_correlation/`
and do not reload the 554-MB FASH object or rerun the gene bootstraps.

```{bash regenerate-pairwise-lfdr-threshold-correlation, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/covariance_estimation/run_pairwise_lfdr_threshold_exploration.R \
  --lfdr-thresholds 0.97,0.96,0.95 \
  --bootstrap-reps 1000 \
  --bootstrap-seed-start 20260821 \
  --output-id pairwise_lfdr_threshold_correlation
```

The focused helper tests can be rerun with:

```{bash test-pairwise-lfdr-threshold-correlation, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/covariance_estimation/test_pairwise_lfdr_threshold_helpers.R
```

The underlying pairwise estimator test can be rerun with:

```{bash test-real-data-correlation-estimators, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/r4_correlated_errors/test_real_data_correlation_estimators.R
```

The matched donor-permutation variogram and covariance summary can be
regenerated with:

```{bash regenerate-permuted-variogram-covariance, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/covariance_estimation/summarize_permuted_variogram_covariance.R
```
