---
title: "Internal experiment: R1 with real-genotype locus blocks"
author: "Ziang Zhang"
date: "2026-08-06"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9,
  fig.height = 6.5,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source("code/revision_simulations/shared/simulation_functions.R")
source(
  "code/revision_simulations/internal/r1_real_genotype/reporting.R"
)
```

<style>
.r1-real-internal-banner {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border: 1px solid #d9a441;
  border-left: 6px solid #d9a441;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.r1-real-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.r1-real-table-scroll table {
  margin-bottom: 1em;
}
.r1-real-inline-path {
  white-space: normal;
  overflow-wrap: anywhere;
  word-break: break-word;
}
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: hidden;
}
</style>

<div class="r1-real-internal-banner">
<strong>Internal experimental page.</strong> This analysis is not linked from
the workflowr homepage and does not replace the retained
<a href="revision_genotype_level_simulation.html">R1 random B-spline page</a>.
It changes only the genotype source; the current R1 page, driver, and cache are
unchanged.
</div>

## Overview

The retained R1 experiment independently simulates each genotype column from a
Binomial distribution. This internal variant asks what happens when those
columns are instead sampled from the YRI genotype dosages used in the paper.
For each Monte Carlo seed, it samples 20 paper gene loci and one
position-contiguous block of 50 common variants per locus, giving the same
$J=1000$ genotype columns and $N=19$ donors as R1. The purpose is to retain
observed dosage distributions and local linkage disequilibrium (LD) while
leaving the R1 expression truth and fitting pipeline fixed.

For donor $i$, synthetic gene-variant unit $j$, and time $t$, the model remains

$$
Y_{ijt} = \alpha_{jt} + G_{ij}\beta_j(t) +
X_i^T\gamma_{jt} + \epsilon_{ijt},
\qquad \epsilon_{ijt} \sim N(0,1),
$$

but $G_{ij}$ is now a real YRI dosage rather than a Binomial draw. As in R1,
the current simulation sets $\alpha_{jt}=0$, includes an intercept in every
fitted regression, uses five standardized PC-like covariates, and generates
the covariate effects and expression errors independently.

## Real-genotype sampling

### Inputs and provenance

The genotype source is the 19-sample dosage field (`DS`) in the paper VCF. The
candidate universe is defined by the named gene-variant pairs in the paper's
`eqtl_summary.rds`. The report validates the cached file names, sizes, and MD5
checksums against the simulation configuration; routine page builds do not
open either raw input.

```{r input-provenance-table, results='asis'}
render_scrollable_table(
  input_provenance_table,
  align = "llrll",
  minimum_width = "980px"
)
```

The VCF sample order is:

`r paste(expected_sample_ids, collapse = ", ")`.

The single VCF pass requested and matched
`r format(extraction_summary$requested_variants, big.mark = ",")` candidate
variants. All were finite, biallelic, polymorphic, and above the observed MAF
threshold after matching.

```{r extraction-table, results='asis'}
render_scrollable_table(
  extraction_table,
  align = "lr",
  minimum_width = "620px"
)
```

### Sampling design

For each seed, genes with at least 80 paper gene-variant pairs are randomly
prioritized. The union of the first 120 candidate genes per seed is extracted
in one VCF pass. Dosage MAF is calculated as

$$
\widehat{\mathrm{MAF}}_v=
\min\!\left\{
\frac{1}{2N}\sum_{i=1}^{N}G_{iv},
1-\frac{1}{2N}\sum_{i=1}^{N}G_{iv}
\right\},
$$

and variants with observed MAF below 0.10 are excluded. The first 20 priority
genes with at least 50 retained variants are selected. Within each gene, the
variants are ordered by chromosome, position, and variant ID, and one random
contiguous 50-variant window is sampled.

```{r real-genotype-sampling-code, eval=FALSE}
pair_ids <- names(readRDS(
  "output/dynamic_eQTL_real/eqtl_summary.rds"
))

real_source <- prepare_real_genotype_source(
  pair_ids = pair_ids,
  vcf_path = "../genotype-data/YRI_genotype.vcf.gz",
  seed_list = c(12345, 22345, 32345, 42345, 52345),
  n_loci = 20,
  variants_per_locus = 50,
  candidate_genes_per_seed = 120,
  minimum_raw_variants = 80,
  maf_min = 0.10
)
real_sample <- sample_real_genotype_blocks(real_source, seed = 12345)
G <- real_sample$G
```

```{r sampling-design-table, results='asis'}
render_scrollable_table(
  sampling_design_table,
  align = "ll",
  minimum_width = "760px"
)
```

The observed MAF across the 5,000 sampled columns ranges from
`r format_decimal(maf_range[1])` to `r format_decimal(maf_range[2])`.
Across the 100 seed-by-block selections there are 98 unique genes. The two
repeats below are independent-seed random coincidences and were retained rather
than resampled.

```{r repeated-loci-table, results='asis'}
render_scrollable_table(
  repeated_loci_table,
  align = "llr",
  minimum_width = "680px"
)
```

## LD diagnostics

The seed-12345 heatmap shows the signed correlation among all 1,000 dosage
columns. White block boundaries mark the 20 sampled loci. Strong diagonal
blocks are the local LD structure this experiment was designed to preserve.

```{r real-genotype-ld-heatmap, fig.width=9.5, fig.height=8.3}
plot_real_genotype_ld_heatmap()
```

Within each seed, all within-locus variant pairs are summarized. As a negative
control, the same number of variant pairs is sampled deterministically across
different loci. This is a design diagnostic, not a population-level estimate
of LD decay.

```{r ld-seed-table, results='asis'}
render_scrollable_table(
  ld_seed_table,
  align = "rrrrrrr",
  minimum_width = "1180px"
)
```

```{r ld-summary-table, results='asis'}
render_scrollable_table(
  ld_summary_table,
  align = "lr",
  minimum_width = "760px"
)
```

Across the five samples, the mean pooled within-locus median absolute
correlation is `r format_decimal(mean_within_ld)`, compared with
`r format_decimal(mean_across_ld)` for matched across-locus pairs. The average
within-minus-across gap is `r format_decimal(mean_ld_gap)`, and
`r format_decimal(100 * mean_high_ld_fraction, 1)`% of within-locus pairs have
$|r|\geq 0.8$.

```{r real-genotype-ld-distribution, fig.width=9, fig.height=6.2}
plot_real_genotype_ld_distribution()
```

These checks establish that the sampled genotype matrices contain substantial
local LD. They do not establish that the synthetic expression outcomes have
all dependence structures present in the paper data; that distinction is
addressed in the limitations below.

## R1 expression truth and inference

Only the genotype source changes. Each dataset still has 19 donors, 16 time
points, 1,000 gene-variant units, and five covariates. The exact truth mixture
is:

- 200 dynamic eQTLs with random cubic B-spline effects;
- 400 non-dynamic eQTLs with constant effects drawn from $N(0,1)$;
- 400 zero-effect variants.

For a dynamic unit,

$$
\beta_j(t)=\mu_j+h_j(t),
\qquad
\mu_j\sim N(0,1),
\qquad
h_j(t)=\sum_{k=1}^{6}b_{jk}B_k(t),
\qquad
b_{jk}\stackrel{\mathrm{iid}}{\sim}N(0,1).
$$

The time-varying deviation $h_j(t)$ is centered over the 16 observed times and
scaled to have maximum absolute value 2. The independent main effect $\mu_j$
is then added, so the dynamic trajectory is not constrained to average zero.
The true dynamic-null proportion remains $\pi_0=0.8$.

```{r real-genotype-r1-engine, eval=FALSE}
out <- run_genotype_level_bspline_eqtl_simulation(
  G = real_sample$G,
  time_grid = 0:15,
  n_covariates = 5,
  class_probs = c(
    dynamic_bspline = 0.20,
    constant = 0.40,
    zero = 0.40
  ),
  expression_noise_sd = 1,
  covariate_effect_sd = 0.5,
  intercept_sd = 0,
  dynamic_amplitude = 2,
  bspline_df = 6,
  bspline_coefficient_sd = 1,
  constant_sd = 1,
  dynamic_main_effect_sd = 1,
  alpha = 0.05,
  seed = 12345,
  num_basis = 20,
  num_cores = 4,
  exact_class_counts = TRUE,
  apply_t_se_correction = TRUE,
  scenario = "r1_real_genotype_locus_blocks_random_bspline",
  save_outputs = FALSE,
  verbose = FALSE
)
```

At every time point, the genetic effect and its standard error are estimated
from `Y ~ G + PC1 + PC2 + PC3 + PC4 + PC5`. The t-statistic standard-error
correction is identical to R1 and uses

$$
\mathrm{df}=19-7=12,
$$

corresponding to the intercept, genotype, and five covariates. Every fitted
design in the cache has rank 7.

```{r real-genotype-r1-fits, eval=FALSE}
eqtl_summary <- estimate_eqtl_summaries_from_genotypes(
  G = real_sample$G,
  expression = expression_sim$expression,
  covariates = covariates,
  apply_t_se_correction = TRUE
)
datasets <- make_fash_datasets_from_eqtl_summary(
  beta_hat = eqtl_summary$beta_hat,
  se = eqtl_summary$se,
  true_beta = effect_sim$beta_matrix,
  time_grid = 0:15,
  unit_info = effect_sim$unit_info,
  scenario = "r1_real_genotype_locus_blocks_random_bspline"
)

fash_iwp1_raw <- fashr::fash(
  Y = "y",
  smooth_var = "x",
  S = "sd",
  data_list = datasets,
  order = 1,
  grid = default_revision_grid(),
  num_basis = 20,
  penalty = 10,
  num_cores = 4,
  verbose = FALSE
)
fash_iwp1_bf <- fashr::BF_update(fash_iwp1_raw, plot = FALSE)

fash_linear_raw <- fit_simplified_fash(
  datasets = datasets,
  sigma_beta = 1,
  estimate_sigma = FALSE,
  scale_time = TRUE
)
fash_linear_bf <- BF_update_simplified_fash(fash_linear_raw)
```

The reported curves and tables average five independent genotype samples and
expression simulations with seeds `12345`, `22345`, `32345`, `42345`, and
`52345`. The complete seed-12345 object is used only for representative
trajectories.

## Example simulated eQTLs

The panels show three zero-effect units, three constant eQTLs, and three random
B-spline dynamic eQTLs. Points are per-time genetic effect estimates, bars show
two corrected standard errors, and red lines are the true effects.

```{r example-trajectories, fig.width=13, fig.height=11}
plot_genotype_eqtl_examples(
  out,
  classes = c("zero", "constant", "dynamic_bspline"),
  n_per_class = 3,
  seed = 1,
  true_curve_n = 500
)
```

## Comparison 1: IWP versus linear FASH

This comparison changes only the alternative prior. `FASH-IWP1` uses the
first-order IWP alternative. `FASH-linear` uses the simplified alternative
$\beta(t)=\beta t$ with fixed $\sigma_\beta=1$. Both raw and BF-corrected fits
are included.

### Estimated dynamic-null proportion

The table reports the fitted prior weight on each method's constant component.
The true dynamic-null proportion is $\pi_0=0.8$.

```{r pi0-table, results='asis'}
render_scrollable_table(
  pi0_table,
  align = "lll",
  minimum_width = "720px"
)
```

### Results at alpha 0.05

```{r fash-alpha-table, results='asis'}
render_scrollable_table(
  fash_table,
  align = "lrrr",
  minimum_width = "1020px"
)
```

### Power as a function of alpha

```{r fash-power}
plot_mc_alpha_curves(
  fash_alpha,
  metric = "power",
  title = "Real-genotype R1: IWP versus linear FASH power",
  subtitle = "Five seeds; 20 paper loci x 50 variants; N = 19, J = 1000",
  style_profile = "combined"
)
```

### Empirical FDR as a function of alpha

```{r fash-fdr}
plot_mc_alpha_curves(
  fash_alpha,
  metric = "fdr",
  title = "Real-genotype R1: IWP versus linear FASH empirical FDR",
  subtitle = "Five seeds; 20 paper loci x 50 variants; N = 19, J = 1000",
  legend_position = "topleft",
  style_profile = "combined"
)
```

At $\alpha=0.05$, IWP-Raw and IWP-BF have mean power
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-IWP1-Raw", "mean_power"))`
and
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-IWP1-BF", "mean_power"))`,
compared with
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-linear-Raw", "mean_power"))`
and
`r format_decimal(get_mc_metric(fash_alpha_005, "FASH-linear-BF", "mean_power"))`
for the corresponding linear fits. The IWP empirical FDR estimates are
`r fash_iwp_raw_fdr_interval` for Raw and
`r fash_iwp_bf_fdr_interval`
for BF. In these five samples, the nonlinear IWP alternative retains the same
substantial power advantage over the linear alternative seen in R1, while BF
is more conservative.

## Comparison 2: IWP FASH versus direct interaction tests

The direct baselines fit individual-level expression using linear or
linear-plus-quadratic genotype-by-time interactions and use permutation-based
eFDR. As in R1, the direct eFDR calculation is supplied with the true
$\pi_0=0.8$ and uses 100 donor-level permutations per replicate.

```{r real-genotype-direct-efdr, eval=FALSE}
out <- add_direct_interaction_efdr_results_to_genotype_output(
  out = out,
  n_permutations = 100,
  alpha = 0.05,
  seed = seed + 10000L,
  lambda = 0.5,
  pi0_method = "conservative",
  true_pi0 = 0.8,
  include_true_pi0 = TRUE,
  permute_covariates_with_expression = TRUE,
  num_cores = 4,
  overwrite = TRUE,
  verbose = FALSE
)
```

### Results at alpha 0.05

```{r direct-alpha-table, results='asis'}
render_scrollable_table(
  direct_table,
  align = "lrrr",
  minimum_width = "1060px"
)
```

### Power as a function of alpha

```{r direct-power}
plot_mc_alpha_curves(
  direct_alpha,
  metric = "power",
  title = "Real-genotype R1: IWP FASH versus direct interaction power",
  subtitle = "Five seeds; direct eFDR uses true pi0 = 0.8 and B = 100 permutations",
  style_profile = "combined"
)
```

### Empirical FDR as a function of alpha

```{r direct-fdr}
plot_mc_alpha_curves(
  direct_alpha,
  metric = "fdr",
  title = "Real-genotype R1: IWP FASH versus direct interaction empirical FDR",
  subtitle = "Five seeds; direct eFDR uses true pi0 = 0.8 and B = 100 permutations",
  legend_position = "bottomright",
  style_profile = "combined"
)
```

At $\alpha=0.05$, IWP-BF has mean power
`r format_decimal(direct_iwp_bf_power)`, compared with
`r format_decimal(direct_linear_power)` for the direct linear test and
`r format_decimal(direct_quadratic_power)` for the direct quadratic test.
Their respective mean empirical FDR estimates are
`r format_decimal(direct_iwp_bf_fdr)`,
`r format_decimal(direct_linear_fdr)`, and
`r format_decimal(direct_quadratic_fdr)`.

## Internal interpretation

This five-seed pilot supports a focused conclusion. Replacing independent
Binomial genotype columns with common YRI dosages sampled in local blocks does
not change the qualitative R1 method ordering: IWP FASH remains more powerful
than linear FASH and the direct interaction baselines for the broad random
B-spline truth used here. The BF-updated IWP fit has mean power 0.902 and mean
empirical FDR 0.025 at $\alpha=0.05$.

The LD diagnostic confirms that the external genotype matrices contain the
intended local dependence: within-locus median $|r|$ exceeds its matched
across-locus reference in every seed. This supports the narrow claim that the
regressions use real dosage and local-LD patterns. With only five Monte Carlo
replicates, the intervals describe Monte Carlo variation in this pilot and do
not justify a general claim for other populations, MAF filters, locus widths,
or LD architectures.

## Important limitation: genotype-only substitution

This page implements the selected genotype-only design. Although each block
contains 50 real variants from one paper gene locus, the simulation creates a
separate synthetic expression outcome for every gene-variant unit. The 50
variants in a block therefore do **not** test the same realized expression
vector for that gene. Expression errors and effect trajectories are generated
independently across units, exactly as in R1.

Consequently, this experiment preserves the real genotype dosage distribution
and local LD in $G$, but it does not reproduce shared-expression dependence,
tag-SNP competition, or correlated summary statistics created when many SNPs
are tested against one observed gene-expression outcome. A separate
shared-expression simulation would be required to study those features. The
results here should be interpreted as a real-design-matrix sensitivity check,
not as a full reconstruction of the paper's multi-SNP-per-gene analysis.

## Reproducibility

Routine builds read the completed cache at

```text
output/revision_simulations/internal/r1_real_genotype_locus_blocks_pilot5/
```

and do not scan the VCF, reload the paper pair universe, or rerun the
simulation. The sampling helper, driver, and cache-validation code are in
<code class="r1-real-inline-path">code/revision_simulations/internal/r1_real_genotype/</code>.

The full five-seed cache can be regenerated with:

```{bash regenerate-real-genotype-r1-cache, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/r1_real_genotype/run_real_genotype_mc_replication.R \
  --n-loci 20 \
  --variants-per-locus 50 \
  --maf-min 0.10 \
  --candidate-genes-per-seed 120 \
  --minimum-raw-variants 80 \
  --n-covariates 5 \
  --noise-sd 1 \
  --dynamic-main-effect-sd 1 \
  --efdr-permutations 100 \
  --seed-list 12345,22345,32345,42345,52345 \
  --output-id r1_real_genotype_locus_blocks_pilot5 \
  --num-cores 4
```

The deterministic sampling tests can be rerun with:

```{bash test-real-genotype-r1-sampling, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/r1_real_genotype/test_real_genotype_sampling.R
```
