---
title: "Internal sensitivity analysis: correlated FASH likelihoods"
author: "Ziang Zhang"
date: "2026-08-07"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 10,
  fig.height = 5.8,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(paste0(
  "code/revision_simulations/internal/correlated_likelihood_sensitivity/",
  "reporting.R"
))
```

<style>
.correlated-likelihood-banner {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border: 1px solid #d9a441;
  border-left: 6px solid #d9a441;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.correlated-likelihood-result {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.correlated-likelihood-caution {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #d55e00;
  border-radius: 4px;
  background: #fff4ec;
  color: #59331e;
}
.correlated-likelihood-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.correlated-likelihood-table-scroll table {
  margin-bottom: 1em;
}
.correlated-likelihood-inline-path {
  white-space: normal;
  overflow-wrap: anywhere;
  word-break: break-word;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
pre {
  max-width: 100%;
  overflow-x: auto;
}
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: hidden;
}
</style>

<div class="correlated-likelihood-banner">
<strong>Internal sensitivity page.</strong> This page is intentionally
unlinked. It does not modify or replace the R4 analysis, the covariance-screen
page, the existing one-variant-per-gene page, the manuscript, or the response
letter.
</div>

## Question and answer

On the exact one-random-variant-per-gene subset used to estimate the two
empirical time-correlation matrices, how much does the FASH result change if
the likelihood uses either matrix instead of conditionally independent time
points?

The three raw fits are:

1. **Diagonal SE:** the current conditional-independence likelihood;
2. **C1:** the empirical z-score correlation from the
   `r format_integer(c1_estimating_pairs)` Screen-1 pairs satisfying
   $\max_t |z_j(t)|<2$;
3. **C2:** the empirical z-score correlation from the
   `r format_integer(c2_estimating_pairs)` more selective Screen-2 pairs also
   satisfying $|\sqrt{16}\,\overline z_j|<2$.

All three fits use the same `r format_integer(expected_n_units)` gene--variant
pairs, one pair per gene; the same beta estimates, adjusted standard errors,
time grid, PSD grid, basis settings, and `penalty = 10`; and differ only in the
likelihood error correlation.

<div class="correlated-likelihood-result">
<strong>Main numerical result.</strong> The sensitivity is large. The raw
diagonal fit estimates $\widehat\pi_0=
`r format_decimal(raw_diagonal_pi0, 4)`$ and makes
`r format_integer(raw_diagonal_calls)` cumulative-lfdr calls at
$\alpha=0.05$. C1 raises $\widehat\pi_0$ to
`r format_decimal(raw_c1_pi0, 4)` and makes no calls; C2 estimates
$\widehat\pi_0=`r format_decimal(raw_c2_pi0, 4)`$, assigns lfdr 1 to every
unit, and also makes no calls. After BF adjustment, the diagonal fit has
$\widehat\pi_0=`r format_decimal(bf_diagonal_pi0, 4)`$ and
`r format_integer(bf_diagonal_calls)` calls, whereas C1 has
$\widehat\pi_0=`r format_decimal(bf_c1_pi0, 4)`$ and no calls.
</div>

<div class="correlated-likelihood-caution">
<strong>C2 BF result is unavailable, not an all-null BF-adjusted fit.</strong>
The raw C2 fit assigns exactly zero total prior mass to all non-null
components. `BF_update()` requires a conditional mixture over those non-null
components, so its normalization is undefined. The attempted update and its
warning are retained in the audit cache, but the returned unchanged raw object
is deliberately not reported as a successful BF-adjusted result.
</div>

This experiment therefore does **not** support a robustness claim with respect
to C1 or C2. It shows strong sensitivity if either empirical matrix is treated
as the true residual correlation. It also does not establish that C1 or C2 is
the true residual correlation: both are estimated from selected z-score
trajectories in the same observed data and can retain biological signal or
selection-induced structure.

## Fixed thinning and empirical correlation inputs

The thinning was performed before either screen by drawing one tested variant
uniformly within each gene with seed `r configuration$thinning_seed`. The
retained universe has exactly `r format_integer(configuration$n_genes)` unique
genes and `r format_integer(configuration$n_units)` unique gene--variant pairs.
Thus the comparison changes neither the units nor their order.

C1 and C2 below are the original empirical correlation matrices. Both are
positive definite without nearest-PD projection.

```{r correlation-diagnostics, results='asis'}
render_scrollable_table(
  correlation_diagnostics_display,
  align = "lrrrr",
  minimum_width = "940px"
)
```

```{r correlation-heatmaps, fig.width=11.5, fig.height=5.4, fig.cap="The two empirical z-score correlation matrices supplied to the sensitivity fits. Both panels use the same off-diagonal color scale."}
plot_correlation_heatmaps()
```

The matrices are not external controls and are not estimated from independent
replicate data. C1 uses a conventional mashr-style maximum-z screen. C2
additionally screens the independence-scaled mean z score. The extra C2 rule
removes obvious common effects, but it also conditions on a function of the
same time trajectory whose correlation is being estimated.

## Likelihood construction

Let $\widehat{\boldsymbol\beta}_j$ denote the 16-vector of estimated effects
for unit $j$, and let

$$
D_j=\operatorname{diag}(s_{j1},\ldots,s_{j,16})
$$

contain its time-specific adjusted standard errors. For a shared correlation
matrix $C$, the sensitivity likelihood uses

$$
\widehat{\boldsymbol\beta}_j\mid\boldsymbol\beta_j
\sim N\!\left(\boldsymbol\beta_j,\;D_j C D_j\right),
$$

or equivalently the unit-specific precision

$$
\Omega_j
=\left(D_j C D_j\right)^{-1}
=D_j^{-1}C^{-1}D_j^{-1}.
$$

The implementation passes this complete $\Omega_j$ to `fashr`. Passing only
$C^{-1}$ would incorrectly discard the fact that standard errors vary across
units and time points.

```{r precision-construction, eval=FALSE}
correlation_precision <- solve(C)

omega <- lapply(seq_len(nrow(adjusted_se)), function(j) {
  inverse_se <- 1 / adjusted_se[j, ]
  correlation_precision * tcrossprod(inverse_se)
})

correlated_fit <- fashr::fash(
  data = data_list,
  Omega = omega,
  likelihood = "gaussian",
  num_basis = 20,
  betaprec = 0,
  order = 1,
  pred_step = 1,
  psd_grid = psd_grid,
  penalty = 10
)
```

### Identity-path validation

Before either correlated fit was run, 24 real-data units were evaluated twice:
once using the ordinary diagonal-SE path and once using the full-precision path
with $C=I$. The complete 52-component likelihood rows were compared after
subtracting each row's null-component log likelihood, because a component-free
row constant cannot affect empirical-Bayes weights or posterior probabilities.

```{r identity-validation, results='asis'}
render_scrollable_table(
  identity_validation_display,
  align = "rrrr",
  minimum_width = "980px"
)
```

The agreement is at floating-point precision, confirming that the correlated
route reduces to the existing diagonal likelihood when $C=I$.

## Fitted results

```{r method-stage-summary, results='asis'}
render_scrollable_table(
  method_stage_summary_display,
  align = "llllrrrr",
  minimum_width = "1320px"
)
```

The diagonal raw fit reuses the exact selected likelihood rows from the current
full-data FASH object and re-estimates the mixture on the thinned universe.
The C1 and C2 likelihoods must be recomputed because their precision matrices
are non-diagonal. Their elapsed times were approximately 26 and 36 minutes,
respectively.

## Prior-weight sensitivity

```{r prior-weight-curves, fig.width=11.5, fig.height=5.5, fig.cap="Estimated prior weights on the common FASH PSD grid. The horizontal axis shows the square root of the PSD-grid value. C2 has no BF-adjusted curve because its raw non-null prior mass is exactly zero."}
plot_prior_weights()
```

```{r prior-pairwise-metrics, results='asis'}
render_scrollable_table(
  prior_pairwise_metrics_display,
  align = "llrrrr",
  minimum_width = "1160px"
)
```

At the raw stage, the total-variation distance between the diagonal prior and
either correlated prior is about 0.517. C1 and C2 themselves are much closer:
their prior total-variation distance is about 0.010 because both concentrate
almost all mass at the null. BF adjustment reduces the diagonal-versus-C1
prior total variation to about 0.063, but does not make the two fitted
analyses equivalent.

## Unit-level lfdr sensitivity

```{r lfdr-comparison-scatterplots, fig.width=11.5, fig.height=10.2, fig.cap="Paired unit-level lfdr values. Each panel uses exactly the same 6,362 gene-variant pairs. C2 raw lfdr is identically 1; C2 BF is unavailable."}
plot_lfdr_comparisons()
```

```{r lfdr-pairwise-metrics, results='asis'}
render_scrollable_table(
  lfdr_pairwise_metrics_display,
  align = "llrrrrrrrr",
  minimum_width = "1660px"
)
```

For raw diagonal versus C1, the lfdr Spearman correlation is 0.773 but the
mean absolute difference is 0.508: ranking is partly retained while absolute
calibration shifts sharply toward the null. Because C2 raw lfdr is constant at
1, Pearson and Spearman correlations with C2 are undefined rather than zero.

After BF adjustment, diagonal versus C1 has Spearman correlation 0.789 and
mean absolute difference 0.089. Thus BF adjustment substantially reduces the
average lfdr displacement, but it does not restore the diagonal discoveries:
C1 still makes no calls at the same cumulative-lfdr threshold.

```{r lfdr-difference-distributions, fig.width=11.5, fig.height=9.2, fig.cap="Distributions of paired lfdr differences. Positive values mean that the correlated-likelihood method assigns greater posterior null probability than the reference method."}
plot_lfdr_difference_distributions()
```

### Most discrepant units

The following table shows the five largest absolute lfdr changes for each
available pairwise comparison. These are paired changes, not a comparison of
different sampled variants.

```{r top-lfdr-discrepancies, results='asis'}
render_scrollable_table(
  top_discrepancy_display,
  align = "llrllrrr",
  minimum_width = "1460px"
)
```

## Discovery consequences

```{r discovery-counts, fig.width=10.8, fig.height=5.3, fig.cap="Numbers of cumulative-lfdr discoveries at alpha 0.05 on the fixed thinned universe. The C2 BF result is unavailable and is labeled accordingly."}
plot_discovery_counts()
```

The raw diagonal analysis calls 268 of the 6,362 pairs. Both correlated raw
fits call none. The BF-adjusted diagonal analysis calls 57, whereas the
available BF-adjusted C1 analysis calls none. Therefore the discovery-set
Jaccard is zero for every diagonal-versus-correlated comparison with an
available result. C1 and C2 raw both make no calls, so their empty-set Jaccard
is defined as one by the analysis code.

## Why the C2 BF update is unavailable

```{r bf-update-status, results='asis'}
render_scrollable_table(
  bf_update_status_display,
  align = "lrll",
  minimum_width = "1220px"
)
```

`BF_update()` first needs a conditional distribution over the alternative
components, obtained by normalizing their fitted raw prior weights. For C2,

$$
\sum_{k:\,\mathrm{PSD}_k>0}\widehat\pi_k=0,
$$

so this normalization is $0/0$. The package warning is therefore a consequence
of an undefined conditional alternative mixture, not evidence that the raw C2
object was successfully BF-adjusted.

For the diagonal and C1 fits, row-centering the likelihood before the update
is numerically safe and reproduces the original successful updates extremely
closely: their maximum lfdr differences are
`r paste(format_scientific(bf_rebuild_validation$lfdr_maximum_difference),
collapse = " and ")`, respectively. The final cache stores only these two
successful BF-adjusted fits.

## Interpretation

The direct empirical finding is clear:

> Conditional on treating C1 or C2 as the likelihood's residual correlation,
> the fitted null weight, unit-level lfdr calibration, and discovery set change
> substantially relative to the diagonal analysis. BF adjustment attenuates
> the average diagonal-versus-C1 lfdr difference, but it does not recover any
> C1 discoveries.

The causal interpretation is not clear. A null residual-correlation matrix is
supposed to describe cross-time dependence in estimation noise conditional on
the true trajectory. C1 and C2 are instead empirical correlations of selected
observed z-score vectors. Even after one-variant-per-gene thinning and
null-like screening, those vectors may contain constant or dynamic effects,
gene-specific artifacts, and structure induced by selecting on the same z
scores. Supplying such a matrix to the likelihood can make smooth trajectories
look more compatible with noise and can absorb signal into the null model.

Accordingly, this page should be read as a **stress test of two plausible but
not identified covariance inputs**, not as evidence that either correlated fit
is better calibrated than the diagonal fit.

## Limitations

- C1 and C2 are estimated from the same thinned data to which they are later
  applied; there is no independent covariance-estimation sample.
- The maximum-z and mean-z screens do not guarantee that every retained pair
  has a zero true trajectory.
- One variant per gene avoids repeated variants within a gene but does not
  remove LD between variants assigned to different genes.
- Correlation is assumed shared across genes, while adjusted marginal standard
  errors remain unit- and time-specific through $D_j$.
- This experiment compares three likelihood assumptions at one fixed thinning
  seed. It does not propagate uncertainty in the estimated C1 or C2 matrices.
- C2 BF adjustment is mathematically undefined for the fitted raw prior, so no
  valid three-way BF comparison exists for this run.

## Source provenance

The analysis cache records immutable input metadata and SHA-256 hashes. Routine
page builds validate these recorded values and current file sizes but do not
reload the two full FASH objects.

```{r source-provenance, results='asis'}
render_scrollable_table(
  source_provenance_display,
  align = "llrl",
  minimum_width = "1760px"
)
```

## Reproducibility

Routine builds read only:

<span class="correlated-likelihood-inline-path">`output/revision_simulations/internal/correlated_likelihood_sensitivity/`</span>

To rerun the full likelihood analysis without overwriting the retained cache,
use a new output ID. On the current machine the two correlated fits together
required about 62 minutes with eight cores.

```{bash regenerate-correlated-likelihood-sensitivity, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/correlated_likelihood_sensitivity/run_correlated_likelihood_sensitivity.R \
  --num-cores 8 \
  --validation-units 24 \
  --thinning-seed 20260811 \
  --output-id correlated_likelihood_sensitivity_reproduction
```

The focused tests are:

```{bash test-correlated-likelihood-sensitivity, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/correlated_likelihood_sensitivity/test_correlated_likelihood_helpers.R

Rscript --vanilla \
  code/revision_simulations/internal/correlated_likelihood_sensitivity/test_reporting.R
```

The internal page can be rebuilt from the retained cache with:

```{bash rebuild-correlated-likelihood-page, eval=FALSE}
Rscript --vanilla -e \
  'workflowr::wflow_build(
    "analysis/revision_internal_correlated_likelihood_sensitivity.rmd"
  )'
```
