---
title: "Revision Real-data Analysis R6: enhancer enrichment of FASH dynamic eQTLs"
author: "Ziang Zhang"
date: "2026-08-26"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 2
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 8.6,
  fig.height = 6.4,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(file.path(
  "code",
  "revision_simulations",
  "internal",
  "fash_strober_enhancer_comparison",
  "reporting.R"
))
```

<style>
html,
body {
  max-width: 100%;
  overflow-x: hidden;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
pre {
  max-width: 100%;
  overflow-x: auto;
}
</style>

## Question

Do the dynamic eQTLs that FASH discovers fall in functionally relevant
enhancers? As an orthogonal biological check, we ask whether one lead variant
per discovered gene overlaps established Roadmap enhancer annotations more often
than the variants tested in this study do. We consider Roadmap epigenomes
relevant to pluripotent, germ-layer, cardiac, and related vascular states, and
use the enhancer definition adopted by Strober et al. (2019). The estimates
below are descriptive and exploratory: they are not a fully adjusted functional
enrichment analysis, and they are not evidence that any individual discovery is
correct.

Two sections, both using one lead variant per discovered gene:

* **Section 1** compares current FASH with Strober's linear and
  quadratic/nonlinear analyses.
* **Section 2** compares all FASH lead variants with the subset at rsIDs that
  neither Strober analysis identified.

## Roadmap enhancer annotations

Following Strober et al. (2019), enhancer regions were defined using the Roadmap
15-state ChromHMM enhancer annotations (`Enh`, `EnhG`, `EnhBiv`, and `BivFlnk`)
(Roadmap Epigenomics Consortium et al. 2015). Using an established annotation
procedure avoids introducing an enhancer definition of our own. Thirteen
epigenomes are displayed, in four groups:

```{r panel, echo=FALSE}
panel_table()
```

The iPSC and heart epigenome choices follow Strober et al. (2019); the three
hESC-derived germ-layer epigenomes are included because cardiomyocytes arise
through mesoderm. E065 aorta is a vascular rather than a cardiac tissue, so it
is kept as a separate, related vascular annotation and is never counted among
the cardiac epigenomes. Each tested variant is treated as a single base and
intersected with the published Roadmap enhancer intervals, giving one indicator
per variant per epigenome over all
`r format_integer(background_variant_count)` tested variants.

## Section 1: FASH versus Strober

```{r section1, fig.cap="Figure 1. Enhancer enrichment for current FASH, Strober linear, and Strober quadratic/nonlinear discoveries, using one lead variant per discovered gene. Enhancer regions are the Strober et al. (2019) Roadmap 15-state ChromHMM enhancer states, across thirteen epigenomes grouped as iPSC, germ layer, cardiac, and related vascular. Fold enrichment is the overlap rate in the variant set divided by the overlap rate among the remaining tested variants; the dashed line marks no enrichment. Bars are leave-one-autosome-out jackknife 95% intervals. One cell has too few overlapping variants to estimate and is absent: E065 (Strober linear)."}
plot_method_comparison()
```

FASH's `r set_size("current_lead")` lead variants are enriched in
`r count_above_one("current_lead")` displayed annotations, over a narrow range of `r range_above_one("current_lead")`. The annotations whose intervals
lie entirely above 1 are `r labels_above_one("current_lead")` — that is, two
iPSC epigenomes, all three germ layers, and left ventricle. The three remaining
cardiac annotations point the same way with intervals that reach or cross 1
(E083 `r fold_value("current_lead", "E083")`,
E105 `r fold_value("current_lead", "E105")`,
E104 `r fold_value("current_lead", "E104")`).

Both Strober lead sets point the same way. Their intervals are wider, which is
what their smaller size would predict — `r set_size("linear_lead")` and
`r set_size("quadratic_lead")` lead variants against FASH's
`r set_size("current_lead")` — and where they do clear 1 it is at annotations
drawn from the same groups: `r labels_above_one("linear_lead")` for linear and
`r labels_above_one("quadratic_lead")` for quadratic/nonlinear. The three
discovery sets differ in size, membership and gene composition, so this is not a
formal comparison between methods.

The enrichment is spread across pluripotent, germ-layer and cardiac annotations
rather than concentrated in any one of them, and within each group some members
clear 1 while others do not. These lead-per-gene estimates therefore support the
general statement that lead variants identified by FASH are enriched in several
established Roadmap enhancer annotations relevant to this differentiation
system, and they do not support any claim of developmental-stage or tissue
specificity. E065 aorta, the sparsest annotation in the panel and the only
related vascular one, carries the single below-1 point estimate,
`r describe_fold("current_lead", "E065")`, and rests on the fewest overlapping
variants, so it should be read as neither support nor contradiction.

## Section 2: FASH-specific discoveries

An rsID is counted as FASH-only if it appears in neither Strober analysis,
regardless of which gene it was paired with; this is the conservative reading,
since an enhancer annotation is a property of a variant rather than of a
gene-variant pair. FASH's lowest-lfdr variant is chosen for each gene first and
the exclusion is applied afterwards, so a gene whose best variant Strober also
found drops out rather than falling back to a weaker one. That leaves
`r set_size("current_only_lead")` of FASH's `r set_size("current_lead")` lead
variants.

```{r section2, fig.cap="Figure 2. Enhancer enrichment for FASH lead variants at rsIDs that neither Strober analysis identified (blue), against all FASH lead variants as a reference (grey), using one lead variant per discovered gene. Same thirteen Roadmap epigenomes, same Strober et al. (2019) enhancer definition, and same estimator as Figure 1. Bars are leave-one-autosome-out jackknife 95% intervals."}
plot_fash_only()
```

The enrichment survives the exclusion essentially intact. The same
`r count_above_one("current_only_lead")` annotations clear 1 —
`r labels_above_one("current_only_lead")` — at
`r range_above_one("current_only_lead")`, against
`r range_above_one("current_lead")` for the unfiltered FASH lead set, and the
point estimates barely move
(E013 `r fold_value("current_only_lead", "E013")` against
`r fold_value("current_lead", "E013")`;
E022 `r fold_value("current_only_lead", "E022")` against
`r fold_value("current_lead", "E022")`;
E095 `r fold_value("current_only_lead", "E095")` against
`r fold_value("current_lead", "E095")`).

So the enhancer-enrichment pattern is largely retained among FASH lead variants
at rsIDs that neither Strober analysis identified. Taken with Section 1, this is
orthogonal support for the regulatory relevance of these discoveries as a set,
resting on externally established Roadmap and Strober annotations rather than on
anything fitted here. It is not validation of any individual discovery, and it
does not establish specificity to a particular differentiation stage or tissue.

## Methods

**Background.** All `r format_integer(background_variant_count)` variants
tested in this study that lie on an autosome and participate in at least one
tested gene-variant pair. No external reference panel enters, so the background
is exactly the cis-eQTL testing universe from which the discoveries were drawn.

**Discovery sets and lead variants.** FASH uses cumulative lfdr at FDR 0.05
after the BF correction; both Strober analyses use their reported eFDR 0.05,
with `non_linear_dynamic_eqtls_5_pc.txt` labelled quadratic/nonlinear
throughout. Every set displayed here keeps, for each discovered gene, one lead
variant: the variant with the lowest lfdr (FASH) or the smallest p-value
(Strober). Using one lead variant per gene reduces the repeated contribution of
multiple correlated variants associated with the same gene.

**Enhancer annotation.** Roadmap Epigenomics 15-state ChromHMM segmentations
for the thirteen epigenomes above, with enhancer regions defined by the states
`Enh`, `EnhG`, `EnhBiv` and `BivFlnk` as in Strober et al. (2019). The
iPSC and heart epigenome choices follow the same source.

**Fold enrichment.** For a variant set and an enhancer annotation,

$$\text{fold enrichment} = \frac{\text{proportion of selected variants overlapping the enhancer}}{\text{proportion of the remaining tested variants overlapping the enhancer}}$$

The background excludes the set being tested, so a set is never part of its own
comparator. No matched controls are used: the purpose here is a simple
exploratory check, and every background variant already comes from the same
testing universe. A set-annotation pair with fewer than `r minimum_overlap`
overlapping variants is reported as not estimable rather than estimated.

**Uncertainty.** Uncertainty was assessed using a leave-one-autosome-out
jackknife to account approximately for local dependence among variants. All
intervals shown are 95% jackknife intervals; nominal p-values are available in
the output CSVs but are not used on this page.

## Reproducibility

The figures on this page read the validated thirteen-epigenome indicator table
and estimates built by
[`run_roadmap_enhancer_exploration.R`](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/internal/roadmap_enhancer_exploration/run_roadmap_enhancer_exploration.R),
through the shared estimator in
[`enrichment_api.R`](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/internal/fash_strober_enhancer_comparison/enrichment_api.R).
The annotation intervals are not rebuilt here. Exactly the calls below reproduce
every estimate shown above:

```{r api-demo, eval=FALSE}
source(
  "code/revision_simulations/internal/fash_strober_enhancer_comparison/enrichment_api.R"
)
source(
  "code/revision_simulations/internal/roadmap_enhancer_exploration/roadmap_enhancer_exploration_helpers.R"
)

# 1. Background: every autosomal variant tested in this study.
background <- load_background()

# 2. Discovery sets, one lead variant per discovered gene.
sets <- load_variant_sets()[c(
  "current_lead", "linear_lead", "quadratic_lead", "current_only_lead"
)]

# 3. The thirteen Roadmap epigenomes, as the indicator table built by the
#    exploration run script.
annotations <- readRDS(file.path(
  roadmap_output_directory(), "roadmap_enhancer_annotation_matrix.rds"
))

# 4. The Strober four-state enhancer definition, one column per epigenome.
strober <- grep(
  "[Strober EnhG+Enh+EnhBiv+BivFlnk]", names(annotations),
  fixed = TRUE, value = TRUE
)

# 5. Fold enrichment against the tested-variant background, with
# 6. leave-one-autosome-out jackknife 95% intervals.
compute_enrichment(
  sets,
  background,
  annotations[, c("variant_id", "chromosome", strober)],
  controls = "background"
)
```

`compute_enrichment()` takes any named list of variant vectors, so substituting
your own sets needs no other change:

```{r api-own-sets, eval=FALSE}
compute_enrichment(
  list(my_hits = readLines("my_variants.txt"), fash_only = sets$current_only_lead),
  background,
  annotations[, c("variant_id", "chromosome", strober)],
  controls = "background"
)
```

Rebuilding the inputs from source — the tested variant universe, then the
discovery sets, then the thirteen Roadmap segmentations and the estimates this
page displays:

```{bash rebuild, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/variant_annotation_enrichment/download_variant_annotations.R
Rscript --vanilla \
  code/revision_simulations/internal/variant_annotation_enrichment/run_variant_annotation_enrichment.R
Rscript --vanilla \
  code/revision_simulations/internal/fash_strober_enhancer_comparison/run_fash_strober_enhancer_comparison.R
Rscript --vanilla \
  code/revision_simulations/internal/roadmap_enhancer_exploration/run_roadmap_enhancer_exploration.R
```

Intermediate data stays under `output/`, which is not tracked in git; the
repository carries code, not data. Rebuilding this page:

```{bash reproduction-command, eval=FALSE}
Rscript --vanilla -e \
  'workflowr::wflow_build("analysis/revision_enhancer_enrichment.rmd")'
```

## References

Roadmap Epigenomics Consortium, Kundaje, A., Meuleman, W., Ernst, J.,
Bilenky, M., Yen, A., et al. (2015). Integrative analysis of 111 reference
human epigenomes. *Nature* **518**, 317–330.
<https://doi.org/10.1038/nature14248>

Strober, B., Elorbany, R., Rhodes, K., Krishnan, N., Tayeb, K., Battle, A., and
Gilad, Y. (2019). Dynamic genetic regulation of gene expression during cellular
differentiation. *Science* **364**, 1287–1290.
<https://doi.org/10.1126/science.aaw0040>
