---
title: "Revision Simulation R3: functional testing under broad and transient effects"
author: "Ziang Zhang"
date: "2026-08-23"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 10,
  fig.height = 8,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source("code/revision_simulations/shared/simulation_functions.R")
source("code/revision_simulations/shared/real_genotype_one_per_gene.R")
source("code/revision_simulations/r3_functional_testing/reporting.R")
```

<style>
.r3-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.r3-table-scroll table {
  margin-bottom: 1em;
}
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: hidden;
}
</style>

## Overview

This simulation evaluates FASH functional testing after global dynamic-eQTL
discovery. It uses two non-IWP truth mechanisms matched on sample size,
variant classes, genetic main effects, functional labels, and analysis.

All numerical results on this page come from the completed, versioned Midway3
cache generated with `fashr` 0.1.43 at Git SHA
`bf223df75da6e41ae48607a56b4cd12d7c3b24e7`. Before formatting any result,
the reporting layer requires the completion flag and manifest, validates the
package, result identity, Middle definition, and genotype provenance, and
checks that every seed uses the exact same gene-variant pairs and canonical
genotype content as the formal R1 analysis. The separate local validator also
checks every artifact and source SHA-256 before the page is rebuilt.

For donor $i$, variant $j$, and time $t$,

$$
Y_{ijt} = \alpha_{jt} + G_{ij}\beta_j(t) +
X_i^T\gamma_{jt} + \epsilon_{ijt},
\qquad \epsilon_{ijt}\sim N(0,1).
$$

The genotype matrix is taken from the real YRI data used in the manuscript.
For each simulation seed, one tested cis variant is sampled uniformly from
each of the `6,362` tested genes, and its observed dosage vector is retained
for the same `19` donors. A tested variant assigned to multiple genes may
therefore appear more than once. R1, R2, R3A, and R3B use the exact same
sampled gene-variant pairs and genotype matrix for a given seed.

The five columns of $X$ remain simulated independently from $N(0,1)$ and are
standardized across donors. Independently for each covariate, gene, and time,
$\gamma_{kjt}\sim N(0,0.5^2)$. The current simulation sets $\alpha_{jt}=0$;
each fitted regression still includes an intercept. Effect trajectories,
covariates, covariate effects, and expression noise remain simulated.

```{r r3-simulate-genotypes-and-covariates, eval=FALSE}
seed <- 12345
component_seeds <- revision_component_seeds(seed)

genotype_cache <- readRDS(
  "output/revision_simulations/shared/real_genotype_one_per_gene_J6362_pilot5/genotype_samples.rds"
)
genotype_sample <- genotype_cache$samples[[as.character(seed)]]
G <- genotype_sample$G
covariates <- simulate_covariate_matrix(
  n_donors = 19,
  n_covariates = 5,
  seed = component_seeds[["covariates"]]
)
```

### Real-genotype design

<div class="r3-table-scroll">

```{r genotype-sampling-table}
knitr::kable(
  genotype_sampling_table,
  align = "l",
  caption = paste(
    "Table 1. Per-seed genotype sampling summary. Repeated assignments",
    "refer to the same tested variant selected for more than one gene."
  )
)
```

</div>

Each replicate contains `19` donors, `16` time points, five PC-like
covariates, `1,272` dynamic eQTLs, `2,545` constant eQTLs, and `2,545`
zero-effect units. Truth classes are assigned within empirical-MAF deciles so
that the dynamic, constant, and zero-effect groups have closely matched MAF
distributions.

<div class="r3-table-scroll">

```{r truth-maf-balance-table}
knitr::kable(
  truth_maf_balance_table,
  align = "l",
  caption = "Table 2. Empirical MAF balance across truth classes over five simulation seeds."
)
```

</div>

For every dynamic candidate,

$$
\beta_j(t)=\mu_j+h_j(t),
\qquad \mu_j\sim N(0,1).
$$

The `1,272` dynamic eQTLs are divided exactly across six preassigned truth
cells (`212` per cell): early, middle, or late crossed with switch or
non-switch. Each
candidate draws a new $\mu_j$ from $N(0,1)$; rejection sampling retains it only
when the final trajectory $\mu_j+h_j(t)$ satisfies its assigned cell.

### R3A: broad random B-spline effects

The deviation $h_j(t)$ is a cubic B-spline with six iid $N(0,1)$ coefficients.
It is centered over the 16 observed times and scaled to maximum absolute value
`2` before adding $\mu_j$.

### R3B: compact raised-cosine effects

The deviation $h_j(t)$ contains one, two, or three compact raised-cosine peaks
with half-width `1.5`. Primary peak centers are sampled from `1.5`--`2.5` for
Early, `4.5`--`10.5` for Middle, and `12.5`--`13.5` for Late. Secondary peak
centers use the same three aligned ranges, their amplitudes are random, and
the deviation is scaled to centered RMS `0.9` before adding $\mu_j$. These
ranges prevent a full raised-cosine peak from straddling a revised timing
boundary solely because its center was sampled too close to that boundary.

### Truth definitions

Early, middle, and late use the same supremum-contrast functionals as the
manuscript, with the revised temporal partition defined exactly as Early
$t\leq 3$, Middle $3<t<12$, and Late $t\geq 12$. On the dense evaluation grid
`seq(0, 15, by = 0.1)`, the Middle functional therefore uses the 89 points
from `3.1` through `11.9`; times `3` and `12` belong only to Early and Late,
respectively. These are continuous-time definitions. The simulation evaluates
their discrete maximum counterparts on the dense grid. Writing
$R_j=\sup_t\beta_j(t)-\inf_t\beta_j(t)$, the scale-adaptive location margin is

$$
m_j=\max(0.1,\,0.1R_j).
$$

The assigned Early, Middle, or Late contrast must be at least $m_j$, and both
competing location contrasts must be at most $-m_j$. Thus the same relative
clearance rule is applied symmetrically to all three location categories.

For the switch threshold $c=0.25$,

$$
T_{\mathrm{switch}}\{\beta\}
=
\min\left[
\sup_t\{\beta(t):\beta(t)>0\},
\sup_t\{|\beta(t)|:\beta(t)<0\}
\right]-c.
$$

A switch truth must have $T_{\mathrm{switch}}\geq 0.1$; equivalently, it has
both a positive and a negative effect of magnitude at least `0.35`. A
non-switch truth remains the same sign over the full dense grid. Its clearance
from zero must satisfy

$$
\min_t|\beta_j(t)|\geq\max(0.1,\,0.1R_j).
$$

Thus, a larger-amplitude non-switch curve must remain proportionally farther
from zero. These margins make the simulation labels unambiguous.

```{r r3-simulate-functional-truth, eval=FALSE}
truth_mechanism <- "random_bspline" # Use "raised_cosine" for R3B.
scenario <- paste0(
  if (truth_mechanism == "random_bspline") "r3a_" else "r3b_",
  "real_genotype_one_per_gene_matched_functional_",
  "open_middle_3_12_",
  truth_mechanism,
  "_relative_clearance_main_effect"
)
time_grid <- 0:15
evaluation_grid <- seq(0, 15, by = 0.1)
class_probs <- c(
  dynamic_bspline = 0.20,
  constant = 0.40,
  zero = 0.40
)

effect_sim <- simulate_matched_functional_effect_set(
  n_variants = 6362,
  truth_mechanism = truth_mechanism,
  time_grid = time_grid,
  evaluation_grid = evaluation_grid,
  class_probs = class_probs,
  dynamic_main_effect_sd = 1,
  constant_sd = 1,
  bspline_amplitude = 2,
  bspline_df = 6,
  bspline_coefficient_sd = 1,
  cosine_width_half = 1.5,
  cosine_spike_counts = 1:3,
  cosine_relative_amplitude_range = c(0.35, 0.75),
  cosine_target_centered_rms = 0.9,
  switch_threshold = 0.25,
  middle_window = c(3, 12),
  middle_boundary = "open",
  location_truth_margin = 0.10,
  location_truth_min_range_fraction = 0.10,
  switch_truth_margin = 0.10,
  non_switch_min_abs = 0.10,
  non_switch_min_range_fraction = 0.10,
  exact_class_counts = TRUE,
  seed = seed,
  class_seed = component_seeds[["classes"]],
  constant_seed = component_seeds[["constant_effects"]],
  shape_seed = component_seeds[["functional_truth"]],
  scenario = scenario
)
effect_sim <- reassign_effect_simulation_by_maf(
  effect_sim = effect_sim,
  maf = genotype_sample$variant_info$observed_maf,
  class_probs = class_probs,
  seed = component_seeds[["classes"]],
  n_strata = 10L
)
expression_sim <- simulate_eqtl_expression_from_genotypes(
  G = G,
  beta_matrix = effect_sim$beta_matrix,
  time_grid = time_grid,
  covariates = covariates,
  expression_noise_sd = 1,
  covariate_effect_sd = 0.5,
  intercept_sd = 0,
  seed = component_seeds[["expression"]]
)
```

At each observed time, the genetic effect and standard error are estimated
from `Y ~ G + PC1 + PC2 + PC3 + PC4 + PC5`. The t-statistic correction uses
`df = 19 - 7 = 12`; FASH is fit to these estimates and corrected standard
errors.

## Example truth curves

Two examples are shown for each of the six functional truth cells. Points and
bars are the per-time estimates and two corrected standard errors used by
FASH; orange lines are the continuous truths. Shading marks the target timing
window.

### R3A examples

```{r r3a-examples, fig.width=11, fig.height=23, fig.cap="Figure 1. Representative broad random B-spline truth curves for each functional target and timing cell. Points and bars show estimated effects and two corrected standard errors, orange lines are continuous truths, and shading marks the target timing window."}
plot_truth_examples(
  example_curves$random_bspline,
  mechanism_labels[["random_bspline"]]
)
```

### R3B examples

```{r r3b-examples, fig.width=11, fig.height=23, fig.cap="Figure 2. Representative compact raised-cosine truth curves for each functional target and timing cell. Points and bars show estimated effects and two corrected standard errors, orange lines are continuous truths, and shading marks the target timing window."}
plot_truth_examples(
  example_curves$raised_cosine,
  mechanism_labels[["raised_cosine"]]
)
```

## Inference procedure

At each nominal level $\alpha$, the same procedure is applied to
`FASH-IWP1-Raw` and `FASH-IWP1-BF`:

1. Apply global dynamic-eQTL FDR control at level $\alpha$.
2. For each dynamically selected variant, evaluate the early, middle, late,
   and switch functionals using posterior samples.
3. Rank functional lfsr values and call variants whose cumulative FSR is at
   most $\alpha$.

Power is the fraction of true positive functionals that are called. Empirical
FSR is the fraction of functional calls whose known true functional is
non-positive. The figures report five-seed Monte Carlo means and 95% t-based
intervals across the full alpha grid.

```{r r3-fit-and-test-functionals, eval=FALSE}
out <- run_genotype_level_dynamic_eqtl_simulation(
  G = G,
  time_grid = time_grid,
  covariates = covariates,
  class_probs = class_probs,
  expression_noise_sd = 1,
  covariate_effect_sd = 0.5,
  intercept_sd = 0,
  dynamic_main_effect_sd = 1,
  alpha = 0.05,
  seed = seed,
  num_cores = 4,
  num_basis = 20,
  scenario = scenario,
  save_outputs = FALSE,
  effect_sim = effect_sim,
  expression_sim = expression_sim
)

true_functionals <- effect_sim$true_functionals[
  , c("early", "middle", "late", "switch"), drop = FALSE
]
true_dynamic <- out$unit_info$effect_class == "dynamic_bspline"
functional_posterior_seed <- component_seeds[["functional_posterior"]]

raw_results <- evaluate_fash_functional_testing(
  fit = out$fash_fits$fash_iwp1_raw,
  true_functionals = true_functionals,
  evaluation_grid = evaluation_grid,
  alpha_grid = seq(0.005, 0.20, by = 0.005),
  method = "FASH-IWP1-Raw",
  switch_threshold = 0.25,
  middle_window = c(3, 12),
  middle_boundary = "open",
  true_dynamic = true_dynamic,
  num_cores = 4,
  seed = functional_posterior_seed
)
bf_results <- evaluate_fash_functional_testing(
  fit = out$fash_fits$fash_iwp1_bf,
  true_functionals = true_functionals,
  evaluation_grid = evaluation_grid,
  alpha_grid = seq(0.005, 0.20, by = 0.005),
  method = "FASH-IWP1-BF",
  switch_threshold = 0.25,
  middle_window = c(3, 12),
  middle_boundary = "open",
  true_dynamic = true_dynamic,
  num_cores = 4,
  seed = functional_posterior_seed
)
```

Raw and BF therefore use common random numbers for posterior functional
summaries; their comparison is not perturbed by an avoidable difference in
Monte Carlo draws.

## R3A: broad random B-spline results

### Results at alpha 0.05

<div class="r3-table-scroll">

```{r r3a-table}
knitr::kable(
  r3a_table,
  align = "l",
  caption = paste(
    "Table 3. Functional-testing power and empirical false sign rate at",
    "alpha 0.05 under broad random B-spline truth."
  )
)
```

</div>

### Power as a function of alpha

```{r r3a-power, fig.height=9, fig.cap="Figure 3. Monte Carlo functional-testing power under broad random B-spline truth. Panels identify early, middle, late, and switch targets; curves identify raw and BF-updated FASH in the in-figure legend, with 95% intervals across five seeds."}
plot_mc_functional_curve_grid(
  mc_curve = r3a_alpha,
  metric = "power",
  title = "Functional-testing power: broad random B-spline truth"
)
```

### Empirical FSR as a function of alpha

```{r r3a-fsr, fig.height=9, fig.cap="Figure 4. Monte Carlo empirical false sign rate under broad random B-spline truth. Panels identify functional targets, curves identify raw and BF-updated FASH, and the diagonal reference denotes nominal alpha."}
plot_mc_functional_curve_grid(
  mc_curve = r3a_alpha,
  metric = "empirical_fsr",
  title = "Empirical FSR: broad random B-spline truth"
)
```

## R3B: compact raised-cosine results

### Results at alpha 0.05

<div class="r3-table-scroll">

```{r r3b-table}
knitr::kable(
  r3b_table,
  align = "l",
  caption = paste(
    "Table 4. Functional-testing power and empirical false sign rate at",
    "alpha 0.05 under compact raised-cosine truth."
  )
)
```

</div>

### Power as a function of alpha

```{r r3b-power, fig.height=9, fig.cap="Figure 5. Monte Carlo functional-testing power under compact raised-cosine truth. Panels identify early, middle, late, and switch targets; curves identify raw and BF-updated FASH in the in-figure legend, with 95% intervals across five seeds."}
plot_mc_functional_curve_grid(
  mc_curve = r3b_alpha,
  metric = "power",
  title = "Functional-testing power: compact raised-cosine truth"
)
```

### Empirical FSR as a function of alpha

```{r r3b-fsr, fig.height=9, fig.cap="Figure 6. Monte Carlo empirical false sign rate under compact raised-cosine truth. Panels identify functional targets, curves identify raw and BF-updated FASH, and the diagonal reference denotes nominal alpha."}
plot_mc_functional_curve_grid(
  mc_curve = r3b_alpha,
  metric = "empirical_fsr",
  title = "Empirical FSR: compact raised-cosine truth"
)
```

## Dynamic-null proportion

```{r r3-dynamic-null-reference}
formal_class_counts <- c(dynamic = 1272L, constant = 2545L, zero = 2545L)
formal_true_pi0 <- sum(formal_class_counts[c("constant", "zero")]) /
  sum(formal_class_counts)
```

The exact first-stage dynamic-null proportion is
$\pi_0=5090/6362\approx$ `r format_decimal(formal_true_pi0, 4)`. The table
reports the fitted constant-component weights for the raw and BF-adjusted IWP
fits.

<div class="r3-table-scroll">

```{r pi0-table}
knitr::kable(
  pi0_table,
  align = "l",
  caption = paste(
    "Table 5. Estimated first-stage dynamic-null proportions for raw and",
    "BF-updated IWP FASH under both functional truth mechanisms."
  )
)
```

</div>

## Interpretation

For R3A, BF-updated FASH has mean empirical FSR
`r format_decimal(max(r3a_alpha_005$mean_empirical_fsr[
  r3a_alpha_005$method == "FASH-IWP1-BF"
]), 3)` or lower for all four targets at $\alpha=0.05$. Its switch power is
`r format_decimal(metric_at(
  r3a_alpha_005, "switch", "FASH-IWP1-BF", "mean_power"
), 3)`.

For the more localized R3B trajectories, BF-updated FASH has low timing power:
`r format_decimal(metric_at(
  r3b_alpha_005, "early", "FASH-IWP1-BF", "mean_power"
), 3)`,
`r format_decimal(metric_at(
  r3b_alpha_005, "middle", "FASH-IWP1-BF", "mean_power"
), 3)`, and
`r format_decimal(metric_at(
  r3b_alpha_005, "late", "FASH-IWP1-BF", "mean_power"
), 3)` for early, middle, and late, respectively. Early has no false timing
calls; late and middle have mean empirical FSR
`r format_decimal(metric_at(
  r3b_alpha_005, "late", "FASH-IWP1-BF", "mean_empirical_fsr"
), 3)` and
`r format_decimal(metric_at(
  r3b_alpha_005, "middle", "FASH-IWP1-BF", "mean_empirical_fsr"
), 3)`, respectively. Its switch power is
`r format_decimal(metric_at(
  r3b_alpha_005, "switch", "FASH-IWP1-BF", "mean_power"
), 3)`, but its mean switch FSR is
`r format_decimal(metric_at(
  r3b_alpha_005, "switch", "FASH-IWP1-BF", "mean_empirical_fsr"
), 3)` at nominal $\alpha=0.05$. In this five-seed pilot, the scale-relative
clearance condition separates switch and non-switch truths in both mechanisms
without removing the compact transient behavior. Aligning the compact-peak
proposal with the revised timing windows removes the suspected
boundary-straddling construction, but it does not improve the R3B Middle
result: the mean empirical FSR is still above the nominal level in this run.
The proposal change is therefore a diagnostic clarification, not evidence that
the Middle calibration issue has been repaired. These results support
functional testing for the stated margin-separated designs; they do not imply
that BF adjustment alone guarantees FSR control for every functional
hypothesis or every truth geometry.

## Reproducibility

The
[R3 simulation driver](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/r3_functional_testing/run_matched_functional_testing_mc_replication.R)
uses the
[shared simulation functions](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/shared/simulation_functions.R).
Real-genotype selection, dosage extraction, validation, and MAF-balanced truth
assignment use the
[shared real-genotype helper](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/shared/real_genotype_one_per_gene.R)
and
[shared cache builder](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/shared/build_real_genotype_one_per_gene_cache.R).
The page-specific cache validation and formatting are in
[R3 reporting code](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/r3_functional_testing/reporting.R).
The formal
[center-aligned fashr 0.1.43 entry point](https://github.com/stephenslab/fashr-paper/blob/master/code/revision_simulations/r3_r4_fashr0143/16_r3_center_aligned_fashr0143.R)
pins the package commit, source hashes, five-seed genotype cache, design, and
output ID. It resumes valid seed-level work in a partial directory and only
promotes a cache after all ten mechanism-by-seed replicates pass its final
contract. On the staged Midway3 workspace, the exact entry point is:

```{bash regenerate-r3-cache-on-midway3, eval=FALSE}
sbatch slurm/16_r3_center_aligned_fashr0143.sbatch
```

The downloaded cache and its exact R1 pairing can be rechecked locally with:

```{bash validate-downloaded-r3-cache, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/r3_r4_fashr0143/validate_downloaded_r3_r4_fashr0143.R
```

These expensive commands are intentionally not evaluated during workflowr
builds; all tables and figures above read the validated saved cache.

Cached replicate metrics, call-level diagnostics, examples, summaries, and
figures are stored under:

```text
output/revision_simulations/mc/r3_real_genotype_one_per_gene_J6362_matched_functional_open_middle_3_12_center_aligned_relative_clearance_main_effect_fashr0143_pilot5/
```
