---
title: "Revision Simulation R3: functional testing under broad and transient effects"
author: "Ziang Zhang"
date: "2026-08-26"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 10,
  fig.height = 8,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source("code/revision_simulations/shared/simulation_functions.R")
source("code/revision_simulations/shared/real_genotype_one_per_gene.R")
source("code/revision_simulations/r3_functional_testing/reporting.R")
```

<style>
.r3-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.r3-table-scroll table {
  margin-bottom: 1em;
}
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: hidden;
}
</style>

## Overview

This simulation evaluates FASH functional testing directly over all `6,362`
simulated units. It uses two non-IWP truth mechanisms matched on sample size,
variant classes, genetic main effects, functional labels, and analysis.

Results are read from a completed five-seed Midway3 cache generated with
`fashr` 0.1.43 at Git SHA
`bf223df75da6e41ae48607a56b4cd12d7c3b24e7`.

For donor $i$, variant $j$, and time $t$,

$$
Y_{ijt} = \alpha_{jt} + G_{ij}\beta_j(t) +
X_i^T\gamma_{jt} + \epsilon_{ijt},
\qquad \epsilon_{ijt}\sim N(0,1).
$$

The genotype matrix is taken from the real YRI data used in the manuscript.
For each simulation seed, one tested cis variant is sampled uniformly from
each of the `6,362` tested genes, and its observed dosage vector is retained
for the same `19` donors. A tested variant assigned to multiple genes may
therefore appear more than once. R1, R2, R3A, and R3B use the exact same
sampled gene-variant pairs and genotype matrix for a given seed.

The five columns of $X$ remain simulated independently from $N(0,1)$ and are
standardized across donors. Independently for each covariate, gene, and time,
$\gamma_{kjt}\sim N(0,0.5^2)$. The current simulation sets $\alpha_{jt}=0$;
each fitted regression still includes an intercept. Effect trajectories,
covariates, covariate effects, and expression noise remain simulated.

```{r r3-simulate-genotypes-and-covariates, eval=FALSE}
seed <- 12345
component_seeds <- revision_component_seeds(seed)

genotype_cache <- readRDS(
  "output/revision_simulations/shared/real_genotype_one_per_gene_J6362_pilot5/genotype_samples.rds"
)
genotype_sample <- genotype_cache$samples[[as.character(seed)]]
G <- genotype_sample$G
covariates <- simulate_covariate_matrix(
  n_donors = 19,
  n_covariates = 5,
  seed = component_seeds[["covariates"]]
)
```

### Real-genotype design

<div class="r3-table-scroll">

```{r genotype-sampling-table}
knitr::kable(
  genotype_sampling_table,
  align = "l",
  caption = paste(
    "Table 1. Per-seed genotype sampling summary. Repeated assignments",
    "refer to the same tested variant selected for more than one gene."
  )
)
```

</div>

Each replicate contains `19` donors, `16` time points, five PC-like
covariates, `1,272` dynamic eQTLs, `2,545` constant eQTLs, and `2,545`
zero-effect units. Truth classes are assigned within empirical-MAF deciles so
that the dynamic, constant, and zero-effect groups have closely matched MAF
distributions.

<div class="r3-table-scroll">

```{r truth-maf-balance-table}
knitr::kable(
  truth_maf_balance_table,
  align = "l",
  caption = "Table 2. Empirical MAF balance across truth classes over five simulation seeds."
)
```

</div>

For every dynamic candidate,

$$
\beta_j(t)=\mu_j+h_j(t),
\qquad \mu_j\sim N(0,1).
$$

The `1,272` dynamic eQTLs are divided exactly across six preassigned truth
cells (`212` per cell): early, middle, or late crossed with switch or
non-switch. Each
candidate draws a new $\mu_j$ from $N(0,1)$; rejection sampling retains it only
when the final trajectory $\mu_j+h_j(t)$ satisfies its assigned cell.

### R3A: broad random B-spline effects

The deviation $h_j(t)$ is a cubic B-spline with six iid $N(0,1)$ coefficients.
It is centered over the 16 observed times and scaled to maximum absolute value
`2` before adding $\mu_j$.

### R3B: compact raised-cosine effects

The deviation $h_j(t)$ contains one, two, or three compact raised-cosine peaks
with half-width `1.5`. Primary peak centers are sampled from `1.5`--`2.5` for
Early, `4.5`--`10.5` for Middle, and `12.5`--`13.5` for Late. Secondary peak
centers use the same three aligned ranges, their amplitudes are random, and
the deviation is scaled to centered RMS `0.9` before adding $\mu_j$. As in
R3A, the final dense-grid trajectory must satisfy its assigned functional
label and clearance margin.

### Truth definitions

Early, middle, and late use the same supremum-contrast functionals as the
manuscript, with the revised temporal partition defined exactly as Early
$t\leq 3$, Middle $3<t<12$, and Late $t\geq 12$. On the dense evaluation grid
`seq(0, 15, by = 0.1)`, the Middle functional therefore uses the 89 points
from `3.1` through `11.9`; times `3` and `12` belong only to Early and Late,
respectively. These are continuous-time definitions. The simulation evaluates
their dense-grid analogues. Writing
$R_j=\sup_t\beta_j(t)-\inf_t\beta_j(t)$, the scale-adaptive location margin is

$$
m_j=\max(0.1,\,0.1R_j).
$$

The assigned Early, Middle, or Late contrast must be at least $m_j$, and both
competing location contrasts must be at most $-m_j$. Thus the same relative
clearance rule is applied symmetrically to all three location categories.

For the switch threshold $c=0.25$,

$$
T_{\mathrm{switch}}\{\beta\}
=
\min\left[
\sup_t\{\beta(t):\beta(t)>0\},
\sup_t\{|\beta(t)|:\beta(t)<0\}
\right]-c.
$$

A switch truth must have $T_{\mathrm{switch}}\geq 0.1$; equivalently, it has
both a positive and a negative effect of magnitude at least `0.35`. A
non-switch truth remains the same sign over the full dense grid. Its clearance
from zero must satisfy

$$
\min_t|\beta_j(t)|\geq\max(0.1,\,0.1R_j).
$$

Thus, a larger-amplitude non-switch curve must remain proportionally farther
from zero. These margins make the simulation labels unambiguous.

```{r r3-simulate-functional-truth, eval=FALSE}
truth_mechanism <- "random_bspline" # Use "raised_cosine" for R3B.
scenario <- paste0(
  if (truth_mechanism == "random_bspline") "r3a_" else "r3b_",
  "real_genotype_one_per_gene_matched_functional_",
  truth_mechanism,
  "_open_middle_3_12_center_aligned_equal_cells_",
  paste0(
    "relative_location_clearance_full_universe_",
    "paired_posterior_main_effect"
  )
)
time_grid <- 0:15
evaluation_grid <- seq(0, 15, by = 0.1)
class_probs <- c(
  dynamic_bspline = 0.20,
  constant = 0.40,
  zero = 0.40
)

effect_sim <- simulate_matched_functional_effect_set(
  n_variants = 6362,
  truth_mechanism = truth_mechanism,
  time_grid = time_grid,
  evaluation_grid = evaluation_grid,
  class_probs = class_probs,
  dynamic_main_effect_sd = 1,
  constant_sd = 1,
  bspline_amplitude = 2,
  bspline_df = 6,
  bspline_coefficient_sd = 1,
  cosine_width_half = 1.5,
  cosine_spike_counts = 1:3,
  cosine_relative_amplitude_range = c(0.35, 0.75),
  cosine_target_centered_rms = 0.9,
  cosine_center_ranges = list(
    early = c(1.5, 2.5),
    middle = c(4.5, 10.5),
    late = c(12.5, 13.5)
  ),
  switch_threshold = 0.25,
  middle_window = c(3, 12),
  middle_boundary = "open",
  location_truth_margin = 0.10,
  location_truth_min_range_fraction = 0.10,
  switch_truth_margin = 0.10,
  non_switch_min_abs = 0.10,
  non_switch_min_range_fraction = 0.10,
  temporal_category_probs = setNames(
    rep(1 / 3, 3),
    c("early", "middle", "late")
  ),
  exact_class_counts = TRUE,
  seed = seed,
  class_seed = component_seeds[["classes"]],
  constant_seed = component_seeds[["constant_effects"]],
  shape_seed = component_seeds[["functional_truth"]],
  scenario = scenario
)
effect_sim <- reassign_effect_simulation_by_maf(
  effect_sim = effect_sim,
  maf = genotype_sample$variant_info$observed_maf,
  class_probs = class_probs,
  seed = component_seeds[["classes"]],
  n_strata = 10L
)
expression_sim <- simulate_eqtl_expression_from_genotypes(
  G = G,
  beta_matrix = effect_sim$beta_matrix,
  time_grid = time_grid,
  covariates = covariates,
  expression_noise_sd = 1,
  covariate_effect_sd = 0.5,
  intercept_sd = 0,
  seed = component_seeds[["expression"]]
)
```

At each observed time, the genetic effect and standard error are estimated
from `Y ~ G + PC1 + PC2 + PC3 + PC4 + PC5`. The t-statistic correction uses
`df = 19 - 7 = 12`; FASH is fit to these estimates and corrected standard
errors.

## Example truth curves

Two examples are shown for each of the six functional truth cells. Points and
bars are the per-time estimates and two corrected standard errors used by
FASH; orange lines are the continuous truths. Shading marks the target timing
window.

### R3A examples

```{r r3a-examples, fig.width=11, fig.height=23, fig.cap="Figure 1. Representative broad random B-spline truth curves for each functional target and timing cell. Points and bars show estimated effects and two corrected standard errors, orange lines are continuous truths, and shading marks the target timing window."}
plot_truth_examples(
  example_curves$random_bspline,
  mechanism_labels[["random_bspline"]]
)
```

### R3B examples

```{r r3b-examples, fig.width=11, fig.height=23, fig.cap="Figure 2. Representative compact raised-cosine truth curves for each functional target and timing cell. Points and bars show estimated effects and two corrected standard errors, orange lines are continuous truths, and shading marks the target timing window."}
plot_truth_examples(
  example_curves$raised_cosine,
  mechanism_labels[["raised_cosine"]]
)
```

## Inference procedure

At each nominal level $\alpha$, the same procedure is applied to
`FASH-IWP1-Raw` and `FASH-IWP1-BF`:

1. Evaluate the early, middle, late, and switch functionals from posterior
   samples for every simulated unit.
2. Compute the functional lfsr for each unit and target.
3. Rank the lfsr values and call variants whose cumulative FSR is at
   most $\alpha$.

Power is the fraction of true positive functionals that are called. Empirical
FSR is the fraction of functional calls whose known true functional is
non-positive. Curves show five-replication Monte Carlo means over the complete
saved alpha grid from `0.005` through `0.20`. Power panels show Raw and BF;
empirical-FSR panels show BF only. BF ribbons are pointwise minima and maxima
across five replications.

```{r r3-fit-and-test-functionals, eval=FALSE}
out <- run_genotype_level_dynamic_eqtl_simulation(
  G = G,
  time_grid = time_grid,
  covariates = covariates,
  class_probs = class_probs,
  expression_noise_sd = 1,
  covariate_effect_sd = 0.5,
  intercept_sd = 0,
  dynamic_main_effect_sd = 1,
  alpha = 0.05,
  seed = seed,
  num_cores = 4,
  num_basis = 20,
  scenario = scenario,
  save_outputs = FALSE,
  effect_sim = effect_sim,
  expression_sim = expression_sim
)

true_functionals <- effect_sim$true_functionals[
  , c("early", "middle", "late", "switch"), drop = FALSE
]
true_dynamic <- out$unit_info$effect_class == "dynamic_bspline"
functional_posterior_seed <- component_seeds[["functional_posterior"]]

raw_results <- evaluate_fash_functional_testing(
  fit = out$fash_fits$fash_iwp1_raw,
  true_functionals = true_functionals,
  evaluation_grid = evaluation_grid,
  alpha_grid = seq(0.005, 0.20, by = 0.005),
  method = "FASH-IWP1-Raw",
  switch_threshold = 0.25,
  middle_window = c(3, 12),
  middle_boundary = "open",
  true_dynamic = true_dynamic,
  candidate_scope = "full_universe",
  num_cores = 4,
  seed = functional_posterior_seed
)
bf_results <- evaluate_fash_functional_testing(
  fit = out$fash_fits$fash_iwp1_bf,
  true_functionals = true_functionals,
  evaluation_grid = evaluation_grid,
  alpha_grid = seq(0.005, 0.20, by = 0.005),
  method = "FASH-IWP1-BF",
  switch_threshold = 0.25,
  middle_window = c(3, 12),
  middle_boundary = "open",
  true_dynamic = true_dynamic,
  candidate_scope = "full_universe",
  num_cores = 4,
  seed = functional_posterior_seed
)
```

Raw and BF therefore use common random numbers for posterior functional
summaries; their comparison is not perturbed by an avoidable difference in
Monte Carlo draws.

## R3A: broad random B-spline results

### Results at alpha 0.05

<div class="r3-table-scroll">

```{r r3a-table}
knitr::kable(
  r3a_table,
  align = "l",
  caption = paste(
    "Table 3. Functional-testing power and empirical false sign rate at",
    "alpha 0.05 under broad random B-spline truth."
  )
)
```

</div>

### Power as a function of alpha

```{r r3a-power, fig.height=9, fig.cap="Figure 3. Monte Carlo functional-testing power through nominal alpha 0.20 under broad random B-spline truth. The BF ribbon shows the pointwise minimum and maximum across five replications; Raw is shown without a ribbon."}
plot_mc_functional_curve_grid(
  mc_curve = r3a_alpha,
  metric = "power",
  target_order = c("early", "middle", "late", "switch"),
  alpha_reference = 0.05,
  title = "Functional-testing power: broad random B-spline truth",
  x_limits = plot_alpha_limits,
  interval_columns = c(
    "power_replication_min",
    "power_replication_max"
  ),
  interval_methods = "FASH-IWP1-BF"
)
```

### Empirical FSR as a function of alpha

```{r r3a-fsr, fig.height=9, fig.cap="Figure 4. BF-adjusted Monte Carlo empirical false sign rate through nominal alpha 0.20 under broad random B-spline truth. The diagonal denotes nominal alpha, and the ribbon shows the pointwise minimum and maximum across five replications."}
plot_mc_functional_curve_grid(
  mc_curve = r3a_alpha[
    r3a_alpha$method == "FASH-IWP1-BF",
    ,
    drop = FALSE
  ],
  metric = "empirical_fsr",
  target_order = c("early", "middle", "late", "switch"),
  alpha_reference = 0.05,
  title = "Empirical FSR: broad random B-spline truth",
  x_limits = plot_alpha_limits,
  interval_columns = c(
    "empirical_fsr_replication_min",
    "empirical_fsr_replication_max"
  ),
  interval_methods = "FASH-IWP1-BF"
)
```

## R3B: compact raised-cosine results

### Results at alpha 0.05

<div class="r3-table-scroll">

```{r r3b-table}
knitr::kable(
  r3b_table,
  align = "l",
  caption = paste(
    "Table 4. Functional-testing power and empirical false sign rate at",
    "alpha 0.05 under compact raised-cosine truth."
  )
)
```

</div>

### Power as a function of alpha

```{r r3b-power, fig.height=9, fig.cap="Figure 5. Monte Carlo functional-testing power through nominal alpha 0.20 under compact raised-cosine truth. The BF ribbon shows the pointwise minimum and maximum across five replications; Raw is shown without a ribbon."}
plot_mc_functional_curve_grid(
  mc_curve = r3b_alpha,
  metric = "power",
  target_order = c("early", "middle", "late", "switch"),
  alpha_reference = 0.05,
  title = "Functional-testing power: compact raised-cosine truth",
  x_limits = plot_alpha_limits,
  interval_columns = c(
    "power_replication_min",
    "power_replication_max"
  ),
  interval_methods = "FASH-IWP1-BF"
)
```

### Empirical FSR as a function of alpha

```{r r3b-fsr, fig.height=9, fig.cap="Figure 6. BF-adjusted Monte Carlo empirical false sign rate through nominal alpha 0.20 under compact raised-cosine truth. The diagonal denotes nominal alpha, and the ribbon shows the pointwise minimum and maximum across five replications."}
plot_mc_functional_curve_grid(
  mc_curve = r3b_alpha[
    r3b_alpha$method == "FASH-IWP1-BF",
    ,
    drop = FALSE
  ],
  metric = "empirical_fsr",
  target_order = c("early", "middle", "late", "switch"),
  alpha_reference = 0.05,
  title = "Empirical FSR: compact raised-cosine truth",
  x_limits = plot_alpha_limits,
  interval_columns = c(
    "empirical_fsr_replication_min",
    "empirical_fsr_replication_max"
  ),
  interval_methods = "FASH-IWP1-BF"
)
```

## Non-dynamic proportion

```{r r3-dynamic-null-reference}
formal_class_counts <- c(dynamic = 1272L, constant = 2545L, zero = 2545L)
formal_true_pi0 <- sum(formal_class_counts[c("constant", "zero")]) /
  sum(formal_class_counts)
```

The exact proportion of constant or zero-effect units is
$\pi_0=5090/6362\approx$ `r format_decimal(formal_true_pi0, 4)`. The table
reports the fitted constant-component weights for the raw and BF-adjusted IWP
fits.

<div class="r3-table-scroll">

```{r pi0-table}
knitr::kable(
  pi0_table,
  align = "l",
  caption = paste(
    "Table 5. Estimated non-dynamic proportions for raw and",
    "BF-updated IWP FASH under both functional truth mechanisms."
  )
)
```

</div>

## Reproducibility

The versioned cache was generated by
`code/revision_simulations/r3_r4_fashr0143/22_r3_full_universe_functional_fashr0143.R`
using the frozen driver and simulation functions under
`code/revision_simulations/r3_r4_fashr0143/source_snapshots/`.
Page-specific cache loading, validation, and formatting are in
`code/revision_simulations/r3_functional_testing/reporting.R`.

From the Midway3 workspace, rerun the simulation with:

```{bash regenerate-r3-cache-on-midway3, eval=FALSE}
sbatch slurm/22_r3_full_universe_functional_fashr0143.sbatch
```

Routine workflowr builds read the saved cache and do not rerun the simulation:

```text
output/revision_simulations/mc/r3_real_genotype_one_per_gene_J6362_matched_functional_open_middle_3_12_center_aligned_equal_cells_relative_location_clearance_full_universe_paired_posterior_fashr0143_pilot5/
```
