---
title: "Internal analysis: small-M clean working-model null sensitivity"
author: "Ziang Zhang"
date: "2026-08-19"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9.4,
  fig.height = 6.4,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(file.path(
  "code",
  "revision_simulations",
  "internal",
  "selected_signal_genotype_permutation",
  "report_small_m_clean_working_null_sensitivity.R"
))
```

<style>
.clean-null-banner {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border: 1px solid #d9a441;
  border-left: 6px solid #d9a441;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.clean-null-definition {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.clean-null-result {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #009e73;
  border-radius: 4px;
  background: #edf8f3;
  color: #164d3a;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
pre,
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
}
</style>

<div class="clean-null-banner">
<strong>Internal single-draw benchmark.</strong> This page is not linked from
the workflowr homepage, manuscript, or response letter. It is a clean
working-model diagnostic rather than a repeated-simulation calibration claim.
</div>

## Question and design

The preceding small-$M$ analysis used synchronized signal-stripped residual
permutations. This benchmark instead generates exact null observations under
the diagonal Gaussian working likelihood:

\[
\widehat\beta_{jt}^{(0)}
=
s_{jt} Z_{jt},
\qquad
Z_{jt}\overset{\mathrm{iid}}{\sim}N(0,1).
\]

<div class="clean-null-definition">
<ul>
<li><strong>Targets:</strong> the same 6,362 one-random-variant-per-gene units
selected with seed 20260817.</li>
<li><strong>Preserved inputs:</strong> each target unit's SE vector, 16-point
time grid, IWP grid, FASH settings, and cached target likelihood row.</li>
<li><strong>Clean-null draw:</strong> one independent Gaussian pool of 6,362
units generated with seed 20260819.</li>
<li><strong>Nested ratios:</strong> (M/J=0.05,0.10,0.20,1.00), using subset
seed 2026081901.</li>
<li><strong>Full procedure:</strong> raw empirical-Bayes fitting and
<code>BF_update()</code> are re-estimated for every merged universe; all lfdr
values come from that universe-specific fit.</li>
<li><strong>Uncertainty:</strong> this is one clean-null draw. No Monte Carlo
interval is reported.</li>
</ul>
</div>

The validated likelihood construction plus the resumed downstream fit took
`r format_decimal(clean_configuration$total_seconds, 1L)` seconds in total.
The likelihood construction itself took
`r format_decimal(clean_configuration$null_likelihood_seconds, 1L)` seconds.

## BF-adjusted result at alpha = 0.05

```{r clean-bf-table, results='asis'}
render_clean_table(
  clean_bf_alpha005_table,
  align = "rrrrrrrrrrr"
)
```

<div class="clean-null-result">
At nominal (\alpha=0.05), no clean working-model null was selected by the
BF-adjusted procedure at any tested (M/J). Therefore (V/R_{\mathrm{merged}}),
(V/M), and both plug-in FDR diagnostics are exactly zero for this one draw.
</div>

This zero is retained without a continuity correction. It does not establish
that the corresponding probability is exactly zero; it says that the observed
single-draw count is zero.

## Comparison with synchronized residual controls

The synchronized-residual entries below are the earlier median [IQR] across 20
nested subsets of one fixed residual-permutation pool. The clean-null entries
are single-draw values and therefore have no interval.

```{r residual-comparison-table, results='asis'}
render_clean_table(
  clean_residual_comparison_table,
  align = "rrrr"
)
```

The contrast is large: the full synchronized-residual benchmark had
$V/R_{\mathrm{merged}}=40.21\%$, whereas the clean working-model null draw has
zero selected BF-adjusted controls. Thus the severe synchronized-residual
result is not reproduced when the appended controls follow the fitted diagonal
Gaussian working model exactly.

## Raw-fit sanity check

```{r clean-raw-table, results='asis'}
render_clean_table(
  clean_raw_alpha005_table,
  align = "rrrrrrrrrrr"
)
```

The raw procedure does select clean null controls: two at $M/J=0.20$ and three
at $M/J=1.00$. Their known-null FDP lower bounds are 0.90% and 2.07%,
respectively. This confirms that the simulated rows enter the selection and
counting pipeline normally; the BF-adjusted zero is not a hard-coded or
denominator artifact.

## Known-null FDP lower-bound curve

```{r clean-known-null-fdp, fig.height=6.4, fig.cap="BF-adjusted single-draw known-null FDP lower bound V/R_merged versus nominal alpha for directly simulated working-model null controls."}
plot_clean_known_null_fdp()
```

If every appended unit is a valid known null, $V/R_{\mathrm{merged}}$ is a
lower bound on the merged FDP because false target discoveries are not counted
in $V$.

## Plug-in FDR curves

The plug-in diagnostics remain distinct from the direct lower bound:

\[
\widehat{\mathrm{FDR}}_T
=
\widehat\pi_{0,T}
\frac{V/M}{R_T/J},
\qquad
\widehat{\mathrm{FDR}}_{\mathrm{merged}}
=
\widehat\pi_{0,\mathrm{merged}}
\frac{V/M}{R_{\mathrm{merged}}/(J+M)}.
\]

```{r clean-plugin-fdr, fig.height=6.6, fig.cap="BF-adjusted target-pi0 and merged-pi0 plug-in FDR diagnostics for one clean working-model null draw."}
plot_clean_plugin_fdr()
```

The formulas use $V/M$ to estimate a null selection probability. The directly
counted discovery-set fraction remains $V/R_{\mathrm{merged}}$.

## Estimated null weights

```{r clean-pi0, fig.height=8.0, fig.cap="Raw and BF-adjusted merged and recovered-target null-proportion estimates after appending nested clean working-model null controls."}
plot_clean_pi0()
```

For the BF-adjusted fit, merged $\widehat\pi_0$ rises from 0.930 in the
target-only reference to 0.934, 0.938, 0.946, and 0.976 as $M/J$ increases.
The recovered target estimate is comparatively stable, moving from 0.930 to
0.931, 0.932, 0.935, and 0.952. Target calls decrease from 75 at $M/J=0.05$
to 46 at $M/J=1.00$, showing that even valid clean null controls affect the
re-estimated EB procedure when appended in large numbers.

## Null-generation diagnostics

The simulated null z-scores have overall mean
`r format_decimal(clean_null_z_overall$overall_mean, 4L)`, overall SD
`r format_decimal(clean_null_z_overall$overall_sd, 4L)`, and mean
off-diagonal time correlation
`r format_decimal(clean_null_z_overall$mean_off_diagonal_correlation, 4L)`.
The mean absolute off-diagonal correlation is
`r format_decimal(clean_null_z_overall$mean_absolute_off_diagonal_correlation, 4L)`.
All 12 numerical validation checks passed with zero genuine FASH warnings.

## Interpretation

1. **The severe synchronized-residual inflation disappears under the clean
working null.** No BF-adjusted clean null control is selected at
$\alpha=0.05$ in this draw.
2. **This is a working-model benchmark, not a real-data temporal-null model.**
The iid Gaussian construction deliberately removes temporal dependence,
signal leakage, heavy tails, and any other null misspecification.
3. **The zero count is finite-sample evidence, not an exact probability.** A
repeated simulation would be needed to estimate a small nonzero null selection
rate precisely.
4. **Large (M) still changes the EB fit.** Merged null weight increases and
target discoveries decrease even though the appended units are valid nulls.
Thus choosing (M\ll J) remains useful for limiting perturbation of the target
procedure.

## Reproduction and provenance

From the workflowr repository root:

```bash
Rscript --vanilla code/revision_simulations/internal/selected_signal_genotype_permutation/test_small_m_clean_working_null_helpers.R
Rscript --vanilla code/revision_simulations/internal/selected_signal_genotype_permutation/run_small_m_clean_working_null_sensitivity.R
Rscript --vanilla -e 'workflowr::wflow_build("analysis/revision_internal_small_m_clean_working_null_sensitivity.rmd")'
```

The run used the existing target likelihood cache but computed every clean-null
likelihood row from the simulated effect estimates and retained SE vectors. It
did not reuse residual-permutation null likelihood rows.
