---
title: "Internal analysis: FASH temporal classes across the Roadmap panel"
author: "Ziang Zhang"
date: "2026-08-26"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 3
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9.5,
  fig.height = 8.4,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

source(file.path(
  "code",
  "revision_simulations",
  "internal",
  "temporal_class_roadmap_enrichment",
  "reporting.R"
))
```

<style>
html,
body {
  max-width: 100%;
  overflow-x: hidden;
}
.tcr-internal-banner {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border: 1px solid #d9a441;
  border-left: 6px solid #d9a441;
  border-radius: 4px;
  background: #fff8e8;
  color: #4a3b1c;
}
.tcr-note {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.tcr-table-scroll {
  width: 100%;
  overflow-x: auto;
  -webkit-overflow-scrolling: touch;
}
.tcr-table-scroll table {
  margin-bottom: 1em;
}
.main-container img {
  max-width: 100%;
  height: auto;
}
pre,
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
}
</style>

<div class="tcr-internal-banner">
<strong>Internal analysis page.</strong> Not linked from the workflowr homepage,
the manuscript, or the response letter.
</div>

## Question

Two internal pages already exist, and this one is their product:

* [The three-epigenome temporal page](revision_internal_fash_temporal_class_enhancer_enrichment.html)
  splits FASH's discoveries into early / middle / late and tests them against
  E020, E013 and E095.
* [The thirteen-epigenome panel page](revision_internal_roadmap_enhancer_exploration.html)
  widens the annotation panel, but only for whole discovery sets.

Crossing them asks the question neither can answer alone: **do FASH's temporal
categories enrich differently across the wider Roadmap panel, and does the
heart-enhancer signal belong to a particular category?** The **switch** class is
new here — the three-epigenome page covers only the timing windows.

Both selection strategies are shown throughout: all discovered variants, and one
lead variant per gene chosen by **lfsr**.

Nothing is refit. Every classification comes from the cached
`testing_functional()` output under `output/dynamic_eQTL_real/`, which the page
verifies is the current open-$(3,12)$ middle window rather than the stale
$[4,11]$ historical estimand (see below). The discovery set is R6's and the
estimator is `enrichment_api.R` unchanged.

## Section 1: the four classes

The manuscript defines four functionals on the posterior of $\beta(t)$ over the
15-day course. Three locate the peak of $|\beta(t)|$:

$$
\text{early: } \sup_{t\le 3}|\beta(t)| - \sup_{t>3}|\beta(t)| > 0, \qquad
\text{middle: } \sup_{3<t<12}|\beta(t)| - \sup_{t\le 3\ \text{or}\ t\ge 12}|\beta(t)| > 0,
$$
$$
\text{late: } \sup_{t\ge 12}|\beta(t)| - \sup_{t<12}|\beta(t)| > 0.
$$

So $[0,3]$, $(3,12)$ and $[12,15]$ **partition** the time course exactly. Among
the sixteen integer-valued measurement days, the open Middle interval contains
days 4 through 11; inference is evaluated on the full 0.1-day grid.

The fourth asks whether the effect changes sign with appreciable magnitude on
both sides:

$$
\text{switch: } \min\Big(\sup_t \beta(t)^{+},\ \sup_t \beta(t)^{-}\Big) - `r format_decimal(configuration$switch_threshold, 2)` > 0 .
$$

<div class="tcr-note">
**Switch is not a fourth mutually exclusive class, and this page never treats it
as one.** The three timing windows partition "where does the peak fall"; switch
asks a different question, "does the sign flip". A pair can be both late and
switch, and `r format_integer(switch_crosstab$switch_pairs[switch_crosstab$timing_class == "all"])`
of the `r format_integer(discovered_pairs)` discovered pairs are switches spread
across all three windows.
</div>

Timing membership is the **argmax of the three window probabilities**, following
the three-epigenome page. A strict per-window call is not an option: at
lfsr $\le$ 0.05 the three windows hold 85 / 30 / 13 pairs, all at or below the
`r minimum_overlap`-overlap floor an estimate needs. Switch is the exception —
it is the one class where a strict call has enough variants, so it uses the
manuscript's own cumulative rule, `cfsr` $\le$
`r format_decimal(configuration$switch_cfsr_alpha, 2)`.

```{r class-table, echo=FALSE}
class_table()
```

Switch against the timing windows:

```{r switch-crosstab, echo=FALSE}
switch_crosstab_table()
```

Switch is spread almost uniformly — `r format_decimal(100 * switch_crosstab$switch_share[switch_crosstab$timing_class == "early"], 1)`%
of early pairs, `r format_decimal(100 * switch_crosstab$switch_share[switch_crosstab$timing_class == "middle"], 1)`%
of middle, `r format_decimal(100 * switch_crosstab$switch_share[switch_crosstab$timing_class == "late"], 1)`%
of late, against `r format_decimal(100 * switch_crosstab$switch_share[switch_crosstab$timing_class == "all"], 1)`%
overall. It is not a disguised version of any timing window.

### Verifying the middle window

The middle window is the open interval, so $[0,3]$, $(3,12)$ and $[12,15]$
partition the time course. That matters because a closed $[4,11]$ variant used to
circulate in this repository; it selected the same sixteen measured days but left
$(3,4)$ and $(11,12)$ uncovered on the 0.1-resolution grid the functionals are
evaluated on. The two are indistinguishable from the saved `lfsr` alone, so this
page tests which one produced its input rather than assuming.

Only the open window makes the three windows exhaustive, hence only it satisfies
$P(\text{early})+P(\text{middle})+P(\text{late}) = 1-\pi_0$:

* mean residual against $1-\pi_0$:
  `r format(middle_definition$mean_residual_open, digits = 2)`
* median residual: `r format(middle_definition$median_residual_open, digits = 2)`
* verdict: **`r middle_definition$window`**

That is Monte Carlo noise at 3,000 posterior draws. The comparison has to be
against $1-\pi_0$ and not against $1$: the residual against $1$ is
`r format(middle_definition$mean_residual_versus_one, digits = 2)`, centred on
$-\pi_0$ (median $\pi_0$ =
`r format_decimal(middle_definition$median_pi0, 3)`), and reading that deficit as
uncovered time is exactly how a correct cache gets mistaken for a stale one.

`verify_middle_definition()` runs the check on every rebuild and refuses to write
a cache if it fails, so a classification regenerated under a different window
would be caught rather than silently mixed in.

### Power, before looking at any result

A `r minimum_overlap`-overlap estimate needs roughly
`r minimum_overlap` / (background rate) variants, and background enhancer rates
across this panel span more than fourfold:

<div class="tcr-table-scroll">
```{r estimability, echo=FALSE}
estimability_table()
```
</div>

So the small lead sets — early with `r class_size("early", "lead")` variants and
switch with `r class_size("switch", "lead")` — cannot be estimated at the sparse
end of the panel whatever the truth is. Those cells are blank in the figures and
listed in Section 6. **Blank means unpowered, not null.**

## Section 2: the four classes against the tested background

```{r figure-a, fig.cap="Figure A. Enhancer enrichment for FASH's four temporal categories across thirteen Roadmap epigenomes, against all tested variants. All discovered variants on the left, one lead variant per gene (by class-specific lfsr) on the right. R6 enhancer definition (ChromHMM EnhG + Enh). Rows grouped by biology; filled circles are the three epigenomes the sister page shows. The x-axis is log2 so that depletion and enrichment are visually symmetric; the dashed line marks no enrichment. Bars are leave-one-autosome-out jackknife 95% intervals."}
plot_classes_vs_background("R6", "lfsr")
```

The all-variants column separates the classes about as cleanly as this kind of
analysis ever does.

**Early is depleted, everywhere.** All `r class_size("early", "all")` early
variants give `r fold_range("early", strategy = "all")` across the panel, with
`r count_below_one("early", strategy = "all")` estimable intervals lying entirely
*below* 1 and not one above it. The depletion is deepest at the germ-layer and
iPSC annotations (`r fold_list("early", strategy = "all", group = "Germ layer")`;
`r fold_list("early", strategy = "all", group = "iPSC")`) and shallowest at the
cardiac ones (`r fold_list("early", strategy = "all", group = "Heart")`). The
three-epigenome page reported early depletion at E020 and E013; it is not
specific to those two.

**Middle is flat.** `r fold_range("middle", strategy = "all")`, with
`r count_above_one("middle", strategy = "all")` intervals above 1 and
`r count_below_one("middle", strategy = "all")` below. Middle discoveries are
indistinguishable from the tested background at every annotation on the panel.

**Late carries the signal, and it is the only class with a tissue gradient.**
iPSC `r fold_range("late", strategy = "all", group = "iPSC")`
(`r count_above_one("late", strategy = "all", group = "iPSC")` above 1), germ
layer `r fold_range("late", strategy = "all", group = "Germ layer")`
(`r count_above_one("late", strategy = "all", group = "Germ layer")`), cardiac
`r fold_range("late", strategy = "all", group = "Heart")`
(`r count_above_one("late", strategy = "all", group = "Heart")`). The four
cardiac epigenomes are `r fold_list("late", strategy = "all", group = "Heart")` —
a tight block, every interval clear of 1.

**Switch is elevated but underpowered.** `r fold_range("switch", strategy = "all")`
over all `r class_size("switch", "all")` switch variants, every point estimate
above 1, but only `r count_above_one("switch", strategy = "all")` intervals clear
it — `r describe_fold("switch", "E013", strategy = "all")` at E013 mesoderm. With
a tenth of the discoveries the intervals are roughly twice as wide as late's.

```{r heatmap-a, fig.height=7.4, fig.cap="Figure A2. Fold enrichment for every class and selection strategy against the tested background, R6 enhancer definition. Bold entries have a jackknife 95% interval entirely clear of 1 in either direction. Blank cells fall below the minimum-overlap threshold. Colour is symmetric on the log2 scale."}
plot_fold_heatmap("Tested background", "R6", "lfsr")
```

The `Early all` column is uniformly blue and the `Late all` / `Switch` columns
uniformly red. That is the whole result in one picture.

### The lead-per-gene column

Leading by lfsr keeps the late-class enrichment —
`r count_above_one("late", strategy = "lead")` of the panel above 1, the most of
any cell on this page — but again flattens the tissue gradient: germ layer
`r fold_list("late", strategy = "lead", group = "Germ layer")` is now as high as
cardiac `r fold_list("late", strategy = "lead", group = "Heart")`. The same
flattening under lead selection appears on the thirteen-epigenome page for the
whole discovery set, so it is a property of lead selection rather than of the
timing classes.

Early's lead set is `r count_estimable("early", strategy = "lead")` estimable, so
that column says nothing either way. Middle's one clear cell is
`r describe_fold("middle", "E021", strategy = "lead")` at E021, an iPSC line, out
of `r count_estimable("middle", strategy = "lead")` estimable cells — a single
cell in a large search space, and not something to build on.

## Section 3: are the classes special *among* the discoveries?

Comparing a class to all tested variants conflates two things: whether FASH's
discoveries as a whole sit in enhancers, and whether this class does so more than
its siblings. Restricting the comparator to the other
`r format_integer(discovered_variants)` FASH-discovered variants isolates the
second.

```{r figure-b, fig.cap="Figure B. The same four categories, now against the other FASH discoveries rather than against all tested variants. This isolates class specificity from the enhancer propensity of the discovery set as a whole. All variants on the left, lead per gene (by lfsr) on the right."}
plot_classes_vs_discoveries("R6")
```

The within-discovery view is flatter across tissues and sharper across classes:

* **Early** is depleted relative to its siblings at
  `r count_below_one("early", strategy = "all", comparator = "Other FASH discoveries")`
  estimable epigenomes, over
  `r fold_range("early", strategy = "all", comparator = "Other FASH discoveries")`.
* **Middle** is
  `r fold_range("middle", strategy = "all", comparator = "Other FASH discoveries")`
  with `r count_above_one("middle", strategy = "all", comparator = "Other FASH discoveries")`
  above 1 and `r count_below_one("middle", strategy = "all", comparator = "Other FASH discoveries")`
  below — genuinely indistinguishable from the rest of the discoveries.
* **Late** is enriched at every epigenome,
  `r fold_range("late", strategy = "all", comparator = "Other FASH discoveries")`,
  with `r count_above_one("late", strategy = "all", comparator = "Other FASH discoveries")`
  intervals clear of 1 — and, crucially, **with no tissue gradient at all**.

<div class="tcr-note">
**A two-factor reading, which is the main thing this page adds.** The tissue
gradient and the class effect come apart. Against the tested background the late
class shows iPSC < germ layer < heart; against the other discoveries it is
uniformly elevated with no gradient. The gradient therefore belongs to the
**discovery set as a whole** — the thirteen-epigenome page shows exactly that
profile for `current_all`, cardiac 1.21–1.25 and iPSC 0.86–1.10 — while the
**timing class scales the overall enhancer propensity up or down**: early
suppresses it, late amplifies it, middle leaves it alone. The two effects
multiply; neither is evidence for the other.
</div>

### Switch against its own complement

```{r figure-b2, fig.cap="Figure B2. Switch variants against non-switch variants, both compared with the other FASH discoveries. The mirror-image behaviour is the same contrast read from either side."}
plot_switch_contrast("R6")
```

Switch runs
`r fold_range("switch", strategy = "all", comparator = "Other FASH discoveries")`
and non-switch
`r fold_range("no_switch", strategy = "all", comparator = "Other FASH discoveries")` —
a real contrast in the expected direction, but with only
`r count_above_one("switch", strategy = "all", comparator = "Other FASH discoveries")`
switch intervals clear of 1. Since switch is spread evenly across the timing
windows (Section 1), this is not the late-class effect in disguise; but 981
variants is not enough to say much more than "same direction, weaker evidence".

## Section 4: the summary questions, answered

**Is each class enriched, and where?** Early is *depleted* across the whole panel
(`r count_below_one("early", strategy = "all")` intervals below 1, none above).
Middle is flat everywhere. Late is enriched, most strongly and most consistently
at the four cardiac epigenomes
(`r count_above_one("late", strategy = "all", group = "Heart")`). Switch is
elevated everywhere but clears 1 at only
`r count_above_one("switch", strategy = "all")` epigenome.

**Does the three-epigenome page's finding survive the wider panel?** Yes, and it
generalises. That page reported early depleted at E020/E013 and the R6 signal
concentrated in late. Here early is depleted at
`r count_below_one("early", strategy = "all")` epigenomes spanning all four
biological groups, and late is the only class carrying the cardiac block. Neither
result was an artefact of the three-epigenome choice.

**Is switch distinguishable from non-switch discoveries?** Weakly. The direction
is consistent and the mirror contrast is clean, but the intervals mostly include
1. This needs either more discoveries or a pre-registered single contrast, not
more cells.

**Does any class track its own developmental stage?** **No.** The stage-matched
prediction — early at iPSC enhancers, middle at mesoderm, late at heart — fails
at the first step: early is depleted at iPSC enhancers rather than enriched
(`r fold_list("early", strategy = "all", group = "iPSC")`), which is the opposite
of the prediction. Middle is flat at E013. Only late-at-heart holds, and Section
3 shows that even that is the discovery set's tissue profile amplified by the
late class, not a late-specific preference for cardiac enhancers.

**Do the two selection strategies agree?** On the class ordering, yes — late
highest, middle flat, early lowest wherever it is estimable. On the tissue
gradient, no: it is present in the all-variants view and absent in the lead view.

**Does leading by lfsr rather than lfdr matter?** More than the enhancer
definition does. See Section 5.

**Is any of this sensitive to the enhancer-state definition?** No. See Section 5.

## Section 5: the two sensitivity axes

```{r sensitivity-numbers, echo=FALSE}
lead_sensitivity <- lead_rule_sensitivity()
definition <- definition_agreement()
```

### Lead-selection rule: this one matters

The two rules pick genuinely different variants. Within a class and gene, the
lfsr rule takes the variant most confidently *in that class*; the lfdr rule takes
the variant with the strongest overall discovery evidence, which is what the
three-epigenome sister page uses.

<div class="tcr-table-scroll">
```{r lead-rule-table, echo=FALSE}
lead_rule_table()
```
</div>

They share only
`r format_decimal(100 * lead_sensitivity$min_jaccard, 0)`%–`r format_decimal(100 * lead_sensitivity$max_jaccard, 0)`%
of their variants (Jaccard). Over the `r lead_sensitivity$n_cells` cells estimable
under both, fold enrichment correlates at
`r format_decimal(lead_sensitivity$correlation, 3)`, the largest single shift is
`r format_decimal(lead_sensitivity$max_absolute_shift, 2)`-fold, and
`r lead_sensitivity$point_crossings` cells move their point estimate across 1.

```{r figure-c, fig.height=8.8, fig.cap="Figure C. The two lead-selection rules side by side, lead sets only, against the tested background. Blue is the lfsr rule this page uses; orange is the overall-lfdr rule the three-epigenome page uses. Columns are the four classes."}
plot_lead_rule_sensitivity("R6")
```

That is a real dependence, and it is worth stating plainly: **the lead-per-gene
numbers on this page are not directly comparable with the lead-per-gene numbers
on the three-epigenome page**, because the two pages choose leads differently.
The all-variants numbers are comparable, and are identical by construction.

### Enhancer-state definition: this one does not

`r definition$n_cells` cells are estimable under both the R6 definition
(`EnhG + Enh`) and the Strober definition
(`EnhG + Enh + EnhBiv + BivFlnk`, verified from their source on the
thirteen-epigenome page). They correlate at
`r format_decimal(definition$correlation, 3)`, the largest shift is
`r format_decimal(definition$max_absolute_shift, 2)`-fold, and
`r definition$point_crossings` cells cross 1.

```{r figure-d, fig.height=8.8, fig.cap="Figure D. The two ChromHMM enhancer-state definitions side by side, all-variant sets, against the tested background. Same estimator, same sets; only the state list differs."}
plot_definition_sensitivity("all")
```

No conclusion in Section 4 changes. The state definition is the less consequential
of the two choices by a wide margin.

## Section 6: cells with too few overlapping variants

```{r not-estimable, echo=FALSE}
not_estimable_table()
```

`r nrow(not_estimable)` of the `r nrow(enrichment)` cells in the full design
(both definitions, both comparators, both lead rules) fall below the
`r minimum_overlap`-variant floor, concentrated in `early_lead`
(`r class_size("early", "lead")` variants) and `switch_lead`
(`r class_size("switch", "lead")`). Every one is a small set against a sparse
annotation, exactly as forecast in Section 1.

## Section 7: full estimate tables

Against the tested background, R6 definition, lfsr leads:

<div class="tcr-table-scroll">
```{r table-background, echo=FALSE}
full_summary_table("Tested background", "R6", "lfsr")
```
</div>

Against the other FASH discoveries, R6 definition, lfsr leads:

<div class="tcr-table-scroll">
```{r table-within, echo=FALSE}
full_summary_table("Other FASH discoveries", "R6", "lfsr")
```
</div>

## Methods

**Classifications.** The four cached `testing_functional()` results under
`output/dynamic_eQTL_real/`, computed on the BF-updated FASH-IWP1 fit over the
`r format_integer(discovered_pairs)` discovered gene-variant pairs. The middle
window is the open $(3,12)$ one, verified rather than assumed and reproducible
from the provenance-safe `code/02_dyn_lfsr.R` entrypoint.
`lfsr` is the
posterior probability that the class functional is $\le 0$; `cfsr` is its
cumulative counterpart. Nothing is refit here.

**Class membership.** Timing: argmax over the three window probabilities
$1 - \text{lfsr}$. The three windows partition the time course, so exactly one
of the three functionals is positive whenever the pair has an effect at all. The
functional is exactly $0$ under the null ($\beta(t)\equiv 0$ gives $0-0$) and
`lfsr_cal = mean(x <= 0)` counts that against membership, so

$$P(\text{early}) + P(\text{middle}) + P(\text{late}) = 1 - \pi_0,$$

with $\pi_0$ the posterior null probability (median
`r format_decimal(middle_definition$median_pi0, 3)` over the discovered pairs).
The residual is that null mass, not uncovered time. Switch: `cfsr` $\le$
`r format_decimal(configuration$switch_cfsr_alpha, 2)`, independent of timing.

**Lead selection.** Within a class, for each gene the variant with the smallest
class-specific lfsr (primary) or the smallest overall cumulative lfdr
(sensitivity). A gene whose variants split across classes contributes one lead to
each class it appears in.

**Annotation.** The thirteen-epigenome, two-definition indicator table built by
the [thirteen-epigenome experiment](revision_internal_roadmap_enhancer_exploration.html),
reused as-is. Nothing is re-downloaded or re-intersected.

**Comparators.** *Tested background*: all
`r format_integer(background_variant_count)` tested autosomal variants.
*Other FASH discoveries*: the same table restricted to the
`r format_integer(discovered_variants)` discovered variants. In both cases the
set under test is excluded from its own comparator, so `Comparator prop.` varies
slightly by row.

**Estimator.** Fold enrichment as a ratio of overlap proportions, with
leave-one-autosome-out jackknife 95% intervals over 22 blocks, minimum overlap
`r minimum_overlap`, no matched controls. `enrichment_api.R` is used unmodified.

**No formal tests.** No estimate here is compared with another by a hypothesis
test. The API's nominal p-values are in the output CSV, uncorrected and not
displayed.

**Consistency with the sister pages.** The union of the three timing classes is
the whole R6 discovery set, so pooling them must reproduce the thirteen-epigenome
page's `current_all` estimates. The run script checks this and refuses to write a
cache otherwise: `r nrow(pooled_crosscheck)` annotations, maximum absolute
difference
`r format(max(abs(pooled_crosscheck$fold_enrichment_pooled - pooled_crosscheck$fold_enrichment_published), na.rm = TRUE), digits = 3)`.
The timing counts also match the three-epigenome page exactly
(`r paste(paste0(class_summary$class[1:3], " ", format_integer(class_summary$pairs[1:3])), collapse = " / ")` pairs).

## Reproducibility

```{bash rebuild, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/temporal_class_roadmap_enrichment/run_temporal_class_roadmap_enrichment.R
```

Runs in about `r format_decimal(configuration$elapsed_minutes, 1)` minutes and
writes only its own output directory. It requires three caches it never modifies:

* `output/dynamic_eQTL_real/classify_dyn_eQTLs_{early,middle,late,switch}.RData`
  (currently the open-$(3,12)$ middle window; reproducible from
  `code/02_dyn_lfsr.R`);
* `output/.../fash_strober_enhancer_comparison_fashr0143/discovery_pair_tables.rds`
  (from the R6 experiment);
* `output/.../roadmap_enhancer_exploration/roadmap_enhancer_annotation_matrix.rds`
  (from the thirteen-epigenome experiment).

Tests:

```{bash tests, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/internal/temporal_class_roadmap_enrichment/test_temporal_class_roadmap_helpers.R
Rscript --vanilla \
  code/revision_simulations/internal/temporal_class_roadmap_enrichment/test_reporting.R
```

Rebuild this page:

```{bash reproduction-command, eval=FALSE}
Rscript --vanilla -e \
  'workflowr::wflow_build("analysis/revision_internal_temporal_class_roadmap_enrichment.rmd", local = TRUE)'
```

Substituting your own class definition needs no new code — the API takes any
named list of variant vectors:

```{r api-demo, eval=FALSE}
source(
  "code/revision_simulations/internal/fash_strober_enhancer_comparison/enrichment_api.R"
)
source(paste0(
  "code/revision_simulations/internal/temporal_class_roadmap_enrichment/",
  "temporal_class_roadmap_helpers.R"
))

classifications <- load_all_classifications()
pairs <- readRDS(paste0(
  "output/revision_simulations/internal/",
  "fash_strober_enhancer_comparison_fashr0143/discovery_pair_tables.rds"
))$current_all
membership <- build_class_membership(pairs, classifications)

# A stricter switch call, say, and lead by class lfsr.
strict <- build_class_variant_sets(
  build_class_membership(pairs, classifications, switch_alpha = 0.01),
  lead_rule = "lfsr"
)

annotations <- readRDS(file.path(
  roadmap_output_directory(), "roadmap_enhancer_annotation_matrix.rds"
))
compute_enrichment(
  strict["switch_all"], load_background(), annotations,
  controls = "background"
)
```

Output CSVs under
`output/revision_simulations/internal/temporal_class_roadmap_enrichment/`:
`enrichment_results.csv`, `set_sizes.csv`, `lead_rule_agreement.csv`,
`estimability_forecast.csv`, `pooled_timing_crosscheck.csv`,
`not_estimable_cells.csv`, `analysis_cache.rds`. Not tracked in git.

Built `r configuration$built_at_utc` with `r configuration$r_version`.

## References

Roadmap Epigenomics Consortium, Kundaje, A., Meuleman, W., Ernst, J.,
Bilenky, M., Yen, A., et al. (2015). Integrative analysis of 111 reference
human epigenomes. *Nature* **518**, 317–330.
<https://doi.org/10.1038/nature14248>

Strober, B. J., Elorbany, R., Rhodes, K., Krishnan, N., Tayeb, K.,
Battle, A., Gilad, Y. (2019). Dynamic genetic regulation of gene expression
during cellular differentiation. *Science* **364**, 1287–1290.
<https://doi.org/10.1126/science.aaw0040>
