---
title: "Revision Real-data Sensitivity R5: one random variant per gene"
author: "Ziang Zhang"
date: "2026-08-21"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 8,
  fig.height = 4.8,
  dpi = 180,
  fig.retina = 2,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

# This script loads the retained cache, validates its provenance and dimensions,
# and exposes the cache-backed data frames used below.
source(
  "code/revision_simulations/r5_one_variant_per_gene/reporting.R"
)
```

<style>
html,
body {
  max-width: 100%;
  overflow-x: hidden;
}
.r5-key-result {
  margin: 1.2em 0 1.5em;
  padding: 1em 1.15em;
  border-left: 6px solid #0072b2;
  border-radius: 4px;
  background: #eef7fc;
  color: #173f54;
}
.r5-table-scroll {
  box-sizing: border-box;
  width: 100%;
  max-width: 100%;
  margin: 0.8em 0 1.4em;
  overflow-x: auto;
  overflow-y: hidden;
  -webkit-overflow-scrolling: touch;
}
.r5-table-scroll table {
  margin-bottom: 0;
}
.r5-inline-path {
  white-space: normal;
  overflow-wrap: anywhere;
  word-break: break-word;
}
.main-container .figure {
  max-width: 820px;
  margin: 1.4em auto;
  text-align: center;
}
.main-container .figure img {
  display: block;
  max-width: 100%;
  height: auto;
  margin: 0 auto;
}
pre {
  box-sizing: border-box;
  max-width: 100%;
  overflow-x: auto;
}
.MathJax_Display {
  max-width: 100%;
  overflow-x: auto;
  overflow-y: hidden;
}
</style>

## Question and analysis

The complete FASH(1) analysis treats each tested gene--variant pair as one
empirical-Bayes unit. Its 1,009,173 pairs come from 6,362 genes, so genes
contribute unequal numbers of units. Because that imbalance could drive the
headline summaries on its own, we repeat them with one representative variant
per gene.

For each of 100 prespecified seeds, we uniformly sampled one tested variant
from every gene without using beta estimates, standard errors, likelihoods,
lfdr values, or discovery calls. We then re-estimated the FASH mixture weights
on the 6,362 selected likelihood rows with the original penalty of 10 and
applied the same BF prior update used in the paper.

<div class="r5-key-result">
<strong>Observed summaries.</strong> Across the 100 representative-variant
selections, the BF-updated null weight ranges from
`r format_decimal(min(seed_summary$thinned_pi0), 4)` to
`r format_decimal(max(seed_summary$thinned_pi0), 4)`; the complete-data value is
`r format_decimal(full_pi0, 4)`. Paired lfdr Spearman correlations range from
`r format_decimal(min(seed_summary$spearman_lfdr), 4)` to
`r format_decimal(max(seed_summary$spearman_lfdr), 4)`, and paired lfdr mean
absolute differences range from
`r format_decimal(min(seed_summary$mean_absolute_lfdr_difference), 4)` to
`r format_decimal(max(seed_summary$mean_absolute_lfdr_difference), 4)`.
Within-sampled-universe discovery Jaccard values range from
`r format_decimal(min(seed_summary$fdr_call_jaccard), 4)` to
`r format_decimal(max(seed_summary$fdr_call_jaccard), 4)`.
</div>

This page reports the BF-updated fit. The 100 seeds repeat the random choice of
representative variants; they do not resample cell lines.

## One-variant-per-gene design

The complete fit is pair-weighted: a gene contributes once for every tested
cis variant. A thinned fit contains one uniformly sampled pair from each gene,
so each gene contributes one empirical-Bayes unit.

```{r design-table, results='asis'}
# Record the fixed analysis choices shown in Table 1.
design_summary <- data.frame(
  Item = c(
    "Full-data empirical-Bayes units",
    "Genes represented in every refit",
    "Representative variants per gene",
    "Prespecified thinning seeds",
    "Selection rule",
    "Reported fit",
    "Raw empirical-Bayes penalty",
    "Discovery threshold",
    "Meaning of seed variation"
  ),
  Value = c(
    paste0(format_integer(expected_full_units), " gene-variant pairs"),
    format_integer(expected_full_genes),
    "Exactly one",
    paste0(
      length(expected_seeds), " (", min(expected_seeds), " to ",
      max(expected_seeds), ")"
    ),
    paste(
      "Uniform within gene and independent of beta, SE, likelihood, lfdr,",
      "and calls"
    ),
    "BF-updated FASH(1) only",
    format_integer(expected_penalty),
    "Cumulative-lfdr FDR 0.05 within each specified universe",
    "Monte Carlo variation over representative-variant choice"
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

render_scrollable_table(
  design_summary,
  caption = paste(
    "Table 1. Fixed design of the one-variant-per-gene analysis.",
    "Every refit reports the BF-updated FASH(1) result."
  ),
  align = "ll",
  minimum_width = "820px"
)
```

### Exact sampling and refitting steps

The retained analysis used the code path shown below. The expensive
gene--variant likelihood rows were computed once in the complete FASH fit. For
each seed, the analysis selected one row per gene, re-estimated the mixture
weights on those rows, and then applied `BF_update()`.

```{r one-per-gene-refit, eval=FALSE}
# Identify every tested pair and partition row indices by gene.
pair_keys <- names(full_raw_fit$fash_data$data_list)
gene_id <- sub("_.*$", "", pair_keys)
gene_index <- split(
  seq_along(pair_keys),
  factor(gene_id, levels = sort(unique(gene_id), method = "radix")
)

seeds <- seq(12345L, by = 10000L, length.out = 100L)

for (seed in seeds) {
  # Uniformly choose one tested pair from every gene.
  set.seed(seed)
  selected_indices <- vapply(
    gene_index,
    function(indices) sample(indices, size = 1L, replace = FALSE),
    integer(1)
  )
  selected_pair_keys <- pair_keys[selected_indices]

  stopifnot(
    length(selected_indices) == 6362L,
    anyDuplicated(selected_indices) == 0L
  )

  # Subset the exact cached likelihood rows for the selected pairs.
  selected_likelihood <- full_raw_fit$L_matrix[
    selected_indices,
    ,
    drop = FALSE
  ]
  rownames(selected_likelihood) <- selected_pair_keys

  # Re-estimate mixture weights with the complete fit's grid and penalty.
  thinned_eb <- fashr::fash_eb_est(
    L_matrix = selected_likelihood,
    grid = full_raw_fit$psd_grid,
    penalty = 10
  )
  rownames(thinned_eb$posterior_weight) <- selected_pair_keys
  null_column <- which(thinned_eb$prior_weight$psd == 0)

  # Construct the minimal FASH object required by the paper's BF update.
  thinned_raw_fit <- structure(
    list(
      prior_weights = thinned_eb$prior_weight,
      posterior_weights = thinned_eb$posterior_weight,
      psd_grid = full_raw_fit$psd_grid,
      lfdr = thinned_eb$posterior_weight[, null_column],
      L_matrix = selected_likelihood
    ),
    class = "fash"
  )
  thinned_bf_fit <- fashr::BF_update(thinned_raw_fit, plot = FALSE)

  # Compare the refitted lfdr with the complete-data lfdr on the same pairs.
  paired_lfdr <- data.frame(
    pair_key = selected_pair_keys,
    full_lfdr = full_bf_fit$lfdr[selected_indices],
    thinned_lfdr = thinned_bf_fit$lfdr
  )
}
```

The complete and thinned fits use the same 52-component PSD grid, Gaussian
likelihood, 20 basis functions, order 1, `betaprec = 0`, `pred_step = 1`, and
raw empirical-Bayes `penalty = 10`.

## Unequal numbers of tested variants

```{r variant-count-summary, results='asis'}
# Summarize how many tested variants each of the 6,362 genes contributes.
variant_count_quantiles <- stats::quantile(
  variant_count_by_gene$n_tested_variants,
  probs = c(0, 0.25, 0.5, 0.75, 1),
  names = FALSE,
  type = 7
)
variant_count_summary <- data.frame(
  Statistic = c("Minimum", "Q1", "Median", "Mean", "Q3", "Maximum"),
  `Tested variants per gene` = c(
    variant_count_quantiles[1:3],
    mean(variant_count_by_gene$n_tested_variants),
    variant_count_quantiles[4:5]
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
variant_count_summary$`Tested variants per gene` <- vapply(
  variant_count_summary$`Tested variants per gene`,
  function(value) {
    if (abs(value - round(value)) < 1e-10) {
      format_integer(value)
    } else {
      format_decimal(value, 1)
    }
  },
  character(1)
)

render_scrollable_table(
  variant_count_summary,
  caption = "Table 2. Tested variants per gene in the complete analysis.",
  align = "lr",
  minimum_width = "520px"
)
```

```{r variants-per-gene-histogram, fig.width=7.4, fig.height=4.5, out.width='82%', fig.cap="Figure 1. Distribution of tested variants per gene in the complete FASH analysis. The horizontal axis is logarithmic; every sensitivity refit uses one uniformly selected pair from each gene."}
# Plot one row per gene using the cached tested-variant counts.
variant_histogram_data <- variant_count_by_gene

ggplot2::ggplot(
  variant_histogram_data,
  ggplot2::aes(x = n_tested_variants)
) +
  ggplot2::geom_histogram(
    bins = 45,
    fill = r5_blue,
    color = "white",
    linewidth = 0.25
  ) +
  ggplot2::scale_x_log10(
    breaks = c(2, 5, 10, 20, 50, 100, 250, 500, 1000, 2000),
    labels = scales::label_comma()
  ) +
  ggplot2::labs(
    title = "Tested variants per gene in the complete fit",
    subtitle = "Each of the 6,362 genes contributes one count",
    x = "Tested variants per gene (log scale)",
    y = "Number of genes"
  ) +
  r5_theme
```

## A concrete draw: seed 12345

```{r seed-12345-table, results='asis'}
# Extract the retained summary for the first prespecified seed.
seed_12345_summary <- seed_summary[
  seed_summary$seed == 12345L,
  ,
  drop = FALSE
]
seed_12345_display <- data.frame(
  Metric = c(
    "Complete-data BF-updated null weight",
    "Thinned BF-updated null weight",
    "Paired lfdr mean absolute difference",
    "Paired lfdr Spearman correlation",
    "Reference calls on the sampled pairs",
    "Thinned-refit calls on the sampled pairs",
    "Within-sampled-universe discovery Jaccard"
  ),
  Value = c(
    format_decimal(seed_12345_summary$full_pi0, 4),
    format_decimal(seed_12345_summary$thinned_pi0, 4),
    format_decimal(seed_12345_summary$mean_absolute_lfdr_difference, 4),
    format_decimal(seed_12345_summary$spearman_lfdr, 4),
    format_integer(seed_12345_summary$full_fdr_calls),
    format_integer(seed_12345_summary$thinned_fdr_calls),
    format_decimal(seed_12345_summary$fdr_call_jaccard, 4)
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

render_scrollable_table(
  seed_12345_display,
  caption = paste(
    "Table 3. Seed-12345 comparison. Both call sets apply",
    "cumulative-lfdr FDR 0.05 to the same 6,362 sampled pairs."
  ),
  align = "lr",
  minimum_width = "700px"
)
```

```{r seed-12345-lfdr, fig.width=6.3, fig.height=5.4, out.width='70%', fig.cap="Figure 2. Paired lfdr values for seed 12345. Each point is one sampled gene--variant pair. Both axes use the identical 6,362 pairs; the dashed line denotes equality."}
# The cache stores complete-fit and refitted lfdr for the same selected pairs.
seed_12345_plot_data <- seed_12345_lfdr

ggplot2::ggplot(
  seed_12345_plot_data,
  ggplot2::aes(x = full_lfdr, y = thinned_lfdr)
) +
  ggplot2::geom_abline(
    intercept = 0,
    slope = 1,
    color = "grey45",
    linetype = 2,
    linewidth = 0.7
  ) +
  ggplot2::geom_point(
    color = r5_blue,
    alpha = 0.45,
    size = 1.25
  ) +
  ggplot2::coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  ggplot2::labs(
    title = "Seed 12345: paired lfdr values",
    subtitle = "Complete-fit and refitted lfdr on the same 6,362 pairs",
    x = "Complete-data BF-updated lfdr",
    y = "Thinned-refit BF-updated lfdr"
  ) +
  r5_theme
```

## Summaries across 100 seeds

```{r null-weight-stability, fig.width=6.4, fig.height=4.2, out.width='70%', fig.cap="Figure 3. BF-updated null-weight estimates across 100 one-variant-per-gene refits. Each point is one seed; the dashed line is the complete-data estimate."}
# Make one plotting row for each thinned refit.
null_weight_plot_data <- data.frame(
  Analysis = "One variant per gene",
  seed = seed_summary$seed,
  null_weight = seed_summary$thinned_pi0,
  stringsAsFactors = FALSE
)

ggplot2::ggplot(
  null_weight_plot_data,
  ggplot2::aes(x = Analysis, y = null_weight)
) +
  ggplot2::geom_hline(
    yintercept = full_pi0,
    color = r5_orange,
    linetype = 2,
    linewidth = 0.8
  ) +
  ggplot2::geom_boxplot(
    width = 0.32,
    outlier.shape = NA,
    fill = scales::alpha(r5_blue, 0.25),
    color = r5_blue
  ) +
  ggplot2::geom_jitter(
    width = 0.12,
    height = 0,
    color = r5_blue,
    alpha = 0.65,
    size = 1.5
  ) +
  ggplot2::scale_y_continuous(
    labels = scales::label_number(accuracy = 0.002)
  ) +
  ggplot2::labs(
    title = "BF-updated null weights",
    subtitle = paste0(
      "100 seeds; dashed line = complete-data estimate (",
      format_decimal(full_pi0, 4), ")"
    ),
    x = NULL,
    y = "Estimated null weight"
  ) +
  r5_theme
```

The maximum absolute difference from the complete-data null weight is
`r format_decimal(max(abs(seed_summary$pi0_difference)), 4)`.

```{r paired-lfdr-agreement, fig.width=7.8, fig.height=4.4, out.width='84%', fig.cap="Figure 4. Paired lfdr summaries across 100 refits. Each point summarizes one seed's 6,362 matched pairs. The two panels show mean absolute difference and Spearman correlation."}
# Stack the two seed-level lfdr diagnostics for faceted plotting.
lfdr_agreement_plot_data <- rbind(
  data.frame(
    seed = seed_summary$seed,
    Metric = "Mean absolute lfdr difference",
    Value = seed_summary$mean_absolute_lfdr_difference
  ),
  data.frame(
    seed = seed_summary$seed,
    Metric = "Spearman lfdr correlation",
    Value = seed_summary$spearman_lfdr
  )
)

ggplot2::ggplot(
  lfdr_agreement_plot_data,
  ggplot2::aes(x = Metric, y = Value)
) +
  ggplot2::geom_boxplot(
    width = 0.32,
    outlier.shape = NA,
    fill = scales::alpha(r5_blue, 0.25),
    color = r5_blue
  ) +
  ggplot2::geom_jitter(
    width = 0.12,
    height = 0,
    color = r5_blue,
    alpha = 0.65,
    size = 1.35
  ) +
  ggplot2::facet_wrap(~ Metric, scales = "free_y") +
  ggplot2::labs(
    title = "Paired lfdr summaries across 100 seeds",
    subtitle = "Each point compares complete-fit and refitted lfdr on the same pairs",
    x = NULL,
    y = NULL
  ) +
  r5_theme +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(face = "bold")
  )
```

```{r across-seed-summary, results='asis'}
# Calculate the same five-number summary for each seed-level diagnostic.
summarize_metric <- function(metric, values, digits) {
  five_numbers <- stats::quantile(
    values,
    probs = c(0, 0.25, 0.5, 0.75, 1),
    names = FALSE,
    type = 7
  )
  formatted <- vapply(
    five_numbers,
    format_decimal,
    character(1),
    digits = digits
  )
  data.frame(
    Metric = metric,
    Minimum = formatted[1],
    Q1 = formatted[2],
    Median = formatted[3],
    Q3 = formatted[4],
    Maximum = formatted[5],
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
}

across_seed_summary <- do.call(rbind, list(
  summarize_metric(
    "Thinned BF-updated null weight",
    seed_summary$thinned_pi0,
    4
  ),
  summarize_metric(
    "Absolute null-weight change",
    abs(seed_summary$pi0_difference),
    4
  ),
  summarize_metric(
    "Paired lfdr mean absolute difference",
    seed_summary$mean_absolute_lfdr_difference,
    4
  ),
  summarize_metric(
    "Paired lfdr Spearman correlation",
    seed_summary$spearman_lfdr,
    4
  ),
  summarize_metric(
    "Within-sampled-universe discovery Jaccard",
    seed_summary$fdr_call_jaccard,
    4
  ),
  summarize_metric(
    "Reference calls on sampled pairs",
    seed_summary$full_fdr_calls,
    1
  ),
  summarize_metric(
    "Thinned-refit calls / genes",
    seed_summary$thinned_fdr_calls,
    1
  )
))

render_scrollable_table(
  across_seed_summary,
  caption = paste(
    "Table 4. Five-number summaries across 100",
    "representative-variant selections."
  ),
  align = "lrrrrr",
  minimum_width = "980px"
)
```

The median paired lfdr mean absolute difference is
`r format_decimal(median(seed_summary$mean_absolute_lfdr_difference), 4)`; the
minimum paired lfdr Spearman correlation is
`r format_decimal(min(seed_summary$spearman_lfdr), 4)`.

## Discovery calls within the same sampled universe

The complete fit has `r format_integer(expected_full_discovered_pairs)` called
pairs from `r format_integer(expected_full_discovered_genes)` genes among all
1,009,173 tested pairs. The thinned fits test 6,362 selected pairs, so their
`r format_integer(thinned_discovery_range[1])`--`r format_integer(thinned_discovery_range[2])`
calls are not compared directly with the complete-data total.

For a matched comparison, both call sets are recomputed on the same 6,362
selected pairs. The within-sampled-universe discovery Jaccard is the number of
pairs called by both fits divided by the number called by either fit.

```{r matched-discovery-code, eval=FALSE}
# Apply cumulative-lfdr FDR 0.05 separately to the two lfdr vectors.
cumulative_fdr_calls <- function(lfdr, alpha = 0.05) {
  lfdr_order <- order(lfdr, method = "radix")
  cumulative_fdr <- cumsum(lfdr[lfdr_order]) / seq_along(lfdr_order)
  accepted_ranks <- which(cumulative_fdr <= alpha)
  lfdr_order[accepted_ranks]
}

reference_calls <- cumulative_fdr_calls(paired_lfdr$full_lfdr)
thinned_calls <- cumulative_fdr_calls(paired_lfdr$thinned_lfdr)

call_intersection <- intersect(reference_calls, thinned_calls)
call_union <- union(reference_calls, thinned_calls)
discovery_jaccard <- if (length(call_union) == 0L) {
  1
} else {
  length(call_intersection) / length(call_union)
}
```

```{r discovery-robustness, fig.width=8.4, fig.height=4.4, out.width='90%', fig.cap="Figure 5. Discovery-call summaries within each seed's matched 6,362-pair universe. The first two panels show call counts after cumulative-lfdr FDR 0.05; the third shows intersection divided by union."}
# Stack the reference count, refit count, and Jaccard for each seed.
discovery_plot_data <- rbind(
  data.frame(
    seed = seed_summary$seed,
    Metric = "Reference calls\non sampled pairs",
    Value = seed_summary$full_fdr_calls
  ),
  data.frame(
    seed = seed_summary$seed,
    Metric = "Thinned-refit calls\non sampled pairs",
    Value = seed_summary$thinned_fdr_calls
  ),
  data.frame(
    seed = seed_summary$seed,
    Metric = "Within-sampled-universe\ndiscovery Jaccard",
    Value = seed_summary$fdr_call_jaccard
  )
)
discovery_plot_data$Metric <- factor(
  discovery_plot_data$Metric,
  levels = c(
    "Reference calls\non sampled pairs",
    "Thinned-refit calls\non sampled pairs",
    "Within-sampled-universe\ndiscovery Jaccard"
  )
)
metric_colors <- c(
  "Reference calls\non sampled pairs" = r5_orange,
  "Thinned-refit calls\non sampled pairs" = r5_blue,
  "Within-sampled-universe\ndiscovery Jaccard" = r5_green
)

ggplot2::ggplot(
  discovery_plot_data,
  ggplot2::aes(x = Metric, y = Value, color = Metric, fill = Metric)
) +
  ggplot2::geom_boxplot(
    width = 0.32,
    outlier.shape = NA,
    alpha = 0.18
  ) +
  ggplot2::geom_jitter(
    width = 0.12,
    height = 0,
    alpha = 0.65,
    size = 1.3
  ) +
  ggplot2::facet_wrap(~ Metric, scales = "free_y", nrow = 1) +
  ggplot2::scale_color_manual(values = metric_colors) +
  ggplot2::scale_fill_manual(values = metric_colors) +
  ggplot2::labs(
    title = "Discovery calls on matched sampled pairs",
    subtitle = "Each point summarizes one of the 100 representative-variant seeds",
    x = NULL,
    y = NULL
  ) +
  r5_theme +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(face = "bold"),
    legend.position = "none"
  )
```

```{r discovery-frequency-summary, results='asis'}
# Count how often each gene appears in a thinned-refit discovery set.
full_data_status <- gene_discovery_frequency$full_data_discovered_gene
discovery_frequency_summary <- data.frame(
  `Gene set` = c(
    "All tested genes",
    "Complete-data discovered genes",
    "Genes not discovered in the complete data"
  ),
  Genes = c(
    nrow(gene_discovery_frequency),
    sum(full_data_status),
    sum(!full_data_status)
  ),
  `Discovered in at least one thinning seed` = c(
    sum(gene_discovery_frequency$n_seeds_discovered > 0L),
    sum(
      gene_discovery_frequency$n_seeds_discovered > 0L &
        full_data_status
    ),
    sum(
      gene_discovery_frequency$n_seeds_discovered > 0L &
        !full_data_status
    )
  ),
  `Mean seed discovery frequency` = c(
    mean(gene_discovery_frequency$discovery_frequency),
    mean(gene_discovery_frequency$discovery_frequency[full_data_status]),
    mean(gene_discovery_frequency$discovery_frequency[!full_data_status])
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
discovery_frequency_summary$Genes <- format_integer(
  discovery_frequency_summary$Genes
)
discovery_frequency_summary$`Discovered in at least one thinning seed` <-
  format_integer(
    discovery_frequency_summary$`Discovered in at least one thinning seed`
  )
discovery_frequency_summary$`Mean seed discovery frequency` <-
  scales::label_percent(accuracy = 0.1)(
    discovery_frequency_summary$`Mean seed discovery frequency`
  )

render_scrollable_table(
  discovery_frequency_summary,
  caption = paste(
    "Table 5. Frequency with which genes are called across the 100",
    "representative-variant selections."
  ),
  align = "lrrr",
  minimum_width = "920px"
)
```

Across the 100 seeds, `r format_integer(ever_discovered)` genes are called at
least once: `r format_integer(ever_discovered_full)` are among the
`r format_integer(expected_full_discovered_genes)` complete-data discovered
genes and `r format_integer(ever_discovered_additional)` are not. These are
repeated-seed counts, not a new discovery list.

## Reproducibility

Routine builds read the compact retained cache at:

<span class="r5-inline-path">`output/revision_simulations/r5_one_variant_per_gene_100seed_fashr0143/`</span>

The cache records the source-object SHA-256 hashes, the exact 52-component PSD
grid, `penalty = 10`, package versions, warnings, and elapsed time. It contains
100 compact seed directories and does not store full thinned FASH objects.

The former 20-variant analysis is retained as an
[unlinked internal supporting page](revision_internal_twenty_variant_per_gene_refit.html).

```{bash reproduction-command, eval=FALSE}
Rscript --vanilla \
  code/revision_simulations/r5_one_variant_per_gene/run_one_variant_per_gene.R

Rscript --vanilla \
  code/revision_simulations/r5_one_variant_per_gene/test_one_variant_per_gene_helpers.R

Rscript --vanilla \
  code/revision_simulations/r5_one_variant_per_gene/test_reporting.R

Rscript --vanilla -e \
  'workflowr::wflow_build("analysis/revision_balanced_variant_thinning.rmd")'
```
